---
title: "Validating SaoMNK: Reproducing Classical NK Landscape Dynamics"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Validating SaoMNK Against Classical NK}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Overview

This vignette validates the SaoMNK framework against canonical results
from the NK fitness landscape literature, particularly Levinthal (1997).
We demonstrate that SaoMNK reproduces the classical dynamics of search
on rugged landscapes---local optima trapping, the effect of epistasis on
ruggedness, and the exploration--exploitation tradeoff---while also
showing how SaoMNK's endogenous network structure yields richer dynamics
unavailable in the standard NK model.

The validation proceeds in three parts:

1.  **Single-agent rugged landscape**: An $M=2$ analogue of Levinthal
    (1997), showing that XWX-mediated epistasis creates local optima.
2.  **Multi-agent search**: An $M=6$ extension showing how endogenous
    sociality ($K_{AA}$) and endogenous epistasis ($K_{CC}$) emerge from
    the coupled bipartite process.
3.  **Parameter mapping**: A systematic correspondence between NK
    parameters and SaoMNK configuration.

```{r load-saomnk}
if (requireNamespace("searchnet", quietly = TRUE)) {
  library(searchnet)
} else {
  dir_r <- file.path(getwd(), "R")
  if (!file.exists(file.path(dir_r, "utils.R")))
    dir_r <- "D:/Search_networks/SaoMNK/R"
  .saomnk_dir <- dir_r
  .saomnk_dir <- dir_r; source(file.path(dirname(dir_r), "inst", "saomnk-loader.R"))
}

library(Matrix)
```

# 1. Single-Agent Rugged Landscape (Levinthal 1997 Analogue)

In the classical NK model, a minimal agent searches an $N$-dimensional
binary landscape whose ruggedness is governed by $K$, the number of
epistatic interactions per component. Levinthal (1997) showed that as
$K$ increases, the number of local optima grows and the agent is
increasingly likely to become trapped at suboptimal peaks.

SaoMNK reproduces this by setting $M=2$ (single actor) with $N=8$
components. The epistasis structure is specified via a block-diagonal
interaction matrix passed to the `XWX` effect, which captures the
component--component interdependence mediated by actor affiliations. The
`density` parameter controls the baseline propensity to adopt
components: negative values make the landscape more rugged (higher cost
of exploration), while positive values flatten it.

## 1.1 Environment Configuration

```{r single-agent-config}
DV_NAME <- "self$bipartite_rsienaDV"
steps_per_actor <- 30

## Modular epistasis: 2 blocks of 4 components each
## This is analogous to K=3 within each module (full intra-block epistasis)
epistasis_mat <- create_block_diag(8, 2)
cat("Epistasis matrix (2 blocks of 4):\n")
print(epistasis_mat)
```

## 1.2 Rugged Landscape (Negative Density)

With negative density, the agent faces a cost for each additional
component adoption. Combined with positive XWX (reward for epistatic
complementarities), this creates a rugged landscape with local optima:
the agent must navigate tradeoffs between the cost of scope and the
benefit of complementarity.

```{r single-rugged-setup}
env_rugged <- SaomNkRSienaBiEnv$new(list(
  M = 2, N = 8, BI_PROB = 0, rand_seed = 42,
  name = "_nk_single_rugged_"
))

structure_rugged <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -1.5, dv_name = DV_NAME, fix = TRUE)
    ),
    coCovars = list(),
    varCovars = list(),
    coDyadCovars = list(
      list(
        effect = "XWX", parameter = 1.5, dv_name = DV_NAME, fix = TRUE,
        nodeSet = c("COMPONENTS", "COMPONENTS"),
        interaction1 = "self$component_1_coDyadCovar",
        x = epistasis_mat
      )
    ),
    varDyadCovars = list()
  )
)

env_rugged$search_rsiena(
  structure_model = structure_rugged,
  iterations_per_actor = steps_per_actor,
  run_seed = 12345,
  verbose = TRUE
)
```

The agent searches the rugged landscape for `r steps_per_actor` steps.
With strong negative density and strong XWX, the actor faces a classic
exploration--exploitation dilemma: adopting components is costly unless
they form complementary bundles.

```{r single-rugged-summary}
env_rugged$search_rsiena_model_summary()
```

```{r single-rugged-snapshots, fig.height=5, fig.width=11}
snapshot_ids <- seq(1, env_rugged$M * steps_per_actor)
env_rugged$plot_snapshots(snapshot_ids)
```

```{r single-rugged-k4, fig.height=10, fig.width=10}
env_rugged$plot_degree_4panel(loess_span = 0.5)
```

```{r single-rugged-utility, fig.height=10, fig.width=9}
env_rugged$plot_utility_contributions(loess_span = 0.5, use_thetas = TRUE)
```

## 1.3 Smooth Landscape (Less Negative Density)

Reducing the magnitude of the density penalty makes the landscape
smoother. The agent can explore more freely, and the search trajectory
is less constrained by local optima.

```{r single-smooth-setup}
env_smooth <- SaomNkRSienaBiEnv$new(list(
  M = 2, N = 8, BI_PROB = 0, rand_seed = 42,
  name = "_nk_single_smooth_"
))

structure_smooth <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -0.3, dv_name = DV_NAME, fix = TRUE)
    ),
    coCovars = list(),
    varCovars = list(),
    coDyadCovars = list(
      list(
        effect = "XWX", parameter = 1.5, dv_name = DV_NAME, fix = TRUE,
        nodeSet = c("COMPONENTS", "COMPONENTS"),
        interaction1 = "self$component_1_coDyadCovar",
        x = epistasis_mat
      )
    ),
    varDyadCovars = list()
  )
)

env_smooth$search_rsiena(
  structure_model = structure_smooth,
  iterations_per_actor = steps_per_actor,
  run_seed = 12345,
  verbose = TRUE
)
```

```{r single-smooth-k4, fig.height=10, fig.width=10}
env_smooth$plot_degree_4panel(loess_span = 0.5)
```

```{r single-smooth-utility, fig.height=10, fig.width=9}
env_smooth$plot_utility_contributions(loess_span = 0.5, use_thetas = TRUE)
```

## 1.4 Comparing Ruggedness Levels

The contrast between the two density conditions reproduces the core
Levinthal (1997) finding: more rugged landscapes (stronger negative
density) constrain search and increase the likelihood of premature
convergence to local optima, while smoother landscapes (weaker negative
density) permit broader exploration and higher attained fitness.

# 2. Multi-Agent Search: Endogenous Sociality and Epistasis

The classical NK model treats agents as isolated searchers on a fixed
landscape. SaoMNK extends this by embedding multiple actors ($M > 1$) in
the same component space. When actors share components, two endogenous
projection networks emerge:

-   $K_{AA}$ (Actor sociality): actors become indirectly connected
    through shared components, creating an emergent social structure.
-   $K_{CC}$ (Component epistasis): components become indirectly
    connected through shared actors, endogenously determining the
    effective $K$.

## 2.1 Baseline: Endogenous Effects Off

First, we run a multi-agent simulation with all structural (endogenous)
effects set to zero. This isolates the XWX epistasis channel and serves
as the NK baseline for comparison.

```{r multi-baseline-setup}
env_base <- SaomNkRSienaBiEnv$new(list(
  M = 6, N = 8, BI_PROB = 0, rand_seed = 99,
  name = "_nk_multi_baseline_"
))

structure_base <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -1, dv_name = DV_NAME, fix = TRUE),
      list(effect = "inPop",   parameter = 0, dv_name = DV_NAME, fix = TRUE),
      list(effect = "outAct",  parameter = 0, dv_name = DV_NAME, fix = TRUE)
    ),
    coCovars = list(),
    varCovars = list(),
    coDyadCovars = list(
      list(
        effect = "XWX", parameter = 1, dv_name = DV_NAME, fix = TRUE,
        nodeSet = c("COMPONENTS", "COMPONENTS"),
        interaction1 = "self$component_1_coDyadCovar",
        x = epistasis_mat
      )
    ),
    varDyadCovars = list()
  )
)

env_base$search_rsiena(
  structure_model = structure_base,
  iterations_per_actor = steps_per_actor,
  run_seed = 12345,
  verbose = TRUE
)
```

```{r multi-baseline-summary}
env_base$search_rsiena_model_summary()
```

```{r multi-baseline-snapshots, fig.height=5, fig.width=11}
env_base$plot_snapshots(seq(1, env_base$M * steps_per_actor))
```

```{r multi-baseline-k4, fig.height=10, fig.width=10}
env_base$plot_degree_4panel(loess_span = 0.3)
```

With endogenous effects zeroed out, the $K_{AA}$ and $K_{CC}$
projections evolve as byproducts of independent actor search---there is
no feedback from the social or epistasis structure back into the search
process. This is the multi-agent NK analogue.

## 2.2 Endogenous Effects Activated

Now we activate the endogenous structural effects. The `inPop` effect
rewards components that are popular (adopted by many actors), and the
`outAct` effect rewards actors who adopt many components. These create
feedback loops: actors who share popular components become socially
connected ($K_{AA}$ increases), and components adopted by the same
actors become epistatically linked ($K_{CC}$ is endogenously amplified).

```{r multi-endog-setup}
env_endog <- SaomNkRSienaBiEnv$new(list(
  M = 6, N = 8, BI_PROB = 0, rand_seed = 99,
  name = "_nk_multi_endogenous_"
))

structure_endog <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -1, dv_name = DV_NAME, fix = TRUE),
      list(effect = "inPop",   parameter = 0.3, dv_name = DV_NAME, fix = TRUE),
      list(effect = "outAct",  parameter = 0.2, dv_name = DV_NAME, fix = TRUE)
    ),
    coCovars = list(),
    varCovars = list(),
    coDyadCovars = list(
      list(
        effect = "XWX", parameter = 1, dv_name = DV_NAME, fix = TRUE,
        nodeSet = c("COMPONENTS", "COMPONENTS"),
        interaction1 = "self$component_1_coDyadCovar",
        x = epistasis_mat
      )
    ),
    varDyadCovars = list()
  )
)

env_endog$search_rsiena(
  structure_model = structure_endog,
  iterations_per_actor = steps_per_actor,
  run_seed = 12345,
  verbose = TRUE
)
```

```{r multi-endog-summary}
env_endog$search_rsiena_model_summary()
```

```{r multi-endog-snapshots, fig.height=5, fig.width=11}
env_endog$plot_snapshots(seq(1, env_endog$M * steps_per_actor))
```

```{r multi-endog-k4, fig.height=10, fig.width=10}
env_endog$plot_degree_4panel(loess_span = 0.3)
```

```{r multi-endog-utility, fig.height=10, fig.width=9}
env_endog$plot_utility_contributions(loess_span = 0.35, use_thetas = TRUE)
```

## 2.3 Comparing Baseline and Endogenous Dynamics

The comparison reveals the key insight: when endogenous effects are
activated, $K_{AA}$ (sociality) and $K_{CC}$ (epistasis) are no longer
mere byproducts of independent search but become coupled drivers of the
search process. The `inPop` effect creates preferential attachment to
popular components, generating convergent social structure. The `outAct`
effect rewards broader scope, which in turn amplifies the component
epistasis projection. These feedback loops produce dynamics that the
classical NK model cannot capture: emergent coordination, herding, and
endogenous landscape deformation.

# 3. Parameter Mapping: NK to SaoMNK

The following table provides a systematic mapping between the classical
NK model and the SaoMNK framework:

| NK Concept | NK Parameter | SaoMNK Parameter | Notes |
|:-----------------|:-----------------|:-----------------|:-----------------|
| Components | $N$ | `N` | Dimension of the binary string / component space |
| Epistasis | $K$ | `influence_matrix` + `XWX` | Block-diagonal matrix specifies which components interact; `XWX` parameter scales the strength |
| Landscape ruggedness | $K/N$ ratio | `density` parameter | Negative density increases ruggedness (cost of exploration); more negative = more local optima |
| Fitness | Random table lookup | SAOM objective function | Fitness is the actor's total utility from network statistics |
| Search rule | Local (1-mutant) | SAOM ministep | Each ministep toggles one tie (add or drop one component) |
| Long-jump search | $> 1$ bit flip | `outAct` parameter | Positive `outAct` rewards broader scope, enabling wider exploration |
| Single agent | Fixed | `M = 2` | Recovers classical NK when $M = 1$ and endogenous effects are zero |
| Multi-agent | Not in NK | `M > 1` | Enables endogenous sociality ($K_{AA}$) and endogenous epistasis ($K_{CC}$) |
| Social structure | Not in NK | `inPop` effect | Preferential attachment to popular components creates emergent $K_{AA}$ |
| Agent heterogeneity | Not in NK | `egoX`, strategy covariates | Actor-level covariates differentiate search strategies |

# 4. Key Insight: NK as a Special Case of SaoMNK

The validation demonstrates that the classical NK model is a special
case of SaoMNK, obtained by:

1.  Setting $M = 1$ (minimal agent),
2.  Zeroing all endogenous structural effects (`inPop = 0`,
    `outAct = 0`),
3.  Specifying the epistasis structure via a block-diagonal `XWX`
    matrix.

Under these restrictions, SaoMNK reproduces the canonical NK dynamics:

-   **Rugged landscapes** emerge from the interaction of negative
    density (cost of scope) and positive XWX (benefit of
    complementarity).
-   **Local optima trapping** occurs when the agent reaches a
    configuration where no single-component change improves utility.
-   **Ruggedness scales with density**: more negative density creates
    more local optima, consistent with Levinthal (1997).

When the restrictions are relaxed ($M > 1$, endogenous effects
$\neq 0$), SaoMNK produces dynamics that go beyond the NK framework:

-   **Endogenous** $K_{CC}$: The effective epistasis ($K$) is no longer
    a fixed parameter but emerges from the coupled actor--component
    process.
-   **Endogenous** $K_{AA}$: A social structure among actors emerges
    without being directly modeled, driven by overlapping component
    affiliations.
-   **Landscape co-evolution**: The fitness landscape is deformed by the
    collective search behavior of all actors, creating path dependencies
    and coordination effects absent from the classical NK model.

This validation establishes SaoMNK as a proper generalization of the NK
framework, grounding its novel contributions in the well-understood
dynamics of the canonical model.

# References

-   Kauffman, S. A., & Levin, S. (1987). Towards a general theory of
    adaptive walks on rugged landscapes. *Journal of Theoretical
    Biology*, 128(1), 11--45.
-   Kauffman, S. A., & Weinberger, E. D. (1989). The NK model of rugged
    fitness landscapes and its application to maturation of the immune
    response. *Journal of Theoretical Biology*, 141(2), 211--245.
-   Levinthal, D. A. (1997). Adaptation on rugged landscapes.
    *Management Science*, 43(7), 934--950.
-   Snijders, T. A. B. (1996). Stochastic actor-oriented models for
    network change. *Journal of Mathematical Sociology*, 21(1--2),
    149--172.

```{r session-info}
sessionInfo()
```
