---
title: "Empirical Calibration: SAOM Parameters to SaoMNK Simulations"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Empirical Calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE, purl=FALSE}
## This vignette is a design document, not a worked example: its code
## operates on empirical panel data the package does not ship, and part of
## the interface it sketches is not implemented yet. eval = FALSE below stops
## the chunks running at build time; every chunk ALSO carries purl = FALSE in
## its own header, which is what stops `R CMD check`'s vignette-code step
## tangling and sourcing them. purl cannot be set here: in tangle mode knitr
## never evaluates this chunk, so a global purl = FALSE is silently ignored.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  eval = FALSE,
  purl = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

> **How to read this vignette.** None of the code below is executed when the
> vignette is built, and the `#>` blocks are illustrative rather than captured
> output. Two groups of functions appear. `saom_to_saomnk()`,
> `empirical_to_saomnk_env()` and `run_calibrated_counterfactual()` are
> implemented and exported; see `?saom_to_saomnk` for the current signatures.
> `saom_to_saomnk_manual()`, `run_all_counterfactuals()`,
> `compute_empirical_k()`, `run_calibrated_waves()`,
> `plot_trajectory_comparison()`, `validate_calibration()`,
> `calibration_gof()` and `saomnk_app()` are **not implemented**; the sections
> that use them describe the intended interface and will not run.

# 1. Introduction

Simulation models are most useful when they are grounded in empirical
reality.  The SaoMNK framework bridges two traditions---empirical network
estimation via Stochastic Actor-Oriented Models (SAOMs) and computational
fitness landscape simulation via NK models---and the `searchnet` package
provides a calibration pipeline that connects them.

The workflow proceeds in three stages:

1. **Empirical estimation**: Fit a SAOM to observed network panel data
   using `RSiena` (Snijders, 2001; Ripley et al., 2023).
2. **Parameter bridge**: Convert the fitted $\hat{\theta}$ vector to
   calibrated SaoMNK model effects via `saom_to_saomnk()`.
3. **Counterfactual simulation**: Run calibrated simulations to explore
   "what if" scenarios grounded in the estimated parameters.

This approach enables researchers to move beyond illustrative simulations
with arbitrary parameters toward validated counterfactual analysis.
Applications include supply chain network restructuring, airline alliance
formation, technology platform competition, and investor search behavior.


# 2. The Parameter Bridge

## 2.1 Conceptual foundation

Both RSiena and `searchnet` specify an actor-oriented model in which
actor $i$ evaluates potential network changes via a conditional logit
utility:

$$U_i(x) = \sum_k \theta_k \, s_{ik}(x)$$

where $s_{ik}(x)$ is the network statistic for effect $k$ evaluated at
the current network state $x$.  The shared functional form means that
parameter estimates from RSiena can be mapped directly to SaoMNK's
structure model specification.

## 2.2 Mapping table

The `saom_to_saomnk()` function maps RSiena effect names to their
SaoMNK equivalents and applies an optional scale factor $\alpha$:

| SAOM Effect   | SaoMNK Effect   | Transformation               | Strategic Interpretation       |
|---------------|-----------------|------------------------------|-------------------------------|
| `density`     | `density`       | $\beta = \theta \cdot \alpha$ | Cost of scope expansion        |
| `inPopSqrt`   | `inPop`         | $\beta = \theta \cdot \alpha$ | Preferential attachment        |
| `gwespFF`     | `transTriads`   | $\beta = \theta \cdot \alpha$ | Triadic closure / clustering   |
| `simX`        | `egoX`          | $\beta = \theta \cdot \alpha$ | Attribute-based homophily      |
| `cycle3`      | `cycle4`        | $\beta = \theta \cdot \alpha$ | Bipartite cycle analog         |

The scale factor $\alpha$ (default 1.0) allows the researcher to
accommodate differences in network size or observation granularity
between the empirical sample and the simulation environment.

## 2.3 Using `saom_to_saomnk()`

```{r bridge-basic, purl=FALSE}
library(searchnet)

## Suppose saom_fit is a fitted RSiena model object
bridge <- saom_to_saomnk(saom_fit, scale_factor = 1.0)

## Inspect the mapping
print(bridge$mapping_table)
#>   saom_effect saomnk_effect  theta   beta
#> 1    density       density  -0.50  -0.50
#> 2  inPopSqrt         inPop   0.30   0.30
#> 3    gwespFF    transTriads   0.45   0.45
#> 4       simX          egoX   0.20   0.20

## Access the converted parameter vector
bridge$saomnk_params
```

The function validates that all RSiena effects have known SaoMNK
counterparts and issues warnings for effects that are dropped during
conversion.

## 2.4 Scale factor considerations

The scale factor serves two purposes:

- **Network size adjustment**: When the empirical network has $M = 50$
  actors and the simulation uses $M = 10$, effect magnitudes may need
  rescaling to preserve comparable dynamics.
- **Temporal granularity**: RSiena estimates parameters over observation
  waves that may span years; SaoMNK steps represent finer-grained
  micro-steps.

A reasonable starting point is $\alpha = 1.0$ (no rescaling), followed
by sensitivity analysis across $\alpha \in \{0.5, 0.75, 1.0, 1.25, 1.5\}$.


# 3. Environment from Data

## 3.1 The `empirical_to_saomnk_env()` function

Real-world network data typically arrives as a one-mode adjacency matrix
(firm-to-firm ties) or as an edgelist with node attributes.  The
`empirical_to_saomnk_env()` function converts this to the bipartite
actor-component incidence matrix required by `searchnet`.

```{r env-from-data, purl=FALSE}
## mi_data: a list with adjacency matrices and node attributes per wave
## (e.g., from the airline ORM study or supply chain panel data)
env_data <- empirical_to_saomnk_env(mi_data, wave = 5)

## Inspect the resulting environment
cat("Actors:", env_data$M, "\n")
cat("Components:", env_data$N, "\n")
cat("Initial density:", mean(env_data$incidence_matrix), "\n")
```

## 3.2 MI data structure

The input `mi_data` object is a list containing:

- `$networks`: A list of adjacency or incidence matrices, one per wave.
- `$attributes`: A data frame of node-level covariates (e.g., firm size,
  R&D intensity, geographic region).
- `$wave_info`: Metadata including wave labels and time intervals.

For one-mode networks (firm-to-firm alliances, supply chain links), the
function performs bipartite projection using a shared-partner decomposition:
actors are the original nodes, and components are extracted as latent
activity dimensions via spectral clustering.

```{r mi-structure, purl=FALSE}
## Example: constructing mi_data from raw alliance data
mi_data <- list(
  networks = list(
    wave1 = alliance_matrix_2015,
    wave2 = alliance_matrix_2017,
    wave3 = alliance_matrix_2019,
    wave4 = alliance_matrix_2021,
    wave5 = alliance_matrix_2023
  ),
  attributes = firm_attributes,
  wave_info = data.frame(
    wave = 1:5,
    year = seq(2015, 2023, by = 2)
  )
)
```

## 3.3 Bipartite conversion for one-mode data

When the empirical data is a one-mode network (e.g., airline code-share
agreements), the conversion involves:

1. **Spectral decomposition** of the one-mode adjacency matrix to
   identify latent component dimensions.
2. **Threshold assignment** to create binary actor-component affiliations.
3. **Validation** that the reconstructed bipartite network preserves the
   degree distribution and clustering of the original one-mode projection.

```{r bipartite-convert, purl=FALSE}
## Convert one-mode airline alliance data to bipartite
env_data <- empirical_to_saomnk_env(
  mi_data, wave = 5,
  mode = "one-mode",
  n_components = 8,
  threshold = "median"
)

## Verify reconstruction quality
cat("Degree correlation:", env_data$diagnostics$degree_cor, "\n")
cat("Clustering preservation:", env_data$diagnostics$clustering_cor, "\n")
```


# 4. Worked Example

This section walks through a complete calibration using example SAOM
parameter estimates from a hypothetical airline competitive dynamics study.

## 4.1 Set up the parameter bridge

```{r worked-bridge, purl=FALSE}
## Example: manually specify SAOM estimates
## (in practice these come from a fitted RSiena model)
example_theta <- c(
  density    = -0.50,
  inPopSqrt  =  0.30,
  gwespFF    =  0.45,
  simX       =  0.20,
  cycle3     = -0.10
)

## Create the bridge with default scale factor
bridge <- saom_to_saomnk_manual(example_theta, scale_factor = 1.0)
print(bridge$mapping_table)
```

## 4.2 Create the simulation environment

```{r worked-env, purl=FALSE}
## Create an environment matching the empirical network dimensions
env <- saomnk_env(
  M = 8,                  # 8 airlines
  N = 12,                 # 12 route components
  K_AC = 4,               # initial scope
  epistasis = "modular",  # 3 blocks of 4 components
  seed = 42
)

## Apply calibrated parameters
model <- saomnk_model(
  env,
  effects = bridge$saomnk_params,
  steps = 100
)
```

## 4.3 Run the calibrated baseline

```{r worked-baseline, purl=FALSE}
## Run calibrated simulation
baseline <- saomnk_run(model)

## Inspect the coupled degree trajectory
summary(baseline)
#> SaoMNK simulation: 8 actors, 12 components, 100 steps
#> Final K_AC: 3.75  K_CA: 2.67  K_AA: 1.50  K_CC: 3.00
#> Mean fitness: 0.623 (sd = 0.082)

## Plot the {K}-4 panel
plot(baseline, type = "k4")
```

## 4.4 Compare to observed data

```{r worked-validate, purl=FALSE}
## Compare simulated degree distributions to empirical wave 5
obs_degrees <- colSums(mi_data$networks$wave5)
sim_degrees <- colSums(baseline$final_incidence)

## Kolmogorov-Smirnov test
ks_result <- ks.test(obs_degrees, sim_degrees)
cat(sprintf("KS statistic: %.3f, p-value: %.3f\n",
            ks_result$statistic, ks_result$p.value))
```


# 5. Counterfactual Scenarios

The calibration bridge enables structured counterfactual analysis by
modifying specific parameters in the calibrated vector and comparing
outcomes.  The `run_calibrated_counterfactual()` function automates this
workflow.

## 5.1 Predefined scenarios

The package includes six predefined scenarios, each motivated by a
substantive research question:

| # | Scenario               | Modification                   | Research Question                                     |
|---|------------------------|-------------------------------|-------------------------------------------------------|
| 1 | Density shock          | `density` $\times$ 0.5        | What if market entry barriers double?                 |
| 2 | Double closure         | `transTriads` $\times$ 2.0    | What if alliance clustering intensifies?              |
| 3 | Popularity concentration | `inPop` $\times$ 2.0        | What if preferential attachment strengthens?           |
| 4 | Scope expansion        | `density` $\times$ 1.5        | What if firms diversify more broadly?                 |
| 5 | Epistasis rewiring     | Modify $\mathbf{E}$ matrix    | What if modular architectures become integral?        |
| 6 | Attribute homophily    | `egoX` $\times$ 2.0           | What if firms increasingly sort by similarity?        |

## 5.2 Running a counterfactual

```{r counterfactual-run, purl=FALSE}
## Scenario 2: Double the transitive closure parameter
result <- run_calibrated_counterfactual(
  env_data, bridge,
  scenario = list(
    name = "double_closure",
    modify = list(transTriads = 2.0)
  ),
  iterations = 50
)

## Examine the comparison
print(result$comparison)
#>         metric   baseline counterfactual   delta  pct_change
#> 1         K_AC      3.750          4.125   0.375       10.0%
#> 2         K_CA      2.667          3.000   0.333       12.5%
#> 3         K_AA      1.500          2.250   0.750       50.0%
#> 4         K_CC      3.000          3.500   0.500       16.7%
#> 5 mean_fitness      0.623          0.591  -0.032       -5.1%
```

## 5.3 Interpreting results

The comparison structure reports changes across all four $\{K\}$
dimensions, mean fitness, and basin geometry metrics.  In the double
closure example above:

- **$K_{AA}$ increases by 50%**: Stronger triadic closure creates denser
  actor-actor co-affiliation, meaning firms increasingly share the same
  components as their alliance partners.
- **Mean fitness declines by 5.1%**: The increased clustering traps firms
  in suboptimal regions of the landscape, consistent with the
  competence-trap literature (Levitt & March, 1988).

## 5.4 Running all predefined scenarios

```{r counterfactual-all, purl=FALSE}
## Run all six predefined scenarios
all_results <- run_all_counterfactuals(env_data, bridge, iterations = 50)

## Summary table
print(all_results$summary)

## Visualization: {K}-4 comparison across scenarios
plot(all_results, type = "k4_comparison")
```


# 6. Validation

Calibrated simulations require validation against the empirical data that
generated the parameter estimates.  The package provides three validation
approaches.

## 6.1 Trajectory comparison

Compare the simulated coupled degree trajectories to observed wave-level
statistics from the empirical panel data.

```{r validation-trajectory, purl=FALSE}
## Extract observed {K} statistics per wave
obs_k <- compute_empirical_k(mi_data)

## Run calibrated simulation with wave-matching
sim_k <- run_calibrated_waves(env_data, bridge, waves = 1:5)

## Plot observed vs. simulated trajectories
plot_trajectory_comparison(obs_k, sim_k)
```

## 6.2 Distribution matching

Compare the full distribution of actor-level outcomes (fitness, scope,
sociality) between simulation and data.

```{r validation-distribution, purl=FALSE}
## QQ-plot of degree distributions
validation <- validate_calibration(baseline, mi_data, wave = 5)
plot(validation, type = "qq")
cat(sprintf("Overall fit (RMSE): %.4f\n", validation$rmse))
```

## 6.3 Goodness-of-fit statistics

Following RSiena's goodness-of-fit framework (Lospinoso et al., 2011),
the package computes Mahalanobis distance between simulated and observed
network statistics across multiple dimensions.

```{r validation-gof, purl=FALSE}
## Goodness-of-fit assessment
gof <- calibration_gof(baseline, mi_data, wave = 5,
                       statistics = c("degree", "triad.census",
                                      "geodesic.distribution"))
print(gof)
plot(gof)
```


# 7. Integration with the GUI

The `searchnet` Shiny application (accessible via `saomnk_app()`) includes
an interactive calibration page at the `/calibrate` route.

## 7.1 Features

The calibration interface provides:

- **Parameter upload**: Import fitted RSiena objects (`.RData` or `.rds`
  files) or enter parameter estimates manually.
- **Interactive mapping table**: Review and adjust the SAOM-to-SaoMNK
  effect mapping with real-time preview.
- **Scale factor slider**: Adjust $\alpha$ and immediately see its effect
  on the simulated $\{K\}$-4 trajectory.
- **Scenario builder**: Select predefined counterfactual scenarios or
  define custom parameter modifications.
- **Comparison dashboard**: Side-by-side visualization of baseline and
  counterfactual outcomes with statistical tests.

## 7.2 Launching the calibration page

```{r gui-launch, purl=FALSE}
## Launch the full application
saomnk_app()

## Or navigate directly to the calibration page
saomnk_app(page = "calibrate")
```

## 7.3 Supply chain example

For supply chain networks, the calibration workflow is identical but the
interpretation changes.  Actors are firms in the supply chain, and
components are product categories or capabilities.  The $\{K\}$-4 panel
then tracks:

- $K_{AC}$: Firm **diversification** across product categories.
- $K_{CA}$: Product category **concentration** among suppliers.
- $K_{AA}$: Supplier **clustering** via shared product portfolios.
- $K_{CC}$: Product **interdependence** via shared supplier bases.

Counterfactual scenarios such as "What if a key supplier exits?" or "What
if two product categories merge?" can be expressed as parameter
modifications to the calibrated bridge.


# 8. References

- Kauffman, S. A. (1993). *The Origins of Order*. Oxford University Press.
- Levinthal, D. A. (1997). Adaptation on rugged landscapes. *Management
  Science*, 43(7), 934--950.
- Levitt, B. & March, J. G. (1988). Organizational learning. *Annual
  Review of Sociology*, 14, 319--340.
- Lospinoso, J. A., Schweinberger, M., Snijders, T. A. B., & Ripley,
  R. M. (2011). Assessing and accounting for time heterogeneity in
  stochastic actor oriented models. *Advances in Data Analysis and
  Classification*, 5, 147--176.
- Ripley, R. M., Snijders, T. A. B., Boda, Z., Voros, A., & Preciado,
  P. (2023). *Manual for RSiena*. University of Oxford.
- Snijders, T. A. B. (2001). The statistical evaluation of social
  network dynamics. *Sociological Methodology*, 31(1), 361--395.
- Teece, D. J., Pisano, G., & Shuen, A. (1997). Dynamic capabilities
  and strategic management. *Strategic Management Journal*, 18(7),
  509--533.
