---
title: "Empirical Calibration: SAOM Parameters to SAOM-NK Simulations"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Empirical Calibration: SAOM Parameters to SAOM-NK Simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
## Chunks run at build time on a small synthetic panel built in section 3.
## Chunks that need a fitted RSiena model, or take minutes, carry
## eval = FALSE and purl = FALSE in their own headers and say so in the text.
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

> **How to read this vignette.** The code runs end to end on a small
> synthetic panel (15 firms, 5 industry categories, 3 waves) that the vignette
> builds itself; no empirical data ships with the package. The functions used
> are all implemented and exported: `saom_to_saomnk()`,
> `empirical_to_saomnk_env()`, `run_calibrated_counterfactual()`,
> `searchnet_k_readings()`, `gof_battery()`,
> `searchnet_recovery()` and `searchnet_chain_calibrate()`. The last three need
> a fitted RSiena model or several minutes, so their chunks are shown with
> `eval = FALSE`. Interfaces that are planned but not implemented are listed
> in one place, section 7.

# 1. Introduction

Simulation models are most useful when they are grounded in empirical
reality.  The SAOM-NK framework bridges two traditions---empirical network
estimation via Stochastic Actor-Oriented Models (SAOMs) and computational
fitness landscape simulation via NK models---and the `searchnet` package
provides a calibration pipeline that connects them.

The workflow proceeds in three stages:

1. **Empirical estimation**: Fit a SAOM to observed network panel data
   using `RSiena` (Snijders, 2001; Ripley et al., 2023).
2. **Parameter bridge**: Convert the fitted $\hat{\theta}$ vector to
   calibrated SAOM-NK model effects via `saom_to_saomnk()`.
3. **Counterfactual simulation**: Run calibrated simulations to explore
   "what if" scenarios grounded in the estimated parameters.

This approach enables researchers to move beyond illustrative simulations
with arbitrary parameters toward validated counterfactual analysis.
Applications include supply chain network restructuring, airline alliance
formation, technology platform competition, and investor search behavior.


# 2. The Parameter Bridge

## 2.1 Conceptual foundation

Both RSiena and `searchnet` specify an actor-oriented model in which
actor $i$ evaluates potential network changes via a conditional logit
utility:

$$U_i(x) = \sum_k \theta_k \, s_{ik}(x)$$

where $s_{ik}(x)$ is the network statistic for effect $k$ evaluated at
the current network state $x$.  The shared functional form means that
parameter estimates from RSiena can be mapped directly to SAOM-NK's
structure model specification.

## 2.2 Mapping table

The `saom_to_saomnk()` function maps RSiena effect names to their
SAOM-NK equivalents and applies an optional scale factor $\alpha$. Each
entry has a status: *exact* (the SAOM-NK effect is the estimated effect),
*approximate* (a different statistic stands in; refused unless
`strict = FALSE`), or *unavailable* (no counterpart for a bipartite
dependent variable; always refused). A selection:

| SAOM Effect   | SAOM-NK Effect  | Status       | Transformation                 | Strategic Interpretation       |
|---------------|-----------------|--------------|--------------------------------|--------------------------------|
| `density`     | `density`       | exact        | $\beta = \theta \cdot \alpha$  | Cost of scope expansion        |
| `inPop`       | `inPop`         | exact        | $\beta = \theta \cdot \alpha$  | Preferential attachment        |
| `inPopSqrt`   | `inPopSqrt`     | exact        | $\beta = \theta \cdot \alpha$  | Preferential attachment (sqrt) |
| `outAct`      | `outAct`        | exact        | $\beta = \theta \cdot \alpha$  | Scope expansion                |
| `cycle3`      | `cycle4`        | approximate  | $\beta = \theta \cdot \alpha$  | Bipartite cycle analog         |
| `recip`       | `cycle4`        | approximate  | $\beta = 0.5\,\theta \cdot \alpha$ | Bipartite cycle analog     |
| `simX`        | `egoX`          | approximate  | $\beta = \theta \cdot \alpha$  | Attribute-based homophily      |
| `gwespFF`     | `transTriads`   | unavailable  | (refused)                      | Triadic closure / clustering   |

`?saom_to_saomnk` documents the full crosswalk.

The scale factor $\alpha$ (default 1.0) allows the researcher to
accommodate differences in network size or observation granularity
between the empirical sample and the simulation environment.

## 2.3 Using `saom_to_saomnk()`

```{r bridge-basic}
library(searchnet)

## A named vector of estimates stands in for a fitted RSiena model here;
## a sienaFit object is accepted in the same position.
example_theta <- c(density = -1.2, inPop = 0.3, outAct = -0.2)
bridge <- saom_to_saomnk(example_theta, scale_factor = 1.0, verbose = FALSE)

## Inspect the mapping
bridge$mapping_table[, c("saom_effect", "saom_theta", "saomnk_effect",
                         "saomnk_parameter", "mapping_status")]

## The converted effect specifications
str(bridge$effects[[1]])
```

The function refuses approximate mappings unless `strict = FALSE`, refuses
unavailable ones regardless, and lists effects it cannot map in
`$unmapped`:

```{r bridge-strict}
try(saom_to_saomnk(c(density = -1.2, cycle3 = -0.1), verbose = FALSE))
```

## 2.4 Scale factor considerations

The scale factor serves two purposes:

- **Network size adjustment**: When the empirical network has $M = 50$
  actors and the simulation uses $M = 10$, effect magnitudes may need
  rescaling to preserve comparable dynamics.
- **Temporal granularity**: RSiena estimates parameters over observation
  waves that may span years; SAOM-NK steps represent finer-grained
  micro-steps.

A reasonable starting point is $\alpha = 1.0$ (no rescaling), followed
by sensitivity analysis across $\alpha \in \{0.5, 0.75, 1.0, 1.25, 1.5\}$.


# 3. Environment from Data

## 3.1 MI data structure

`empirical_to_saomnk_env()` reads a multiply imputed panel of one-mode
firm-to-firm networks. The input `mi_data` object is a list in which
`mi_data$imputations[[imp]][[wave]]` holds:

- `$network`: an $M \times M$ firm-to-firm adjacency matrix.
- `$covariates`: a named list with at least `$sic`, each firm's industry
  (SIC) code; optional covariates such as `is_seed`, `assets` and
  `revenue` are carried into the actor attributes.

The synthetic panel used throughout this vignette:

```{r mi-structure}
set.seed(7)
M   <- 15
sic <- rep(c("2011", "2834", "3571", "3674", "4512"), each = 3)
make_wave <- function(p) {
  net <- matrix(rbinom(M * M, 1, p), M, M)
  diag(net) <- 0
  list(network = net, covariates = list(sic = sic))
}
mi_data <- list(imputations = list(
  list(make_wave(0.06), make_wave(0.08), make_wave(0.10))
))
```

## 3.2 The `empirical_to_saomnk_env()` function

The function converts one wave to the bipartite actor-component incidence
matrix required by `searchnet`. Components are the SIC categories held by
at least `min_firms_per_sic` firms. Firm $i$ is tied to its own category
and to the category of every firm it has a tie to. The influence matrix
$W$ is the normalized co-occurrence of categories within firms' rows.

```{r env-from-data}
bridge_env <- empirical_to_saomnk_env(mi_data, wave = 3)

## Inspect the resulting environment
dim(bridge_env$bi_matrix)
mean(bridge_env$bi_matrix)          # initial density
round(bridge_env$W, 2)              # influence matrix from co-occurrence
```


# 4. Worked Example

This section walks through a complete calibration on the synthetic panel,
using the parameter bridge from section 2.3.

## 4.1 Observed {K} readings

`searchnet_k_readings()` reads the four $\{K\}$ dimensions off each
observed wave, by the same definitions the simulation engine uses, so the
observed panel and a simulated run can be compared directly.

```{r worked-observed}
obs_waves <- lapply(1:3, function(w) {
  utils::capture.output(e <- empirical_to_saomnk_env(mi_data, wave = w))
  e$bi_matrix
})
obs_k <- searchnet_k_readings(obs_waves)
attr(obs_k, "summary")[, c("period", "density", "mean_K_AC", "mean_K_CA",
                           "mean_K_AA", "mean_K_CC", "tie_jaccard")]
```

## 4.2 Run the calibrated baseline and a counterfactual

`run_calibrated_counterfactual()` runs a matched pair of simulations,
baseline and counterfactual, `n_reps` times from the empirical
environment. The settings below are small so that the vignette builds in
seconds; use more replications and iterations for reported results.

```{r worked-baseline}
result <- run_calibrated_counterfactual(
  bridge_env, bridge,
  scenario   = list(name = "double_popularity", modify = list(inPop = 2.0)),
  iterations = 20,
  n_reps     = 3,
  seed       = 42,
  verbose    = FALSE
)

## Baseline arm, first replication: mean {K} degrees
unlist(result$baseline$k4[c("mean_K_AC", "mean_K_CA", "mean_K_AA", "mean_K_CC")])
```

## 4.3 Compare to observed data

```{r worked-validate}
## Compare simulated degree distributions to the observed final wave
obs_degrees <- colSums(obs_waves[[3]])
sim_degrees <- colSums(result$baseline$env$bipartite_matrix)

## Kolmogorov-Smirnov test
ks_result <- suppressWarnings(ks.test(obs_degrees, sim_degrees))
cat(sprintf("KS statistic: %.3f, p-value: %.3f\n",
            ks_result$statistic, ks_result$p.value))
```

With five components the test has almost no power; it is shown for the
mechanics only.


# 5. Counterfactual Scenarios

The calibration bridge enables structured counterfactual analysis by
modifying specific parameters in the calibrated vector and comparing
outcomes.  The `run_calibrated_counterfactual()` function automates this
workflow.

## 5.1 Scenarios

A scenario is a list with a `name` and a `modify` element: a named
multiplier on each SAOM-NK effect it changes. Six examples:

| # | Scenario               | Modification                   | Research Question                                     |
|---|------------------------|--------------------------------|-------------------------------------------------------|
| 1 | `double_closure`       | `cycle4` $\times$ 2.0          | What if closure tendency were twice as strong?        |
| 2 | `remove_homophily`     | `egoX` $\times$ 0              | What if firms ignored industry similarity?            |
| 3 | `double_popularity`    | `inPop` $\times$ 2.0           | What if preferential attachment were stronger?        |
| 4 | `density_shock`        | `density` $\times$ 1.5         | What if tie formation became much costlier?           |
| 5 | `remove_rivalry`       | `cycle4` $\times$ 0            | What if competitive avoidance disappeared?            |
| 6 | `epistasis_boost`      | `XWX` $\times$ 2.0             | What if component interdependencies doubled?          |

A scenario whose effect is not in the calibrated bridge is an error, not a
silent no-op. The bridge in section 2.3 carries `density`, `inPop` and
`outAct`, so scenarios 3 and 4 apply to it.

## 5.2 Running a counterfactual

The run in section 4.2 doubled the `inPop` parameter. Its comparison
reports the mean change in each $\{K\}$ dimension over replications:

```{r counterfactual-run}
unlist(result$comparison)
```

## 5.3 Interpreting results

The `summary` element reports, for each $\{K\}$ dimension, the mean delta
over replications with its Monte Carlo standard error, a paired t-test and
a paired quantile interval. Matched seeds fix the initial state, not the
realized path, so a single pair is not a counterfactual; read the mean
delta with its error (see `result$caveat`).

```{r counterfactual-summary}
result$summary[, c("measure", "mean_baseline", "mean_cf", "mean_delta",
                   "mc_se", "p_value")]
```

Per-actor and per-component deltas are in `result$actor_deltas` and
`result$component_deltas` for heterogeneous effects.

## 5.4 Running several scenarios

```{r counterfactual-all}
applicable <- list(
  double_popularity = list(name = "double_popularity", modify = list(inPop = 2.0)),
  density_shock     = list(name = "density_shock",     modify = list(density = 1.5))
)
all_results <- lapply(applicable, function(s)
  run_calibrated_counterfactual(bridge_env, bridge, scenario = s,
                                iterations = 20, n_reps = 3, seed = 42,
                                verbose = FALSE)$summary)

## Summary table: mean delta per scenario and {K} dimension
sapply(all_results, function(s) setNames(s$mean_delta, s$measure))
```

## 5.5 Supply chain interpretation

For supply chain networks, the calibration workflow is identical but the
interpretation changes.  Actors are firms in the supply chain, and
components are product categories or capabilities.  The $\{K\}$-4 panel
then tracks:

- $K_{AC}$: Firm **diversification** across product categories.
- $K_{CA}$: Product category **concentration** among suppliers.
- $K_{AA}$: Supplier **clustering** via shared product portfolios.
- $K_{CC}$: Product **interdependence** via shared supplier bases.

Counterfactual scenarios such as "What if a key supplier exits?" or "What
if two product categories merge?" can be expressed as parameter
modifications to the calibrated bridge.


# 6. Validation

Calibrated simulations require validation against the empirical data that
generated the parameter estimates.  Four checks are implemented.

## 6.1 Trajectory comparison

Compare the simulated $\{K\}$ readings to the observed wave-level
readings from section 4.1.

```{r validation-trajectory}
sim_k <- searchnet_k_readings(result$baseline$env$bipartite_matrix)
rbind(observed_wave3 = attr(obs_k, "summary")[3, c("density", "mean_K_AC",
                                                   "mean_K_CA", "mean_K_AA",
                                                   "mean_K_CC")],
      simulated      = attr(sim_k, "summary")[1, c("density", "mean_K_AC",
                                                   "mean_K_CA", "mean_K_AA",
                                                   "mean_K_CC")])
```

## 6.2 Distribution matching

Compare the full distribution of actor-level scope ($K_{AC}$) between
simulation and data.

```{r validation-distribution}
obs_kac <- obs_k$K_AC[obs_k$level == "actor" & obs_k$period == 3]
sim_kac <- sim_k$K_AC[sim_k$level == "actor"]
qqplot(obs_kac, sim_kac, xlab = "Observed K_AC (wave 3)",
       ylab = "Simulated K_AC (baseline)")
abline(0, 1, lty = 2)
```

## 6.3 Goodness-of-fit statistics

Following RSiena's goodness-of-fit framework (Lospinoso et al., 2011),
`gof_battery()` runs `RSiena::sienaGOF()` on five auxiliary statistics of a
bipartite SAOM fit and returns a verdict for each. It needs a `sienaFit`
estimated with `returnDeps = TRUE`, so the chunk is not run here.

```{r validation-gof, eval=FALSE, purl=FALSE}
## Goodness-of-fit assessment of the empirical SAOM fit
fit <- RSiena::siena07(alg, data = dat, effects = eff, returnDeps = TRUE)
gof <- gof_battery(fit, varName = "dv")
print(gof)
```

## 6.4 Recovery and test calibration

Before trusting a calibrated design, check that it recovers parameters
planted at the estimated values, and that the generative-sufficiency test
holds its nominal size on the model's own chains. Both take minutes, so
the chunk is not run here; `vignette("searchnet-recovery")` shows stored
results.

```{r validation-recovery, eval=FALSE, purl=FALSE}
## Planted-truth recovery at the design's size
rec <- searchnet_recovery(c("density", "inPop", "outAct"),
                          theta_true = c(rate = 4, example_theta),
                          reps = 50, M = 15, N = 5, waves = 3, seed = 1)
summary(rec)

## Leave-one-out size check of the chain-statistics test
chains <- searchnet_chain_from_fit(fit_with_chains, dat)
searchnet_chain_calibrate(chains)
```


# 7. Planned Interfaces

The following are planned and **not implemented**; no exported function
of these names exists:

- `saom_to_saomnk_manual()`: superseded, since `saom_to_saomnk()` already
  accepts a named vector.
- `run_all_counterfactuals()`: run a list of scenarios in one call; section 5.4
  shows the implemented route.
- `compute_empirical_k()`, `run_calibrated_waves()` and
  `plot_trajectory_comparison()`: wave-matched trajectory comparison;
  section 6.1 shows the implemented route with `searchnet_k_readings()`.
- `validate_calibration()` and `calibration_gof()`: a single validation
  object; sections 6.2 and 6.3 show the implemented routes.
- One-mode conversion options for `empirical_to_saomnk_env()` (spectral
  decomposition into latent components, threshold assignment, and
  reconstruction diagnostics).
- `saomnk_app()`: a Shiny calibration page with parameter upload, an
  interactive mapping table, a scale factor slider, a scenario builder and
  a comparison dashboard. No GUI ships with the package.


# 8. References

- Kauffman, S. A. (1993). *The Origins of Order*. Oxford University Press.
- Levinthal, D. A. (1997). Adaptation on rugged landscapes. *Management
  Science*, 43(7), 934--950.
- Levitt, B. & March, J. G. (1988). Organizational learning. *Annual
  Review of Sociology*, 14, 319--340.
- Lospinoso, J. A., Schweinberger, M., Snijders, T. A. B., & Ripley,
  R. M. (2011). Assessing and accounting for time heterogeneity in
  stochastic actor oriented models. *Advances in Data Analysis and
  Classification*, 5, 147--176.
- Ripley, R. M., Snijders, T. A. B., Boda, Z., Voros, A., & Preciado,
  P. (2023). *Manual for RSiena*. University of Oxford.
- Snijders, T. A. B. (2001). The statistical evaluation of social
  network dynamics. *Sociological Methodology*, 31(1), 361--395.
- Teece, D. J., Pisano, G., & Shuen, A. (1997). Dynamic capabilities
  and strategic management. *Strategic Management Journal*, 18(7),
  509--533.
