---
title: "Introduction to searchnet: Stochastic Actor-Oriented MNK Search Simulation"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Introduction to searchnet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# Overview

The `searchnet` package implements a Stochastic Actor-Oriented MNK model for
simulating strategic search in complex environments. It endogenizes the NK
fitness landscape (Kauffman & Levin, 1987; Kauffman & Weinberger, 1989;
Levinthal, 1997) by representing the search environment as a bipartite network
of $M$ actors affiliating with $N$ components. The resulting network data
generation process---using the bipartite network as the dependent variable---
enables dynamic network analysis of the search system as a Stochastic
Actor-Oriented Model (SAOM; Snijders, 1996).

The SaoMNK framework reveals that the conventional $K$ term (complexity,
or degree of epistatic interdependence) is just one part of a coupled process
of four network **degree** terms:

**2-mode** (Bipartite, direct ties):

1. $K_{AC}$ = Actor **scope** (Actor-Component degree)
2. $K_{CA}$ = Component **popularity** (Component-Actor degree)

**1-mode** (Unipartite Projection, indirect ties):

3. $K_{AA}$ = Actor **sociality** (Actor-[Component]-Actor degree)
4. $K_{CC}$ = Component **epistasis** (Component-[Actor]-Component degree),
   the traditional $K$

## Installation

```{r load-saomnk}
if (requireNamespace("searchnet", quietly = TRUE)) {
  library(searchnet)
} else {
  dir_r <- file.path(getwd(), "R")
  if (!file.exists(file.path(dir_r, "utils.R")))
    dir_r <- "D:/Search_networks/SaoMNK/R"
  .saomnk_dir <- dir_r
  .saomnk_dir <- dir_r; source(file.path(dirname(dir_r), "inst", "saomnk-loader.R"))
}

library(Matrix)
```

# 1. Hello World: A Minimal SaoMNK Simulation

## 1.1 Environment Configuration

The environment is configured by specifying:

- `M`: number of actors (firms, agents)
- `N`: number of components (resources, technologies, markets)
- `BI_PROB`: initial density of the bipartite network (0 = empty start)
- `rand_seed`: for reproducibility

```{r env-config}
environment_parameters <- list(
  M = 3,
  N = 6,
  BI_PROB = 0,
  rand_seed = 1234,
  name = "_hello_world_"
)

env <- SaomNkRSienaBiEnv$new(environment_parameters)
cat("Environment created:", env$M, "actors x", env$N, "components\n")
```

## 1.2 Structure Model

The structure model defines the SAOM objective function---the stochastic
actor's utility function for strategic search. Effects are specified as a
nested list:

```{r structure-model}
DV_NAME <- "self$bipartite_rsienaDV"

## Define a modular influence matrix (2 blocks of 3 components)
component_int_mat <- create_block_diag(environment_parameters$N, 2)

structure_model <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = 0, dv_name = DV_NAME, fix = TRUE)
    ),
    coDyadCovars = list(
      list(
        effect = "XWX", parameter = 1, dv_name = DV_NAME, fix = TRUE,
        nodeSet = c("COMPONENTS", "COMPONENTS"),
        interaction1 = "self$component_1_coDyadCovar",
        x = component_int_mat
      )
    )
  )
)
```

## 1.3 Run the Simulation

RSiena is the R interface to SIENA (Simulation Investigation for Empirical
Network Analysis). SaoMNK is formalized as a SAOM with one bipartite network
as the dependent variable:

```{r run-sim}
steps_per_actor <- 5

env$search_rsiena(
  structure_model = structure_model,
  iterations_per_actor = steps_per_actor,
  run_seed = 12345,
  verbose = TRUE
)
```

## 1.4 Model Summary

```{r model-summary}
env$search_rsiena_model_summary()
```

# 2. Visualizing the Search System

## 2.1 Network Snapshots

Snapshots of the bipartite network (DGP) and the social and component epistasis
projections show the evolving multidimensional coupled search environment:

```{r snapshots, fig.height=5, fig.width=11}
## Sample the trajectory evenly rather than plotting every ministep: one
## panel per step is both slow and unreadable at this figure size.
snapshot_ids <- unique(round(
  seq(1, env$M * steps_per_actor, length.out = 12)))
env$plot_snapshots(snapshot_ids)
```

## 2.2 K-4 Panel: Degree Trajectories

The K-4 panel shows the four coupled degree processes over simulated
decision-chain steps:

```{r k4-panel, fig.height=10, fig.width=10}
env$plot_degree_4panel(loess_span = 0.25)
```

## 2.3 Utility Contributions

Decompose actor utility into contributions from each network statistic:

```{r utility-contributions, fig.height=10, fig.width=9}
env$plot_utility_contributions(loess_span = 0.35, use_thetas = FALSE)
```

With $\theta$ weights applied:

```{r utility-contributions-theta, fig.height=10, fig.width=9}
env$plot_utility_contributions(loess_span = 0.35, use_thetas = TRUE)
```

# 3. A Longer Simulation

Increase `iterations_per_actor` for richer dynamics:

```{r longer-sim, fig.height=5, fig.width=11}
env2 <- SaomNkRSienaBiEnv$new(environment_parameters)

env2$search_rsiena(
  structure_model,
  iterations_per_actor = 100,
  run_seed = 12345
)

env2$plot_snapshots(c(1:5, env2$get_step()))
```

```{r longer-sim-plots, fig.height=10, fig.width=10}
env2$plot_degree_4panel(loess_span = 0.3)
env2$plot_utility_contributions(loess_span = 0.35)
```

# References

- Kauffman, S. A., & Levin, S. (1987). Towards a general theory of adaptive
  walks on rugged landscapes. *Journal of Theoretical Biology*, 128(1), 11--45.
- Kauffman, S. A., & Weinberger, E. D. (1989). The NK model of rugged fitness
  landscapes and its application to maturation of the immune response.
  *Journal of Theoretical Biology*, 141(2), 211--245.
- Levinthal, D. A. (1997). Adaptation on rugged landscapes.
  *Management Science*, 43(7), 934--950.
- Snijders, T. A. B. (1996). Stochastic actor-oriented models for network
  change. *Journal of Mathematical Sociology*, 21(1--2), 149--172.

```{r session-info}
sessionInfo()
```
