---
title: "Theoretical Foundations: From NK to SaoMNK"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Theoretical Foundations: From NK to SaoMNK}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

# From NK to SaoMNK

## The NK Model

The NK model (Kauffman & Levin, 1987; Kauffman & Weinberger, 1989)
characterizes fitness landscapes with two parameters:

- $N$: the number of components (loci, decisions, design choices)
- $K$: the degree of epistatic interdependence among components

Higher $K$ increases the ruggedness of the fitness landscape, creating
more local optima and making search more difficult. The NK model has become
a workhorse in strategy and organization theory (Levinthal, 1997; Rivkin, 2000)
for studying adaptation, complexity, and search.

## Limitations of Classical NK

The classical NK model has several limitations that `searchnet` addresses:

1. **Single-agent**: NK models typically feature a single searcher navigating
   the landscape. Multi-agent extensions exist but treat agents independently.
2. **Exogenous landscape**: The fitness landscape is fixed---it does not
   respond to agents' positions or to the structure of affiliation.
3. **No network structure**: There is no explicit representation of the
   *social* ties among actors or the *epistatic* ties among components as
   emergent network processes.
4. **No endogeneity**: The conventional $K$ is set exogenously. In reality,
   epistatic coupling emerges from actors' overlapping affiliations.

## The SaoMNK Generalization

`searchnet` endogenizes the NK landscape by reformulating it as a **bipartite
network** of $M$ actors affiliating with $N$ components.

### The $\{K\}$ Interpretive Framework

The four statistics that compose the $\{K\}$ framework are individually
well-known: $K_{AC}$ and $K_{CA}$ are standard bipartite (two-mode) degree
measures, while $K_{AA}$ and $K_{CC}$ are standard one-mode projection
degree measures. None of these is new to network science (see, e.g.,
Latapy, Magnien, & Del Vecchio, 2008; Borgatti & Everett, 1997).

The contribution of the $\{K\}$ framework is not statistical but
**interpretive**: it provides a *theoretical bridge* between the NK
landscape tradition in strategy and the bipartite network tradition in
sociology. Specifically, the framework offers a structured lens for reading
established network degree measures as dimensions of an interdependence
landscape:

| Degree | Symbol | Interpretation | Network | Standard Network Measure |
|--------|--------|----------------|---------|--------------------------|
| Actor Scope | $K_{AC}$ | How many components each actor affiliates with | Bipartite (2-mode) | Row degree of incidence matrix |
| Component Popularity | $K_{CA}$ | How many actors affiliate with each component | Bipartite (2-mode) | Column degree of incidence matrix |
| Actor Sociality | $K_{AA}$ | How many other actors share components with each actor | Projection (1-mode) | Degree of actor co-membership projection |
| Component Epistasis | $K_{CC}$ | How many other components share actors with each component | Projection (1-mode) | Degree of component co-occurrence projection |

Formally, given the bipartite incidence matrix
$\mathbf{B} \in \{0,1\}^{M \times N}$, the four $\{K\}$ projections are:

$$K_{AC} = \text{rowSums}(\mathbf{B}), \quad K_{CA} = \text{colSums}(\mathbf{B})$$

$$K_{AA} = \text{deg}(\mathbf{B}\mathbf{B}^T), \quad K_{CC} = \text{deg}(\mathbf{B}^T\mathbf{B})$$

where $\text{deg}(\cdot)$ denotes the degree vector of the (binary)
one-mode projection, with the diagonal zeroed. $K_{AA}$ captures
actor--actor co-affiliation (sociality) and $K_{CC}$ captures
component--component co-occurrence (epistasis). A change to any single
cell $b_{ij}$ of $\mathbf{B}$ propagates to all four projections
simultaneously.

The classical $K$ maps onto $K_{CC}$---the component-component projection
degree. But `searchnet` shows that $K_{CC}$ is endogenously determined by the
bipartite network, and it co-evolves with $K_{AC}$, $K_{CA}$, and $K_{AA}$.

### Why Group These Four Measures?

Individually, each $K$ measure is a standard degree statistic that could be
analyzed in isolation. The analytical leverage of the $\{K\}$ framework
comes from treating them as a **coupled system**. Because all four measures
derive from the same bipartite incidence matrix, they are structurally
linked: a change in actor scope ($K_{AC}$) mechanically alters component
popularity ($K_{CA}$), which in turn reshapes both the actor co-membership
network ($K_{AA}$) and the component co-occurrence network ($K_{CC}$).
Analyzing any single measure in isolation obscures these cross-domain
dependencies. Grouping them makes visible the transmission mechanism
through which micro-level affiliation decisions propagate into macro-level
landscape ruggedness---a mechanism that is invisible when $K$ is treated as
a scalar parameter or when bipartite and projection degree are examined in
separate analyses.

### Coupling and Endogenization as the Core Novelty

The novelty of `searchnet` therefore lies not in the individual statistics but
in two analytical moves. First, **coupling**: by tracking all four degree
processes simultaneously, the framework exposes the structural constraints
that link actor behavior to landscape topology. Second,
**endogenization**: embedding these coupled degrees within a stochastic
actor-oriented model (SAOM) allows the interdependence structure to emerge
from actors' forward-looking affiliation decisions rather than being imposed
exogenously by the modeler. Together, coupling and endogenization transform
$K$ from a fixed model parameter into a dynamic, emergent property of the
system.

### Why SAOM over ERGM?

A natural question is why `searchnet` adopts Snijders' stochastic
actor-oriented model (SAOM; Snijders, 1996; Snijders, van de Bunt, &
Steglich, 2010) rather than the exponential random graph model (ERGM;
Lusher, Koskinen, & Robins, 2013) as its network engine. The choice follows
directly from the theoretical premise.

The NK tradition models **agents making strategic decisions**: an actor
evaluates alternative configurations and selects the one that maximizes a
fitness criterion. The SAOM's rate-function architecture is the natural
formalization of this process. At each micro-step, a single actor is
selected (rate function), evaluates the marginal utility of each possible
tie change (objective function), and chooses probabilistically according to
a multinomial logit rule. This actor-centric, sequential-decision structure
maps directly onto the local search process in NK models.

ERGMs, by contrast, characterize the equilibrium distribution of network
configurations without specifying a generative decision process. An ERGM
answers the question "what network structures are over-represented relative
to chance?" but does not model *who* makes *which* decision *when*. For a
framework whose purpose is to endogenize landscape ruggedness through
actors' strategic choices, this equilibrium focus omits precisely the
mechanism of interest.

In short, the SAOM is chosen because it models the same object that NK
theory models---an agent's adaptive search---and extends it to a setting
where the landscape co-evolves with the searcher's network position.

## Formal Framework

The following definitions and theorems establish SaoMNK as a strict
generalization of the NK model within a stochastic actor-oriented
architecture.

### Definition 1: NK Landscape

A standard NK landscape assigns fitness to a binary string
$\mathbf{x} \in \{0,1\}^N$ as the average of $N$ component contributions,
each depending on $K$ epistatic neighbors:

$$W(\mathbf{x}) = \frac{1}{N} \sum_{d=1}^{N} f_d(\mathbf{x}_{N(d)})$$

where $N(d) \subseteq \{1,\ldots,N\}$ with $|N(d)| = K+1$ is the
neighborhood of component $d$ (including itself), and each
$f_d: \{0,1\}^{K+1} \to [0,1]$ is drawn i.i.d.\ from $\text{Uniform}(0,1)$
(Kauffman, 1993).

### Definition 2: SaoMNK Utility

For actor $i$ with affiliation vector $\mathbf{b}_i \in \{0,1\}^N$
in the context of all other actors' affiliations $\mathbf{B}_{-i}$ and
influence matrix $E \in \{0,1\}^{N \times N}$, the SaoMNK utility function
is:

$$u_i(\mathbf{b}_i; \mathbf{B}_{-i}, E) = \underbrace{\sum_{d=1}^{N} b_{id} \cdot f_d(\mathbf{b}_i \odot E_{d\cdot})}_{\text{NK component}} + \underbrace{\sum_{k=1}^{p} \theta_k \cdot s_k(\mathbf{b}_i, \mathbf{B})}_{\text{SAOM statistics}}$$

The first term is the epistatic fitness contribution, where $\odot$
denotes element-wise masking by the $d$-th row of the influence matrix.
The second term captures endogenous network effects (density, popularity,
scope, etc.) weighted by structural parameters $\theta_k$.

### Definition 3: Choice Probability (McFadden Conditional Logit)

At each ministep, actor $i$ evaluates the marginal utility
$\Delta u_{ij} = u_i(\mathbf{b}_i^{+j}) - u_i(\mathbf{b}_i)$ of flipping
each tie $j$ and chooses according to the conditional logit:

$$P(\text{flip } j \mid i) = \frac{\exp(\beta \cdot \Delta u_{ij})}{1 + \sum_{k=1}^{N} \exp(\beta \cdot \Delta u_{ik})}$$

where $\beta > 0$ is the rationality parameter governing the precision of
choice. The $1$ in the denominator represents the option of making no change.

### Theorem 1: Reduction (NK as Special Case)

*Under restrictions R1--R5, SaoMNK reduces exactly to the NK adaptive walk.*

The restrictions are: (R1) $M=1$ actor, (R2) influence matrix
$E = A$ fixed exogenously, (R3) all structural parameters $\theta_k = 0$,
(R4) rationality $\beta \to \infty$ (greedy deterministic choice), and
(R5) constant ministep rate. Under these conditions the choice rule
selects the fitness-maximizing single-bit flip at each step, which is
precisely the NK adaptive walk (Kauffman, 1993).

### Theorem 2: Generalization

*SaoMNK strictly extends NK along three independent dimensions:*

(a) **Multi-actor interaction** ($M > 1$): Multiple actors search
simultaneously on overlapping landscapes, introducing competitive and
cooperative externalities absent from single-agent NK.

(b) **Endogenous landscape co-evolution** ($\theta \neq \mathbf{0}$):
Structural parameters allow the landscape topology to change as actors
modify their affiliations, coupling search dynamics to network evolution.

(c) **Bounded rationality** ($\beta < \infty$): Finite $\beta$ introduces
stochastic choice, spanning the continuum from random search ($\beta = 0$)
to deterministic greedy search ($\beta \to \infty$).

### Theorem 3: Approximation (Mixed Logit Universality)

*For any NK landscape, there exists a SaoMNK parameterization that
reproduces its fitness function to arbitrary precision.*

This follows from the McFadden and Train (2000) universality result: any
random utility model (specifically, any discrete choice model in the
McFadden, 1974 conditional-logit family, including the social-interaction
extension of Brock and Durlauf, 2001) can be approximated arbitrarily well
by a mixed logit specification. Since the SaoMNK objective function admits
mixed logit form, it inherits this approximation guarantee.

### Theorem 4: SAOM--QRE Equivalence

*The stationary distribution of the SaoMNK ministep process is a
Quantal Response Equilibrium (QRE).*

The logit choice rule (Definition 3) applied to a finite strategic game
defines a logit QRE (McKelvey & Palfrey, 1995). Under standard regularity
conditions (finite action space, continuous payoffs, positive ministep
rates), the SaoMNK Markov chain is ergodic and its stationary distribution
corresponds to a logit QRE of the underlying network game (Blume, 1993).
This connects the stochastic dynamics to a well-characterized equilibrium
concept.

### Theorem 5: Brock--Durlauf Recovery (Economic Foundation)

*The Brock and Durlauf (2001) binary discrete-choice-with-social-interactions
model is recovered as a degenerate special case of SaoMNK in the
$M \to \infty$, congestion-only, mean-field limit.*

Theorem 5 is the *complementary* reduction to Theorem 1 (NK Reduction),
which takes the single-actor / greedy / no-network limit. Together,
Theorems 1 and 5 establish that SaoMNK contains both Kauffman (1993)
NK landscape search and Brock & Durlauf (2001) discrete choice with
social interactions as special cases of a single CTMC. The five
restrictions Rb1--Rb5 (many-actor mean-field; component independence;
mean-field rescaling of the social-interaction coefficient; explicit
$\{0,1\} \leftrightarrow \{-1,+1\}$ spin recoding with the factor-of-4
correction; finite logit precision $\beta$) and the formal proof are
detailed in PROOF_TABLE.md Part L (rows L1--L15). Computational
verification is provided by `verify_brock_durlauf_reduction()`. See
`vignette("saomnk-brock-durlauf")` for the full tutorial walkthrough,
including the bifurcation diagram showing the $\beta J > 1$ multiplicity
threshold (Corollary 2, L11) and the social-multiplier formula
(Corollary 3, L12). The reflection-problem identification result of
Manski (1993) is resolved in SaoMNK via the logit nonlinearity (Remark
L13).

### Equilibrium-Concept Typology

Theorem 4 places SaoMNK within a broader landscape of equilibrium
concepts. Equilibrium solutions to strategic games can be organized along
two dimensions: the rationality assumption (full vs. bounded) and the
temporal structure (static vs. dynamic). The following typology clarifies
where the SAOM--QRE connection sits relative to classical alternatives:

|                        | **Static**                        | **Dynamic**                             |
|:-----------------------|:----------------------------------|:----------------------------------------|
| **Full rationality**   | Nash equilibrium                  | MPE (Ericson & Pakes, 1995; Wibbens et al., 2025) |
| **Bounded rationality**| Static QRE (McKelvey & Palfrey, 1995) | Dynamic QRE / Logit MPE (SaoMNK)   |

In the top-left cell, Nash equilibrium assumes perfectly rational
players in a one-shot or stage game. The top-right cell houses the
Markov perfect equilibrium (MPE), which retains full rationality but
embeds it in a dynamic state-transition framework (Ericson & Pakes,
1995; Wibbens, 2023; Wibbens et al., 2025). Repeated-game refinements such as
Bernheim and Whinston (1990) also inhabit the full-rationality row.

The bottom row relaxes perfect rationality. In the bottom-left cell,
McKelvey and Palfrey's (1995) static QRE replaces best response with
logit probabilistic response but retains a one-shot structure. SaoMNK
occupies the bottom-right cell: it combines bounded rationality (the
logit choice rule from Definition 3) with dynamic, sequential
decision-making over an evolving network landscape. Doraszelski and
Satterthwaite (2010) formalize a closely related object---the logit
Markov perfect equilibrium---as an intra-cell refinement.
Eibelsh\"auser and Poensgen (2019) prove that logit Markov QRE
converges to MPE as the precision parameter $\lambda \to \infty$ for
all finite dynamic stochastic games, establishing a formal homotopy
between the bottom-right and top-right cells with $\beta$ as the
bridging parameter. SaoMNK thus nests both bounded (finite $\beta$)
and full ($\beta \to \infty$) rationality within a single framework.

For strategic search on rugged NK landscapes, bounded rationality is
not merely a tractability device but a foundational modeling premise.
Agents cannot survey the full landscape; they evaluate local
alternatives and choose probabilistically. This makes the bottom-right
cell the natural equilibrium home for SaoMNK, with Blume's (1993)
convergence result providing the bridge from the SAOM ministep process
to the logit QRE stationary distribution.

### Bounded Rationality Spectrum

The rationality parameter $\beta$ governs search precision and connects
SaoMNK to classical NK behavior at the limit:

| $\beta$ | Behavior | NK Equivalent | Equilibrium Concept |
|---------|----------|---------------|---------------------|
| $0$ | Random search (uniform over actions) | --- | Uniform distribution |
| $(0, \infty)$ | Noisy best response (logit QRE) | --- | Markov QRE (Eibelsh\"auser \& Poensgen, 2019) |
| $\to \infty$ | Greedy deterministic choice | NK adaptive walk | MPE (Ericson \& Pakes, 1995) |

The intermediate regime $\beta \in (0, \infty)$ has no NK analogue---it
is a genuinely new behavioral domain enabled by the SaoMNK generalization.

Crucially, the interaction of landscape ruggedness $K$ and rationality
$\beta$ jointly determines equilibrium selection. When $K$ is high
(rugged landscape with many local peaks) and $\beta$ is also high, agents
lock into the nearest peak rapidly---an exploitation trap. When $K$ is
high but $\beta$ is moderate, agents maintain positive probability of
crossing fitness valleys, enabling cross-basin exploration and preserving
strategic optionality (Trigeorgis \& Reuer, 2017). When $K$ is low
(smooth landscape), a single global peak dominates regardless of $\beta$.
This $K \times \beta$ interaction is absent from both the NK literature
(which assumes $\beta = \infty$) and the static QRE literature (which
lacks landscape structure).

```{r load-saomnk, include=FALSE}
if (requireNamespace("searchnet", quietly = TRUE)) {
  library(searchnet)
} else {
  dir_r <- file.path(getwd(), "R")
  if (!file.exists(file.path(dir_r, "utils.R")))
    dir_r <- "D:/Search_networks/SaoMNK/R"
  .saomnk_dir <- dir_r
  .saomnk_dir <- dir_r; source(file.path(dirname(dir_r), "inst", "saomnk-loader.R"))
}

library(Matrix)

DV_NAME <- "self$bipartite_rsienaDV"
```

# Illustration: NK vs SaoMNK

## Case 1: No Endogenous Effects (Pure Epistasis)

When all structural effects are zero, actors search purely based on the
exogenous influence matrix---mimicking the classical NK model:

```{r case1-no-endogeneity}
environ_params <- list(
  M = 12, N = 12,
  BI_PROB = 0,
  rand_seed = 1234,
  name = "_nk_comparison_"
)

env_exog <- SaomNkRSienaBiEnv$new(environ_params)

component_int_mat <- create_block_diag(environ_params$N,
                                        round(environ_params$N / 3))

structure_model_exog <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = 0, dv_name = DV_NAME, fix = TRUE),
      list(effect = "inPop",   parameter = 0, dv_name = DV_NAME, fix = TRUE),
      list(effect = "outAct",  parameter = 0, dv_name = DV_NAME, fix = TRUE)
    ),
    coDyadCovars = list(
      list(
        effect = "XWX", parameter = 0.1, dv_name = DV_NAME, fix = TRUE,
        nodeSet = c("COMPONENTS", "COMPONENTS"),
        interaction1 = "self$component_1_coDyadCovar",
        x = component_int_mat
      )
    )
  )
)

env_exog$search_rsiena(
  structure_model_exog,
  iterations_per_actor = 30,
  run_seed = 12345
)
```

```{r case1-plots, fig.height=6, fig.width=6.5}
env_exog$plot_actor_degrees(loess_span = 0.35)
env_exog$plot_component_degrees(loess_span = 0.35)
```

## Case 2: Endogenous Network Effects

Now we activate structural effects. Negative `density` creates sparsity
pressure; positive `inPop` creates preferential attachment to popular
components; positive `outAct` creates scope expansion:

```{r case2-endogenous}
structure_model_endog <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -0.3, dv_name = DV_NAME, fix = TRUE),
      list(effect = "inPop",   parameter = 0.2,  dv_name = DV_NAME, fix = TRUE),
      list(effect = "outAct",  parameter = 0.1,  dv_name = DV_NAME, fix = TRUE)
    ),
    coDyadCovars = list(
      list(
        effect = "XWX", parameter = 0.02, dv_name = DV_NAME, fix = TRUE,
        nodeSet = c("COMPONENTS", "COMPONENTS"),
        interaction1 = "self$component_1_coDyadCovar",
        x = component_int_mat
      )
    )
  )
)

env_endog <- SaomNkRSienaBiEnv$new(environ_params)
env_endog$search_rsiena(
  structure_model_endog,
  iterations_per_actor = 30,
  run_seed = 12345
)
```

```{r case2-plots, fig.height=10, fig.width=10}
env_endog$plot_degree_4panel(loess_span = 0.3)
```

## Comparing the Two Cases

The exogenous case shows uncoupled degree evolution---each dimension changes
independently. The endogenous case shows **coupled** evolution: density pressure
constrains actor scope ($K_{AC}$), while preferential attachment creates
heterogeneous component popularity ($K_{CA}$). The resulting $K_{CC}$
(traditional epistasis) emerges endogenously from the combination of these
forces rather than being set by the modeler.

# The SAOM Objective Function

The SAOM objective function in `searchnet` takes the general form:

$$
f_i(\mathbf{x}) = \sum_k \theta_k s_{ik}(\mathbf{x})
$$

where $s_{ik}(\mathbf{x})$ are network statistics computed from the bipartite
network $\mathbf{x}$, and $\theta_k$ are the corresponding parameters. The
key statistics include:

| Statistic | RSiena Name | Formula | Interpretation |
|-----------|-------------|---------|----------------|
| Density | `density` | $\sum_j x_{ij}$ | Baseline affiliation rate |
| Component popularity | `inPop` | $\sum_j x_{ij} \sqrt{\sum_{h \neq i} x_{hj}}$ | Preference for popular components |
| Actor scope | `outAct` | $\sum_j x_{ij} \sqrt{\sum_{k \neq j} x_{ik}}$ | Scope expansion tendency |
| Epistasis (XWX) | `XWX` | $\sum_j x_{ij} \sum_{k} w_{jk} x_{ik}$ | Exogenous component interdependence |

# References

- Blume, L. E. (1993). The statistical mechanics of strategic interaction.
  *Games and Economic Behavior*, 5(3), 387--424.
- Borgatti, S. P., & Everett, M. G. (1997). Network analysis of 2-mode
  data. *Social Networks*, 19(3), 243--269.
- Kauffman, S. A. (1993). *The Origins of Order: Self-Organization and
  Selection in Evolution*. Oxford University Press.
- Kauffman, S. A., & Levin, S. (1987). Towards a general theory of adaptive
  walks on rugged landscapes. *Journal of Theoretical Biology*, 128(1), 11--45.
- Kauffman, S. A., & Weinberger, E. D. (1989). The NK model of rugged fitness
  landscapes and its application to maturation of the immune response.
  *Journal of Theoretical Biology*, 141(2), 211--245.
- Latapy, M., Magnien, C., & Del Vecchio, N. (2008). Basic notions for
  the analysis of large two-mode networks. *Social Networks*, 30(1), 31--48.
- Levinthal, D. A. (1997). Adaptation on rugged landscapes.
  *Management Science*, 43(7), 934--950.
- Lusher, D., Koskinen, J., & Robins, G. (Eds.). (2013). *Exponential
  Random Graph Models for Social Networks*. Cambridge University Press.
- McFadden, D. (1974). Conditional logit analysis of qualitative choice
  behavior. In P. Zarembka (Ed.), *Frontiers in Econometrics* (pp. 105--142).
  Academic Press.
- McFadden, D., & Train, K. (2000). Mixed MNL models for discrete response.
  *Journal of Applied Econometrics*, 15(5), 447--470.
- McKelvey, R. D., & Palfrey, T. R. (1995). Quantal response equilibria
  for normal form games. *Games and Economic Behavior*, 10(1), 6--38.
- Rivkin, J. W. (2000). Imitation of complex strategies.
  *Management Science*, 46(6), 824--844.
- Snijders, T. A. B. (1996). Stochastic actor-oriented models for network
  change. *Journal of Mathematical Sociology*, 21(1--2), 149--172.
- Snijders, T. A. B. (2001). The statistical evaluation of social network
  dynamics. *Sociological Methodology*, 31(1), 361--395.
- Snijders, T. A. B., van de Bunt, G. G., & Steglich, C. E. G. (2010).
  Introduction to stochastic actor-based models for network dynamics.
  *Social Networks*, 32(1), 44--60.

```{r session-info}
sessionInfo()
```
