---
title: "Foundational Search Models as Restrictions of SAOM-NK"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Foundational Search Models as Restrictions of SAOM-NK}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 11, fig.height = 9,
                      out.width = "100%")
library(searchnet)
have_yaml <- requireNamespace("yaml", quietly = TRUE)
knitr::opts_chunk$set(eval = have_yaml)
```

The SAOM-NK objective is a sum over classes of effects. Each class is one
summand: complementarity (the fitness term under an influence matrix `W`),
scope (actor degree), popularity / crowding (components other actors hold),
contact (four-cycles between actor pairs), and imitation (co-holder performance
similarity). Each class stands for a construct and has two \{K\} fields: the
dimension its change statistic reads, and the dimension its target statistic
moves.

Many foundational models of search are this objective with some classes
switched off and some restrictions imposed. The translation registry records
those statements, one entry per model, with the restriction, the term each
class becomes, the Lean declarations behind the statement, and what the
statement does not cover. This vignette shows how to browse the registry,
generate a runnable special case, translate your own model, compare models,
draw the translation, add an entry, and reach the Lean proofs.

## The classes

```{r classes}
rosetta_classes()[, c("id", "effects", "reads", "moves", "lean_stat")]
```

The four \{K\} dimensions (Expansiveness K_AC, Popularity K_CA, Sociality
K_AA, Epistasis K_CC) and the two fields are defined once, in
`inst/rosetta/K_DIMENSIONS.md`
(`system.file("rosetta", "K_DIMENSIONS.md", package = "searchnet")`). `reads`
is derived from the change statistic, the change in actor `i`'s statistic when
it toggles component `j`: its inputs are perturbed one at a time, and the first
it depends on names the dimension (who the other holders are, Popularity; else
other actors' holdings, Sociality; else W, Epistasis; else the actor's own
portfolio or attributes, Expansiveness). `moves` is the dimension of the target
statistic (the statistic summed over actors): what estimation matches and what
the coefficient moves first. `k_channel` is a deprecated alias of `moves`.
`searchnet_effect_dimensions()` gives both fields effect by effect, with the
formulas:

```{r dimensions}
searchnet_effect_dimensions(c("density", "outAct", "inPop", "XWX", "cycle4",
                              "egoX", "altX", "simEgoInDist2"))[,
  c("effect", "reads", "moves", "change_statistic")]
```

The package's K_AA and K_CC degrees are distinct-partner counts; their weighted
forms, Sociality and Epistasis strength, are what the identities speak to: the
second moment of actor degree is the total Epistasis strength, so scope effects
read Expansiveness and move Epistasis strength. `moves` is the moment moved
first, not the full equilibrium response, and the sign of the coefficient
changes the reading (`inPop` below zero spreads actors out, above zero herds
them), never the dimension.

A statistic of your own classifies itself by the same rules. A toy example,
ties weighted by the number of other holders of each component:

```{r classify}
toy <- function(B, cov) c(B %*% (colSums(B) - 1))
searchnet_classify_effect(toy)[c("reads", "moves")]
```

## Browse the foundational models

```{r entries}
e <- rosetta_entries()
e[, c("id", "relation", "active", "off")]
```

Relations follow a fixed vocabulary: a `special-case` is recovered exactly
under its restrictions, a `special-case-linearized` after a stated
approximation, an `equilibrium-characterization` is a solution concept the
stationary law satisfies, a `comparison` is related but not nested, a
`translation` maps a theory construct to statistics with no reduction proved,
and `not-carried` records a construct the framework does not represent.

```{r one-entry}
rosetta_entry("nk-adaptive-walk")
```

## Generate a special case and run it

`rosetta_model()` turns an entry's restriction into a specification that
searchnet runs. The NK restriction has one actor and greedy choice; searchnet
runs it as the classical adaptive walk on the landscape with the same `N`,
`K` and seed.

```{r nk-run}
nk <- rosetta_model("nk-adaptive-walk", N = 6, K = 2, seed = 3)
nk
r <- rosetta_run(nk)
r$engine
r$B
```

A multi-actor restriction runs on the SAOM engine. Here, the two-mode SAOM
with the influence matrix set to zero:

```{r saom-run}
s <- rosetta_model("plain-saom-two-mode", M = 3, N = 5)
r2 <- rosetta_run(s, steps_per_actor = 3)
r2$B
```

The same model in ordinary searchnet code, to use without the registry:

```{r r-code}
cat(rosetta_r_code(s))
```

## Translate your own model

Write a model the usual way. `rosetta_translate()` decomposes it class by
class and gives a nesting verdict for every registry entry. A class is
included in your model when it has an effect of that class not held fixed at
zero. Your model nests an entry when it includes every class the entry has
active, linearized or fixed, and uses the entry's revision rule if the entry
sets one. The verdict is `nests`, `nests_in_limit` (reached as M or beta
tends to infinity), `conditional` (the classes are there but the revision
rule differs), or `does_not_nest`.

```{r translate}
mine <- saomnk_model(density = -1, popularity = -0.2,
                     influence_matrix = saomnk_block_diagonal(6, 2),
                     c4 = list(effect = "cycle4", parameter = 0.1))
tr <- rosetta_translate(mine)
tr
tr$nesting[, c("id", "verdict", "missing", "conditions", "limits")]
tr$contains[, c("id", "differences")]
```

Read `differences` as the switches that take your model to the entry: the
classes to set to zero and the restrictions to impose. A model built with
`saomnk_model()` runs RSiena's multinomial ministep, so entries proved for
single-flip revision come out `conditional`.

## Compare models

```{r compare}
rosetta_compare(list("nk-adaptive-walk", "brock-durlauf-social-interactions",
                     "plain-saom-two-mode", mine = mine))
```

## Draw the translation

Row I writes the objective by class: one chip per class, the effects it
contains, and its construct; classes on the same \{K\} dimension sit side
by side under one black badge naming it. By default the badges show what the
classes move (popularity / crowding, contact and imitation share
"Sociality K_AA"); `view = "reads"` shows what each class reads, and
`view = "both"` draws the two rows, reads above moves. Row II
rewrites a foundational model term by term in the same classes, with `+ 0`
and an outlined chip for each class it switches off. Row III shows the
bipartite state of a short seeded run of the restriction (circles are
actors, squares are components), the influence matrix `W`, and its binary
pattern `1[W != 0]` off the diagonal (the engine's XWX sums over
`j != h`), whose row counts are the K that a binary NK influence
pattern records.

```{r plot-nk}
rosetta_plot(NULL, compare = "nk-adaptive-walk")
```

```{r plot-both, fig.height = 3.6}
rosetta_plot(NULL, glyph = FALSE, view = "both")
```

```{r plot-mine}
rosetta_plot(mine, compare = "brock-durlauf-social-interactions")
```

## Add your own entry

`rosetta_new_entry()` writes a template from a model. Fill in the citation,
relation, statement and what the statement does not cover, then validate.
Validation reports errors (schema, ids, Lean names, unregistered classes,
em-dashes, British spelling) separately from advisory measurements (for
example, an entry drafted by a model and not yet reviewed).

```{r new-entry}
d <- file.path(tempdir(), "my_registry")
dir.create(file.path(d, "entries"), recursive = TRUE, showWarnings = FALSE)
file.copy(system.file("rosetta", "classes.yaml", package = "searchnet"), d)
f <- rosetta_new_entry("my-congestion-theory", from = mine, dir = file.path(d, "entries"))
rosetta_validate(path = d)
```

A new statistic can join as its own class for the session:

```{r register}
rosetta_register_class("prestige", effects = "altX", k_channel = "K_CA",
                       construct = "status of components")
rosetta_classes()$id
rosetta_reset_classes()
```

## Link to Lean

`rosetta_lean()` lists the declarations an entry names, with their statements
from the Lean registry, and writes an instance file with
`lean_export_model()` where the library supports the model. For the NK
restriction the instance proves that the NK term is NK fitness and that greedy
search is the zero-noise limit; check it with `lean_check()` when Lean is
installed (see `vignette("saomnk-lean")`).

```{r lean}
l <- rosetta_lean("nk-adaptive-walk")
l$declarations[, c("decl", "statement")]
basename(l$instance)
```

## Export for other tools

`rosetta_export_json()` writes the classes and entries as JSON, and
`rosetta_model_to_json()` and `rosetta_model_from_json()` carry a model
specification between R and a front end.

```{r json}
js <- rosetta_model_to_json(s, pretty = FALSE)
js
identical(rosetta_model_from_json(js)$N, s$N)
```
