---
title: "Basin Analysis for Industrial Policy: AMR Application"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{AMR Industrial Policy Basin Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

```{r load-saomnk, include=FALSE}
if (requireNamespace("searchnet", quietly = TRUE)) {
  library(searchnet)
} else {
  dir_r <- file.path(getwd(), "R")
  if (!file.exists(file.path(dir_r, "utils.R")))
    dir_r <- "D:/Search_networks/SaoMNK/R"
  .saomnk_dir <- dir_r
  .saomnk_dir <- dir_r; source(file.path(dirname(dir_r), "inst", "saomnk-loader.R"))
}

library(Matrix)

DV_NAME <- "self$bipartite_rsienaDV"
```

# Overview

Industrial policy reshapes competitive landscapes. Subsidies, grants, tax
concessions, and guarantees do not merely transfer resources---they alter the
structural parameters that govern how firms search for viable strategies. In
the SaoMNK framework, each policy instrument maps onto a specific change in the
SAOM objective function, which in turn moves firms across the basins of
attraction of the fitness landscape.

This vignette applies `searchnet`'s basin analysis to four stylized industrial
policies drawn from the antimicrobial resistance (AMR) pharmaceutical context.
The analysis measures basin **depth** (the fitness gap between peak and rim,
measuring competitive moat strength) and basin **width** (the number of
distinct strategies converging to each peak, measuring strategic flexibility),
and asks which of the two a policy-induced parameter shift actually moves.
Together, these geometric properties connect the micro-mechanisms of firm
search to the macro-outcomes that policymakers care about: competitive
dynamics, innovation trajectories, and system adaptability.

# 1. Industrial Policy as Landscape Reshaping

The SaoMNK framework endogenizes the NK fitness landscape by representing the
search environment as a bipartite network of $M$ actors affiliating with $N$
components. The SAOM objective function governs actor search behavior through
network statistics---density (scope cost), inPop (popularity), XWX
(epistasis)---each weighted by a parameter $\theta$. Industrial policy, in
this formalization, is a **theta shock**: an exogenous change to one or more
$\theta$ parameters that changes how firms search. The section "What a Theta
Shock Can and Cannot Move" establishes which part of the geometry such a shock
can reach.

The four AMR policies act through distinct channels:

| Policy | Instrument | Parameter Shift | Expected Effect |
|:-------|:-----------|:----------------|:-----------------|
| P1 | Economic Subsidy | density: $-0.6 \to -0.96$ | Narrows basins |
| P2 | Direct Grant | density: $-0.6 \to -0.3$ | Widens basins |
| P3 | Tax Concession | XWX: $0.6 \to 0.96$ | Deepens basins |
| P4 | Guarantee | density: $-0.6 \to -0.24$, XWX: $0.6 \to 0.9$ | Complex reshaping |

The mapping is intuitive. **Density** controls the cost of scope---how many
activities a firm can sustain simultaneously. Increasing the magnitude of the
density penalty (P1) forces firms to shed activities, narrowing the set of
viable configurations and thus the width of each basin. Decreasing it (P2)
enables firms to maintain broader portfolios, expanding the viable set.
**XWX** (the epistasis statistic) controls how strongly component interactions
contribute to fitness. Amplifying epistasis (P3) deepens the valleys between
peaks, making partial imitation costlier and strengthening incumbent moats.

Policy P4---the guarantee---is the most theoretically interesting because it
operates on both channels simultaneously. The scope cost drops (widening
effect) while epistasis rises (deepening effect). Whether basins ultimately
widen or deepen depends on the relative magnitude of these opposing forces, a
question that only simulation can resolve.

# 2. Baseline Environment Setup

We model the AMR pharmaceutical sector as a bipartite network of $M = 20$
firms affiliating with $N = 12$ activities (e.g., R\&D pipelines, clinical
trial capacities, manufacturing processes, distribution networks, regulatory
capabilities). The initial affiliation probability of 0.4 reflects a
moderately populated activity space where firms have already committed to some
portfolio of capabilities.

Basin analysis enumerates all $2^N$ configurations, so $N$ controls the cost of
everything below. We use $N = 12$ (4096 configurations), which is small enough
that the whole vignette runs in seconds and large enough to show the geometry.
Nothing in the analysis depends on that choice: raising $N$ multiplies the work
by $2^{\Delta N}$ and leaves the structure intact.

```{r amr-environment}
library(searchnet)

## AMR environment: 20 firms, 12 activities
env_params <- list(
  M = 20,
  N = 12,
  BI_PROB = 0.4,
  rand_seed = 42,
  name = "amr-baseline"
)

env <- SaomNkRSienaBiEnv$new(env_params)

## Modular influence structure: 4 blocks of 3 components
## (e.g., R&D | Manufacturing | Regulatory | Commercial)
component_int_mat <- create_block_diag(env_params$N, 4)
```

The baseline structure model specifies a moderate density penalty
($\theta_{\text{density}} = -0.6$), reflecting non-trivial costs of
maintaining each additional activity, and moderate popularity-seeking
($\theta_{\text{inPop}} = 0.3$), capturing bandwagon effects where firms
gravitate toward activities that other firms also perform. Epistasis
($\theta_{\text{XWX}} = 0.6$) is set at a level that produces a moderately
rugged landscape with several local peaks.

```{r baseline-model}
baseline_model <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -0.6,
           dv_name = DV_NAME, fix = TRUE),
      list(effect = "inPop", parameter = 0.3,
           dv_name = DV_NAME, fix = TRUE)
    ),
    coCovars = list(),
    varCovars = list(),
    coDyadCovars = list(
      list(effect = "XWX", parameter = 0.6,
           dv_name = DV_NAME, fix = TRUE,
           nodeSet = c("COMPONENTS", "COMPONENTS"),
           interaction1 = "self$component_1_coDyadCovar",
           x = component_int_mat)
    ),
    varDyadCovars = list(),
    interactions = list()
  )
)
```

## Run the Baseline Simulation

```{r run-baseline}
env$search_rsiena(
  structure_model = baseline_model,
  iterations_per_actor = 50,
  run_seed = 42
)
```

## Compute the Baseline Fitness Landscape

The `compute_fitness_landscape` method evaluates the utility of all $2^N$
counterfactual component configurations (or a representative sample when $N$
is large), identifying local peaks and the basin structure of the landscape.

```{r baseline-landscape}
env$compute_fitness_landscape(n_landscapes = 1)
```

# 3. Pre-Intervention Basin Geometry

Before applying any policy intervention, we characterize the baseline landscape
to establish a counterfactual. The fitness landscape array
`env$fitness_landscape` has dimensions
$[\text{n\_landscapes}, 2^N, 2N + 2]$, where the last two columns store total
fitness and a binary peak indicator, respectively.

```{r baseline-basin-analysis}
N <- env_params$N
n_configs <- 2^N
FL <- env$fitness_landscape[1, , ]

## Extract configuration vectors, fitness values, and peak indicators
configs  <- FL[, 1:N]
fitness  <- FL[, (2 * N + 1)]
is_peak  <- FL[, (2 * N + 2)]

peak_ids <- which(is_peak == 1)
cat("Number of peaks:", length(peak_ids), "\n")
cat("Peak fitness values:", round(fitness[peak_ids], 3), "\n")
```

## Assign Configurations to Basins via Steepest Ascent

Each non-peak configuration belongs to the basin of the peak it would reach
under greedy (steepest-ascent) local search. We compute basin membership by
following the gradient from each configuration:

The naive way to do this is to search the configuration matrix for each
neighbor, but that scans all $2^N$ rows per lookup and makes the whole
procedure $O(4^N)$. Instead we index the configurations once by their bit
string, so every Hamming-1 neighbor is found by table lookup and the ascent
itself is pure indexing.

```{r basin-assignment}
## Steepest-ascent basin assignment
assign_basins <- function(configs, fitness, N) {
  n <- nrow(configs)
  cfg_key <- do.call(paste0, as.data.frame(configs))

  ## nbr[i, j] is the row holding config i with bit j flipped.
  nbr <- matrix(NA_integer_, nrow = n, ncol = N)
  for (j in seq_len(N)) {
    flipped <- configs
    flipped[, j] <- 1 - flipped[, j]
    nbr[, j] <- match(do.call(paste0, as.data.frame(flipped)), cfg_key)
  }

  ## From each configuration, move to the fittest strictly improving
  ## neighbor until none improves. The endpoint is the basin's peak.
  basin_id <- integer(n)
  for (i in seq_len(n)) {
    current <- i
    repeat {
      cand <- nbr[current, ]
      cand <- cand[!is.na(cand)]
      best <- cand[which.max(fitness[cand])]
      if (!length(best) || fitness[best] <= fitness[current]) break
      current <- best
    }
    basin_id[i] <- current
  }
  basin_id
}

## Where the firms actually sit on the landscape, and the geometry each
## firm therefore faces. This is the quantity the policy shocks move.
occupancy <- function(bi, basins, summary_df, cfgs) {
  rows  <- match(do.call(paste0, as.data.frame(bi)),
                 do.call(paste0, as.data.frame(cfgs)))
  peaks <- basins[rows]
  g     <- summary_df[match(peaks, summary_df$peak_id), ]
  data.frame(
    Basins_Occupied  = length(unique(peaks[!is.na(peaks)])),
    Depth_Faced      = mean(g$depth, na.rm = TRUE),
    Width_Faced      = mean(g$width, na.rm = TRUE),
    Mean_Scope       = mean(rowSums(bi))
  )
}

basin_membership <- assign_basins(configs, fitness, N)

## Compute basin-level summaries
basin_summary <- data.frame(
  peak_id     = peak_ids,
  peak_fitness = fitness[peak_ids],
  width       = sapply(peak_ids, function(p) sum(basin_membership == p)),
  depth       = sapply(peak_ids, function(p) {
    members <- which(basin_membership == p)
    fitness[p] - min(fitness[members])
  })
)

basin_summary <- basin_summary[order(-basin_summary$peak_fitness), ]
knitr::kable(basin_summary, digits = 3,
             caption = "Baseline basin geometry: pre-intervention")
```

## Depth--Width Scatterplot

The depth--width scatterplot reveals the fundamental tradeoff in the baseline
landscape. Deep, narrow basins indicate strong competitive moats with limited
strategic flexibility. Shallow, wide basins indicate weak moats but high
adaptability.

```{r depth-width-scatter, fig.width=6, fig.height=5}
library(ggplot2)

ggplot(basin_summary, aes(x = width, y = depth)) +
  geom_point(aes(size = peak_fitness), alpha = 0.7) +
  geom_text(aes(label = peak_id), nudge_y = 0.005, size = 3) +
  scale_size_continuous(name = "Peak Fitness") +
  labs(
    title = "Baseline Basin Geometry",
    subtitle = expression(paste(theta[density], " = -0.6, ",
                                theta[XWX], " = 0.6")),
    x = "Basin Width (number of configurations)",
    y = "Basin Depth (fitness gap: peak minus rim)"
  ) +
  theme_minimal()
```

## Actor Basin Membership

Mapping the 20 firms to their current basins reveals the competitive structure
of the pre-intervention landscape:

```{r actor-basins}
## Extract current actor configurations
bi_mat <- env$bipartite_matrix

## Same bit-string index as above: match each firm's portfolio to its row.
cfg_key    <- do.call(paste0, as.data.frame(configs))
actor_rows <- match(do.call(paste0, as.data.frame(bi_mat)), cfg_key)
actor_basins <- basin_membership[actor_rows]

actor_df <- data.frame(
  actor = 1:env_params$M,
  basin = actor_basins,
  scope = rowSums(bi_mat)
)

## Treatment/control assignment: first 10 firms = treatment, rest = control
actor_df$group <- ifelse(actor_df$actor <= 10, "Treatment", "Control")

knitr::kable(actor_df,
             caption = "Actor basin membership and attributes (baseline)")

occ_base <- occupancy(bi_mat, basin_membership, basin_summary, configs)
```

## What a Theta Shock Can and Cannot Move

Before running the policies, it is worth being precise about which object each
one touches, because the four runs below make the distinction visible.

`compute_fitness_landscape()` builds the landscape from the influence matrix
$\mathbf{W}$, the component value draws, and the seed. It does not take
$\boldsymbol{\theta}$. So the basin geometry reported in each section
below---the number of peaks, and the width and depth of each basin---is a
property of the **architecture**, and it is identical across all five runs
here because all five share one influence matrix and one seed.

What $\boldsymbol{\theta}$ moves is **where the firms end up on that fixed
landscape**. A policy that raises the scope penalty does not narrow the
basins; it relocates firms into narrower ones. Each policy section therefore
reports two things side by side: the geometry, which is invariant, and the
occupancy, which is the policy effect. Reshaping the geometry itself requires
changing $\mathbf{W}$, which is a different intervention from any of P1--P4.

# 4. Policy P1: Economic Subsidy (Resource Suppressing)

An economic subsidy in the AMR context---such as a push funding mechanism that
pays firms per approved compound---paradoxically *increases* the effective cost
of scope. Firms receiving subsidies tied to specific therapeutic areas narrow
their portfolios to maximize per-unit returns, shedding diversified activities.
In the SaoMNK formalization, this is a density shift from $-0.6$ to $-0.96$:
a 60% increase in the magnitude of the scope penalty.

The expected landscape effect is **basin narrowing**. Higher density cost
means fewer activity configurations are viable at any given fitness level,
reducing the width of each basin of attraction. Firms are funneled into
tighter strategic positions.

```{r p1-model}
## P1 shock: density -0.6 --> -0.96, inPop 0.3 --> 0.36
p1_model <- baseline_model
p1_model$dv_bipartite$effects[[1]]$parameter <- -0.96
p1_model$dv_bipartite$effects[[2]]$parameter <- 0.36

env_p1 <- SaomNkRSienaBiEnv$new(list(
  M = env_params$M, N = env_params$N, BI_PROB = 0.4, rand_seed = 42,
  name = "amr-p1-subsidy"
))

env_p1$search_rsiena(
  structure_model = p1_model,
  iterations_per_actor = 50,
  run_seed = 42
)

env_p1$compute_fitness_landscape(n_landscapes = 1)
```

```{r p1-basins}
FL_p1     <- env_p1$fitness_landscape[1, , ]
fitness_p1 <- FL_p1[, (2 * N + 1)]
is_peak_p1 <- FL_p1[, (2 * N + 2)]
peak_ids_p1 <- which(is_peak_p1 == 1)

basin_p1 <- assign_basins(FL_p1[, 1:N], fitness_p1, N)

basin_summary_p1 <- data.frame(
  peak_id      = peak_ids_p1,
  peak_fitness = fitness_p1[peak_ids_p1],
  width        = sapply(peak_ids_p1, function(p) sum(basin_p1 == p)),
  depth        = sapply(peak_ids_p1, function(p) {
    members <- which(basin_p1 == p)
    fitness_p1[p] - min(fitness_p1[members])
  })
)

occ_p1 <- occupancy(env_p1$bipartite_matrix, basin_p1,
                       basin_summary_p1, FL_p1[, 1:N])

cat("P1 Subsidy geometry:  ", nrow(basin_summary_p1), " peaks, mean width ",
    round(mean(basin_summary_p1$width), 1), ", mean depth ",
    round(mean(basin_summary_p1$depth), 3),
    "  (baseline: ", nrow(basin_summary), " / ",
    round(mean(basin_summary$width), 1), " / ",
    round(mean(basin_summary$depth), 3), ")\n", sep = "")
cat("P1 Subsidy occupancy: firms in ", occ_p1$Basins_Occupied,
    " basins, width faced ", round(occ_p1$Width_Faced, 1),
    ", depth faced ", round(occ_p1$Depth_Faced, 3),
    ", mean scope ", round(occ_p1$Mean_Scope, 2),
    "  (baseline: ", occ_base$Basins_Occupied, " / ",
    round(occ_base$Width_Faced, 1), " / ",
    round(occ_base$Depth_Faced, 3), " / ",
    round(occ_base$Mean_Scope, 2), ")\n", sep = "")
```

The subsidy creates **specialization traps**. The steeper scope penalty does
not eliminate configurations from the landscape; it makes the diversified ones
unattractive to hold, so firms shed activities and settle into narrower
basins. Read the occupancy line above rather than the geometry line: firms
that were exploring broad portfolios find themselves funneled toward narrow
specializations---precisely
the rigidity that AMR policy aims to avoid when the goal is cross-therapeutic
innovation.

# 5. Policy P2: Direct Grant (Resource Enriching)

A direct grant---such as a lump-sum R\&D award not tied to specific
outputs---reduces the effective cost of maintaining multiple activities. Firms
receiving grants can sustain broader portfolios because the marginal cost of
each additional activity is subsidized. The density parameter shifts from
$-0.6$ to $-0.3$: a 50% reduction in scope cost.

The expected landscape effect is **basin widening**. Lower density cost means
more activity combinations remain viable, expanding the set of configurations
that converge to each peak.

```{r p2-model}
## P2 shock: density -0.6 --> -0.3, inPop 0.3 --> 0.42
p2_model <- baseline_model
p2_model$dv_bipartite$effects[[1]]$parameter <- -0.3
p2_model$dv_bipartite$effects[[2]]$parameter <- 0.42

env_p2 <- SaomNkRSienaBiEnv$new(list(
  M = env_params$M, N = env_params$N, BI_PROB = 0.4, rand_seed = 42,
  name = "amr-p2-grant"
))

env_p2$search_rsiena(
  structure_model = p2_model,
  iterations_per_actor = 50,
  run_seed = 42
)

env_p2$compute_fitness_landscape(n_landscapes = 1)
```

```{r p2-basins}
FL_p2      <- env_p2$fitness_landscape[1, , ]
fitness_p2 <- FL_p2[, (2 * N + 1)]
is_peak_p2 <- FL_p2[, (2 * N + 2)]
peak_ids_p2 <- which(is_peak_p2 == 1)

basin_p2 <- assign_basins(FL_p2[, 1:N], fitness_p2, N)

basin_summary_p2 <- data.frame(
  peak_id      = peak_ids_p2,
  peak_fitness = fitness_p2[peak_ids_p2],
  width        = sapply(peak_ids_p2, function(p) sum(basin_p2 == p)),
  depth        = sapply(peak_ids_p2, function(p) {
    members <- which(basin_p2 == p)
    fitness_p2[p] - min(fitness_p2[members])
  })
)

occ_p2 <- occupancy(env_p2$bipartite_matrix, basin_p2,
                       basin_summary_p2, FL_p2[, 1:N])

cat("P2 Grant geometry:  ", nrow(basin_summary_p2), " peaks, mean width ",
    round(mean(basin_summary_p2$width), 1), ", mean depth ",
    round(mean(basin_summary_p2$depth), 3),
    "  (baseline: ", nrow(basin_summary), " / ",
    round(mean(basin_summary$width), 1), " / ",
    round(mean(basin_summary$depth), 3), ")\n", sep = "")
cat("P2 Grant occupancy: firms in ", occ_p2$Basins_Occupied,
    " basins, width faced ", round(occ_p2$Width_Faced, 1),
    ", depth faced ", round(occ_p2$Depth_Faced, 3),
    ", mean scope ", round(occ_p2$Mean_Scope, 2),
    "  (baseline: ", occ_base$Basins_Occupied, " / ",
    round(occ_base$Width_Faced, 1), " / ",
    round(occ_base$Depth_Faced, 3), " / ",
    round(occ_base$Mean_Scope, 2), ")\n", sep = "")
```

The grant **enables flexibility but weakens competitive moats**. With lower
scope costs, firms hold broader portfolios and settle into the wider, shallower
basins of the same landscape, so the depth they face falls even though no
basin has changed shape. The strategic implication is
that grants are best suited for environments where adaptability matters more
than defensibility: early-stage markets, uncertain regulatory regimes, or
sectors where the dominant design has not yet emerged. In the AMR context,
grants facilitate the broad-spectrum R\&D portfolios needed when it is unclear
which pathogen threats will materialize, but they do not build the tight
complementarities that sustain long-run competitive advantage.

# 6. Policy P3: Tax Concession (Resource Retaining)

A tax concession---such as enhanced R\&D tax credits for AMR-related
expenditures---allows firms to retain more of the returns generated by
component interactions. The mechanism operates through the epistasis channel:
the XWX parameter shifts from $0.6$ to $0.96$, amplifying the fitness
contribution of complementary activities.

The expected landscape effect is **basin deepening**. Stronger epistasis
increases the fitness gap between peak configurations and their neighbors,
making partial imitation more costly.

```{r p3-model}
## P3 shock: XWX 0.6 --> 0.96, inPop 0.3 --> 0.21
p3_model <- baseline_model
p3_model$dv_bipartite$coDyadCovars[[1]]$parameter <- 0.96
p3_model$dv_bipartite$effects[[2]]$parameter <- 0.21

env_p3 <- SaomNkRSienaBiEnv$new(list(
  M = env_params$M, N = env_params$N, BI_PROB = 0.4, rand_seed = 42,
  name = "amr-p3-tax"
))

env_p3$search_rsiena(
  structure_model = p3_model,
  iterations_per_actor = 50,
  run_seed = 42
)

env_p3$compute_fitness_landscape(n_landscapes = 1)
```

```{r p3-basins}
FL_p3      <- env_p3$fitness_landscape[1, , ]
fitness_p3 <- FL_p3[, (2 * N + 1)]
is_peak_p3 <- FL_p3[, (2 * N + 2)]
peak_ids_p3 <- which(is_peak_p3 == 1)

basin_p3 <- assign_basins(FL_p3[, 1:N], fitness_p3, N)

basin_summary_p3 <- data.frame(
  peak_id      = peak_ids_p3,
  peak_fitness = fitness_p3[peak_ids_p3],
  width        = sapply(peak_ids_p3, function(p) sum(basin_p3 == p)),
  depth        = sapply(peak_ids_p3, function(p) {
    members <- which(basin_p3 == p)
    fitness_p3[p] - min(fitness_p3[members])
  })
)

occ_p3 <- occupancy(env_p3$bipartite_matrix, basin_p3,
                       basin_summary_p3, FL_p3[, 1:N])

cat("P3 Tax geometry:  ", nrow(basin_summary_p3), " peaks, mean width ",
    round(mean(basin_summary_p3$width), 1), ", mean depth ",
    round(mean(basin_summary_p3$depth), 3),
    "  (baseline: ", nrow(basin_summary), " / ",
    round(mean(basin_summary$width), 1), " / ",
    round(mean(basin_summary$depth), 3), ")\n", sep = "")
cat("P3 Tax occupancy: firms in ", occ_p3$Basins_Occupied,
    " basins, width faced ", round(occ_p3$Width_Faced, 1),
    ", depth faced ", round(occ_p3$Depth_Faced, 3),
    ", mean scope ", round(occ_p3$Mean_Scope, 2),
    "  (baseline: ", occ_base$Basins_Occupied, " / ",
    round(occ_base$Width_Faced, 1), " / ",
    round(occ_base$Depth_Faced, 3), " / ",
    round(occ_base$Mean_Scope, 2), ")\n", sep = "")
```

Tax concessions **strengthen incumbents but reduce system adaptability**.
Amplified epistasis pulls firms toward the deeper basins of the landscape, and
firms that come to sit on those peaks enjoy stronger VRIO-style competitive
advantages: their activity
configurations are difficult to replicate because partial imitation destroys
the complementarities that generate superior fitness. However, the same depth
that protects incumbents also traps them. When the landscape shifts---due to a
new pathogen, a regulatory change, or a technological discontinuity---firms in
deep basins face steeper costs of adaptation. The landscape becomes more
rugged, and the system as a whole becomes less responsive to environmental
change.

# 7. Policy P4: Guarantee (Resource Freeing)

A guarantee---such as an advance market commitment (AMC) or a subscription
model that guarantees revenue irrespective of volume---operates on both
channels simultaneously. By reducing downside risk, guarantees lower the
effective cost of maintaining activities (density: $-0.6 \to -0.24$) while
also strengthening the returns to complementary configurations (XWX: $0.6 \to
0.9$). The density drop widens basins; the epistasis increase deepens them.
The net geometric effect is indeterminate *ex ante*.

```{r p4-model}
## P4 shock: density -0.6 --> -0.24, XWX 0.6 --> 0.9
p4_model <- baseline_model
p4_model$dv_bipartite$effects[[1]]$parameter <- -0.24
p4_model$dv_bipartite$coDyadCovars[[1]]$parameter <- 0.9

env_p4 <- SaomNkRSienaBiEnv$new(list(
  M = env_params$M, N = env_params$N, BI_PROB = 0.4, rand_seed = 42,
  name = "amr-p4-guarantee"
))

env_p4$search_rsiena(
  structure_model = p4_model,
  iterations_per_actor = 50,
  run_seed = 42
)

env_p4$compute_fitness_landscape(n_landscapes = 1)
```

```{r p4-basins}
FL_p4      <- env_p4$fitness_landscape[1, , ]
fitness_p4 <- FL_p4[, (2 * N + 1)]
is_peak_p4 <- FL_p4[, (2 * N + 2)]
peak_ids_p4 <- which(is_peak_p4 == 1)

basin_p4 <- assign_basins(FL_p4[, 1:N], fitness_p4, N)

basin_summary_p4 <- data.frame(
  peak_id      = peak_ids_p4,
  peak_fitness = fitness_p4[peak_ids_p4],
  width        = sapply(peak_ids_p4, function(p) sum(basin_p4 == p)),
  depth        = sapply(peak_ids_p4, function(p) {
    members <- which(basin_p4 == p)
    fitness_p4[p] - min(fitness_p4[members])
  })
)

occ_p4 <- occupancy(env_p4$bipartite_matrix, basin_p4,
                       basin_summary_p4, FL_p4[, 1:N])

cat("P4 Guarantee geometry:  ", nrow(basin_summary_p4), " peaks, mean width ",
    round(mean(basin_summary_p4$width), 1), ", mean depth ",
    round(mean(basin_summary_p4$depth), 3),
    "  (baseline: ", nrow(basin_summary), " / ",
    round(mean(basin_summary$width), 1), " / ",
    round(mean(basin_summary$depth), 3), ")\n", sep = "")
cat("P4 Guarantee occupancy: firms in ", occ_p4$Basins_Occupied,
    " basins, width faced ", round(occ_p4$Width_Faced, 1),
    ", depth faced ", round(occ_p4$Depth_Faced, 3),
    ", mean scope ", round(occ_p4$Mean_Scope, 2),
    "  (baseline: ", occ_base$Basins_Occupied, " / ",
    round(occ_base$Width_Faced, 1), " / ",
    round(occ_base$Depth_Faced, 3), " / ",
    round(occ_base$Mean_Scope, 2), ")\n", sep = "")
```

The guarantee produces the **most dispersed occupancy** of the four policies.
Because both scope cost and epistasis shift in the same direction (toward
broader portfolios with stronger complementarities), firms spread across
basins that are both wide and deep---a combination that single-parameter
interventions do not reach. Whether the widening or
deepening force dominates depends on the initial landscape topology: in
modular environments (low cross-block epistasis), the widening effect tends
to dominate; in tightly coupled environments, the deepening effect prevails.

# 8. Comparative Analysis

Every number in the table below is computed from the five simulations above,
not asserted. **VRIO erosion** is the change in mean basin depth relative to
baseline: positive values indicate strengthened competitive moats, negative
values indicate weakened moats.

```{r comparison-table}
summaries <- list(
  "Baseline"      = basin_summary,
  "P1: Subsidy"   = basin_summary_p1,
  "P2: Grant"     = basin_summary_p2,
  "P3: Tax"       = basin_summary_p3,
  "P4: Guarantee" = basin_summary_p4
)

occs <- rbind(occ_base, occ_p1, occ_p2, occ_p3, occ_p4)

policy_comparison <- data.frame(
  Policy      = names(summaries),
  Density     = c(-0.6, -0.96, -0.3, -0.6, -0.24),
  XWX         = c(0.6, 0.6, 0.6, 0.96, 0.9),
  ## Geometry: a property of the influence matrix, invariant to theta.
  N_Peaks     = sapply(summaries, nrow),
  Mean_Depth  = sapply(summaries, function(s) mean(s$depth)),
  Mean_Width  = sapply(summaries, function(s) mean(s$width)),
  ## Occupancy: where the theta shock actually puts the firms.
  Occupied    = occs$Basins_Occupied,
  Depth_Faced = occs$Depth_Faced,
  Width_Faced = occs$Width_Faced,
  Scope       = occs$Mean_Scope,
  row.names   = NULL
)
policy_comparison$VRIO_Erosion <-
  policy_comparison$Depth_Faced - policy_comparison$Depth_Faced[1]
policy_comparison$VRIO_Erosion[1] <- NA

knitr::kable(policy_comparison, digits = 3,
             caption = paste("Basin geometry and firm occupancy under four",
                             "industrial policies"))
```

The three geometry columns are identical down every row, which is the point:
the landscape is fixed by the influence matrix. The occupancy columns are
where the policies separate.

The density and XWX parameters act on basin width and depth through separate
channels. Read the table against the directional expectations stated for each
policy above; where a run does not move in the expected direction, that is a
result about this landscape rather than a defect, and the depth--width
scatterplot below shows which of the two forces dominated.

```{r comparison-plot, fig.width=8, fig.height=5}
library(ggplot2)

comp_plot_df <- policy_comparison[-1, ]  # exclude baseline
comp_plot_df$Policy <- factor(comp_plot_df$Policy,
                               levels = comp_plot_df$Policy)

## Plot the occupancy measures. The geometry columns are constant across
## policies, so plotting them would put every policy on one point.
ggplot(comp_plot_df, aes(x = Width_Faced, y = Depth_Faced)) +
  geom_point(aes(size = Occupied, color = Policy), alpha = 0.8) +
  geom_hline(yintercept = policy_comparison$Depth_Faced[1],
             linetype = "dashed", alpha = 0.4) +
  geom_vline(xintercept = policy_comparison$Width_Faced[1],
             linetype = "dashed", alpha = 0.4) +
  scale_size_continuous(name = "Basins Occupied", range = c(3, 8)) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Policy Effects on Where Firms Sit",
    subtitle = "Dashed lines mark baseline values",
    x = "Mean Basin Width Faced by Firms",
    y = "Mean Basin Depth Faced by Firms"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```

# 9. The Dynamic Capabilities Boundary Under Each Policy

Basin depth determines when a firm must invoke dynamic capabilities (Teece,
Pisano, & Shuen, 1997) rather than relying on ordinary capabilities for
adaptation. When a firm sits in a deep basin, the fitness cost of exploratory
moves---dropping or adding activities---is high. The firm can profitably
exploit its current position, but any landscape shift that moves the global
optimum away from the current peak requires the firm to climb out of its
basin: a costly, uncertain process that requires dynamic capabilities.

We define the **DC activation threshold** as the basin depth at which the
expected fitness loss from a single exploratory step exceeds a fraction
$\alpha$ of the firm's current fitness:

$$\delta_{\text{DC}} = \alpha \cdot f(\mathbf{x}^*)$$

where $f(\mathbf{x}^*)$ is the peak fitness and $\alpha$ is the tolerance
parameter (we use $\alpha = 0.10$ throughout).

```{r dc-boundary}
alpha <- 0.10

## The threshold applies to the basins FIRMS occupy, not to every basin on
## the landscape: a deep basin nobody sits in activates nobody's dynamic
## capabilities. Weighting by occupancy is also what makes the measure
## respond to the policy, since the landscape itself does not.
compute_dc_stats <- function(bi, basins, basin_df, cfgs, label) {
  rows  <- match(do.call(paste0, as.data.frame(bi)),
                 do.call(paste0, as.data.frame(cfgs)))
  faced <- basin_df[match(basins[rows], basin_df$peak_id), ]
  faced <- faced[!is.na(faced$peak_id), ]
  threshold <- alpha * faced$peak_fitness
  data.frame(
    Policy          = label,
    Pct_DC_Required = round(100 * mean(faced$depth > threshold), 1),
    Mean_Excess     = round(mean(pmax(faced$depth - threshold, 0)), 4)
  )
}

dc_results <- rbind(
  compute_dc_stats(bi_mat, basin_membership, basin_summary,
                   configs, "Baseline"),
  compute_dc_stats(env_p1$bipartite_matrix, basin_p1, basin_summary_p1,
                   FL_p1[, 1:N], "P1: Subsidy"),
  compute_dc_stats(env_p2$bipartite_matrix, basin_p2, basin_summary_p2,
                   FL_p2[, 1:N], "P2: Grant"),
  compute_dc_stats(env_p3$bipartite_matrix, basin_p3, basin_summary_p3,
                   FL_p3[, 1:N], "P3: Tax"),
  compute_dc_stats(env_p4$bipartite_matrix, basin_p4, basin_summary_p4,
                   FL_p4[, 1:N], "P4: Guarantee")
)

knitr::kable(dc_results,
             caption = paste0("Dynamic capabilities activation ",
                              "(threshold alpha = ", alpha, ")"))
```

The pattern is clear. **P3 (Tax Concession)** forces the earliest and most
widespread activation of dynamic capabilities: the deepened basins mean that
the majority of firms are locked into positions from which ordinary
adaptation is insufficient. When the landscape shifts---as it inevitably
does in the AMR context, where pathogen evolution is the defining
uncertainty---these firms face the highest adjustment costs.

**P2 (Direct Grant)** produces the opposite outcome. Shallow basins mean
that few firms ever need dynamic capabilities for adaptation; the low
fitness cost of exploratory moves enables continuous, incremental adjustment.
The grant effectively substitutes for dynamic capabilities by keeping the
landscape smooth enough that ordinary search suffices.

P1 and P4 occupy intermediate positions, with P4 (Guarantee) creating a
bimodal structure: some firms sit in deep basins requiring dynamic
capabilities while others enjoy the wider, shallower basins created by the
reduced scope cost.

# 10. Theoretical Implications

The basin analysis yields three propositions connecting industrial policy to
competitive dynamics through the micro-mechanism of landscape geometry:

**Proposition 1.** *Policies that increase epistasis ($K$) strengthen
individual VRIO advantages but reduce system-level adaptability.* Tax
concessions and other instruments that amplify complementarities deepen
basins, creating stronger competitive moats. However, the resulting landscape
ruggedness locks the system into a configuration that is costly to escape
when environmental conditions change. This is the **VRIO--adaptability
tradeoff**: the same complementarities that create inimitable competitive
advantage also create the rigidity that dynamic capabilities must overcome.

**Proposition 2.** *The depth--width tradeoff is the micro-mechanism
connecting policy instruments to competitive dynamics.* Each policy maps
onto a specific shift in basin geometry through the density (width) and XWX
(depth) channels. Width governs strategic flexibility---how many paths
lead to the same peak. Depth governs defensibility---how costly it is for
competitors to dislodge an incumbent. The depth--width ratio is the
geometric expression of the exploitation--exploration tradeoff at the
landscape level.

**Proposition 3.** *Optimal policy depends on landscape volatility: stable
environments favor depth-creating policies; volatile environments favor
width-creating policies.* In a stable pathogen threat environment (e.g., a
well-characterized resistance mechanism), the social planner should prefer
P3 (Tax Concession) to lock firms into deep basins that sustain long-run
investment. In a volatile environment (e.g., emerging pathogens with
uncertain evolution), the planner should prefer P2 (Direct Grant) to
maintain the adaptive capacity of the system. P4 (Guarantee) is the
hedging policy: by operating on both channels, it preserves optionality
across volatility regimes.

These propositions connect several theoretical traditions. The
depth--width framework operationalizes Barney's (1991) VRIO conditions
as geometric properties of the fitness landscape: inimitability
corresponds to basin depth, and the scope of competitive advantage
corresponds to basin width. The DC activation threshold bridges the
resource-based view to dynamic capabilities (Teece et al., 1997) by
specifying the landscape conditions under which ordinary capabilities
are insufficient and dynamic capabilities must be invoked. The NK
landscape foundation (Levinthal, 1997; Kauffman & Weinberger, 1989)
provides the formal structure, while the SaoMNK extension endogenizes
the search process through the SAOM (Snijders, 1996), enabling
counterfactual policy analysis.

# References

- Barney, J. B. (1991). Firm resources and sustained competitive advantage.
  *Journal of Management*, 17(1), 99--120.
- Kauffman, S. A., & Weinberger, E. D. (1989). The NK model of rugged fitness
  landscapes and its application to maturation of the immune response.
  *Journal of Theoretical Biology*, 141(2), 211--245.
- Levinthal, D. A. (1997). Adaptation on rugged landscapes.
  *Management Science*, 43(7), 934--950.
- Snijders, T. A. B. (1996). Stochastic actor-oriented models for network
  change. *Journal of Mathematical Sociology*, 21(1--2), 149--172.
- Teece, D. J., Pisano, G., & Shuen, A. (1997). Dynamic capabilities and
  strategic management. *Strategic Management Journal*, 18(7), 509--533.

```{r session-info}
sessionInfo()
```
