---
title: "Basin Geometry and the Depth-Width Tradeoff in searchnet"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Basin Geometry and the Depth-Width Tradeoff in searchnet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  eval = TRUE,
  fig.width = 7,
  fig.height = 5
)
```

# 1. Introduction

An NK fitness landscape assigns a scalar fitness value to every binary
configuration of $N$ components.  Think of a mountain range: each point on the
terrain is a particular portfolio of strategic activities ($N$ binary choices),
and the elevation at that point is the portfolio's fitness.  The ruggedness of
the terrain---how many peaks there are and how jagged the ridges between
them---is governed by $K$, the degree of epistatic interdependence among
components (Kauffman, 1993).

Firms navigate this terrain through local search: evaluating neighboring
configurations that differ from the current position by a single component flip,
and moving to any neighbor that offers strictly higher fitness.  Under this
greedy improvement rule, every starting configuration eventually converges to
some local peak---and the set of all configurations that converge to the same
peak is that peak's **basin of attraction** (Levinthal, 1997).

Basins of attraction are the fundamental structural units of a fitness
landscape.  They partition the full $2^N$ configuration space into non-overlapping
regions, each anchored by a local optimum.  The geometry of these basins---how
deep they are, how wide they are, and how they reshape under shocks---connects
directly to core concepts in strategic management: competitive moats (RBV),
strategic flexibility, and the boundary between exploitation and exploration
(dynamic capabilities).

This vignette introduces the formal definitions of basin depth and width,
demonstrates their computation in `searchnet`, and develops the depth-width
tradeoff as an analytical lens for understanding how landscape complexity
shapes competitive advantage.

# 2. Computing the Fitness Landscape

We begin by constructing an NK fitness landscape using `searchnet`'s
`SaomNkRSienaBiEnv` class.  The `compute_fitness_landscape` method evaluates
the utility of all $2^N$ counterfactual component configurations, producing a
three-dimensional array with dimensions `[n_landscapes, 2^N, 2*N+2]`.

The columns of the landscape array are organized as follows:

- Columns `1:N` --- the binary configuration vector
- Columns `(N+1):(2*N)` --- per-component fitness contributions
- Column `2*N+1` --- total fitness (mean of per-component contributions)
- Column `2*N+2` --- local peak indicator (1 if no single-flip neighbor has
  strictly higher fitness, 0 otherwise)

```{r compute-landscape}
library(searchnet)

# Create an 8-actor, 8-component environment
env <- SaomNkRSienaBiEnv$new(list(M = 8, N = 8, BI_PROB = 0.4, rand_seed = 42))

# Compute the fitness landscape (one replicate for exposition)
env$compute_fitness_landscape(n_landscapes = 1)

landscape <- env$fitness_landscape[1, , ]
N <- 8
configs  <- landscape[, 1:N]
fitness  <- landscape[, 2 * N + 1]
is_peak  <- landscape[, 2 * N + 2] == 1
n_peaks  <- sum(is_peak)
cat(sprintf("Landscape has %d local peaks out of %d configurations\n",
            n_peaks, nrow(configs)))
```

The number of peaks increases with $K$: when $K = 0$ (no epistasis) there is a
single global peak; when $K = N - 1$ (full epistasis) the landscape is maximally
rugged, with the expected number of peaks scaling roughly as
$2^N / (N + 1)$ (Kauffman, 1993).

# 3. Basin Depth: The Competitive Moat

## 3.1 Definition

Basin depth measures the fitness gap between a peak and the worst-performing
configuration in its basin:

$$\text{Depth}(p) = F(p) - \min_{x \in B(p)} F(x)$$

where $B(p)$ is the set of all configurations that converge to peak $p$ under
greedy single-flip improvement.  Equivalently, depth can be defined relative to
the basin boundary:

$$\text{Depth}_{\partial}(p) = F(p) - \min_{x \in \partial B(p)} F(x)$$

where $\partial B(p)$ is the set of configurations one Hamming flip from leaving
basin $B(p)$.  Both formulations capture the same strategic intuition.

## 3.2 Strategic Interpretation

Depth is the formal analogue of a **competitive moat** in the resource-based
view (RBV).  A deep basin means that partial imitation---adopting some but not
all of a rival's activity choices---produces substantially lower fitness than
the fully integrated configuration at the peak.  Competitors who approximate
the peak strategy sit far below it on the fitness surface, making the
incumbent's position difficult to replicate.

Consider Apple's integrated ecosystem as an industry example.  Hardware design,
proprietary silicon, operating system, and services are tightly coupled.
Removing any single component---as Samsung does with a hardware-only
approach---drops fitness significantly.  The basin surrounding Apple's peak
configuration is deep: the moat is wide in fitness terms, even if relatively
few alternative configurations converge to the same optimum.

## 3.3 Computing Basin Depth

To compute depth, we first assign every configuration to the basin of its
nearest peak (by Hamming distance), then measure the fitness gap from peak to
basin floor:

```{r basin-assignment}
# Hamming distance between two binary vectors
hamming_dist <- function(a, b) sum(abs(a - b))

peak_configs  <- configs[is_peak, , drop = FALSE]
peak_fitness  <- fitness[is_peak]

# Assign each configuration to the basin of its nearest peak
assign_basin <- function(config, peaks) {
  dists <- apply(peaks, 1, function(p) hamming_dist(config, p))
  which.min(dists)
}
basin_ids <- apply(configs, 1, assign_basin, peaks = peak_configs)
```

```{r compute-depth}
# Compute depth for each basin
basin_depth <- numeric(length(peak_fitness))
for (b in seq_along(peak_fitness)) {
  members <- which(basin_ids == b)
  if (length(members) > 1) {
    basin_depth[b] <- peak_fitness[b] - min(fitness[members])
  } else {
    basin_depth[b] <- 0
  }
}

cat(sprintf("Basin depths: %s\n",
            paste(round(basin_depth, 3), collapse = ", ")))
```

# 4. Basin Width: Strategic Flexibility

## 4.1 Definition

Basin width captures the average Hamming distance from a basin's constituent
configurations to its peak:

$$\text{Width}(p) = \frac{1}{|B(p)|} \sum_{x \in B(p)} d_H(x, p)$$

Width can also be measured simply as basin cardinality $|B(p)|$ (the number of
configurations in the basin), but the mean-distance formulation carries more
strategic content: it tells us not just *how many* paths converge to the peak,
but *how different* they are from the peak configuration itself.

## 4.2 Strategic Interpretation

Width is the formal analogue of **strategic flexibility**.  A wide basin means
that many distinct activity configurations---including some that look very
different from the peak---nevertheless converge to the same local optimum
through greedy improvement.  Organizations in wide basins can tolerate
substantial deviation from the ideal configuration and still reach a
high-fitness outcome.

The contrast between McKinsey and a boutique consulting firm illustrates the
concept.  McKinsey's peak admits many practice-area and geographic
configurations that all converge to a viable general-management-consulting
model: the basin is wide.  A boutique firm occupying a narrow niche---say,
pharmaceutical regulatory strategy in the EU---has a peak that only a handful
of very specific activity configurations can reach.  The basin is narrow:
any deviation from the precise niche destroys the fitness advantage.

## 4.3 Computing Basin Width

```{r compute-width}
# Mean Hamming distance from each basin's members to its peak
basin_width <- numeric(length(peak_fitness))
for (b in seq_along(peak_fitness)) {
  members <- which(basin_ids == b)
  member_configs <- configs[members, , drop = FALSE]
  dists <- apply(member_configs, 1, function(x) hamming_dist(x, peak_configs[b, ]))
  basin_width[b] <- mean(dists)
}

cat(sprintf("Basin widths: %s\n",
            paste(round(basin_width, 3), collapse = ", ")))
```

# 5. The Depth-Width Tradeoff

## 5.1 The Core Result

As epistasis $K$ increases, the fitness landscape becomes more rugged: peaks
proliferate, and each basin contracts.  This produces a fundamental tradeoff
between depth and width:

- **High $K$** $\rightarrow$ deep, narrow basins $\rightarrow$ strong moats but
  low flexibility (the **exploitation trap**)
- **Low $K$** $\rightarrow$ shallow, wide basins $\rightarrow$ weak moats but
  high flexibility

The scaling relationships below are the heuristic, large-$N$ Kauffman-style
expectation under uniformly random epistasis; this vignette computes a single
landscape (Section 2) and one two-point structural comparison (Section 6), not
a systematic sweep over $K$, so the proportionality itself is not demonstrated
by anything computed here:

$$E[\text{Depth}] \propto K/N, \quad E[\text{Width}] \propto N/K$$

Indeed, the two structural comparisons actually computed in this vignette do
not follow the monotone "more coupling $\Rightarrow$ more peaks" intuition
cleanly: block-diagonal $W$ with four size-4 blocks (Section 6's `W_high`,
full coupling within each size-4 block) produced *fewer* peaks (2) than four
size-2 blocks (`W_low`, 9 peaks) despite denser within-block coupling,
because block-diagonal epistasis concentrates ruggedness within blocks rather
than driving the whole-landscape peak count up.  The depth side is more
consistent with the intuition (denser blocks gave deeper basins here), but a
single comparison at $N = 8$ is not evidence for the $K/N$ scaling exponent.

High-epistasis environments (e.g., technology platforms with many tightly coupled
subsystems) produce landscapes where a firm that finds a good configuration
enjoys a deep moat---but sits in a narrow basin that leaves little room for
strategic adjustment when the landscape shifts.  Low-epistasis environments
(e.g., regulated utilities with modular, substitutable assets) produce shallow
basins that offer minimal competitive protection but permit easy reconfiguration.

## 5.2 Visualization

The depth-width scatterplot provides a diagnostic view of the tradeoff across
all basins in a landscape:

```{r depth-width-plot, fig.width=7, fig.height=6}
# Depth-width scatterplot in the package style (theme_searchnet())
library(ggplot2)
dw <- data.frame(width = basin_width, depth = basin_depth,
                 peak = seq_along(peak_fitness), fitness = peak_fitness)
ggplot(dw, aes(width, depth)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_point(aes(color = fitness), size = 4, alpha = 0.85) +
  geom_text(aes(label = peak), size = 2.6, vjust = -1.1, color = "grey25") +
  annotate("text", x = max(basin_width) * 0.2, y = max(basin_depth) * 0.9,
           label = "Strong VRIO\nLow DC", size = 3.2, hjust = 0, color = "grey40") +
  annotate("text", x = max(basin_width) * 0.8, y = max(basin_depth) * 0.1,
           label = "Weak VRIO\nHigh DC", size = 3.2, color = "grey40") +
  scale_color_gradient(low = searchnet_palette()[["sky"]], high = "#023858",
                       name = "Peak fitness F") +
  labs(title = "Depth-Width Tradeoff in NK Landscape Basins",
       subtitle = "Each point is one local peak, labeled by its index",
       x = "Basin Width (mean Hamming distance from peak)",
       y = "Basin Depth (fitness gap to basin floor)") +
  theme_searchnet()
```

The upper-left quadrant (deep, narrow) corresponds to strong VRIO-based
competitive advantage with limited dynamic capabilities---the classic
exploitation position.  The lower-right quadrant (shallow, wide) corresponds to
weak moats but high strategic flexibility---the exploration-ready position.

## 5.3 Industry Implications

The tradeoff connects directly to patterns of disruption.  Technology firms
(high $K$) occupy deep narrow basins.  When the landscape shifts---as it did
with the smartphone revolution for Nokia, or with streaming for Blockbuster---
firms in deep narrow basins lack the flexibility to reconfigure.  Their moat
was real, but it was a moat around a position that the landscape eroded away.

Utilities and commodities firms (low $K$) occupy shallow wide basins.  Their
advantage is never dramatic, but they can absorb regulatory changes and
technology shifts without existential risk.  The cost of low epistasis is that
no configuration is dramatically better than its neighbors.

# 6. Landscape Erosion Under Shocks

## 6.1 Definition

When an exogenous shock modifies the epistasis structure---changing the matrix
$W$ that governs component interdependencies---basin geometry reshapes.  We
define **erosion** as the change in basin depth:

$$\text{Erosion}(p) = \text{Depth}_{W_0}(p) - \text{Depth}_{W_1}(p)$$

where $W_0$ and $W_1$ are the pre-shock and post-shock influence matrices.
Positive erosion means the moat has weakened; negative erosion means the shock
has deepened the basin (rare but possible when the shock reinforces existing
complementarities).

## 6.2 Strategic Interpretation

Erosion formalizes the destruction of competitive advantage through
environmental change.  Blockbuster's basin in the pre-streaming landscape was
deep: physical inventory management, store location optimization, and
late-fee revenue were tightly coupled in a configuration that resisted partial
imitation.  The streaming shock fundamentally altered the $W$-matrix, converting
positive complementarities (physical inventory $\times$ store location) into
negative ones.  The basin did not merely become shallower---it disappeared
entirely, with Blockbuster's former peak becoming a valley in the new landscape.

## 6.3 Computing Erosion in searchnet

In `searchnet`, shocks are implemented as changes to the `coDyadCovars`
specification in the structure model.  To compute erosion, we evaluate the
fitness landscape under both the pre-shock and post-shock influence matrices:

```{r erosion-setup}
# Pre-shock environment with high epistasis
env_pre <- SaomNkRSienaBiEnv$new(list(M = 8, N = 8, BI_PROB = 0.4, rand_seed = 42))

# Dense influence matrix (high K)
W_high <- create_block_diag(8, 4)

structure_model_pre <- list(
  dv_bipartite = list(
    name = "self$bipartite_rsienaDV",
    effects = list(
      list(effect = "density", parameter = -0.6,
           dv_name = "self$bipartite_rsienaDV", fix = TRUE)
    ),
    coDyadCovars = list(
      list(effect = "XWX", parameter = 0.3,
           dv_name = "self$bipartite_rsienaDV", fix = TRUE,
           nodeSet = c("COMPONENTS", "COMPONENTS"),
           interaction1 = "self$component_1_coDyadCovar",
           x = W_high)
    )
  )
)

# Post-shock: sparser epistasis (some complementarities destroyed)
W_low <- create_block_diag(8, 2)

structure_model_post <- list(
  dv_bipartite = list(
    name = "self$bipartite_rsienaDV",
    effects = list(
      list(effect = "density", parameter = -0.6,
           dv_name = "self$bipartite_rsienaDV", fix = TRUE)
    ),
    coDyadCovars = list(
      list(effect = "XWX", parameter = 0.3,
           dv_name = "self$bipartite_rsienaDV", fix = TRUE,
           nodeSet = c("COMPONENTS", "COMPONENTS"),
           interaction1 = "self$component_1_coDyadCovar",
           x = W_low)
    )
  )
)
```

```{r erosion-compute}
# `compute_fitness_landscape()` builds the static NK landscape from the
# STRUCTURE of the interdependency matrix stored in
# `component_1_coDyadCovar` (or, absent that, `search_matrix`); it never
# reads RSiena effect parameters/theta.  The `structure_model_pre` /
# `structure_model_post` lists above declare `XWX` coefficients for use in
# a behavioral `run_simulation()`, but that coefficient plays no role
# here -- what reshapes the *static* landscape is the interdependency
# matrix itself, so we assign `W_high` / `W_low` directly before computing
# each landscape.
env_pre$component_1_coDyadCovar <- W_high
env_pre$compute_fitness_landscape(n_landscapes = 1)
landscape_pre <- env_pre$fitness_landscape[1, , ]

env_post <- SaomNkRSienaBiEnv$new(list(M = 8, N = 8, BI_PROB = 0.4, rand_seed = 42))
env_post$component_1_coDyadCovar <- W_low
env_post$compute_fitness_landscape(n_landscapes = 1)
landscape_post <- env_post$fitness_landscape[1, , ]

# Extract fitness and peak indicators
fitness_pre  <- landscape_pre[, 2 * N + 1]
is_peak_pre  <- landscape_pre[, 2 * N + 2] == 1

fitness_post <- landscape_post[, 2 * N + 1]
is_peak_post <- landscape_post[, 2 * N + 2] == 1

# Compute basin depth under both regimes for pre-shock peaks
peak_configs_pre <- configs[is_peak_pre, , drop = FALSE]
peak_fitness_pre <- fitness_pre[is_peak_pre]

basin_ids_pre  <- apply(configs, 1, assign_basin, peaks = peak_configs_pre)
basin_ids_post <- apply(configs, 1, assign_basin, peaks = peak_configs_pre)

depth_pre  <- numeric(length(peak_fitness_pre))
depth_post <- numeric(length(peak_fitness_pre))
for (b in seq_along(peak_fitness_pre)) {
  members_pre  <- which(basin_ids_pre == b)
  members_post <- which(basin_ids_post == b)
  if (length(members_pre) > 1)
    depth_pre[b]  <- peak_fitness_pre[b] - min(fitness_pre[members_pre])
  if (length(members_post) > 1)
    depth_post[b] <- fitness_post[is_peak_pre][b] - min(fitness_post[members_post])
}

erosion <- depth_pre - depth_post
cat(sprintf("Mean VRIO erosion: %.3f\n", mean(erosion)))
cat(sprintf("Max erosion:       %.3f (Peak %d)\n", max(erosion), which.max(erosion)))
```

Going from the denser interdependency structure (`W_high`, block size 4) to the
sparser one (`W_low`, block size 2) produces positive mean erosion here: both
pre-shock peaks lose depth, not the near-total collapse the Blockbuster analogy
above describes.  This toy example moves the interdependency *structure* itself
(which components are coupled), which is the lever `compute_fitness_landscape()`
actually responds to.  Changing only an `XWX` coefficient value, with the same
`x` matrix, leaves this static landscape and its basin depths unchanged.

# 7. RBV to Dynamic Capabilities: When to Jump

The depth-width tradeoff provides a formal criterion for the boundary between
exploitation and exploration.  An organization should switch from **exploitation**
(deepening its current basin through incremental refinement) to **exploration**
(jumping to a new basin) when the rate of erosion exceeds the marginal returns
to continued exploitation:

$$\frac{d(\text{Depth})}{dt} < -\frac{\text{marginal fitness gain from exploitation}}{\text{remaining exploitation horizon}}$$

This inequality says: if the moat is eroding faster than the firm can deepen it
through incremental improvement, the rational strategy is to abandon the current
peak and search for a new one.  In strategy terms, this is the transition from
RBV-style competitive advantage to dynamic-capabilities-style reconfiguration
(Teece, Pisano, & Shuen, 1997).

IBM illustrates the pattern across three landscape shifts.  In the mainframe
era, IBM occupied a deep basin with strong complementarities among hardware
manufacturing, proprietary architecture, and maintenance contracts.  When
commoditization eroded that basin, IBM jumped to IT services---a new peak in
a reshaped landscape.  When cloud computing eroded the services basin, IBM
jumped again.  Each transition was a basin-to-basin leap triggered by erosion
exceeding the returns to exploitation.

The formal criterion can be computed in `searchnet` by tracking basin depth
across simulation steps in a `theta_shocks` schedule and comparing the depth
trajectory to the within-basin fitness improvement rate.

# 8. Summary

Basin geometry provides a rigorous formal language for core concepts in
strategic management:

| Concept               | Basin Measure      | Strategic Meaning             |
|:----------------------|:-------------------|:------------------------------|
| Competitive moat      | Depth              | Fitness gap to partial imitators |
| Strategic flexibility | Width              | Variety of convergent paths   |
| Disruption            | Erosion            | Depth loss under $W$-shock    |
| DC boundary           | Erosion rate       | When to jump basins           |
| Policy effect         | $\Delta$(Depth, Width) | How interventions reshape competition |

The `searchnet` package makes these measures computable from first principles,
grounding the verbal theory of competitive advantage in the formal mathematics
of fitness landscape geometry.

# References

- Kauffman, S. A. (1993). *The Origins of Order: Self-Organization and
  Selection in Evolution*. Oxford University Press.
- Levinthal, D. A. (1997). Adaptation on rugged landscapes.
  *Management Science*, 43(7), 934--950.
- Rivkin, J. W. (2000). Imitation of complex strategies.
  *Management Science*, 46(6), 824--844.
- Siggelkow, N. (2002). Evolution toward fit. *Administrative Science
  Quarterly*, 47(1), 125--159.
- Teece, D. J., Pisano, G., & Shuen, A. (1997). Dynamic capabilities and
  strategic management. *Strategic Management Journal*, 18(7), 509--533.

```{r session-info, eval=TRUE}
sessionInfo()
```
