---
title: "searchnet Workshop: Building Your First Search Simulation"
subtitle: "A Professional Development Workshop for Strategy & Organization Scholars"
author: "Stephen Downing, University of Missouri"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{PDW Workshop: Building Your First Search Simulation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7
)
```

```{r load-saomnk, include=FALSE}
if (requireNamespace("searchnet", quietly = TRUE)) {
  library(searchnet)
} else {
  dir_r <- file.path(getwd(), "R")
  if (!file.exists(file.path(dir_r, "utils.R")))
    dir_r <- "D:/Search_networks/SaoMNK/R"
  .saomnk_dir <- dir_r
  source(file.path(dirname(dir_r), "inst", "saomnk-loader.R"))
}

library(Matrix)
```

# PART 1: THE SIMULATION WORLD --- What Are We Building? {.tabset}

## The Big Picture: A Factory for "What If?" Questions

Imagine you are building a **miniature economy in a bottle**. Inside that
bottle, firms make choices---entering markets, adopting technologies,
forming alliances---and those choices reshape the competitive landscape
for everyone else. The `searchnet` package lets you build that bottle,
set the rules, shake it up, and watch what happens.

**Why simulate?** Because theory alone gives us propositions, and data
alone gives us correlations, but simulation gives us a *laboratory*
where we can run controlled experiments on complex systems. Each
simulation run is an **alternate reality** with different rules. Change
one parameter---say, how costly diversification is---and compare the
outcomes. That is causal inference by construction.

The simulation multiverse works like this:

1. **Design** a world (firms, activities, initial conditions)
2. **Set the rules** (what drives firm behavior?)
3. **Run** the simulation (let firms make choices over time)
4. **Observe** emergent patterns (who ends up where, and why?)
5. **Compare** across worlds (change one rule, see what shifts)

> **For the skeptical theorist:** Every parameter in `searchnet` maps to
> a testable mechanism. The simulation does not *replace* formal theory;
> it *extends* it into territory where closed-form solutions do not
> exist.

## The Cast of Characters

Every `searchnet` simulation has three building blocks: **actors**,
**components**, and the **ties** between them. Together, they form a
bipartite network---the dependent variable of the entire system.

### Actors (Firms / People / Agents)

Think of $M$ firms competing in an industry. Each firm decides which
activities to pursue and which to abandon. In network language, actors
are the *row nodes* of a bipartite graph.

- **Conceptual:** $M$ strategic decision-makers
- **Technical:** $M$ rows of a binary matrix
  $\mathbf{B} \in \{0,1\}^{M \times N}$
- **Example:** $M = 6$ airlines deciding which routes to fly

### Components (Activities / Resources / Technologies)

Think of $N$ possible activities, products, or technologies. These are
the choices available to every firm. In network language, components are
the *column nodes*.

- **Conceptual:** $N$ discrete strategic options
- **Technical:** $N$ columns of the bipartite matrix
- **Example:** $N = 10$ possible route markets

### Ties (Affiliations / Choices)

A $1$ means the firm does that activity; a $0$ means it does not. This
is the **bipartite adjacency matrix**---the state of the world at any
point in time.

- **Conceptual:** "Airline $i$ flies route $j$" means $b_{ij} = 1$
- **Technical:** $b_{ij} \in \{0,1\}$ for all $i \in \{1,...,M\}$,
  $j \in \{1,...,N\}$

Let us build our first world:

```{r first-world}
# Step 1: Create a world with 6 actors and 10 components
# density = 0.3 means ~30% of possible ties exist at the start
env <- saomnk_env(M = 6, N = 10, density = 0.3, seed = 42)

# Look at the initial bipartite matrix
B <- env$bipartite_matrix
cat("Our world: ", env$M, "actors x", env$N, "components\n")
cat("Initial density:", round(sum(B) / (env$M * env$N), 2), "\n\n")
print(B)
```

> **Plain English:** Each row is a firm. Each column is an activity. A 1
> means "this firm does this activity." We just created a small industry
> with 6 firms and 10 possible activities, where about 30% of
> firm-activity pairs are active.

## The Four Hidden Networks (The {K} Framework)

Here is the key insight that motivates `searchnet`: from **one**
bipartite matrix, **four** networks emerge automatically. You do not
have to construct them---they are mathematically implied by the pattern
of ties. Together, these four *degree* measures form the $\{K\}$
framework.

### K_AC --- Actor Scope ("How much is each firm doing?")

Count the activities per firm. Broad firms have high $K_{AC}$;
specialists have low $K_{AC}$.

- **Technical:** $K_{AC_i} = \sum_j b_{ij} = \text{rowSums}(\mathbf{B})$
- **In strategy terms:** This is *diversification* or *scope*

```{r k-ac}
cat("K_AC (actor scope):", rowSums(B), "\n")
cat("  -> Actor 1 does", rowSums(B)[1], "activities\n")
cat("  -> Actor", which.max(rowSums(B)), "is the most diversified\n")
```

### K_CA --- Component Popularity ("How crowded is each activity?")

Count the firms doing each activity. Popular activities are crowded;
niche activities are uncrowded.

- **Technical:**
  $K_{CA_j} = \sum_i b_{ij} = \text{colSums}(\mathbf{B})$
- **In strategy terms:** This is *competitive density* or *crowding*

```{r k-ca}
cat("K_CA (component popularity):", colSums(B), "\n")
cat("  -> Activity", which.max(colSums(B)), "is the most popular\n")
cat("  -> Activity", which.min(colSums(B)), "is the most niche\n")
```

### K_AA --- Actor Sociality ("Who overlaps with whom?")

Two firms overlap if they do the same activities. The actor projection
$\mathbf{S} = \mathbf{B}\mathbf{B}^T$ reveals "who competes with whom."

- **Technical:**
  $K_{AA_i} = \sum_{j \neq i} \mathbf{1}[S_{ij} > 0]$ where
  $\mathbf{S} = \mathbf{B}\mathbf{B}^T$
- **In strategy terms:** This is *strategic group overlap*

```{r k-aa}
S <- (B > 0) %*% t(B > 0)
diag(S) <- 0
cat("K_AA (actor sociality):", rowSums(S > 0), "\n")
cat("  -> Actor", which.max(rowSums(S > 0)), "overlaps with the most rivals\n")
```

### K_CC --- Component Epistasis ("Which activities go together?")

Two activities "go together" if the same firms tend to do both. The
component projection
$\mathbf{E} = \mathbf{B}^T\mathbf{B}$ reveals "natural bundles."

- **Technical:**
  $K_{CC_j} = \sum_{k \neq j} \mathbf{1}[E_{jk} > 0]$ where
  $\mathbf{E} = \mathbf{B}^T\mathbf{B}$
- **In strategy terms:** This is *complementarity* or *epistasis*---the
  $K$ from the classical NK model

```{r k-cc}
E <- t(B > 0) %*% (B > 0)
diag(E) <- 0
cat("K_CC (component epistasis):", rowSums(E > 0), "\n")
cat("  -> This is the 'K' from classical NK models!\n")
```

### Seeing All Four Together

```{r k-four-summary}
cat("============================================\n")
cat("  THE {K} FRAMEWORK: Four Coupled Degrees\n")
cat("============================================\n\n")
cat("K_AC (scope):      ", rowSums(B), "\n")
cat("K_CA (popularity): ", colSums(B), "\n")
cat("K_AA (sociality):  ", rowSums(S > 0), "\n")
cat("K_CC (epistasis):  ", rowSums(E > 0), "\n")
```

> **KEY INSIGHT:** These four numbers are **coupled**. Change one, and
> you change them all. When a firm enters a new market ($K_{AC}$ rises),
> that market becomes more crowded ($K_{CA}$ rises for that component),
> the firm overlaps with more rivals ($K_{AA}$ rises), and the
> components become more epistatically linked ($K_{CC}$ shifts). This is
> why we track all four simultaneously---and why `searchnet` exists.

## The Rules of the World

### Time: The Ministep Clock

Time in `searchnet` does not flow smoothly. Instead, **one actor at a
time** gets a chance to act. Think of it as a board game where players
take turns, but the turn order is random.

- **Technical:** Continuous-time Markov chain. At each *ministep*, one
  actor is selected stochastically
- **Analogy:** Like a board game where players take turns in random
  order

The total number of ministeps is `M * steps_per_actor`. With $M = 6$
actors and `steps_per_actor = 20`, that gives 120 individual decision
opportunities.

### The Rate Function: Who Acts Next?

Some actors may be more active than others. The rate function controls
how often each actor gets a turn.

- **Simple version:** Equal rates---each actor is equally likely to be
  selected at each ministep (the default)
- **Technical:** Actor $i$ is selected with probability
  $\lambda_i / \sum_j \lambda_j$
- **Advanced:** Rates can depend on the current state (e.g., larger
  firms act more frequently)

```{r rate-explanation}
cat("With M =", env$M, "actors and equal rates:\n")
cat("  Each actor gets selected with probability",
    round(1/env$M, 3), "per ministep\n")
cat("  With steps_per_actor = 20, total ministeps =",
    env$M * 20, "\n")
cat("  Each actor gets ~20 decision opportunities\n")
```

### The Choice Function: What Do They Do?

When it is your turn, you look at all possible moves and pick one.
Better moves are more likely, but you can make mistakes. This is the
McFadden conditional logit (multinomial choice model).

**The choice set** for actor $i$:

- **Add** activity $j$ (for any $j$ not currently held)
- **Drop** activity $j$ (for any $j$ currently held)
- **Do nothing** (status quo / pass)

**The probability** of each choice:

$$P(\text{flip } j) = \frac{\exp(\beta \cdot \Delta U_{ij})}{1 + \sum_{k=1}^{N} \exp(\beta \cdot \Delta U_{ik})}$$

where $\Delta U_{ij}$ is the *change in utility* from flipping activity
$j$ (adding it if absent, dropping it if present). The $1$ in the
denominator represents the option of doing nothing.

> **Plain English:** The actor mentally tries every possible move,
> calculates "how much better (or worse) would I be?", and picks moves
> in proportion to how much they improve things. The $\beta$ parameter
> controls how precise this choice is---more on that below.

## The Objective Function: What Makes a "Good" Move?

Each actor has a **scorecard**. Adding or dropping an activity changes
the score. The actor picks moves that improve the score---usually.

The utility function $U_i$ has multiple components. In a typical
simulation, you select which components to activate and set their
weights:

| Component | Plain English | RSiena Effect | Parameter |
|:----------|:-------------|:-------------|:----------|
| **Density** | "Doing stuff is costly" (negative = sparse) | `density` | $\theta_{\text{dens}}$ |
| **Popularity** | "Popular activities attract" | `inPop` | $\theta_{\text{pop}}$ |
| **Scope** | "Doing more begets doing more" | `outAct` | $\theta_{\text{scope}}$ |
| **Epistasis** | "Some activities complement each other" | `XWX` | $\theta_{\text{epist}}$ |
| **Strategy** | "Actor-level heterogeneity" | `egoX` | $\theta_{\text{ego}}$ |

> **Plain English:** Density is the "cost of doing business" baseline.
> Popularity captures herding (firms flock to crowded markets). Scope
> captures diversification tendency. Epistasis captures synergies
> between related activities. Strategy lets different firms have
> different behavioral tendencies.

## The Bounded Rationality Dial ($\beta$ parameter)

The $\beta$ parameter is your **rationality thermostat**:

| $\beta$ Value | Behavior | Analogy |
|:-------------|:---------|:--------|
| $\beta = 0$ | Pure random search (coin flip) | Blindfolded dart-throwing |
| $\beta \approx 1$ | Noisy optimization (realistic) | Informed but imperfect |
| $\beta \to \infty$ | Perfect optimization (greedy) | Classical NK adaptive walk |

> **Plain English:** In real life, firms are not perfectly rational---they
> experiment, make mistakes, and sometimes stumble into good strategies.
> The $\beta$ dial lets you tune how "smart" your actors are. Most
> realistic simulations live in the middle range.

# PART 2: RUNNING YOUR FIRST SIMULATION {.tabset}

## Three Steps to a Simulation

Every `searchnet` simulation follows the same three-step recipe:

```{r three-steps}
# ---- STEP 1: Create the world ----
env1 <- saomnk_env(M = 6, N = 8, density = 0.3, seed = 42)
cat("Step 1 complete: World created with",
    env1$M, "actors and", env1$N, "components\n")

# ---- STEP 2: Define the rules ----
model1 <- saomnk_model(
  density  = -0.5,       # scope is costly (negative = pressure to drop ties)
  popularity = 0.2,      # popular activities attract firms
  influence_matrix = saomnk_block_diagonal(8, 2),  # 2 clusters of 4
  influence_weight = 0.3 # synergy within clusters matters
)
cat("Step 2 complete: Model defined\n")
print(model1)

# ---- STEP 3: Run! ----
saomnk_run(env1, model1, steps_per_actor = 20, seed = 100)
cat("Step 3 complete: Simulation finished with",
    env1$get_step(), "total ministeps\n")
```

Let us unpack each step.

**Step 1** creates the world. `M = 6` actors, `N = 8` components,
initial density of 0.3 (about 30% of ties exist at the start). The
`seed` ensures reproducibility.

**Step 2** defines the rules:

- `density = -0.5` means "each additional activity carries a cost"---this
  keeps the network from becoming fully connected
- `popularity = 0.2` means "firms are somewhat attracted to activities
  that other firms already do" (herding / legitimacy)
- `influence_matrix` defines which components have synergies with each
  other. The block-diagonal structure creates 2 clusters of 4
  components---activities within a cluster complement each other
- `influence_weight = 0.3` sets the strength of those synergies

**Step 3** runs the simulation for 20 ministeps per actor
($6 \times 20 = 120$ total ministeps).

## Inspecting the State After Simulation

```{r inspect-after}
# What does the bipartite matrix look like now?
cat("Bipartite matrix BEFORE simulation:\n")
print(saomnk_get_bipartite(env1, step = 1))

cat("\nBipartite matrix AFTER simulation:\n")
print(env1$bipartite_matrix)

# How have the four K dimensions changed?
B_final <- env1$bipartite_matrix
S_final <- (B_final > 0) %*% t(B_final > 0); diag(S_final) <- 0
E_final <- t(B_final > 0) %*% (B_final > 0); diag(E_final) <- 0

cat("\nFinal K_AC (scope):     ", rowSums(B_final), "\n")
cat("Final K_CA (popularity):", colSums(B_final), "\n")
cat("Final K_AA (sociality): ", rowSums(S_final > 0), "\n")
cat("Final K_CC (epistasis): ", rowSums(E_final > 0), "\n")
```

## Visualizing What Happened

### The K-4 Panel: Your Dashboard

The K-4 panel shows **all four coupled degree processes** over time.
This is the single most important diagnostic plot in `searchnet`.

```{r k4-panel, fig.height=10, fig.width=10}
saomnk_plot_k4(env1, smooth = 0.3)
```

### Network Snapshots: Watch the World Evolve

Snapshots show the bipartite network at selected time points, along
with the actor-actor and component-component projections.

```{r snapshots, fig.height=5, fig.width=11}
saomnk_plot_snapshots(env1)
```

### Utility Contributions: Who Is Winning, and Why?

Decompose each actor's utility into contributions from each network
effect. This reveals *which mechanism* drives behavior.

```{r utility-plot, fig.height=10, fig.width=9}
saomnk_plot_utility(env1, smooth = 0.35)
```

## Reading the K-4 Panel (Interpretation Guide)

The K-4 panel is your **dashboard**. Here is how to read each quadrant:

| Quadrant | If Rising... | If Falling... |
|:---------|:------------|:-------------|
| $K_{AC}$ (scope) | Firms are expanding, diversifying | Firms are specializing, shedding activities |
| $K_{CA}$ (popularity) | Activities are becoming crowded | Activities are being abandoned |
| $K_{AA}$ (sociality) | Firms are converging (imitation) | Firms are differentiating |
| $K_{CC}$ (epistasis) | Activities are bundling into clusters | Activity bundles are fragmenting |

> **The key question:** When one quadrant moves, do the others move
> *with* it or *against* it? The coupling pattern tells you about the
> fundamental dynamics of the industry.

## Model Summary

The model summary provides a regression-style table of all effects and
their parameter values:

```{r model-summary}
saomnk_summary(env1)
```

## Adding Actor Heterogeneity (Strategies)

Not all firms are alike. Some are aggressive diversifiers; others are
cautious specialists. The `strategies` argument lets you give each
actor a different behavioral tendency.

```{r strategies}
# Create a fresh environment
env_strat <- saomnk_env(M = 6, N = 8, density = 0.3, seed = 42)

# Define a model with actor heterogeneity
model_strat <- saomnk_model(
  density = -0.5,
  popularity = 0.2,
  influence_matrix = saomnk_block_diagonal(8, 2),
  influence_weight = 0.3,
  strategies = list(
    egoX = c(-1, -1, 0, 0, 1, 1)  # 3 types: cautious, neutral, aggressive
  )
)

cat("Strategy types assigned:\n")
cat("  Actors 1-2: cautious (egoX = -1) -- less likely to add ties\n")
cat("  Actors 3-4: neutral  (egoX =  0) -- baseline behavior\n")
cat("  Actors 5-6: aggressive (egoX = 1) -- more likely to add ties\n\n")

saomnk_run(env_strat, model_strat, steps_per_actor = 20, seed = 200)
cat("Simulation complete.\n")

# Compare scope across strategy types
B_strat <- env_strat$bipartite_matrix
cat("\nFinal scope (K_AC) by strategy type:\n")
cat("  Cautious (actors 1-2): ", rowSums(B_strat)[1:2], "\n")
cat("  Neutral  (actors 3-4): ", rowSums(B_strat)[3:4], "\n")
cat("  Aggressive (actors 5-6):", rowSums(B_strat)[5:6], "\n")
```

```{r strategies-k4, fig.height=10, fig.width=10}
saomnk_plot_k4(env_strat, smooth = 0.3)
```

# PART 3: EXPERIMENTS AND SHOCKS {.tabset}

## The Simulation Factory

One simulation is one alternate reality. The power of simulation comes
from **systematic comparison**:

- Change **one** parameter = controlled experiment
- Compare outcomes = causal inference by construction
- Repeat with different seeds = statistical power

```{r factory-baseline}
# Baseline world: moderate scope cost, moderate epistasis
env_base <- saomnk_env(M = 6, N = 8, density = 0.3, seed = 42)
model_base <- saomnk_model(
  density = -0.5,
  popularity = 0.2,
  influence_matrix = saomnk_block_diagonal(8, 2),
  influence_weight = 0.3
)
saomnk_run(env_base, model_base, steps_per_actor = 20, seed = 100)

# Counterfactual: HIGH scope cost
env_high <- saomnk_env(M = 6, N = 8, density = 0.3, seed = 42)
model_high <- saomnk_model(
  density = -2.0,        # <-- much higher cost!
  popularity = 0.2,
  influence_matrix = saomnk_block_diagonal(8, 2),
  influence_weight = 0.3
)
saomnk_run(env_high, model_high, steps_per_actor = 20, seed = 100)

# Compare final scope
cat("Average K_AC (scope):\n")
cat("  Baseline (density = -0.5):", round(mean(rowSums(env_base$bipartite_matrix)), 2), "\n")
cat("  High cost (density = -2.0):", round(mean(rowSums(env_high$bipartite_matrix)), 2), "\n")
cat("\nHigher scope costs lead to more specialization,\n")
cat("which in turn reshapes all four K dimensions.\n")
```

## Exogenous Shocks (The "What If?" Machine)

Shocks let you change the rules *mid-simulation*. The timeline is split
into proportional segments, each with its own parameter regime.

```{r shocks}
# Create a world and define a two-phase shock:
#   Phase 1 (first half): density = -0.5 (normal)
#   Phase 2 (second half): density = -2.0 (austerity shock!)
env_shock <- saomnk_env(M = 6, N = 8, density = 0.3, seed = 42)

model_shock <- saomnk_model(
  density = -0.5,
  popularity = 0.2,
  influence_matrix = saomnk_block_diagonal(8, 2),
  influence_weight = 0.3
)

shock_baseline <- saomnk_shock("density", parameter = -0.5, portion = 1)
shock_austerity <- saomnk_shock("density", parameter = -2.0, portion = 1)

saomnk_run(env_shock, model_shock, steps_per_actor = 30, seed = 100,
           shocks = list(shock_baseline, shock_austerity))

cat("Shock simulation complete.\n")
cat("Total ministeps:", env_shock$get_step(), "\n")
cat("  Phase 1 (normal):    first half of the simulation\n")
cat("  Phase 2 (austerity): second half -- density jumps to -2.0\n")
```

```{r shocks-k4, fig.height=10, fig.width=10}
saomnk_plot_k4(env_shock, smooth = 0.3)
```

> **What to look for:** The K-4 panel should show a *regime change*
> around the midpoint. Before the shock, firms expand normally. After
> the shock, the higher scope cost forces them to shed activities. Watch
> how this propagates across all four K dimensions---that coupling is
> the signature of an endogenous landscape.

## Multiple Shocks: Richer Scenarios

You can chain multiple shocks together to create multi-phase scenarios:

```{r multi-shock}
env_multi <- saomnk_env(M = 6, N = 8, density = 0.3, seed = 42)

model_multi <- saomnk_model(
  density = -0.5,
  popularity = 0.2,
  influence_matrix = saomnk_block_diagonal(8, 2),
  influence_weight = 0.3
)

# Three phases: normal -> crisis -> recovery
s1 <- saomnk_shock("density", parameter = -0.5, portion = 2)  # normal (2 parts)
s2 <- saomnk_shock("density", parameter = -3.0, portion = 1)  # crisis (1 part)
s3 <- saomnk_shock("density", parameter = -0.3, portion = 1)  # recovery (1 part)

saomnk_run(env_multi, model_multi, steps_per_actor = 30, seed = 100,
           shocks = list(s1, s2, s3))

cat("Three-phase shock simulation complete.\n")
cat("  Phase 1 (normal):   2/4 of timeline, density = -0.5\n")
cat("  Phase 2 (crisis):   1/4 of timeline, density = -3.0\n")
cat("  Phase 3 (recovery): 1/4 of timeline, density = -0.3\n")
```

```{r multi-shock-k4, fig.height=10, fig.width=10}
saomnk_plot_k4(env_multi, smooth = 0.3)
```

## From Simulation to Causal Inference

The shock machinery in `searchnet` is designed to feed directly into
causal inference pipelines. Because you control the exact moment the
shock occurs and the exact parameter that changes, you can apply:

- **Difference-in-Differences (DID):** Compare treated vs. untreated
  actors before and after the shock
- **Synthetic Control:** Construct counterfactual trajectories from
  unsimulated baselines
- **Regression Discontinuity:** Exploit the sharp boundary at the shock
  point

These methods are covered in depth in the causal inference vignette
(`vignette("saomnk-causal-inference")`).

# PART 4: CONNECTING TO YOUR RESEARCH {.tabset}

## The Translation Table

The hardest part of simulation-based research is *translation*: mapping
your substantive question onto the simulation primitives. This table
provides a starting point.

| Your Research Context | searchnet Element | API Parameter |
|:---------------------|:-----------------|:-------------|
| Firms in an industry | Actors ($M$) | `saomnk_env(M = ...)` |
| Products / technologies / capabilities | Components ($N$) | `saomnk_env(N = ...)` |
| "Firm X does activity Y" | Tie $b_{ij} = 1$ | `density` in `saomnk_env()` |
| Cost of diversification | Density effect (negative) | `saomnk_model(density = ...)` |
| Competitive pressure / herding | Popularity effect | `saomnk_model(popularity = ...)` |
| Diversification tendency | Scope effect | `saomnk_model(scope = ...)` |
| Complementarities / synergies | Epistasis (XWX) | `saomnk_model(influence_matrix = ...)` |
| Strategic groups | $K_{AA}$ (sociality) | `saomnk_get_degrees(env)$K_AA` |
| Landscape ruggedness | $K_{CC}$ (epistasis) | `saomnk_get_degrees(env)$K_CC` |
| Environmental shock | Parameter regime change | `saomnk_shock(...)` |
| Firm strategy type | Actor covariate | `saomnk_model(strategies = ...)` |

## Exercise: Design Your Own Simulation

Take five minutes to think about a research question you are working
on. Then fill in this template:

```
My research question:
____________________________________________________________

Actors (M) represent:
____________________________________________________________

Components (N) represent:
____________________________________________________________

A tie (b_ij = 1) means:
____________________________________________________________

The key mechanism I want to study maps to:
  [ ] density (scope cost)
  [ ] popularity (herding / competitive pressure)
  [ ] scope (diversification tendency)
  [ ] epistasis (complementarities)
  [ ] shocks (environmental change)
  [ ] strategy heterogeneity (different firm types)

My "what if?" comparison:
  Baseline:  ________________________________________________
  Counterfactual: ___________________________________________
```

Here is a worked example to get you started:

```{r exercise-example}
cat("=== WORKED EXAMPLE: Technology Adoption in Biotech ===\n\n")
cat("Research question: How do complementarities between\n")
cat("  technologies affect industry-level specialization?\n\n")
cat("  M = 5 biotech firms\n")
cat("  N = 8 drug development platforms\n")
cat("  b_ij = 1 means 'firm i uses platform j'\n")
cat("  Key mechanism: epistasis (complementary platforms)\n\n")
cat("  Baseline: weak complementarities (influence_weight = 0.1)\n")
cat("  Counterfactual: strong complementarities (influence_weight = 0.5)\n\n")

# Baseline
env_ex_base <- saomnk_env(M = 5, N = 8, density = 0.2, seed = 99)
mod_ex_base <- saomnk_model(
  density = -0.5,
  influence_matrix = saomnk_block_diagonal(8, 2),
  influence_weight = 0.1   # weak complementarities
)
saomnk_run(env_ex_base, mod_ex_base, steps_per_actor = 20, seed = 100)

# Counterfactual
env_ex_cf <- saomnk_env(M = 5, N = 8, density = 0.2, seed = 99)
mod_ex_cf <- saomnk_model(
  density = -0.5,
  influence_matrix = saomnk_block_diagonal(8, 2),
  influence_weight = 0.5   # strong complementarities
)
saomnk_run(env_ex_cf, mod_ex_cf, steps_per_actor = 20, seed = 100)

cat("Results:\n")
cat("  Avg scope (weak compl.): ",
    round(mean(rowSums(env_ex_base$bipartite_matrix)), 2), "\n")
cat("  Avg scope (strong compl.):",
    round(mean(rowSums(env_ex_cf$bipartite_matrix)), 2), "\n")
cat("\n  Strong complementarities lead to broader scope\n")
cat("  because adopting one platform increases the value\n")
cat("  of adopting its complement.\n")
```

## Where to Go Next

This workshop covered the fundamentals. Here is where to dive deeper:

| Vignette | What You Will Learn |
|:---------|:-------------------|
| **Introduction** (`saomnk-introduction`) | Detailed API walkthrough with richer examples |
| **Theory** (`saomnk-theory`) | Formal mathematical framework, theorems, and proofs |
| **NK Validation** | Classical NK comparison and reduction proof |
| **Causal Inference** | DID / Synthetic Control / RD pipeline on simulated data |

```{r where-next}
cat("To access other vignettes:\n")
cat("  vignette('saomnk-introduction', package = 'searchnet')\n")
cat("  vignette('saomnk-theory', package = 'searchnet')\n")
```

# Appendix: Quick Reference Card

## API Functions at a Glance

```{r quick-ref}
cat("==================================================\n")
cat("  searchnet Quick Reference\n")
cat("==================================================\n\n")

cat("SETUP:\n")
cat("  env   <- saomnk_env(M, N, density, seed)\n")
cat("  model <- saomnk_model(density, popularity, scope,\n")
cat("             influence_matrix, influence_weight, strategies)\n\n")

cat("RUN:\n")
cat("  saomnk_run(env, model, steps_per_actor, seed, shocks)\n\n")

cat("VISUALIZE:\n")
cat("  saomnk_plot_k4(env)          # K-4 coupled degree panel\n")
cat("  saomnk_plot_snapshots(env)   # network snapshots\n")
cat("  saomnk_plot_utility(env)     # utility decomposition\n")
cat("  saomnk_summary(env)          # model summary table\n\n")

cat("EXTRACT:\n")
cat("  saomnk_get_bipartite(env)    # current bipartite matrix\n")
cat("  saomnk_get_bipartite(env, step = k)  # matrix at step k\n")
cat("  saomnk_get_degrees(env)      # list of K_AC, K_CA, K_AA, K_CC\n\n")

cat("SHOCKS:\n")
cat("  shock <- saomnk_shock(effect, parameter, portion)\n")
cat("  saomnk_run(env, model, shocks = list(s1, s2, ...))\n\n")

cat("HELPERS:\n")
cat("  saomnk_block_diagonal(N, blocks)  # modular influence matrix\n")
```

## The {K} Framework at a Glance

```{r k-summary-card}
cat("+----------+------------------+------------------+\n")
cat("|          |   ACTOR (row)    | COMPONENT (col)  |\n")
cat("+----------+------------------+------------------+\n")
cat("| 2-mode   | K_AC = scope     | K_CA = popularity|\n")
cat("| (direct) | rowSums(B)       | colSums(B)       |\n")
cat("+----------+------------------+------------------+\n")
cat("| 1-mode   | K_AA = sociality | K_CC = epistasis |\n")
cat("| (proj.)  | deg(BB')         | deg(B'B)         |\n")
cat("+----------+------------------+------------------+\n")
cat("\n  All four derive from the SAME bipartite matrix B.\n")
cat("  Change one cell of B --> all four shift.\n")
```

```{r session-info}
sessionInfo()
```
