---
title: "Computational Experiments: Batch Simulation and Analysis"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Computational Experiments: Batch Simulation and Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

```{r load-saomnk, include=FALSE}
if (requireNamespace("searchnet", quietly = TRUE)) {
  library(searchnet)
} else {
  dir_r <- file.path(getwd(), "R")
  if (!file.exists(file.path(dir_r, "utils.R")))
    dir_r <- "D:/Search_networks/SaoMNK/R"
  .saomnk_dir <- dir_r
  .saomnk_dir <- dir_r; source(file.path(dirname(dir_r), "inst", "saomnk-loader.R"))
}

library(Matrix)

DV_NAME <- "self$bipartite_rsienaDV"
```

# Overview

This vignette demonstrates how to design and execute computational experiments
with `searchnet`. The package supports systematic variation of model parameters,
batch simulation execution, and comparative analysis of search dynamics across
experimental conditions.

# 1. Experimental Design

## 1.1 Parameter Space

A typical `searchnet` experiment varies one or more parameters systematically:

- **Environment size**: $M$ (actors), $N$ (components)
- **Structural effects**: density, popularity (inPop), scope (outAct)
- **Strategy heterogeneity**: actor covariate weights
- **Epistasis structure**: block-diagonal, random, or empirical
- **Shock timing and magnitude**: when and how parameters change

```{r exp-design}
## Define a small experimental grid
experiment_grid <- expand.grid(
  M = c(6, 12),
  density = c(-0.3, -0.5, -1.0),
  inPop = c(0, 0.2),
  stringsAsFactors = FALSE
)

cat("Experimental conditions:", nrow(experiment_grid), "\n")
print(experiment_grid)
```

## 1.2 Replication Strategy

Each condition should be replicated with different random seeds to capture
stochastic variation:

```{r replication}
n_replications <- 3
seeds <- 1000 + seq_len(n_replications)

## Full design: conditions x replications
full_design <- merge(
  experiment_grid,
  data.frame(seed = seeds),
  all = TRUE
)
cat("Total simulation runs:", nrow(full_design), "\n")
```

# 2. Running Experiments

## 2.1 Single Condition

```{r single-condition}
N <- 12
component_int_mat <- create_block_diag(N, round(N / 3))

run_single_condition <- function(M, density_param, inPop_param,
                                  seed, N = 12, steps_per_actor = 30) {
  env <- SaomNkRSienaBiEnv$new(list(
    M = M, N = N, BI_PROB = 0,
    rand_seed = seed,
    name = sprintf("_exp_M%d_d%.1f_", M, density_param)
  ))

  sm <- list(
    dv_bipartite = list(
      name = DV_NAME,
      effects = list(
        list(effect = "density", parameter = density_param,
             dv_name = DV_NAME, fix = TRUE),
        list(effect = "inPop", parameter = inPop_param,
             dv_name = DV_NAME, fix = TRUE)
      ),
      coDyadCovars = list(
        list(effect = "XWX", parameter = 0.1, dv_name = DV_NAME, fix = TRUE,
             nodeSet = c("COMPONENTS", "COMPONENTS"),
             interaction1 = "self$component_1_coDyadCovar",
             x = component_int_mat[1:N, 1:N])
      )
    )
  )

  env$search_rsiena(
    sm,
    iterations = M * steps_per_actor,
    run_seed = seed
  )

  env
}
```

## 2.2 Batch Execution

```{r batch-run}
## Run a subset of conditions for demonstration
demo_conditions <- experiment_grid[c(1, 3, 5), ]

results <- list()
for (i in seq_len(nrow(demo_conditions))) {
  cond <- demo_conditions[i, ]
  cat(sprintf("Running condition %d: M=%d, density=%.1f, inPop=%.1f\n",
              i, cond$M, cond$density, cond$inPop))

  env <- tryCatch(
    run_single_condition(
      M = cond$M,
      density_param = cond$density,
      inPop_param = cond$inPop,
      seed = 12345,
      N = 12,
      steps_per_actor = 20
    ),
    error = function(e) {
      cat("  Error:", e$message, "\n")
      NULL
    }
  )

  if (!is.null(env)) {
    results[[i]] <- list(
      condition = cond,
      env = env,
      final_step = env$get_step()
    )
  }
}

cat("\nCompleted:", sum(!sapply(results, is.null)), "/",
    nrow(demo_conditions), "conditions\n")
```

# 3. Comparative Analysis

## 3.1 Visual Comparison Across Conditions

```{r compare-k4, fig.height=10, fig.width=10}
for (i in seq_along(results)) {
  if (!is.null(results[[i]])) {
    cond <- results[[i]]$condition
    cat(sprintf("\n--- M=%d, density=%.1f, inPop=%.1f ---\n",
                cond$M, cond$density, cond$inPop))
    results[[i]]$env$plot_degree_4panel(loess_span = 0.3)
  }
}
```

## 3.2 Extracting Summary Statistics

```{r summary-stats}
summary_df <- do.call(rbind, lapply(seq_along(results), function(i) {
  if (is.null(results[[i]])) return(NULL)
  env <- results[[i]]$env
  cond <- results[[i]]$condition

  ## Get final bipartite matrix
  final_mat <- env$bi_env_arr[, , env$get_step()]

  data.frame(
    M = cond$M,
    density_param = cond$density,
    inPop_param = cond$inPop,
    final_density = mean(final_mat),
    mean_actor_scope = mean(rowSums(final_mat)),
    mean_component_pop = mean(colSums(final_mat)),
    n_steps = env$get_step(),
    stringsAsFactors = FALSE
  )
}))

if (!is.null(summary_df)) {
  print(summary_df)
}
```

# 4. Using the SaoMNKexperiments Class

For larger-scale experiments, the `SaoMNKexperiments` class (in
`saomnk-experiments.R`) provides:

- Automated batch execution with progress tracking
- Result aggregation across replications
- Built-in visualization for experiment comparisons

```{r experiments-class, eval=FALSE}
## SaoMNKexperiments is exported by the package (or loaded via saomnk-loader.R above)

## Create experiment
exp <- SaoMNKexperiments$new(
  environ_params = list(M = 6, N = 12, BI_PROB = 0, rand_seed = 42),
  structure_model = sm,
  batch_seeds = c(1001, 1002, 1003),
  name = "density_experiment"
)

## Run all replications
exp$run_batch(iterations_per_actor = 50)

## Process and visualize results
exp$process_results()
```

# 5. Design Considerations

## 5.1 Environment Size

- **Small** ($M \leq 6$, $N \leq 8$): fast iteration, good for debugging
  and parameter exploration
- **Medium** ($M \sim 12$, $N \sim 16$): realistic competitive dynamics,
  suitable for publication-quality experiments
- **Large** ($M \geq 18$, $N \geq 24$): computationally intensive, best
  for robustness checks with cluster computing

## 5.2 Iterations per Actor

- **5--10**: minimal dynamics, good for unit testing
- **30--50**: moderate dynamics, visible K-4 evolution
- **100+**: convergence-scale dynamics, needed for equilibrium analysis

## 5.3 Number of Replications

For statistical inference, use $\geq 30$ replications per condition (Monte
Carlo standard). The `mc30` experiment scripts in the package demonstrate
this pattern.

# References

- Levinthal, D. A. (1997). Adaptation on rugged landscapes.
  *Management Science*, 43(7), 934--950.
- Snijders, T. A. B. (1996). Stochastic actor-oriented models for network
  change. *Journal of Mathematical Sociology*, 21(1--2), 149--172.

```{r session-info}
sessionInfo()
```
