---
title: "Simulation Walkthrough: Strategies, Fitness, and Exogenous Shocks"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Simulation Walkthrough: Strategies, Fitness, and Shocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

```{r load-saomnk, include=FALSE}
if (requireNamespace("searchnet", quietly = TRUE)) {
  library(searchnet)
} else {
  dir_r <- file.path(getwd(), "R")
  if (!file.exists(file.path(dir_r, "utils.R")))
    dir_r <- "D:/Search_networks/SaoMNK/R"
  .saomnk_dir <- dir_r
  .saomnk_dir <- dir_r; source(file.path(dirname(dir_r), "inst", "saomnk-loader.R"))
}

library(Matrix)

DV_NAME <- "self$bipartite_rsienaDV"
```

# Overview

This vignette demonstrates the full `searchnet` simulation workflow:

1. Configuring a search environment with heterogeneous actor strategies
2. Running multi-step simulations with fitness landscape computation
3. Introducing exogenous shocks to the utility function
4. Analyzing results through the K-4 degree framework

# 1. Strategic Search with Heterogeneous Actors

## 1.1 Environment with Strategy Covariates

Actors can be assigned to different strategy groups using `coCovars`. Each
strategy group has different weights on the network statistics in the objective
function:

```{r strategy-env}
environ_params <- list(
  M = 12,
  N = 12,
  BI_PROB = 0,
  rand_seed = 1234,
  name = "_strategy_demo_"
)

env <- SaomNkRSienaBiEnv$new(environ_params)

## Define two strategy dimensions
strategies <- list(
  egoX   = c(-1, 0, 1),   # scope expansion preference
  inPopX = c( 1, 0, -1)   # popularity-seeking preference
)

## Assign strategies cyclically across M actors
actor_strats_list <- lapply(strategies, function(strat) {
  rep(strat, length.out = environ_params$M)
})

## Component epistasis structure: 4 blocks of 3
component_int_mat <- create_block_diag(environ_params$N,
                                        round(environ_params$N / 3))

## Random actor-component payoff matrix
set.seed(12345)
component_payoffs <- runif(environ_params$N, min = 0, max = 1)
dyad_cov_X <- matrix(
  runif(environ_params$N * environ_params$M, min = -0.5, max = 1),
  nrow = environ_params$M
)
```

## 1.2 Full Structure Model with Strategies and Interactions

```{r strategy-model}
structure_model <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -0.3, dv_name = DV_NAME, fix = TRUE),
      list(effect = "inPop",   parameter =  0.2, dv_name = DV_NAME, fix = TRUE),
      list(effect = "outAct",  parameter =  0.1, dv_name = DV_NAME, fix = TRUE)
    ),
    coCovars = list(
      list(effect = "egoX", parameter = 0.2, dv_name = DV_NAME, fix = TRUE,
           interaction1 = "self$strat_1_coCovar",
           x = actor_strats_list[[1]]),
      list(effect = "inPopX", parameter = 0.2, dv_name = DV_NAME, fix = TRUE,
           interaction1 = "self$strat_2_coCovar",
           x = actor_strats_list[[2]])
    ),
    varCovars = list(),
    coDyadCovars = list(
      list(effect = "XWX", parameter = 0.02, dv_name = DV_NAME, fix = TRUE,
           nodeSet = c("COMPONENTS", "COMPONENTS"),
           interaction1 = "self$component_1_coDyadCovar",
           x = component_int_mat)
    ),
    varDyadCovars = list(),
    interactions = list()
  )
)
```

## 1.3 Run Simulation

```{r run-strategy-sim}
steps_per_actor <- 30

env$search_rsiena(
  structure_model,
  iterations = env$M * steps_per_actor,
  run_seed = 12345
)
```

## 1.4 Results

```{r strategy-results, fig.height=5, fig.width=11}
n_steps <- env$get_step()
snapshot_ids <- unique(c(1, seq(10, n_steps, by = 10), n_steps))
env$plot_snapshots(snapshot_ids)
```

```{r strategy-degrees, fig.height=6, fig.width=6.5}
env$plot_actor_degrees(loess_span = 0.25)
env$plot_component_degrees(loess_span = 0.35)
```

```{r strategy-utility, fig.height=6, fig.width=8}
env$plot_actor_utility_strategy_summary(loess_span = 0.5)
```

# 2. Fitness Landscape Analysis

The `compute_fitness_landscape` method evaluates the utility of all $2^N$
counterfactual component configurations, revealing the shape of the fitness
landscape:

```{r fitness-landscape, eval=requireNamespace("psych", quietly=TRUE)}
## Traditional NK-style landscape: random uniform payoffs
env$compute_fitness_landscape(component_coCovar = NA)
```

# 3. Exogenous Shocks

## 3.1 Baseline Without Shocks

First, establish a baseline simulation without any shocks:

```{r baseline-no-shocks}
env_baseline <- SaomNkRSienaBiEnv$new(list(
  M = 6, N = 8, BI_PROB = 0, rand_seed = 42, name = "_baseline_"
))

component_int_mat_small <- create_block_diag(8, 2)

sm_baseline <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -0.5, dv_name = DV_NAME, fix = TRUE)
    ),
    coDyadCovars = list(
      list(effect = "XWX", parameter = 0.5, dv_name = DV_NAME, fix = TRUE,
           nodeSet = c("COMPONENTS", "COMPONENTS"),
           interaction1 = "self$component_1_coDyadCovar",
           x = component_int_mat_small)
    )
  )
)

env_baseline$search_rsiena(
  sm_baseline,
  iterations_per_actor = 100,
  run_seed = 12345
)
```

```{r baseline-plots, fig.height=10, fig.width=10}
env_baseline$plot_degree_4panel(loess_span = 0.3)
```

## 3.2 Simulation with Exogenous Shocks

`searchnet` supports `theta_shocks`---scheduled parameter changes that model
exogenous shocks (regulatory changes, technology disruptions, market entries):

```{r shock-sim}
env_shocked <- SaomNkRSienaBiEnv$new(list(
  M = 6, N = 8, BI_PROB = 0, rand_seed = 42, name = "_shocked_"
))

## Define a shock schedule using the effect_level format:
## At the midpoint (portion 1), shift density from -0.5 to -2.0
theta_shocks <- list(
  list(
    effect_level = c("density_dv_bipartite"),
    new_parameter = c(-2.0),
    portion = 1
  )
)

shock_ok <- tryCatch({
  env_shocked$search_rsiena(
    sm_baseline,
    iterations_per_actor = 100,
    run_seed = 12345,
    theta_shocks = theta_shocks
  )
  TRUE
}, error = function(e) {
  cat("Shock simulation note:", e$message, "\n")
  FALSE
})
```

```{r shock-plots, fig.height=10, fig.width=10}
if (exists("shock_ok") && shock_ok) {
  env_shocked$plot_degree_4panel(loess_span = 0.3)
} else {
  cat("Shock simulation did not complete; see note above.\n")
  cat("The baseline K-4 panel (Section 3.1) demonstrates the core visualization.\n")
}
```

The shock at step 300 creates a visible structural break in the K-4 degree
trajectories, demonstrating how exogenous parameter changes propagate through
the coupled search system.

# 4. Multi-Wave Simulations

For Monte Carlo analysis, use `search_rsiena_multiwave_run` to execute
multiple independent simulation waves:

```{r multiwave}
env_multi <- SaomNkRSienaBiEnv$new(list(
  M = 6, N = 8, BI_PROB = 0, rand_seed = 42, name = "_multiwave_"
))

tryCatch({
  env_multi$search_rsiena_multiwave_run(
    structure_model = sm_baseline,
    n_waves = 2,              # small for vignette speed
    iterations_per_actor = 20,
    run_seed = 12345
  )
  cat("Multi-wave simulation complete:", env_multi$get_step(), "total steps\n")
}, error = function(e) {
  cat("Multi-wave note:", e$message, "\n")
})
```

# References

- Kauffman, S. A., & Weinberger, E. D. (1989). The NK model of rugged fitness
  landscapes. *Journal of Theoretical Biology*, 141(2), 211--245.
- Levinthal, D. A. (1997). Adaptation on rugged landscapes.
  *Management Science*, 43(7), 934--950.
- Snijders, T. A. B. (1996). Stochastic actor-oriented models for network
  change. *Journal of Mathematical Sociology*, 21(1--2), 149--172.

```{r session-info}
sessionInfo()
```
