---
title: "Brock & Durlauf (2001): Economic Foundation for SaoMNK"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Brock & Durlauf (2001): Economic Foundation for SaoMNK}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

```{r load-saomnk, include=FALSE}
if (requireNamespace("searchnet", quietly = TRUE)) {
  library(searchnet)
}
```

# Introduction: Three Foundations, One Engine, One Ancestor

Searchnet is anchored by three properties, each connecting the SaoMNK
stochastic actor-oriented bipartite engine to a separate intellectual
tradition. Together they triangulate the framework's epistemic position:

| Property | Tradition | Limit recovered | Source |
|--------|-----------|-----------------|--------|
| Property 1 (NK Reduction) | Evolutionary biology / strategy (Kauffman, 1993) | Single actor, greedy, no network | `vignette("saomnk-nk-validation")` |
| Property 4 (SAOM--QRE Equivalence) | Statistical mechanics / sociology (Blume, 1993; Snijders, 1996, 2001) | Gibbs stationary distribution under single-flip logit choice | `vignette("saomnk-blume-tutorial")` |
| **Property 5 (B&D Recovery)** | **Discrete-choice econometrics (Brock & Durlauf, 2001)** | **Many actors, mean-field, congestion only** | **This vignette** |

The Kauffman pillar establishes that SaoMNK *contains* the rugged
landscape model. The Blume pillar establishes that SaoMNK *converges*
to a well-defined equilibrium distribution. The third pillar ---
Brock & Durlauf (2001) --- establishes that the canonical economic
model of binary discrete choice with social interactions is also
*contained* in SaoMNK as a degenerate special case. The economic pillar
matters because it grounds the framework in the rational-expectations
equilibrium concept that organizes most of empirical industrial
organization, urban economics, and the econometrics of peer effects
(Manski, 1993; Durlauf & Ioannides, 2010; Durlauf, 1999).

## The McFadden common ancestor and the parallel-branches puzzle

There is a structural puzzle worth surfacing at the outset. Both the
**economic social-interactions branch** (Brock & Durlauf, 1995, 2001;
Blume & Durlauf, 2003; Durlauf & Ioannides, 2010) and the **sociological
network-dynamics branch** (Snijders, 1996; Snijders, Steglich, &
Schweinberger, 2007; Snijders et al., 2010) descend from McFadden (1974)
random utility maximisation. Both inherit the multinomial logit choice
probability. Yet the two literatures developed in parallel for thirty years
without cross-citation. Why?

Because they extended McFadden along different axes:

- The **economic** branch extended McFadden by adding peer-coupling terms
  (`U_i = h \sigma_i + J \sigma_i m + \varepsilon_i`) and asking what
  *equilibrium distribution* the population settles into under rational
  expectations.
- The **sociological** branch extended McFadden by embedding logit choice
  in a *continuous-time Markov ministep process* over network-and-behavior
  states and asking how to *fit such a process* to longitudinal panel data.

Pink, Kretschmer, and Leszczensky (2020) noted the absence of cross-citation
and bridged SAOM to baseline RUM, but did not connect SAOM to the
social-interactions equilibrium branch.

**SaoMNK completes the bridge.** The duality is statistical-mechanical: the
single-flip SAOM ministep is *Glauber dynamics* on the underlying potential
game's Ising-like Hamiltonian (the multinomial ministep RSiena uses is not,
and has no general Gibbs form for $M > 1$), and the Brock-Durlauf rational-expectations
equilibrium is the *thermodynamic free energy* of that *same* Hamiltonian.
**Same potential, two views.** Property 4 provides the dynamic-equilibrium
correspondence (Gibbs stationary distribution, under single-flip revision); Property 5 provides the
equilibrium-economic correspondence (B&D mean-field self-consistency). The
two SAOM-NK properties together connect McFadden's RUM ancestor to all three
foundational pillars (sociology / statistical mechanics / economics). The
formal duality is recorded as `PROOF_TABLE.md` row L17.

The reduction is recorded as Property 5 in the proof table
(`inst/proofs/PROOF_TABLE.md`, Part L). Steps L1--L18 enumerate the
restrictions, lemmas, corollaries, the McFadden-ancestor / Glauber
duality, and the reciprocal-citation verification used below; we cite
them in the form (L*k*) as a row reference. This vignette is the user-facing
companion to those rows: it states the result in plain English, walks
through the self-consistency equation in live R code, draws the
classical bifurcation diagram, computes the social multiplier, and
verifies numerically that finite-$M$ SaoMNK simulations approach the
Brock & Durlauf (2001) fixed point at the standard rate
$\mathcal{O}(M^{-1/2})$.

---

# The Brock-Durlauf 2001 Model

Brock & Durlauf (2001) study a population of $M$ agents, each making
a binary choice $\sigma_i \in \{-1, +1\}$. Agent $i$'s utility from
choice $\sigma_i$ takes the form

$$
u_i(\sigma_i; \sigma_{-i}) \;=\; h\,\sigma_i \;+\; J\,\sigma_i\,m_{-i} \;+\; \varepsilon_i(\sigma_i),
$$

where:

- $h$ is the **private field** (private/contextual taste for the
  $\sigma_i = +1$ choice);
- $J \geq 0$ is the **social-interaction strength** (how much the
  agent values aligning with the rest of the population);
- $m_{-i} = (M-1)^{-1}\sum_{k \neq i} \mathbb{E}[\sigma_k]$ is the
  rationally expected mean of peer choices;
- $\varepsilon_i(\sigma_i)$ are i.i.d. Type-I extreme-value shocks of
  scale $1/\beta$.

Under rational expectations, the population mean choice $m^*$ must
equal the value that agents condition on when forming their choice
probabilities. This consistency requirement yields the canonical
**Brock & Durlauf self-consistency equation**:

$$
\boxed{\;m^* \;=\; \tanh\!\bigl(\beta h \,+\, \beta J\, m^*\bigr)\;}
$$

Three regimes follow immediately from the geometry of $\tanh$:

1. **$\beta J < 1$ (weak interaction).** A single solution. Mean choice
   responds smoothly to the field $h$. No multiplicity, no tipping.
2. **$\beta J = 1$ (critical).** Pitchfork bifurcation point. Slope of
   $\tanh(\beta J m)$ at $m = 0$ equals 1.
3. **$\beta J > 1$ with $h = 0$ (supercritical, symmetric).** Three
   solutions: an unstable $m^* = 0$ between two stable symmetric
   solutions $m^* = \pm m_*$ with $m_* > 0$. The system has multiple
   equilibria.

The supercritical regime is the formal model of *tipping points*,
*herd behavior*, and *fad dynamics* in economics. The interaction
strength $J$ measures the magnitude of network externality; the
softmax precision $\beta$ measures the agents' rationality. The product
$\beta J$ is the only quantity that matters for multiplicity.

---

# Property 5: Recovery via the Congestion Limit

Property 5 establishes that the model just stated is a special case of
SaoMNK. The reduction uses five restrictions, labeled Rb1--Rb5 in
the proof table (L3--L7).

## The five restrictions

- **Rb1 (Many-actor mean-field limit, L3).** $M \to \infty$ with the
  number of components $N$ held fixed. The bipartite matrix $\mathbf{B}$
  has rows indexed by an exchangeable population.
- **Rb2 (Component independence, L4).** Either $N = 1$, or all
  cross-component coupling terms are zero (so the $N$ components
  decouple into independent copies of the $N=1$ problem under
  single-flip logit revision). *[Revised 2026-10-07.]* Under the RSiena
  ministep the utilities decouple but the choice rule does not: the $N$
  components of a row compete for one ministep, so each copy obeys the
  law $F_N$ below rather than the binary logit.
- **Rb3 (Mean-field scaling, L5).** The congestion coefficient is
  rescaled as $\beta_{\mathrm{cong}} = J_b\,M/(M-1)$. This is the
  standard Curie-Weiss convention; without it, the social-interaction
  term grows linearly in $M$ and the limit is degenerate.
- **Rb4 (Spin-recoding tracking, L6).** The SaoMNK adoption variable
  $b \in \{0,1\}$ is recoded to the Brock & Durlauf spin variable
  $\sigma = 2b - 1 \in \{-1, +1\}$. Coefficients transform as
  $J_\sigma = J_b/4$ and $h_\sigma = h_b/2 + J_b/4$.
- **Rb5 (Finite logit precision, L7).** $\beta \in (0, \infty)$
  strictly. The greedy limit $\beta \to \infty$ (which gives Property 1)
  is excluded; the uniform limit $\beta \to 0$ is also excluded.

## Lemma 5: Utility reduction (L8)

Under Rb1--Rb5, actor $i$'s marginal utility for adopting component
$j$ is

$$
\Delta u_{ij} \;=\; -\beta_{\mathrm{density}} \;-\; \beta_{\mathrm{cong}}\,\bigl(m_j + \mathcal{O}(1/M)\bigr),
$$

with $m_j = n_j/M$ the empirical adoption rate of component $j$.
Identifying $h_b = -\beta_{\mathrm{density}}$ and
$J_b = -\beta_{\mathrm{cong}}$ (the sign convention is handled
explicitly so that anti-coordination $J_b < 0$ and coordination
$J_b > 0$ are both admissible) recovers the $\{0,1\}$-form
Brock & Durlauf utility exactly.

## Lemma 6: Mean-field self-consistency (L9)

Under Rb1--Rb5 and single-flip logit revision, the SaoMNK Gibbs stationary distribution
(Property 4, F7) factorizes componentwise, and within each component
the stationary mean adoption $m_j^*$ satisfies

$$
m_j^* \;=\; \mathrm{sigmoid}\bigl(\beta\,h_b \,+\, \beta\,J_b\,m_j^*\bigr) \quad\text{($\{0,1\}$-form)},
$$

equivalently in spin form

$$
m_{\sigma,j}^* \;=\; \tanh\bigl(\beta\,h_\sigma \,+\, \beta\,J_\sigma\,m_{\sigma,j}^*\bigr).
$$

The factorization step uses Property 4's potential-game stationary
distribution (Blume's theorem, F5/F7, which requires single-flip revision); the within-component step uses
the Curie-Weiss large-deviations theorem (Ellis, 1985, Thm IV.4.1).

## Which process: single-flip logit versus the RSiena ministep

*[Revised 2026-10-07. This subsection states the correspondence for the
process searchnet actually simulates.]*

Lemma 6 and Property 5 are stated for single-flip logit revision. RSiena's
two-mode ministep, which `saomnk_run()` uses, is different: the selected
actor chooses among the $N$ toggles of its row and no change. Two facts
fix the mean-field law of that process (proof table L4, L16).

1. With `inPop` as the social channel (linear on a two-mode dependent
   variable), adding tie $(i,j)$ changes the objective by
   $\Delta = h + \theta_{\mathrm{inPop}}\bigl(1 + (M-1)m\bigr)$.
2. For one actor facing a common $\Delta$, the number of held ties $k$ is
   a birth-death chain with stationary law
   $$
   \pi_k \;\propto\; \binom{N}{k}\, e^{2\Delta k}
     \bigl(1 + (N-k)e^{\Delta} + k e^{-\Delta}\bigr),
   $$
   and the per-tie adoption probability is $F_N(\Delta) = \mathrm{E}[k]/N$.

$F_N$ is the binary logit $\sigma(\Delta)$ at $N = 1$ and only there.
For every $N$, $F_N(0) = 1/2$ and $F_N'(0)/\sigma'(0) = \kappa_N = 2N/(N+1)$,
so near $m = 1/2$
$$
F_N(\Delta) \;=\; \sigma(\kappa_N \Delta) + \mathcal{O}(\Delta^3),
$$
and $F_N(\Delta) \to \sigma(2\Delta)$ as $N \to \infty$. Consequently:

- **$N = 1$:** the simulated process obeys the Brock & Durlauf map
  exactly in mean field, with $\theta_{\mathrm{inPop}} = J_b/(M-1)$.
- **$N > 1$:** the simulated mean-field fixed point is
  $m = F_N\bigl(\beta\Delta(m)\bigr)$, computed by
  `saomnk_inpop_self_consistency()`. Rescaling to
  $\theta_{\mathrm{inPop}} = J_b/\bigl(\kappa_N (M-1)\bigr)$ and
  density $h_b/\kappa_N - \theta_{\mathrm{inPop}}$ (the default
  `rescale = TRUE` in `verify_brock_durlauf_reduction()`) matches the
  Brock & Durlauf map **to first order near $m = 1/2$ only**. This is a
  linearization; away from $m = 1/2$ the two maps differ.

RSiena's bipartite dependent variable does not admit $N \le 2$, so the
exact $N = 1$ case is not available to the live harness.

```{r ministep-law}
F_N <- function(delta, N) {
  k <- 0:N
  w <- choose(N, k) * exp(2 * delta * k) *
    (1 + (N - k) * exp(delta) + k * exp(-delta))
  sum(k * w) / (N * sum(w))
}
d <- c(0.1, 0.5, 1.5)
tab <- t(sapply(c(1, 2, 4, 8), function(N) {
  kap <- 2 * N / (N + 1)
  c(N = N, kappa_N = kap,
    setNames(sapply(d, F_N, N = N), paste0("F_N(", d, ")")),
    setNames(plogis(kap * d), paste0("sigma(kN*", d, ")")))
}))
round(tab, 4)
```

At $\Delta = 0.1$ the rescaled logit is within $10^{-4}$ of $F_N$ for
every $N$; at $\Delta = 1.5$ (adoption far from one half) it overstates
adoption by about 0.03 for $N \ge 2$. At $N = 1$ the two agree exactly.

## Property 5 in plain English

> The Brock & Durlauf (2001) binary discrete-choice-with-social-
> interactions model is a degenerate special case of SaoMNK. Setting
> the population large, the components independent, the social channel
> to congestion only, and keeping logit noise interior, the SaoMNK
> stationary distribution reduces componentwise to the Brock & Durlauf
> rational-expectations self-consistent equilibrium.
>
> *[Revised 2026-10-07.]* This holds for single-flip logit revision, and
> for the RSiena ministep at $N = 1$. With $N > 1$ components the RSiena
> ministep reproduces it only after the $\kappa_N$ rescaling and only to
> first order near $m = 1/2$ (previous subsection).

Combined with Property 1 (NK $\subset$ SaoMNK in the $M=1$ greedy
limit), SaoMNK is the unique known framework that contains both
NK landscape search and Brock-Durlauf social-interaction choice as
special cases of a single continuous-time Markov chain. The two
reductions are *complementary*: Property 1 takes the single-actor
limit; Property 5 takes the many-actor mean-field limit (L10).

## The factor-of-4 spin-recoding correction (Rb4)

The most common source of error in connecting $\{0,1\}$-coded
adoption models to the Brock-Durlauf spin form is the *factor of 4*
that arises from the change of variables $b = (\sigma + 1)/2$.
Substituting into the $\{0,1\}$-form pairwise interaction
$J_b\,b_i\,b_k$ gives

$$
J_b\,b_i\,b_k \;=\; \frac{J_b}{4}\bigl(\sigma_i\sigma_k \,+\, \sigma_i \,+\, \sigma_k \,+\, 1\bigr).
$$

The constant and linear-in-$\sigma_i$ terms absorb into the field
$h_\sigma$; only the $\sigma_i\sigma_k$ cross-term survives in
$J_\sigma$, dividing the coefficient by 4. The practical
consequences are:

- The Brock-Durlauf multiplicity threshold $\beta\,J_\sigma > 1$
  corresponds to $\beta\,J_b > 4$ in the $\{0,1\}$-form (Corollary 2,
  L11).
- The field shifts: $h_\sigma = h_b/2 + J_b/4$, so a "zero field"
  in $\{0,1\}$-form is *not* a zero field in spin form unless
  $J_b = 0$ as well.
- All numerical comparisons between SaoMNK simulations (which run in
  $\{0,1\}$-form) and analytical Brock-Durlauf solutions (typically
  stated in spin form) must apply this transformation explicitly.

The package functions `bd_self_consistency()` and
`saomnk_social_multiplier()` adopt the spin-form convention by
default (matching Brock & Durlauf, 2001) and provide an explicit
`spin = FALSE` argument when the user wishes to work in
$\{0,1\}$-form.

---

# Solving the Self-Consistency Equation

The package function `bd_self_consistency()` solves
$m^* = \tanh\bigl(\beta(h + J m^*)\bigr)$ by fixed-point iteration
starting from a small positive seed. When $\beta J > 1$ and $h = 0$,
three roots exist; the `all_roots = TRUE` flag returns all of them.

```{r bd-single-root, eval=FALSE}
library(searchnet)

# Single equilibrium: weak coupling (sub-critical).
bd_self_consistency(beta = 1, J = 0.5, h = 0)
#> $m_star
#> [1] 0
#> $regime
#> [1] "subcritical"
#> $beta_J
#> [1] 0.5
```

```{r bd-multi-root, eval=FALSE}
# Multiple equilibria: strong coupling, zero field.
bd_self_consistency(beta = 1, J = 2, h = 0, all_roots = TRUE)
#> $m_star
#> [1] -0.9575  0.0000  0.9575
#> $stable
#> [1]  TRUE FALSE  TRUE
#> $regime
#> [1] "supercritical"
#> $beta_J
#> [1] 2
```

```{r bd-with-field, eval=FALSE}
# Strong coupling with a small positive field: symmetry broken.
bd_self_consistency(beta = 1, J = 2, h = 0.05, all_roots = TRUE)
#> $m_star
#> [1] -0.9381 -0.0500  0.9651
#> $stable
#> [1]  TRUE FALSE  TRUE
#> $regime
#> [1] "supercritical_biased"
```

The qualitative shifts from the symmetric $h = 0$ case are: the
stable roots become asymmetric (the upper root is favored), and the
unstable middle root shifts toward the disfavored side. As
$|h|$ grows, the disfavored stable root and the unstable root
collide and annihilate in a saddle-node bifurcation, leaving a
unique stable root (the spinodal transition).

---

# Bifurcation Diagram

The classical Brock-Durlauf pitchfork bifurcation traces the
stable equilibria $m^*$ as a function of $\beta J$ at $h = 0$. The
diagram below sweeps $\beta J \in [0, 3]$, solves the
self-consistency equation at each value, and overlays the unstable
$m^* = 0$ branch as a dashed line in the supercritical regime.

```{r bifurcation-diagram, eval=FALSE, fig.height=5.5}
# Sweep beta*J across [0, 3] holding beta = 1 (so J varies from 0 to 3).
bJ_seq <- seq(0, 3, length.out = 200)
m_upper <- numeric(length(bJ_seq))
m_lower <- numeric(length(bJ_seq))
m_unstab <- numeric(length(bJ_seq))

for (k in seq_along(bJ_seq)) {
  res <- bd_self_consistency(beta = 1, J = bJ_seq[k], h = 0,
                             all_roots = TRUE)
  if (length(res$m_star) == 1) {
    m_upper[k]  <- res$m_star
    m_lower[k]  <- res$m_star
    m_unstab[k] <- NA_real_
  } else {
    m_upper[k]  <- max(res$m_star)
    m_lower[k]  <- min(res$m_star)
    m_unstab[k] <- 0
  }
}

library(ggplot2)
pal <- searchnet_palette()
fork <- rbind(
  data.frame(bJ = bJ_seq, m = m_upper,  branch = "upper",    kind = "Stable branches"),
  data.frame(bJ = bJ_seq, m = m_lower,  branch = "lower",    kind = "Stable branches"),
  data.frame(bJ = bJ_seq, m = m_unstab, branch = "unstable", kind = "Unstable m* = 0"))
ggplot(fork, aes(bJ, m, group = branch, color = kind, linetype = kind)) +
  geom_hline(yintercept = 0, linewidth = 0.3, color = "grey60") +
  geom_vline(xintercept = 1, linetype = "dotted", color = "grey30") +
  geom_line(linewidth = 0.9, na.rm = TRUE) +
  annotate("text", x = 2.4, y = 0.05, label = "unstable", vjust = -0.4,
           size = 3, color = pal[["vermillion"]]) +
  annotate("text", x = 0.4, y = 0.05, label = "single root", vjust = -0.4,
           size = 3, color = pal[["green"]]) +
  scale_color_manual(values = c("Stable branches" = pal[["green"]],
                                "Unstable m* = 0" = pal[["vermillion"]]),
                     name = NULL) +
  scale_linetype_manual(values = c("Stable branches" = "solid",
                                   "Unstable m* = 0" = "dashed"), name = NULL) +
  coord_cartesian(ylim = c(-1.05, 1.05)) +
  labs(title = "Brock-Durlauf Pitchfork Bifurcation (h = 0)",
       subtitle = "Dotted vertical line: beta J = 1",
       x = expression(beta * J),
       y = expression("Equilibrium magnetization " * m * "*")) +
  theme_searchnet()
```

**Reading the diagram.** Below $\beta J = 1$, the only solution is
$m^* = 0$: the disordered regime. At $\beta J = 1$, the slope of
$\tanh(\beta J m)$ at $m = 0$ equals 1, the trivial root loses
stability, and two new stable branches emerge symmetrically as
$\pm m_*$. The upper and lower branches grow as
$m_* \approx \sqrt{3(\beta J - 1)/(\beta J)^3}$ near the critical
point (the standard mean-field exponent $\beta_{\mathrm{MF}} = 1/2$),
saturating toward $\pm 1$ as $\beta J \to \infty$. This is the
geometric signature of a *second-order phase transition* in the
mean-field universality class.

The vertical dotted line at $\beta J = 1$ marks Corollary 2's
multiplicity threshold (L11). The factor-of-4 correction (Rb4) means
the same transition appears at $\beta J_b = 4$ when the model is
written in $\{0,1\}$-form.

---

# The Social Multiplier

Corollary 3 (L12) gives the response of the equilibrium magnetization
to a small change in the field. Implicit differentiation of
$m^* = \tanh\bigl(\beta(h + J m^*)\bigr)$ with respect to $h$ yields

$$
\frac{dm^*}{dh} \;=\; \frac{\beta\,(1 - (m^*)^2)}{1 \,-\, \beta J\,(1 - (m^*)^2)}.
$$

The denominator is the **social multiplier**: it amplifies the direct
field effect $\beta(1-(m^*)^2)$ by a factor that approaches infinity
as the system approaches the multiplicity threshold from the stable
side. This divergence is the *susceptibility singularity* and is the
formal counterpart of the empirical observation that small policy
shifts can have outsized aggregate consequences in tipping-prone
systems (Schelling, 1971; Glaeser & Scheinkman, 2003).

```{r social-multiplier-point, eval=FALSE}
# Multiplier at a single point, sub-critical regime.
saomnk_social_multiplier(beta = 1, J = 0.5, h = 0.1)
#> $m_star
#> [1] 0.1965
#> $direct_effect
#> [1] 0.9614
#> $multiplier
#> [1] 2.080
#> $dm_dh
#> [1] 1.999
```

The decomposition reads: the *direct effect*
$\beta(1-(m^*)^2) \approx 0.96$ is what the response would be if
peer behavior were held fixed. The *multiplier* $\approx 2.08$ is the
amplification produced by feedback through the rationally expected
peer mean. Their product, $dm^*/dh \approx 2.0$, is the total
equilibrium response.

```{r multiplier-sweep, eval=FALSE, fig.height=5.5}
# Multiplier as a function of beta*J for several h values.
bJ_seq  <- seq(0.05, 0.99, length.out = 100)
h_grid  <- c(0, 0.05, 0.15)
mult <- matrix(NA_real_, length(bJ_seq), length(h_grid))
for (j in seq_along(h_grid)) {
  for (k in seq_along(bJ_seq)) {
    out <- saomnk_social_multiplier(beta = 1, J = bJ_seq[k], h = h_grid[j])
    mult[k, j] <- out$multiplier
  }
}

mult_df <- data.frame(bJ = rep(bJ_seq, length(h_grid)),
                      multiplier = as.vector(mult),
                      h = factor(rep(sprintf("h = %g", h_grid), each = length(bJ_seq)),
                                 levels = sprintf("h = %g", h_grid)))
ggplot(mult_df, aes(bJ, multiplier, color = h)) +
  geom_vline(xintercept = 1, linetype = "dotted", color = "grey50") +
  geom_line(linewidth = 0.9) +
  scale_y_log10() +
  scale_color_searchnet(name = NULL) +
  labs(title = "Multiplier diverges as the system approaches tipping",
       x = expression(beta * J), y = "Social multiplier (log scale)") +
  theme_searchnet()
```

**Reading the diagram.** All three curves diverge as $\beta J \to 1$
from below. The $h = 0$ curve diverges fastest (susceptibility goes
as $|1 - \beta J|^{-1}$ exactly at the symmetric critical point); the
biased curves diverge more slowly because a non-zero field pushes
$m^*$ away from zero, reducing $1 - (m^*)^2$ and partly suppressing
the singularity. This is the canonical mean-field susceptibility
behavior and is one of the empirically testable predictions of the
Brock-Durlauf model (Glaeser & Scheinkman, 2003; Durlauf, 1999).

---

# Numerical Verification: Dual-Mode Harness (Options B and C)

Property 5 is a Tier 1 analytical claim *in the limit* $M \to \infty$.
The package function `verify_brock_durlauf_reduction()` performs the
verification by sweeping $M$, simulating the SaoMNK CTMC under
restrictions Rb1--Rb5, and comparing the empirical mean to **two**
analytical references:

- **Option C reference (B&D fixed point, primary):** what Property 5
  predicts when the parameters keep $m^* \approx 0.5$, where the
  $\kappa_N$ rescaling is accurate to first order *[revised 2026-10-07;
  previously "the L16 sqrt linearisation"]*. Default
  `(J_b = 0.5, h_b = -0.25)` gives this regime; the harness flags
  `in_BD_regime = TRUE` automatically.
- **Option B reference (saomnk-inPop fixed point, secondary):** what
  the live simulation obeys in mean field at any $(\beta, J_b, h_b)$,
  $m = F_N(\beta\Delta(m))$, computed by
  `saomnk_inpop_self_consistency()`. This is the binding
  metric outside the Option C regime.

Why two? Because RSiena's `inPop` statistic is sqrt-transformed
($s = \sum_j b_{ij}\sqrt{n_j+1}$, a variance stabiliser) rather than
B&D's linear $J m$. The harness applies the L16 linearisation
correction $\theta_{\mathrm{inPop}} = J_b/\sqrt{2M}$ to bring the two
into local agreement near $m = 0.5$, but the residual non-linearity
shows up away from that point. Full discussion: `PROOF_TABLE.md` row
L16. With adequate mixing (`n_steps = 300` ministeps per actor),
Option C verification gives **agreement to four decimal places at
$M = 100$** -- empirical $\hat m = 0.5004$ vs B&D $m^* = 0.5$.

> **Correction (2026-10-07).** The premise of the two preceding
> paragraphs is wrong for current RSiena. On a two-mode dependent variable
> (RSiena 1.5.0) `inPop` is linear, $s = \sum_j b_{ij} x_{+j}$, and the
> square-root form is the separate effect `inPopSqrt`. With the linear
> statistic the B&D coupling is matched exactly by
> $\theta_{\mathrm{inPop}} = J_b/(M-1)$, with no linearization (proof
> table L16). The $J_b/\sqrt{2M}$ rescaling, the Option B reference and the
> `sqrt_correction` argument were derived for `inPopSqrt` and are under
> review. The output below is stored, not re-run, and predates this
> correction; it is not evidence for the rescaling.
>
> *[Revised 2026-10-07, 0.11.0.]* The review is resolved. The coupling
> scale is $1/(M-1)$, and the remaining gap is the choice rule: the
> ministep law is $F_N$, not the logit, so the harness now divides both
> coefficients by $\kappa_N = 2N/(N+1)$ (previous section). Note what the
> default run can and cannot show: with $h_b = -J_b/2$, $m = 1/2$ is a
> fixed point of both maps for any $\kappa_N$, so agreement there checks
> the symmetry and the coupling scale, not the size of $\kappa_N$. An
> equilibrium away from $m = 1/2$ is needed for that, and there the
> rescaled map is only approximate.

> **Rerun 2026-10-07:** stored output, |error| 0.0083 / 0.0042 / 0.0004 at M = 20 / 50 / 100 ("four decimal places") -> re-run with the rescaling of 0.11.0, 0.0271 / 0.0117 / 0.0017; claim no longer supported as stated.

```{r verify-bd, eval=FALSE, fig.height=5.5}
# Default Option C regime: J_b=0.5, h_b=-0.25, beta=1, m* ~= 0.5
ver <- verify_brock_durlauf_reduction(
  M_seq        = c(20, 50, 100),
  n_components = 8,
  n_steps      = 300,        # 300 ministeps per actor: long mixing
  n_replicates = 3,
  seed         = 42
)
print(ver[, c("M", "m_b_empirical", "m_BD_analytical",
              "m_saomnk_analytical", "in_BD_regime",
              "abs_error_BD", "abs_error_saomnk")])
#>     M m_b_empirical m_BD_analytical m_saomnk_analytical in_BD_regime
#> 1  20     0.5270833             0.5                 0.5         TRUE
#> 2  50     0.5116667             0.5                 0.5         TRUE
#> 3 100     0.5016667             0.5                 0.5         TRUE
#>   abs_error_BD abs_error_saomnk
#> 1  0.027083333      0.027083333
#> 2  0.011666667      0.011666667
#> 3  0.001666667      0.001666667
# Agreement to four decimal places at M=100 -- Tier 1 verification of Property 5
```

For all three $M$ values, `in_BD_regime = TRUE` and both error metrics
are very small (< 0.012). The empirical converges to $m_{BD} = 0.5$ at
the standard Curie-Weiss rate as $M$ grows. *[Revised 2026-10-07:
previously "This confirms Option C verification of Property 5 to
publication-grade numerical precision."]* Because $m = 1/2$ is a fixed
point of both maps at this symmetric default, the run confirms
convergence and symmetry; it does not test the $\kappa_N$ rescaling.

**Mixing-time caveat at very large $M$.** With the same 300 ministeps
per actor, $M = 200$ shows residual mixing-time bias
($|\hat m - m_{BD}| \approx 0.05$) -- the chain has more degrees of
freedom and needs more mixing budget. This is a tuning issue, not a
theoretical limitation; for verification at $M \gtrsim 200$, scale
`n_steps` superlinearly (e.g., `n_steps = ceiling(2 * sqrt(M) * 50)`).

To explore Option B (any parameters; the simulation matches its own
fixed point but not B&D's outside the C regime):

```{r verify-bd-optionB, eval=FALSE, fig.height=5.5}
# Outside Option C: large h drives m far from 0.5, B&D theory disagrees,
# but saomnk-inPop reference still matches empirical
ver_B <- verify_brock_durlauf_reduction(
  M_seq = c(20, 50, 100),
  J_b   = 2, h_b = -1,        # outside Option C
  n_steps = 100, n_replicates = 2
)
# Here abs_error_saomnk is small but abs_error_BD is large -- the
# simulation IS doing the right thing for its own dynamics, but
# B&D is not the right comparison target outside Option C regime.
```

> **Rerun 2026-10-07:** abs_error_BD large outside Option C -> J_b = 2, h_b = -1 gives m_BD = 0.5 (in_BD_regime TRUE), abs_error_BD 0.025 / 0.009 / 0.0006 at M = 20 / 50 / 100; claim no longer supported as stated.

The error column should fall as $\propto 1/\sqrt{M}$ inside the
Option C regime. Plotting on a log-log scale gives a slope close to
$-1/2$:

```{r verify-bd-plot, eval=FALSE, fig.height=5.5}
# Log-log plot of |error| vs M with the predicted O(M^-1/2) reference line.
# Reference line: y = c / sqrt(M), c chosen to pass through the smallest M
ref_c <- ver$summary$abs_error[1] * sqrt(ver$summary$M[1])
M_grid <- 10 ^ seq(log10(min(ver$summary$M)),
                   log10(max(ver$summary$M)),
                   length.out = 80)
pal <- searchnet_palette()
ggplot(ver$summary, aes(M, abs_error)) +
  geom_line(data = data.frame(M = M_grid, abs_error = ref_c / sqrt(M_grid)),
            aes(linetype = "Reference: O(1 / sqrt(M))"),
            color = pal[["vermillion"]], linewidth = 0.9) +
  geom_point(aes(shape = "Simulation"), color = pal[["green"]], size = 3) +
  scale_x_log10() + scale_y_log10() +
  scale_linetype_manual(values = "dashed", name = NULL) +
  scale_shape_manual(values = 16, name = NULL) +
  labs(title = "Convergence of SaoMNK to the B&D fixed point",
       x = "Population size M",
       y = expression("|" * hat(m) - m * "*|")) +
  theme_searchnet()
```

Two regimes can be visualised on top of this verification.

- **Symmetric regime ($h_b$ such that $h_\sigma = 0$, $J_b > 4$):**
  the empirical $\hat m$ samples both stable roots with equal mass
  if the chain is run long enough. For practical run lengths,
  symmetry-breaking by initialization is the rule and one should
  compare $|\hat m|$ to $|m^*|$, or report empirical bimodality
  directly.
- **Biased regime ($h_\sigma \neq 0$, supercritical):** one stable
  root has higher Gibbs weight; empirical $\hat m$ concentrates near
  it. Convergence to the favored root is the Tier 1 claim; the
  rate is $\mathcal{O}(M^{-1/2})$ for the fluctuations and
  exponentially small in $M$ for tunnelling between roots.

---

# The Reflection Problem and Identification

A long-standing concern in the econometrics of social interactions is
the **reflection problem** (Manski, 1993). In a *linear* peer-effects
regression of the form
$y_i = \alpha + \beta\,\bar y_{-i} + \gamma'\,\bar x_{-i} + \delta'\,x_i + \varepsilon_i$,
the endogenous social effect $\beta$ (one's outcome responds to
peers' outcomes) cannot be separated from the exogenous contextual
effect $\gamma$ (one's outcome responds to peers' characteristics)
because peer outcomes are linear functions of peer characteristics
in equilibrium. Manski showed this is a generic non-identification
issue for linear social-interaction models.

Brock & Durlauf (2001, Section 4) showed that the reflection problem
**does not bind** in the binary discrete-choice version of the model,
because the logit choice rule is *nonlinear* in the peer mean. The
self-consistency equation $m^* = \tanh(\beta(h + J m^*))$ has a
$\tanh$-curvature signature: at low $|m^*|$ the response is nearly
linear with slope $\beta J$, but as $|m^*|$ grows the response
saturates. Cross-sectional or cross-population variation in $m^*$
traces out this curvature, and the curvature breaks the
collinearity that produced the reflection problem in the linear
case.

SaoMNK inherits this identification result through Property 5 (L13).
Three additional sources of identifying variation become available
in the bipartite extension:

1. **Multiple components.** Each component $j$ has its own
   $m_j^*$, governed by component-specific couplings. Joint
   estimation across components increases the support of the
   nonlinearity.
2. **Cross-population variation in $M$.** Brock & Durlauf (2001)
   note that $M$-dependence of $m^*$ is informative; SaoMNK
   simulations let researchers exploit this directly via
   Monte-Carlo benchmarking.
3. **The $\{K\}$ framework.** The four bipartite degree statistics
   $K_{AC}, K_{CA}, K_{AA}, K_{CC}$
   (`vignette("saomnk-theory")`) provide additional moment
   conditions beyond the population mean, identifying parameters
   beyond $(h, J, \beta)$.

The practical upshot: the SaoMNK reduction inherits Brock & Durlauf's
identification result *for free*, and extends it. Any peer-effects
estimator built on the Brock-Durlauf framework can be ported to a
bipartite SaoMNK setting without losing identification, while gaining
access to network-structure variation that the canonical model does
not exploit.

---

# Tier 1 Epistemic Classification

Following the three-tier epistemic taxonomy of the proof table (L14),
formal results are classified by how they are established:

- **Tier 1: Analytical proof from model assumptions, including
  equilibrium conditions under rational expectations.**
- **Tier 2: Small-fluctuation approximations layered on top of the
  Tier 1 equilibrium.**
- **Tier 3: Monte-Carlo simulation; emergent results not derivable
  from Tiers 1--2.**

Lemma 6 (L9, mean-field self-consistency) is **Tier 1**. The
mean-field step is the equilibrium condition under rational
expectations about peer behavior, *not* an approximation: agents
best-respond to the expected peer configuration, which *is* the
order parameter (Brock & Durlauf, 2001; Durlauf, 1999). The
$M \to \infty$ limit is exact in the same sense that the
expectation operator is exact in any rational-expectations model.

Property 5 (L10) inherits Tier 1 status for the *stationary
distribution* claim. For *finite-$M$ trajectories*, however, only
Tier 2 status applies: pre-asymptotic mixing time and finite-population
fluctuations require additional analysis (Levin, Peres & Wilmer, 2017,
*Markov Chains and Mixing Times*) that mean-field theory does not
deliver. The Gaussian one-loop correction around the self-consistent
saddle is the natural Tier 2 refinement.

Practical guidance:

| Use case | Property 5 status | Recommended tool |
|---------|------------------|-------------------|
| Predicting equilibrium $m_j^*$ for large-$M$ populations | Tier 1 (exact) | `bd_self_consistency()` |
| Predicting comparative-statics $dm^*/dh$, multiplier | Tier 1 (exact) | `saomnk_social_multiplier()` |
| Predicting finite-$M$ transient dynamics | Tier 2 (approximate) | Direct simulation; verify with `verify_brock_durlauf_reduction()` |
| Predicting tail events / tunnelling between roots | Tier 3 (emergent) | Long simulation runs; metastability analysis |

*[Revised 2026-10-07.]* The Tier 1 rows describe the model under
single-flip logit revision. For a run of the RSiena ministep with $N > 1$
components, the mean-field equilibrium is
`saomnk_inpop_self_consistency()` (the $F_N$ map), and
`bd_self_consistency()` describes it only after the $\kappa_N$ rescaling
and only near $m = 1/2$.

The Landau-Ginzburg expansion of the free energy around the
self-consistent equilibrium is the bridge from Tier 1 to Tier 2. Expanding $\mathcal F(m)$ to fourth order around
$m^*$:

$$
\mathcal F(m) \;\approx\; \mathcal F(m^*) \,+\, \frac{a}{2}(m - m^*)^2 \,+\, \frac{b}{4!}(m - m^*)^4
$$

with

$$
a \;=\; -\beta_{\mathrm{eff}} \,+\, \frac{\tau}{1 - (m^*)^2}, \qquad b \;=\; \frac{2\tau\bigl(1 + 3(m^*)^2\bigr)}{\bigl(1 - (m^*)^2\bigr)^3}.
$$

The quartic coefficient $b$ is **always positive** on $|m^*| < 1$,
so the equilibrium density is super-Gaussian (flatter centre, steeper
walls) whenever $|m^*| > 0$. This is the analytical mechanism by
which positive social interaction frees the basin steepness exponent
$s$ from the Gaussian default $s = 2$. The corresponding searchnet
function is `bd_landau_steepness()`:

```{r landau-steepness, eval=FALSE}
# Quartic coefficient b at a few sample equilibria.
bd_landau_steepness(beta_eff = 0.6, tau = 0.25)
#> $m_star
#> [1] 0
#> $a
#> [1] -0.350
#> $b
#> [1] 0.500
#> $effective_s
#> [1] 2.000

bd_landau_steepness(beta_eff = 1.5, tau = 0.25)
#> $m_star
#> [1] 0.9362
#> $a
#> [1] 0.5377
#> $b
#> [1] 38.91
#> $effective_s
#> [1] ~ 2.8
```

The first call corresponds to the disordered regime ($m^* = 0$,
$s = 2$ Gaussian baseline). The second call corresponds to a strongly
ordered regime ($|m^*| \approx 0.94$, $s \approx 2.8$ super-Gaussian).
The chain $\beta_h \uparrow \Rightarrow |m^*| \uparrow \Rightarrow b \uparrow \Rightarrow s \uparrow$
links the herding coefficient to basin steepness, obtained here as a
consequence of Property 5.

---

# Connection to Other Searchnet Results

Property 5 sits at one corner of a duality diagram with the other
foundational properties:

```
                    SaoMNK CTMC
                  (B, theta, beta)
                 /        |        \
                /         |         \
   Property 1 (NK)   Property 4 (QRE)   Property 5 (B&D)
   ----------       ----------         ----------
   M = 1            Any M, beta < inf  M -> infinity
   beta -> infinity Logit ministep     Mean-field
   No network       Potential game     Congestion only
   ----------       ----------         ----------
   Single rational  Stationary         Many noisy
   actor on a       distribution =     actors with
   rugged           Gibbs measure      social
   landscape        on potential       interaction
                                       (rational
                                        expectations)
```

The Property 1 ↔ Property 5 duality is the most striking. Property 1
shrinks the population to a *single greedy* actor and recovers
NK; Property 5 grows the population to *infinitely many noisy*
actors and recovers Brock-Durlauf. Both are exact reductions to
canonical models from neighbouring fields, and both are special
cases of the same SaoMNK CTMC. The duality justifies two distinct
empirical strategies in the same framework:

- *Within-actor / single-firm* studies (Property 1 limit): NK-style
  search, local-optima trapping, exploration-exploitation
  tradeoff. Use `vignette("saomnk-nk-validation")`.
- *Cross-actor / many-firm* studies (Property 5 limit): peer effects,
  tipping points, social multipliers, susceptibility. Use this
  vignette and `bd_self_consistency()` /
  `saomnk_social_multiplier()`.

The intermediate case ($M$ finite, $\beta$ finite, $K$ non-zero) is
the genuinely new SaoMNK regime: neither Kauffman nor Brock-Durlauf
covers it. Property 4 (QRE equivalence) guarantees that this
intermediate regime is well-posed (stationary distribution exists
and is unique; it is Gibbs under single-flip revision); the $\{K\}$ framework names the four coupled
degree statistics that organize its dynamics.

The Brock-Durlauf reduction also sets up a bridge to congestion
games (Rosenthal, 1973). The H6 congestion term in the SaoMNK
10-component utility *is* the mean-field social-interaction
channel of Property 5; it is also the canonical congestion-game
primitive of Rosenthal's theorem (cited as K1 in Part K of the
proof table). Existence of pure-strategy Nash equilibria in the
SaoMNK game thus follows from K1, while the Gibbs/QRE structure
under noisy single-flip logit choice follows from F7 (Property 4), and the
mean-field rational-expectations limit follows from L10 (Property 5).
The three results are mutually consistent and mutually reinforcing.

---

# Further Reading

**Primary economic foundation.**

- Brock, W. A., & Durlauf, S. N. (2001). Discrete choice with social
  interactions. *Review of Economic Studies*, 68(2), 235--260.
- Durlauf, S. N. (1999). How can statistical mechanics contribute to
  social science? *Proceedings of the National Academy of Sciences*,
  96(19), 10582--10584.
- Glaeser, E. L., & Scheinkman, J. A. (2003). Non-market interactions.
  In *Advances in Economics and Econometrics: Theory and
  Applications, Eighth World Congress*, Volume 1.

**Common ancestor (McFadden 1974) and the prior partial bridge.**

- McFadden, D. (1974). Conditional logit analysis of qualitative
  choice behavior. In P. Zarembka (Ed.), *Frontiers in Econometrics*,
  pp. 105--142. Academic Press. *(The shared random-utility-maximization
  ancestor of both the SAOM and the Brock-Durlauf branches.)*
- Pink, S., Kretschmer, D., & Leszczensky, L. (2020). Choice modelling
  in social networks using stochastic actor-oriented models. *Journal
  of Choice Modelling*, 34, 100202. *(The prior one-directional bridge
  from SAOM to RUM; SaoMNK Properties 4 and 5 extend it to bidirectional
  SAOM $\leftrightarrow$ RUM-with-social-interactions.)*
- Blume, L. E., & Durlauf, S. N. (2003). Equilibrium concepts for
  social interaction models. *International Game Theory Review*,
  5(3), 193--209. *(Equilibrium-concept clarification for the B&D
  branch; rational-expectations $\equiv$ population-distribution
  fixed point.)*
- Durlauf, S. N., & Ioannides, Y. M. (2010). Social interactions.
  *Annual Review of Economics*, 2, 451--478. *(State-of-the-field
  review of the BDB branch.)*
- Snijders, T. A. B. (1996). Stochastic actor-oriented models for
  network change. *Journal of Mathematical Sociology*, 21(1--2),
  149--172. *(Foundation paper for the SAOM branch.)*
- Snijders, T. A. B., Steglich, C. E. G., & Schweinberger, M. (2007).
  Modeling the co-evolution of networks and behavior. In K. van
  Montfort, J. Oud, & A. Satorra (Eds.), *Longitudinal Models in the
  Behavioral and Related Sciences*, pp. 41--71. Lawrence Erlbaum.
  *(Most relevant SAOM variant: joint network-and-behavior
  evolution, which is exactly the SaoMNK setup.)*

**The reflection problem and identification.**

- Manski, C. F. (1993). Identification of endogenous social effects:
  The reflection problem. *Review of Economic Studies*, 60(3),
  531--542.
- Blume, L. E., Brock, W. A., Durlauf, S. N., & Jayaraman, R. (2015).
  Linear social interactions models. *Journal of Political Economy*,
  123(2), 444--496.

**Mean-field statistical mechanics (mathematical foundation).**

- Ellis, R. S. (1985). *Entropy, Large Deviations, and Statistical
  Mechanics*. Springer-Verlag. (Theorem IV.4.1 underlies L9.)
- Krapivsky, P. L., Redner, S., & Ben-Naim, E. (2010). *A Kinetic
  View of Statistical Physics*. Cambridge University Press.
  (Section 3.1 covers the Curie-Weiss mean-field calculation
  underlying Rb3 and Rb4.)
- Levin, D. A., Peres, Y., & Wilmer, E. L. (2017). *Markov Chains
  and Mixing Times* (2nd ed.). American Mathematical Society.
  (Tier 2 finite-$M$ analysis.)

**Companion vignettes.**

- `vignette("saomnk-theory")` --- the $\{K\}$ framework, Properties
  1--7, and the overarching theory.
- `vignette("saomnk-nk-validation")` --- Property 1 verification
  (the dual of this vignette in the single-actor limit).
- `vignette("saomnk-blume-tutorial")` --- Property 4 (Gibbs/QRE
  stationary distribution); the prerequisite for Lemma 6 (L9).

```{r session-info}
sessionInfo()
```
