---
title: "Formal Verification of SAOM-NK Models with Lean 4"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Formal Verification of SAOM-NK Models with Lean 4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(searchnet)
## Chunks that call Lean run only when a toolchain and a built library are
## present AND the builder opts in; otherwise the output recorded below, from a
## real run, is shown instead.
lean_ok <- isTRUE(lean_available(check_build = TRUE)) &&
  identical(Sys.getenv("SEARCHNET_LEAN_VIGNETTE"), "true")
```

searchnet ships a Lean 4 library, `SaomNK`, of machine-checked results about
stochastic actor-oriented models on NK landscapes, and R functions that turn a
concrete model into a Lean file that the library then checks. This vignette
says what is proved, how to export and check a model, and how to prove the next
result yourself. Lean is optional: nothing else in the package needs it.

# What is proved

The library proves general statements once, for every model of a class:

* **Potentials.** A unilateral deviation changes the deviator's utility by
  exactly the change in a potential (Monderer and Shapley 1996), for every
  own-row term (NK fitness, density, outAct, XWX, ego and activity
  covariates, actor-specific landscapes), for activity-weighted crowding
  (`inPop`, with the Rosenthal 1973 potential), and for every symmetric
  pairwise statistic (overlap kernels, four-cycles). Exact potentials add,
  which is what makes the library extensible.
* **Equilibria.** Every such model has a pure-strategy Nash equilibrium, and
  its single-flip-stable configurations are exactly the local maxima of the
  potential.
* **Dynamics.** With binary-logit (Glauber) or Metropolis ministeps and any
  state-independent proposal rates, the Gibbs weight `exp(β Φ)` satisfies
  detailed balance and is stationary (Blume 1993), also under structural zeros
  and ones; every configuration is reachable. As `β → ∞` the SAOM ministep
  becomes adaptive search on an NK landscape (Kauffman 1993; Levinthal 1997).
* **Where potentials fail.** Asymmetric coupling between two layers, an ego
  covariate on a one-mode relational degree, heterogeneous weights on a shared
  pairwise term, and a creation/endowment split whose change statistics
  differ: each provably admits no exact potential.
* **Mean field.** For the two-population logit mean field (Brock and Durlauf
  2001), aligned and polarized equilibria exist iff the relevant coupling
  exceeds one, with their linear stability conditions.

`lean_registry()` lists the headline results with the proof-table steps,
tests and R functions each one underwrites:

```{r registry, eval = requireNamespace("yaml", quietly = TRUE)}
reg <- lean_registry()
nrow(reg)
head(reg[, c("decl", "kind", "proof_table")], 8)
```

# From a model to Lean

`lean_spec()` reads a model definition and maps its RSiena effects onto the
library's statistics. Effects with no Lean counterpart are listed and left out,
so a theorem about the export is never silently a theorem about a different
model.

```{r effect-map}
lean_effect_map()[, c("effect", "slot", "potential")]
```

A two-actor model on a small NK landscape, with crowding, four-cycles, an XWX
synergy term and an ego covariate:

```{r spec}
nk <- nk_landscape(N = 3, K = 1, seed = 11)
model <- list(
  M = 2, N = 3, landscape = nk,
  theta = c(nk = 1, density = -0.25, outAct = 0.05, inPop = -0.2,
            cycle4 = 0.1, XWX = 0.3, egoX = 0.5),
  W = matrix(c(0, 1, 0, 1, 0, 1, 0, 1, 0), 3), ego = c(1, 2),
  config = matrix(c(1, 0, 1, 0, 1, 1), 2, byrow = TRUE))
spec <- lean_spec(model)
spec
```

`lean_export_model()` writes the Lean file. Every double is written as the
exact binary fraction it is, so nothing is rounded between R and Lean. The file
name is a hash of the content: the same model always exports to the same file.

```{r export}
out <- file.path(tempdir(), "lean_vignette")
f <- lean_export_model(spec, dir = out)
basename(f)
attr(f, "facts")
```

The file defines the model as a `RatSpec` and states, among others:

```{r show-file}
txt <- readLines(f, encoding = "UTF-8")
cat(grep("^theorem ", txt, value = TRUE), sep = "\n")
```

The general theorems are instantiated, not re-proved: for example
`exact_potential` is `S.exact_potentialQ h`. The numeric facts (the
utilities at the current configuration, the number of local optima) are proved
by kernel evaluation, `decide +kernel`, and a library theorem
(`RatSpec.countLocalOpt_eq_card`) says what the counted number means: the
number of configurations at which no actor gains from a single flip. When the
'gmp' package is installed the utilities are stated exactly; otherwise they are
bracketed to `1e-9`.

The index convention is checked, not assumed: `index_convention` proves that
the binary code of each actor's portfolio, least significant bit first, is the
row of the landscape table R uses (`1 + sum(b * 2^(0:(N-1)))`).

# Checking

```{r check, eval = lean_ok}
res <- lean_check(out)
res[, c("decl", "status", "axioms")]
```

Output of that chunk in a run with Lean 4 (`v4.34.0-rc2`) and Mathlib
(`v4.34.0-rc2`):

```
#>                                    decl status                                axioms
#> 1          SaomNKInstance.h…E_regular     ok propext, Classical.choice, Quot.sound
#> 2       SaomNKInstance.h…C_consistent     ok propext, Classical.choice, Quot.sound
#> 3      SaomNKInstance.h…nk_is_fitness     ok propext, Classical.choice, Quot.sound
#> 4    SaomNKInstance.h…exact_potential     ok propext, Classical.choice, Quot.sound
#> 5        SaomNKInstance.h…exists_nash     ok propext, Classical.choice, Quot.sound
#> 6   SaomNKInstance.h…gibbs_stationary     ok propext, Classical.choice, Quot.sound
#> 7       SaomNKInstance.h…greedy_limit     ok propext, Classical.choice, Quot.sound
#> 8   SaomNKInstance.h…index_convention     ok                               propext
#> 9       SaomNKInstance.h…utility_B0_0     ok propext, Classical.choice, Quot.sound
#> 10      SaomNKInstance.h…utility_B0_1     ok propext, Classical.choice, Quot.sound
#> 11      SaomNKInstance.h…potential_B0     ok propext, Classical.choice, Quot.sound
#> 12   SaomNKInstance.h…count_local_opt     ok propext, Classical.choice, Quot.sound
#> 13 SaomNKInstance.h…count_local_opt_card  ok propext, Classical.choice, Quot.sound
```

`status == "ok"` means the declaration depends on no axiom beyond the three
standard ones. A `sorry` would show as `"sorry"`; a fact proved with
`native_decide`, which trusts the compiler, as `"native"`. In a test suite,
`expect_lean_theorem(res, "exact_potential")` turns this into an expectation.

Size limits: the landscape table has `2^N` rows and the count enumerates
`2^(MN)` configurations, so the export stops above `max_N = 6` and skips the
count above `max_cells = 8` unless `native = TRUE`, which is flagged in the file
and in `lean_check()`.

## Setting up Lean

```{r setup-lean, eval = FALSE}
lean_setup()   # prints the commands; runs them only after you confirm
## reuse a Mathlib already built on this machine instead of downloading one:
lean_setup(packages_dir = "/path/to/project/.lake/packages")
```

# Proving the next result

Every statistic in the library follows one pattern, and a new one is added the
same way. The file `inst/lean/SaomNK/Examples/WeightedOverlap.lean` does it for
a statistic the library did not have, activity-weighted overlap
`Σ_k Σ_j v_j b_ij b_kj`:

```lean
/-- Step 1: the kernel on two rows. -/
def wOverlap (v : Fin N → ℝ) (b b' : Fin N → Bool) : ℝ :=
  ∑ j, v j * (ind (b j) * ind (b' j))

/-- Step 2: it is symmetric. -/
lemma wOverlap_symm (v : Fin N → ℝ) (b b' : Fin N → Bool) :
    wOverlap v b b' = wOverlap v b' b := by
  unfold wOverlap
  exact Finset.sum_congr rfl fun j _ => by ring

/-- Step 3: its exact potential, from the row-pair building block. -/
theorem isExactPotential_wOverlap (v : Fin N → ℝ) :
    IsExactPotential (fun i (B : Config M N) => rowPairStat (wOverlap v) B i)
      (rowPairPotential (wOverlap v)) :=
  isExactPotential_rowPair _ (wOverlap_symm v)

/-- Step 4: add it to the core specification. -/
theorem core_plus_wOverlap (S : CoreSpec M N) (θ : ℝ) (v : Fin N → ℝ) :
    IsExactPotential (fun i B => S.utility i B + θ * rowPairStat (wOverlap v) B i)
      (fun B => S.potential B + θ * rowPairPotential (wOverlap v) B) :=
  (show IsExactPotential S.utility S.potential from
      fun _ _ _ h => S.exact_potential h).add
    ((isExactPotential_wOverlap v).smul θ)
```

Nash existence and Gibbs stationarity of the extended model then follow from
the general theorems with one line each (`exists_nash_of_exactPotential`,
`global_balance_of_potential`); the example file states both. Two steps of
real work, symmetry and the choice of building block, and the rest is
bookkeeping the library already did.

When a statistic has no potential, the same structure gives the negative
result: compute the change statistics around a four-move cycle and show their
sum is nonzero (`mixed_cycle_sum`, `TwoLayer.cycle_affinity`,
`dir_two_cycle`); an exact potential would force it to zero.

To make a new statistic exportable from R, add a computable `ℚ` version in
`SaomNK/Compute/RatSpec.lean`, extend `RatSpec.cast_utility` (the proof that
the computable value is the cast of the real one), map the RSiena effect in
`lean_spec()`, and record the result in `inst/lean/registry.yml`.

# What is not proved

* That a finite irreducible chain has a unique stationary law to which it
  converges. With global balance and reachability, which are proved, this is
  the textbook fact that completes the argument.
* Gibbs stationarity under RSiena's default multinomial ministep (choice among
  all flips and no change). The exporter does not claim it for
  `revision = "multinomial"`.
* State-dependent rate functions, root (`#`-parameter) forms of statistics,
  and statistics such as `outActSqrt` whose values are irrational (own-row, so
  the potential claim holds, but they are not exported numerically).

# References

Blume, L. E. (1993). The statistical mechanics of strategic interaction.
*Games and Economic Behavior*, 5, 387-424.

Brock, W. A., & Durlauf, S. N. (2001). Discrete choice with social
interactions. *Review of Economic Studies*, 68, 235-260.

Kauffman, S. A. (1993). *The Origins of Order*. Oxford University Press.

Levinthal, D. A. (1997). Adaptation on rugged landscapes. *Management Science*,
43, 934-950.

Monderer, D., & Shapley, L. S. (1996). Potential games. *Games and Economic
Behavior*, 14, 124-143.

Rosenthal, R. W. (1973). A class of games possessing pure-strategy Nash
equilibria. *International Journal of Game Theory*, 2, 65-67.

Snijders, T. A. B. (2001). The statistical evaluation of social network
dynamics. *Sociological Methodology*, 31, 361-395.
