---
title: "Causal Inference with searchnet: DID, Synthetic Control, and Regression Discontinuity"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Causal Inference with searchnet: DID, Synthetic Control, and RD}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

```{r load-saomnk, include=FALSE}
if (requireNamespace("searchnet", quietly = TRUE)) {
  library(searchnet)
} else {
  dir_r <- file.path(getwd(), "R")
  if (!file.exists(file.path(dir_r, "utils.R")))
    dir_r <- "D:/Search_networks/SaoMNK/R"
  .saomnk_dir <- dir_r
  .saomnk_dir <- dir_r; source(file.path(dirname(dir_r), "inst", "saomnk-loader.R"))
  source(file.path(dir_r, "searchnet-causal.R"))
}

library(ggplot2)

DV_NAME <- "self$bipartite_rsienaDV"
```

# 1. Overview: Theta Shocks as Natural Experiments

Searchnet's `theta_shocks` mechanism changes the parameters of the SAOM
objective function at a designated simulation step.
Before the shock, actors search and adapt under one parameter regime
(baseline); after the shock, the rules change---a parameter is strengthened,
weakened, or reversed.

This creates a **natural experiment within the simulation**:

- **Treatment onset** is the `shock_step` (the boundary between regimes).
- **Treated actors** can be all actors (aggregate shock) or a designated
  subset (heterogeneous shock via strategy groups).
- **Control actors** are those unaffected by the parameter change, or a
  comparison simulation run without the shock.

The `actor_util_df` and K-4 trajectory data (`K_AC_df`, `K_CA_df`,
`K_AA_df`, `K_CC_df`) are panel data indexed by `(actor_id, chain_step_id)`.
This panel structure maps directly to three causal inference designs:

| Design | Package | When to use |
|--------|---------|-------------|
| Difference-in-Differences | `did` | Multiple treated and control actors with staggered or common treatment timing |
| Synthetic Control | `Synth` | Single treated actor (or small group); construct a data-driven counterfactual from control actors |
| Regression Discontinuity | `rdrobust` | Sharp change at the shock step; estimate the local treatment effect at the discontinuity |

# 2. Setting Up a Shocked Simulation

We use a small environment (M=6 actors, N=8 components) with a two-phase
shock design: the first half of the simulation runs under a baseline density
parameter, and the second half runs under a shocked (more negative) density.

```{r sim-setup}
set.seed(2026)
env <- saomnk_env(M = 6, N = 8, seed = 42)

K <- saomnk_block_diagonal(8, 2)

mod <- saomnk_model(
  density  = -0.5,
  popularity = 0.15,
  influence_matrix = K
)

# Two-phase shock: baseline (portion=1) then shocked (portion=1)
s_baseline <- saomnk_shock("density", parameter = -0.5, portion = 1)
s_shocked  <- saomnk_shock("density", parameter = -2.0, portion = 1)

saomnk_run(env, mod, steps_per_actor = 20, seed = 12345,
           shocks = list(s_baseline, s_shocked))
```

The shock divides the simulation roughly in half.  We compute the shock step
as half the total number of steps:

```{r shock-step}
total_steps <- max(env$actor_util_df$chain_step_id)
shock_step  <- round(total_steps / 2)
cat("Total steps:", total_steps, "\nShock step:", shock_step, "\n")
```


# 3. Difference-in-Differences Example

DID requires both treated and control actors.  In this design, we designate
actors 1--3 as "treated" (most affected by the density shock because of
their initial positions) and actors 4--6 as controls.

```{r did-panel}
panel_did <- searchnet_causal_panel(
  env,
  shock_step     = shock_step,
  outcome        = "utility",
  treated_actors = c(1, 2, 3)
)

head(panel_did)
cat("\nPanel dimensions:", nrow(panel_did), "rows x", ncol(panel_did), "cols\n")
cat("Treated actors:", sum(panel_did$treated == 1) / length(unique(panel_did$step)),
    "\nControl actors:", sum(panel_did$treated == 0) / length(unique(panel_did$step)), "\n")
```

Now we run the Callaway & Sant'Anna (2021) estimator:

```{r did-estimate}
if (requireNamespace("did", quietly = TRUE)) {
  tryCatch({
    att <- searchnet_did(panel_did)
    summary(att)
  }, error = function(e) {
    message("DID estimation encountered an issue: ", e$message)
  })
} else {
  message("Package 'did' not installed. Install with: install.packages('did')")
}
```

The event study plot shows the dynamic treatment effect over time:

```{r did-plot, fig.cap="Event study plot from DID estimation"}
if (requireNamespace("did", quietly = TRUE)) {
  tryCatch({
    att <- searchnet_did(panel_did)
    p <- searchnet_causal_plot(att, type = "did")
    print(p)
  }, error = function(e) {
    message("DID plot skipped: ", e$message)
  })
} else {
  message("Skipping DID plot (package 'did' not installed)")
}
```


# 4. Synthetic Control Example

Synthetic control is ideal when a single actor (or a small group) is treated
and we want to construct a data-driven counterfactual from the remaining
control actors.  Here we treat actor 1 as the sole treated unit:

```{r synth-panel}
panel_sc <- searchnet_causal_panel(
  env,
  shock_step     = shock_step,
  outcome        = "utility",
  treated_actors = c(1)
)

cat("Treated unit: actor 1\n")
cat("Control units:", paste(setdiff(unique(panel_sc$actor_id[panel_sc$treated == 0]),
                                     character(0)), collapse = ", "), "\n")
```

Run the synthetic control estimator:

```{r synth-estimate}
if (requireNamespace("Synth", quietly = TRUE)) {
  tryCatch({
    sc_result <- searchnet_synth(
      panel_sc,
      treated_unit = 1
    )
    cat("Synthetic control weights:\n")
    print(round(sc_result$synth_out$solution.w, 3))
    cat("\nPre-treatment MSPE:",
        mean(sc_result$gap$gap[sc_result$gap$step < shock_step]^2), "\n")
  }, error = function(e) {
    message("Synthetic control encountered an issue: ", e$message)
  })
} else {
  message("Package 'Synth' not installed. Install with: install.packages('Synth')")
}
```

Plot the treated vs. synthetic control trajectory:

```{r synth-plot, fig.cap="Treated actor vs. synthetic counterfactual"}
if (requireNamespace("Synth", quietly = TRUE)) {
  tryCatch({
    sc_result <- searchnet_synth(panel_sc, treated_unit = 1)
    p <- searchnet_causal_plot(sc_result, type = "synth")
    print(p)
  }, error = function(e) {
    message("Synth plot skipped: ", e$message)
  })
} else {
  message("Skipping Synth plot (package 'Synth' not installed)")
}
```


# 5. Regression Discontinuity Example

The RD design treats the shock step as a sharp cutoff in a continuous running
variable (the simulation step).  The local polynomial estimator from
`rdrobust` identifies the causal effect at the boundary by comparing the
outcome just before and just after the shock.

This design works well when all actors are treated simultaneously and we
want to estimate the immediate local effect of the parameter change:

```{r rd-panel}
panel_rd <- searchnet_causal_panel(
  env,
  shock_step = shock_step,
  outcome    = "utility"
)

cat("RD panel: all actors treated at step", shock_step, "\n")
cat("Pre-treatment steps:", sum(unique(panel_rd$step) < shock_step), "\n")
cat("Post-treatment steps:", sum(unique(panel_rd$step) >= shock_step), "\n")
```

Run the RD estimator:

```{r rd-estimate}
if (requireNamespace("rdrobust", quietly = TRUE)) {
  tryCatch({
    rd_result <- searchnet_rd(panel_rd)
    summary(rd_result$rd)
  }, error = function(e) {
    message("RD estimation encountered an issue: ", e$message)
  })
} else {
  message("Package 'rdrobust' not installed. Install with: install.packages('rdrobust')")
}
```

Plot the RD with pre- and post-treatment fitted curves:

```{r rd-plot, fig.cap="Regression discontinuity at the shock step"}
if (requireNamespace("rdrobust", quietly = TRUE)) {
  tryCatch({
    rd_result <- searchnet_rd(panel_rd)
    p <- searchnet_causal_plot(rd_result, type = "rd")
    print(p)
  }, error = function(e) {
    message("RD plot skipped: ", e$message)
  })
} else {
  message("Skipping RD plot (package 'rdrobust' not installed)")
}
```


# 6. Using K-4 Outcomes

All three methods work with any of the K-4 coupled degree trajectories,
not just utility.  For example, to analyze the effect of the shock on
actor scope (K_AC):

```{r k4-example}
panel_kac <- searchnet_causal_panel(
  env,
  shock_step     = shock_step,
  outcome        = "K_AC",
  treated_actors = c(1, 2, 3)
)

cat("K_AC panel dimensions:", nrow(panel_kac), "x", ncol(panel_kac), "\n")
head(panel_kac)
```

This panel can be passed to any of `searchnet_did()`, `searchnet_synth()`,
or `searchnet_rd()` in the same way as the utility panels above.


# 7. Comparison: When to Use Which Method

| Criterion | DID | Synthetic Control | RD |
|-----------|-----|-------------------|-----|
| **Number of treated units** | Multiple | One (or few) | Any |
| **Requires control group** | Yes | Yes | No |
| **Parallel trends assumption** | Required | Relaxed (data-driven weights) | Not needed |
| **Estimates** | ATT over all post-periods | Counterfactual trajectory | Local effect at cutoff |
| **Best for** | Comparing strategy groups | Single-actor counterfactual | Immediate shock impact |
| **searchnet use case** | Strategy group A shocked, B unshocked | One actor's parameter changed | Aggregate shock to all actors |

**Practical guidance:**

- **DID** is the natural choice when `theta_shocks` differentially affect
  actor subgroups (e.g., one strategy group's density parameter is shocked
  while another's is held constant).  The `did` package handles staggered
  treatment timing if shocks arrive at different steps for different groups.

- **Synthetic control** excels when a single actor experiences a unique
  shock.  The method constructs an optimal weighted combination of control
  actors that matches the treated actor's pre-shock trajectory.

- **RD** is appropriate when the shock affects everyone simultaneously and
  the researcher wants to estimate the immediate local effect at the
  discontinuity.  It requires no explicit control group but estimates only
  the effect at the boundary, not the full post-treatment path.
