---
title: "Degree Bounds in SAOM-NK: Floors and Caps on Portfolios"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Degree Bounds: Floors and Caps on Portfolios}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What a degree bound is here

A degree bound restricts how many components an actor may hold (its portfolio
size, the actor's outdegree in the bipartite network) or how many actors may
hold a component (the component's indegree). A floor of 1 on actors says that
no actor may be left holding nothing: an actor that wants to switch from
component A to component B must add B first and drop A afterwards. A cap says
that no actor may hold more than `max` components.

In SAOM-NK a bound is a **fixed modeling assumption**. It holds in forward
simulation and is imposed, unchanged, when the same model is estimated. It is
never an estimated parameter, and searchnet never reports an estimate for it.

## How each bound is carried in RSiena

Everything below was verified on RSiena 1.5.0 with a two-mode dependent
network.

| Bound | RSiena mechanism | Exact? |
|---|---|---|
| actor max `k` | `MaxDegree = k` algorithm setting | exact: options above the cap are removed |
| actor min 1 | `outIso` (number of out-isolates), fixed at `-bound_penalty` | penalty |
| actor min `k > 1` | `outTrunc(k)` (sum of outdegrees truncated at `k`), fixed at `+bound_penalty` | penalty |
| component min 1, 2, 3 | `antiInIso` (+ `in2Plus`, + `in3Plus`), fixed at `+bound_penalty` | penalty |
| component max 1, 2 | `in2Plus` or `in3Plus`, fixed at `-bound_penalty` | penalty (barrier) |
| component min above 3, component max above 2 | none | **non-implementation** |

With the default `bound_penalty = 20`, a move across a penalized bound has a
choice probability of about `exp(-20)`, roughly 2e-9, relative to staying put.
The bound is therefore soft in principle, and searchnet checks it: every
simulated period ends with a gate that aborts the run if any bound was
crossed.

Two consequences of the penalty form are worth knowing.

* **A floor also attracts.** An actor below its floor gains `+20` by adding a
  component, so it adds one at its first opportunity. A start state below the
  floor would be repaired almost at once, but it would not satisfy the stated
  bounds, so searchnet stops by default and lists the violators.
  `repair_initial = TRUE` adds seeded random ties first instead, and says so.
* **A component cap is a barrier only.** The indicator `in3Plus` stops a
  component from gaining a third holder, but a component that already has four
  is not pulled back. A start state above any cap is always an error.

**Non-implementations.** RSiena 1.5.0 offers indegree indicators only up to
`in3Plus`, no indegree analogue of `outTrunc`, and `MaxDegree` applies to
outdegrees only. A component floor above 3 or a component cap above 2 is
refused with an error of class `"searchnet_not_implemented"`. That says
nothing about whether such a bound matters in the world; it says the software
offers no statistic for it.

**Maximum likelihood.** Estimation under bounds uses the method of moments.
RSiena 1.5.0 itself warns that `maxlike` and `MaxDegree` are incompatible, and
an ML fit of a bipartite network with `MaxDegree` stopped with a NaN score;
with only a fixed floor penalty, ML returned a missing standard error for a
period rate. searchnet refuses `maxlike = TRUE` whenever bounds are set.

## Declaring bounds

```{r declare}
library(searchnet)
b <- searchnet_degree_bounds(actor = c(min = 1, max = 3))
b
mod <- saomnk_model(density = -1.5, popularity = 0.1,
                    degree_bounds = c(min = 1, max = 3))
mod
```

The same `degree_bounds` argument is accepted by `searchnet_recovery()`,
`searchnet_coevolve()`, `run_calibrated_counterfactual()` and
`run_counterfactual_with_uncertainty()`.

A simulation under these bounds is an ordinary `saomnk_run()`:

```{r run, eval = FALSE}
env <- saomnk_env(M = 8, N = 6, density = 0.3, seed = 7)
saomnk_run(env, saomnk_model(density = -1.5, degree_bounds = c(min = 1, max = 3),
                             repair_initial = TRUE),
           steps_per_actor = 20, seed = 11)
range(apply(env$bi_env_arr, 3, rowSums))   # 1 3
```

## Checking observed data before estimation

Estimation under a bound assumes the bound held in every observed wave. A
method-of-moments fit cannot reproduce an observed degree the model forbids,
so a violating panel is an error before estimation, not a fit to report.
`searchnet_check_degree_bounds()` is called automatically by every estimation
entry point when bounds are set, and can be called directly:

```{r check}
B1 <- matrix(c(1, 0, 0,
               0, 1, 1), 2, 3, byrow = TRUE)
B2 <- B1; B2[1, 1] <- 0          # actor 1 drops its last component
try(searchnet_check_degree_bounds(list(B1, B2), c(min = 1)))
```

## Landscapes: infeasible portfolios

A landscape plot that enumerates all `2^N` portfolios of an actor would color
an infeasible portfolio by the fixed penalty, a utility near -20 that is a
modeling device rather than a payoff. `searchnet_portfolio_infeasible()` flags
those portfolios so a figure can grey them out:

```{r infeasible}
configs <- as.matrix(expand.grid(rep(list(0:1), 3)))
cbind(configs, infeasible = searchnet_portfolio_infeasible(configs, c(min = 1, max = 2)))
```

No existing figure changes: a plotting script opts in by calling it. The
README landscape script (`tools/make_readme_gif.R`) adds the flag to its JSON
only when its model carries bounds, which the README model does not.

## Choice probabilities under bounds

`env$compute_choice_probabilities()` reports, per actor, the logit
probabilities of each single toggle and of staying put. Under bounds it follows
the two mechanisms, which differ in kind:

* **An actor cap is a hard constraint.** `MaxDegree` removes the add options of
  an actor at the cap from its choice set, so they get probability exactly 0
  and the remaining toggles and the status quo are renormalized. Against
  RSiena's own simulated ministeps from a cap state, an actor never added and
  the frequency of staying put matched the renormalized probability within
  Monte Carlo error.
* **A floor or component cap is a fixed penalty.** The crossing move stays in
  the choice set with its weight multiplied by `exp(-20)`, as in the
  simulation; it is not set to 0, because the model is soft there.

Each element also carries `infeasible` (toggles that would leave any bound) and
`removed` (toggles `MaxDegree` removes). `renormalize = FALSE` returns the
unbounded formal-utility logit for comparison.

## Evidence: recovery with and without the bound

**What this section shows, and what it does not.** It shows recovery at one
design: M = 30 actors, N = 10 components, 3 waves, 100 replications, an actor
floor of 1, and the planted values below. Nothing more. It says nothing about
other sizes, densities, effects, or bounds.

The planted data-generating process has an actor floor of 1 (`outIso` fixed at
-20), `density = -1.8`, `inPop = 0.08` and a basic rate of 4 per period. Wave 1
is a Bernoulli draw at density 0.12, brought within the floor by seeded random
additions. Each simulated panel is estimated twice, on the same panels and
with the same seeds:

* **with the bound**: the panel is checked, `outIso` enters fixed at -20, and
  the method of moments estimates rate, `density` and `inPop`;
* **without the bound**: the same estimation with no `outIso` term, which is
  what an analyst who did not know about the floor would fit.

The runs take a few minutes, so the chunk that runs them is shown with
`eval = FALSE`. The printed output comes from results precomputed with exactly
that code by `tools/make_degree_bounds_vignette_data.R` and stored in
`inst/extdata/searchnet_degree_bounds_recovery.rds`.

```{r heavy, eval = FALSE}
design <- list(
  effects_spec = c("density", "inPop"),
  theta_true = c(rate = 4, density = -1.8, inPop = 0.08),
  M = 30L, N = 10L, waves = 3L, reps = 100L,
  nbrNodes = 1L, seed = 20261010L,
  n3 = 500L, nsub = 4L, init_density = 0.12,
  degree_bounds = c(min = 1),
  criterion = list(max_abs_bias = 0.10, min_coverage = 0.90,
                   min_converged_share = 0.90))
with_bounds    <- do.call(searchnet_recovery, c(design, list(estimate_with_bounds = TRUE)))
without_bounds <- do.call(searchnet_recovery, c(design, list(estimate_with_bounds = FALSE)))
```

```{r load}
path <- system.file("extdata", "searchnet_degree_bounds_recovery.rds",
                    package = "searchnet")
if (!nzchar(path))  # building from a source tree without an installed copy
  path <- file.path("..", "inst", "extdata", "searchnet_degree_bounds_recovery.rds")
stored <- readRDS(path)
```

How often the floor binds in the simulated data (share of actors holding
exactly one component, share holding none, and mean portfolio size, by wave):

```{r binding}
round(stored$floor_binding, 3)
```

Estimated **with** the bound:

```{r with}
stored$with_bounds
```

Estimated **without** the bound, on the same panels:

```{r without}
stored$without_bounds
```

Side by side, for the two structural effects:

```{r compare}
pick <- function(x, arm) {
  t <- x$table[x$table$effect %in% c("density", "inPop"),
               c("effect", "true", "bias", "mcse_bias", "coverage", "rmse")]
  cbind(arm = arm, t)
}
cmp <- rbind(pick(stored$with_bounds, "with bound"),
             pick(stored$without_bounds, "without bound"))
num <- vapply(cmp, is.numeric, logical(1))
cmp[num] <- lapply(cmp[num], round, 3)
cmp
c(with_bound_seconds = stored$with_bounds$elapsed,
  without_bound_seconds = stored$without_bounds$elapsed)
```

**Reading.** The bias, coverage and verdict lines above are the result; the
criterion (absolute bias at most 0.10, coverage at least 0.90, at least 90
percent of replications converged) was fixed in the generating script before
the run. Omitting a floor that generated the data does not leave the model
merely less efficient: the estimator has to explain, through the remaining
effects, why actors at one component never drop it, and the effects it uses
for that absorb the bound. Which effects absorb it, and by how much, is
specific to this design.
