---
title: "From Your Own Data to {K} Readings"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{From Your Own Data to {K} Readings}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4, out.width = "100%")
library(searchnet)
```

SAOM-NK works on `B`, an actor-by-component matrix with one slice per period.
Observed data rarely arrive in that shape. They arrive in long format: one row
per actor, component and period, sometimes with a weight. This vignette goes
from such records to `B`, reads the four {K} dimensions off it, and hands the
result to the estimation layer.

## 1. Long data

The example is synthetic: ten actors, eight components, four periods, and a
numeric weight on each record (an intensity, a count, a share).

```{r data}
set.seed(42)
grid <- expand.grid(actor = sprintf("actor_%02d", 1:10),
                    component = sprintf("comp_%d", 1:8),
                    period = 1:4, stringsAsFactors = FALSE)
long <- grid[runif(nrow(grid)) < 0.35, ]
long$weight <- round(rexp(nrow(long), 2), 2)
head(long)
```

## 2. Long data to B

`searchnet_bipartite_from_long()` builds the array. A tie is a summed weight
above `threshold`; with `weight = NULL` any record is a tie. Passing `actors`
fixes the universe, so an actor with no records in a period, or none at all,
keeps a row of zeros and composition is stable across waves.

```{r build}
obs <- searchnet_bipartite_from_long(
  long, actor = "actor", component = "component", period = "period",
  weight = "weight", threshold = 0.1,
  actors = sprintf("actor_%02d", 1:11),   # actor_11 never appears
  components = sprintf("comp_%d", 1:8))
obs
dim(obs$B)
```

## 3. {K} readings

`searchnet_k_readings()` returns one row per node and period. The definitions
are those of `system.file("rosetta", "K_DIMENSIONS.md", package = "searchnet")`
and are computed exactly as the simulation engine computes its {K}-4 degrees:

- **Expansiveness** `K_AC`: components an actor holds (row sum of `B`).
- **Popularity** `K_CA`: actors holding a component (column sum).
- **Sociality** `K_AA`: other actors sharing at least one component.
- **Epistasis** `K_CC`: other components co-held with a component.

`K_AA` and `K_CC` exclude the node itself. The weighted forms,
`sociality_strength` and `epistasis_strength`, are reported alongside, and the
accounting identities that tie the four together (total ties; second moment of
one margin equals total strength in the other projection) are checked for every
period.

```{r readings}
k <- searchnet_k_readings(obs)
head(k[k$level == "actor", c("period", "id", "K_AC", "K_AA", "sociality_strength")])
head(k[k$level == "component", c("period", "id", "K_CA", "K_CC", "epistasis_strength")])
```

The `"summary"` attribute gives the per-period totals: density, mean degrees,
mean pairwise Jaccard overlap of portfolios, and `tie_jaccard`, the overlap of
each period's tie set with the one before it. Values that are very low warn
that a SAOM will see too much change between waves.

```{r summary}
s <- attr(k, "summary")
s[, c("period", "ties", "density", "mean_K_AC", "mean_K_AA",
      "mean_K_CA", "mean_K_CC", "tie_jaccard")]
```

## 4. Plot

Mean degree per period, one panel per dimension, in the order of the {K}-4
panel.

```{r plot}
long_means <- rbind(
  data.frame(period = s$period, dimension = "Expansiveness (K_AC)", mean = s$mean_K_AC),
  data.frame(period = s$period, dimension = "Popularity (K_CA)",    mean = s$mean_K_CA),
  data.frame(period = s$period, dimension = "Sociality (K_AA)",     mean = s$mean_K_AA),
  data.frame(period = s$period, dimension = "Epistasis (K_CC)",     mean = s$mean_K_CC))
long_means$dimension <- factor(long_means$dimension, levels = unique(long_means$dimension))
ggplot2::ggplot(long_means, ggplot2::aes(period, mean)) +
  ggplot2::geom_line() + ggplot2::geom_point() +
  ggplot2::facet_wrap(~ dimension, nrow = 1, scales = "free_y") +
  ggplot2::labs(x = "Period", y = "Mean degree") +
  ggplot2::theme_minimal()
```

## 5. Into estimation

`obs` goes straight to `searchnet_coevolve_data()` as the bipartite dependent
network, and `searchnet_k_covariate()` reshapes any reading into the
node-by-period matrix it takes for a varying covariate. A degree of the
dependent network entered this way is a lagged reading of the network itself,
a different estimand from the endogenous effects (`outAct`, `inPop`) that model
the same degree as a mechanism.

```{r coevolve, eval = requireNamespace("RSiena", quietly = TRUE)}
arch <- array(rbinom(8 * 8 * 4, 1, 0.25), dim = c(8, 8, 4))
dat <- searchnet_coevolve_data(
  bipartite = obs, component = arch,
  actor_covars = list(scope = searchnet_k_covariate(k, "K_AC")),
  verbose = FALSE)
class(dat)
```
