---
title: "Planted-Truth Recovery and Power for a Bipartite SAOM"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Planted-Truth Recovery and Power}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What this vignette shows, and what it does not

`searchnet_recovery()` plants a parameter vector in an actor-by-component
SAOM, simulates panels from it, re-estimates each panel with RSiena under one
fixed algorithm, and reports how well each planted value comes back. It is the
planted-truth gate to run before trusting a SAOM-NK design: the design should
return zero where nothing is planted, and recover what is planted.

**The results below show recovery at one design: M = 30 actors, N = 12
components, 3 waves, 100 replications, initial density 0.2, and the planted
values stated. Nothing more.** They say nothing about a smaller panel, a
sparser network, more effects, or a different estimation algorithm. Run the
harness at your own design.

**How the results were produced.** The two runs take a few minutes, so the
chunks that run them are shown with `eval = FALSE`. The printed output comes
from results precomputed with exactly that code by
`tools/make_recovery_vignette_data.R` and stored in
`inst/extdata/searchnet_recovery_vignette.rds`, which the vignette loads.

```{r load}
library(searchnet)
path <- system.file("extdata", "searchnet_recovery_vignette.rds",
                    package = "searchnet")
if (!nzchar(path))  # building from a source tree without an installed copy
  path <- file.path("..", "inst", "extdata", "searchnet_recovery_vignette.rds")
stored <- readRDS(path)
```

## The design

Two effects on the bipartite network: outdegree (`density`) and indegree
popularity (`inPop`), the focal effect. The basic rate is 5 per period. These
values keep the network near its starting density of 0.2 over the panel, so
every period carries change for the estimator to use.

The estimation algorithm is the same for every replication: unconditional
method of moments (`cond = FALSE`, so period rates are estimated with standard
errors and compare directly with the planted rate), `n3 = 500`, `nsub = 4`,
`nbrNodes = 1`, and a seed derived per replication from `seed`. Hold
`nbrNodes` and `seed` fixed when comparing two harness runs: `nbrNodes` enters
the algorithm, so changing it changes the estimates.

The criteria were written into the generating script before it was run.

## Recovery

```{r recovery-run, eval = FALSE}
rec <- searchnet_recovery(
  effects_spec = c("density", "inPop"),
  theta_true   = c(rate = 5, density = -1.5, inPop = 0.10),
  M = 30, N = 12, waves = 3, reps = 100,
  nbrNodes = 1, seed = 20261010, n3 = 500, nsub = 4,
  focal = "inPop",
  criterion = list(max_abs_bias = 0.03, min_coverage = 0.90,
                   min_converged_share = 0.90))
```

```{r recovery-print}
rec <- stored$recovery
rec
```

Reading the columns: `bias` and `rmse` are against the planted value;
`coverage` is the share of 95% Wald intervals containing it; `reject` is the
share of fits rejecting H0: theta = 0 at alpha = 0.05, which is power when the
planted value is nonzero. `mde` is the approximate minimum detectable effect
at 80% power, (z_{0.975} + z_{0.80}) times the mean estimated standard error.
It treats the Wald statistic as normal with the SE fixed at its planted-value
mean, so read it as a scale, not an exact threshold.

`summary()` adds the Monte Carlo standard errors, the replication status
table and each verdict check:

```{r recovery-summary}
summary(rec)
```

```{r recovery-plot, fig.width = 6, fig.height = 3}
plot(rec)
```

## Null mode: the size check

With `null = TRUE`, the focal planted value is set to zero and everything else
is unchanged. The rejection rate on the focal effect is then the empirical
size of the test, and should sit near alpha.

```{r null-run, eval = FALSE}
rec0 <- searchnet_recovery(
  effects_spec = c("density", "inPop"),
  theta_true   = c(rate = 5, density = -1.5, inPop = 0.10),
  M = 30, N = 12, waves = 3, reps = 100,
  nbrNodes = 1, seed = 20261010, n3 = 500, nsub = 4,
  focal = "inPop", null = TRUE,
  criterion = list(max_size = 0.10, min_converged_share = 0.90))
```

```{r null-print}
rec0 <- stored$null
rec0
```

The null design did not estimate as cleanly as the recovery design: with the
popularity effect removed, some panels give RSiena a period rate it cannot
bound. Those replications are listed rather than dropped:

```{r null-failures}
rec0$reps[rec0$reps$status != "converged", ]
```

That is a property of this design at this size, and it is part of the result.

## Replications that do not converge

A replication counts as converged when the overall maximum convergence ratio
(`tconv.max`) is below 0.25 and every parameter has a finite standard error.
The others are filed as `not_converged`, `non_identified` (a missing standard
error: a non-identification, not a fit), `error_simulation` or
`error_estimation`. Every status is counted and printed. Summaries use the
converged replications only, and `rec$estimates` keeps the estimates of the
rest for inspection. A design where many replications fail to converge has
answered a question too: it is not estimable at that size.

## Reporting a null with its bound

When an empirical estimate of the focal effect is not distinguishable from
zero, report the harness MDE at the empirical design alongside it: the null
then reads as "no effect larger than about the MDE", a bound on the
association, rather than as evidence of absence.

## A fast check

The same call with tiny settings runs in a few seconds and is what the test
suite uses. It is a smoke test of the machinery, not evidence about any
design.

```{r tiny, eval = FALSE}
searchnet_recovery(c("density", "inPop"),
                   theta_true = c(rate = 3, density = -1.2, inPop = 0.15),
                   reps = 2, M = 8, N = 6, waves = 3,
                   n3 = 60, nsub = 1, seed = 7)
```
