---
title: "Seeing the Landscape for Any Influence Matrix"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Seeing the Landscape for Any Influence Matrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
```

## What this vignette shows, and what it does not

The README animation draws an actor's whole fitness landscape as a 16 x 16
grid: the components held in one module on the rows, the other module on the
columns. That picture is honest only because its influence matrix $W$ has two
blocks. When $W$ is a ring, a random sparse matrix, or a set of nested
modules, there is no natural pair of axes, and a grid built on an arbitrary
split hides the interdependence it is supposed to show.

This vignette presents three views that do not depend on $W$ having two
blocks, and compares them across four architectures:

| View | Question it answers | Scope |
|---|---|---|
| **Move compass** | Which of my $N$ single moves raise $f$, and by how much? | local; any $N$ |
| **W overlay** | Why is $f$ what it is? Which cells of $W$ are switched on? | the held portfolio only; any $N$ |
| **W-fitted grid** | Where are the peaks of the whole landscape? | global; small $N$ only |

Each view is computed from the package's own statistics, so the numbers in
the figures are the numbers the SAOM-NK engine uses when an actor chooses a
move. **The results below are for one simulated run per architecture, at
$N = 8$ components and $M = 6$ actors, with the model and seeds stated.**
They illustrate the views; they are not a study of how architecture shapes
search.

```{r load}
library(searchnet)
library(ggplot2)
```

## Four architectures at equal total weight

The four influence matrices are those of the README architectures figure, at
$N = 8$: two modules, nested modules, a local ring with $K = 2$, and a random
matrix with $K = 2$.

```{r archs}
N <- 8L
M <- 6L
offdiag <- function(W) { W <- abs(W); diag(W) <- 0; sum(W) }
archs <- list(
  modular = saomnk_block_diagonal(N, 2),
  nested  = (saomnk_block_diagonal(N, 2) + saomnk_block_diagonal(N, 4) + diag(N)) / 3,
  ring    = nk_to_saomnk(nk_landscape(N, 2, model = "adjacent", seed = 1))$influence_matrix,
  random  = nk_to_saomnk(nk_landscape(N, 2, model = "random", seed = 1))$influence_matrix
)
archs <- lapply(archs, function(W) { W <- unname(as.matrix(W)); storage.mode(W) <- "double"; W })
```

The matrices differ in how much off-diagonal weight they carry, not only in
where it sits. At one common XWX weight, an architecture with more total
$|w|$ would simply have stronger complementarity, and any difference in its
landscape would mix *where* $W$ puts interdependence with *how much* of it
there is. The comparison here is about the first, so each architecture's XWX
weight is scaled to give every $W$ the same total off-diagonal weight as the
two-module $W$ at weight 0.8:

$$\text{weight} = 0.8 \times \frac{24}{\sum_{h \ne j} |w_{hj}|}.$$

```{r weights}
ref_total <- 0.8 * offdiag(archs$modular)
weights <- vapply(archs, function(W) ref_total / offdiag(W), numeric(1))
data.frame(architecture = names(archs),
           offdiag_total = round(vapply(archs, offdiag, numeric(1)), 2),
           xwx_weight = round(weights, 2),
           weighted_total = round(weights * vapply(archs, offdiag, numeric(1)), 2),
           row.names = NULL)
```

This is also the default of the script that renders the animations,
`tools/make_architecture_gifs.R` in the source repository; its
`--raw-weight` option uses 0.8 for every $W$ instead, for checking how much
the scaling matters.

```{r wplot, fig.width = 7, fig.height = 2.3}
wdf <- do.call(rbind, lapply(names(archs), function(a) {
  W <- archs[[a]]
  data.frame(architecture = factor(a, levels = names(archs)),
             h = rep(LETTERS[1:N], N), j = rep(LETTERS[1:N], each = N),
             w = as.vector(W) * weights[[a]])
}))
wdf$w[wdf$h == wdf$j] <- NA
ggplot(wdf, aes(j, h, fill = w)) +
  geom_tile(color = "grey85", linewidth = 0.2) +
  scale_y_discrete(limits = rev(LETTERS[1:N])) +
  scale_fill_gradient(low = "white", high = "#0b6aa8", na.value = "grey95",
                      name = "weight x w") +
  facet_wrap(~ architecture, nrow = 1) +
  coord_fixed() +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 9) +
  theme(panel.grid = element_blank(), axis.text = element_text(size = 6))
```

## One run per architecture

Each architecture gets the same model (density $-2.5$, popularity $0.2$, XWX
at the weight above), the same starting network (seed 42) and the same
simulation seed (12345). The focal actor in each run is the one that moves
most often.

```{r runs}
run_arch <- function(W, wt) {
  env <- saomnk_env(M = M, N = N, density = 0.15, seed = 42)
  mod <- saomnk_model(density = -2.5, popularity = 0.2,
                      influence_matrix = W, influence_weight = wt)
  invisible(capture.output(suppressMessages(
    saomnk_run(env, mod, steps_per_actor = 10, seed = 12345))))
  env
}
envs <- Map(run_arch, archs, weights)
focal <- vapply(envs, function(env)
  which.max(tabulate(env$bi_env_changes[, "actor_i"], nbins = M)), integer(1))
n_steps <- vapply(envs, function(env) dim(env$bi_env_arr)[3], integer(1))
data.frame(architecture = names(envs), focal_actor = focal, ministeps = n_steps,
           row.names = NULL)
```

## The move compass: the actor's decision view

An actor in a SAOM-NK ministep does not see its landscape. It sees the $N$
portfolios one move away: add a component it lacks, or drop one it holds.
The compass draws exactly that. Each of the $N$ spokes is one move, labeled
`+C` for adding component C and `-C` for dropping it, and colored by the
change in $f$ the move would produce, red for a gain and blue for a loss.
When no spoke is red, the current portfolio is a local peak.

`searchnet_move_compass()` returns the data for one actor at one state:

```{r compass-data}
cp <- searchnet_move_compass(envs$modular, actor = focal[["modular"]], step = 0)
cp
attr(cp, "local_peak")
```

and a static compass is a few lines of `ggplot2`:

```{r compass-fn}
plot_compass <- function(cp, title = NULL) {
  N <- nrow(cp)
  ang <- pi / 2 - (seq_len(N) - 1) * 2 * pi / N
  cp$x <- cos(ang); cp$y <- sin(ang)
  cp$lab <- paste0(ifelse(cp$move == "add", "+", "-"), cp$label)
  lim <- max(abs(cp$delta))
  ggplot(cp) +
    geom_segment(aes(x = 0.42 * x, y = 0.42 * y, xend = 0.85 * x, yend = 0.85 * y,
                     color = delta, linewidth = abs(delta))) +
    annotate("point", 0, 0, shape = 21, size = 26, fill = "white", color = "#d9a21b") +
    geom_point(aes(x, y, fill = delta), shape = 21, size = 11, color = "grey20") +
    geom_text(aes(x, y, label = lab), data = cp[abs(cp$delta) <= 0.55 * lim, ],
              size = 3.2, fontface = "bold", color = "grey10") +
    geom_text(aes(x, y, label = lab), data = cp[abs(cp$delta) > 0.55 * lim, ],
              size = 3.2, fontface = "bold", color = "white") +
    geom_text(aes(1.32 * x, 1.32 * y, label = sprintf("%+.1f", delta)), size = 3) +
    annotate("text", 0, 0, label = sprintf("f = %.1f", cp$f_now[1]), fontface = "bold") +
    scale_fill_gradient2(low = "#2166ac", mid = "grey95", high = "#b2182b",
                         limits = c(-lim, lim), name = "change in f") +
    scale_color_gradient2(low = "#2166ac", mid = "grey85", high = "#b2182b",
                          limits = c(-lim, lim), guide = "none") +
    scale_linewidth(range = c(0.4, 3), guide = "none") +
    coord_fixed(xlim = c(-1.5, 1.5), ylim = c(-1.5, 1.5)) +
    labs(title = title,
         subtitle = if (attr(cp, "local_peak")) "local peak: no single move raises f"
                    else sprintf("%d of %d single moves raise f", sum(cp$delta > 1e-9), N)) +
    theme_void(base_size = 10)
}
```

```{r compass-plot, fig.width = 5, fig.height = 4.2}
plot_compass(cp, sprintf("Modular W: the moves open to actor %d at the start",
                         focal[["modular"]]))
```

The same actor at the end of its run:

```{r compass-end, fig.width = 5, fig.height = 4.2}
plot_compass(searchnet_move_compass(envs$modular, actor = focal[["modular"]]),
             sprintf("Modular W: actor %d at the end of the run", focal[["modular"]]))
```

The animated compass follows the focal actor through the whole run, ringing
the move it takes at each of its ministeps. One frame from each architecture:

```{r compass-stills, echo = FALSE, out.width = "49%", fig.show = "hold"}
knitr::include_graphics(file.path("figures", paste0("arch-", names(archs), "-compass.png")))
```

The full animations (about 2.5 MB each) are in the source repository:
[modular](https://github.com/sdownin/searchnet/blob/public-release/man/figures/readme-arch-modular-compass.gif),
[nested](https://github.com/sdownin/searchnet/blob/public-release/man/figures/readme-arch-nested-compass.gif),
[ring](https://github.com/sdownin/searchnet/blob/public-release/man/figures/readme-arch-ring-compass.gif),
[random](https://github.com/sdownin/searchnet/blob/public-release/man/figures/readme-arch-random-compass.gif).

**What the compass cannot show.** It is local by construction. It says
whether the current portfolio is a peak and which moves lead uphill, but not
how many peaks the landscape has, where they are, or how far the actor is
from the best portfolio. Two portfolios with identical compasses can sit on
very different landscapes. Its cost is $N + 1$ evaluations of $f$, so it
works at any $N$.

## The W overlay: why f changes

The compass reports *that* a move changes $f$; the overlay shows *why*. It
draws $W$ and lights the cells the actor's held portfolio switches on: cell
$(h, j)$ contributes $\theta_{XWX}\, w_{hj}$ to $f$ when the actor holds both
$h$ and $j$. Below the matrix, $f$ is split into its parts (density,
popularity, and the lit cells of $W$), which sum to the value on the compass.
Adding a component lights a new row and column of $W$; how many of those
cells carry weight is the complementarity gain of the move.

```{r overlay-stills, echo = FALSE, out.width = "49%", fig.show = "hold"}
knitr::include_graphics(file.path("figures", paste0("arch-", names(archs), "-overlay.png")))
```

The four frames are the end of each run. Under the two-module $W$ the actor
has filled one module, and every lit cell carries weight. Under the ring, the
held components form a contiguous arc, so the lit cells run along the band.
Under the random $W$, only the few held pairs that happen to be linked
contribute. The full animation for the modular $W$ is
[here](https://github.com/sdownin/searchnet/blob/public-release/man/figures/readme-arch-modular-overlay.gif).

**What the overlay cannot show.** Peaks. It describes the one portfolio held,
not the $2^N$ portfolios the actor could hold, so it cannot say whether a
better portfolio exists or how to reach it. It is the explanation for the
compass, not a replacement for it.

## The W-fitted grid: the global view for small N

For small $N$ every portfolio can be drawn. The W-fitted grid places all
$2^8 = 256$ portfolios on a 16 x 16 grid, with the eight components split
into a row group and a column group of four. The split is chosen from $W$:
of the 35 balanced splits, the one with the least $|w|$ across it. Within
each axis the components are ordered so that strongly linked ones are
neighbors, and the axis runs in Gray-code order, so adjacent rows (or
columns) differ by exactly one component. Local peaks, portfolios that no
single add or drop improves, are ringed.

```{r grid-stills, echo = FALSE, out.width = "49%", fig.show = "hold"}
knitr::include_graphics(file.path("figures", paste0("arch-", names(archs), "-grid.png")))
```

For the modular and nested $W$, the split falls on the module boundary and
no weight crosses it. For the ring and random $W$, even the best balanced
split leaves a quarter of $|W|$ across it; the grid is still a correct
picture of $f$, but its rows and columns no longer separate the
interdependence.

**What the grid cannot show.** It is lossless only while $2^N$ portfolios fit
on a screen: $N \le 10$ or so (1024 cells) before cells become too small to
read, and its cost doubles with each component. Its adjacency is partial: of
the $N$ single moves from a portfolio, at most four are neighboring cells
(one step along each axis in each direction), and the rest jump across the
grid. Two cells far apart can be one move apart, so the grid is a map of
heights, not of paths. The compass is the view of paths.

## Local peaks per architecture

`searchnet_actor_landscape()` evaluates $f$ at all $2^N$ portfolios for one
actor at one state, holding the other actors fixed, and marks the local
peaks. The counts for each focal actor, at the start and at the end of its
run:

```{r peaks}
peaks <- do.call(rbind, lapply(names(envs), function(a) {
  env <- envs[[a]]
  start <- searchnet_actor_landscape(env, focal[[a]], step = 0)
  end <- searchnet_actor_landscape(env, focal[[a]])
  best <- end[which.max(end$f), seq_len(N)]
  data.frame(architecture = a,
             peaks_start = sum(start$local_peak),
             peaks_end = sum(end$local_peak),
             best_f_end = round(max(end$f), 2),
             best_portfolio_end = paste(LETTERS[1:N][unlist(best) == 1], collapse = " "),
             at_peak_end = end$local_peak[end$current])
}))
peaks
```

At equal total weight, the ring has fewer peaks than the other three
architectures in this run. The counts describe one actor's landscape at one
model and one set of seeds; they are not a general claim about ring
architectures, which would need many runs and the planted-truth discipline of
`vignette("searchnet-recovery")`.

## When to use which

| | Move compass | W overlay | W-fitted grid |
|---|---|---|---|
| Shows | the $N$ moves and their $\Delta f$ | the cells of $W$ switched on; the parts of $f$ | $f$ at all $2^N$ portfolios; local peaks |
| Cannot show | peaks elsewhere; global shape | peaks; better portfolios | paths (adjacency is partial) |
| Cost | $N + 1$ evaluations | one evaluation | $2^N$ evaluations |
| Use for | any $N$; the actor's decision | any $N$; explaining a move | $N \lesssim 10$; the global picture |
| Function | `searchnet_move_compass()` | (animation only) | `searchnet_actor_landscape()` |

For $N$ up to about 10, show the grid for the global shape and the compass for
the decision. For larger $N$, the compass and the overlay remain exact, and
the global view has to come from summaries such as peak counts on sampled
landscapes rather than from a picture.

## Reproducing the animations

The stills above are frames from animations rendered by
`tools/make_architecture_gifs.R` in the source repository, which runs the
same models and seeds as this vignette, checks that every recomputed utility
equals the recorded one, and renders with manim (`inst/manim/scene_architecture_views.py`):

```{r gif-cmd, eval = FALSE}
## from the package root of a source checkout
## Rscript tools/make_architecture_gifs.R OUTDIR [--only=ring,random]
##   [--views=grid,compass,overlay] [--raw-weight] [--export-only]
```
