---
title: "{{TITLE}}"
output:
  html_document:
    toc: true
params:
  data: "{{DATA}}"
  seed: 20261010
  tiny: false
  run_recovery: true
---

<!-- Created by searchnet_new_analysis() with searchnet {{SEARCHNET_VERSION}}.
     Render with rmarkdown::render("analysis.Rmd"). For a quick smoke run:
     rmarkdown::render("analysis.Rmd", params = list(tiny = TRUE)). -->

```{r opts, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## 1. Setup

Versions are recorded so the results can be tied to the code that produced
them. Every random step below sets its own seed from `params$seed`, so a
chunk reruns identically on its own.

```{r setup}
library(searchnet)
source(file.path("R", "world.R"))

## Guards against a broken installation being mistaken for a result.
setup <- searchnet_check_setup(run_smoke = FALSE, verbose = FALSE)
setup$checks[, c("check", "status", "detail")]
stopifnot(setup$ok)

versions <- c(searchnet = as.character(packageVersion("searchnet")),
              RSiena    = as.character(packageVersion("RSiena")),
              R         = as.character(getRversion()))
versions
seed <- params$seed
rec <- NULL   # filled by the recovery section when it runs

## Sizes. `tiny` is for a smoke run only; no result from it should be read.
size <- if (isTRUE(params$tiny)) {
  list(M = 20, N = 12, W = 3, gate_reps = 10, placebo_draws = 30,
       cf_W = 6, rec_reps = 1, rec_M = 12, rec_N = 8, rec_n3 = 100,
       rec_nsub = 2)
} else {
  list(M = 30, N = 20, W = 4, gate_reps = 100, placebo_draws = 199,
       cf_W = 8, rec_reps = 50, rec_M = NULL, rec_N = NULL, rec_n3 = 500,
       rec_nsub = 4)
}
```

## 2. Data

Either a synthetic panel or a long-format CSV with one row per
actor-component-period tie. Both go through `searchnet_bipartite_from_long()`,
so the rest of the analysis sees one format.

```{r data}
## `actors` and `components` fix the universe, so a node with no ties in some
## period keeps a row of zeros. With your own data, list the full universe;
## NULL uses the ids that appear in the file.
if (identical(params$data, "long_csv")) {
  long <- utils::read.csv(file.path("data", "example_long.csv"),
                          stringsAsFactors = FALSE)
  actors <- NULL
  components <- NULL
} else {
  set.seed(seed + 1)
  long <- world_to_long(world_redraw_panel(size$M, size$N, size$W,
                                           p = 0.2, redraw = 0.15))
  actors <- sprintf("a%03d", seq_len(size$M))
  components <- sprintf("c%03d", seq_len(size$N))
}
head(long)

obs <- searchnet_bipartite_from_long(long, actor = "actor",
                                     component = "component", period = "period",
                                     actors = actors, components = components)
obs
B_obs <- obs$B
```

## 3. {K} readings

Per-node expansiveness (K_AC), popularity (K_CA), sociality (K_AA) and
epistasis (K_CC), per period. The function asserts the accounting identities,
so a malformed array stops here rather than downstream.

```{r k-readings}
k_obs <- searchnet_k_readings(obs)
attr(k_obs, "summary")
```

## 4. Simulated world and moment gate

### Pre-registered criteria

Stated before any simulation is run. Change them only in a new, dated commit,
never after looking at the gate.

```{r preregistered-criteria}
## Moment gate: largest acceptable gap of the simulated mean from the observed
## value. density and change_rate are calibrated (they check the simulator);
## the degree means and spreads are not, so they test the world.
gate_tolerance <- list(density = 0.01, change_rate = 0.02,
                       mean_K_AC = c(rel = 0.10), mean_K_CA = c(rel = 0.10),
                       sd_K_AC = c(rel = 0.25), sd_K_CA = c(rel = 0.25))
## Recovery: design checks on the estimator at this panel size.
recovery_criterion <- list(max_abs_bias = 0.10, min_coverage = 0.85,
                           min_converged_share = 0.80)
## Tests: two-sided level and the power at which a minimum detectable effect
## (MDE) is reported for every null.
alpha <- 0.05
power <- 0.80
```

### Simulated world

The simplest world with the observed density and rate of change (see
`R/world.R`). A failed gate means conclusions drawn in this world would be
drawn in the wrong world; it guards against tests run on an uncalibrated
simulation.

```{r world}
cal <- world_calibrate(B_obs)
unlist(cal)
set.seed(seed + 2)
world <- lapply(seq_len(size$gate_reps), function(r)
  world_redraw_panel(cal$M, cal$N, cal$W, cal$p, cal$redraw))
gate <- searchnet_moment_gate(B_obs, world, moments = names(gate_tolerance),
                              tolerance = gate_tolerance, on_fail = "warn",
                              theta_source = "redraw_world_calibrated")
gate
```

## 5. Recovery check

Can the estimator recover a planted effect at this panel size? Guards
against reading an estimate, or a null, from a design that cannot identify it.
Set `run_recovery: false` in the header to skip (it is the slowest section).

```{r recovery, eval = isTRUE(params$run_recovery)}
rec <- searchnet_recovery(
  c("density", "inPop"),
  theta_true = c(rate = 3, density = -1.2, inPop = 0.15),
  reps = size$rec_reps,
  M = if (is.null(size$rec_M)) cal$M else size$rec_M,
  N = if (is.null(size$rec_N)) cal$N else size$rec_N,
  waves = 3, nbrNodes = 1,    # nbrNodes and seed held fixed across comparisons
  seed = seed + 3, alpha = alpha, power = power,
  criterion = recovery_criterion, focal = "inPop",
  init_density = min(max(cal$p, 0.05), 0.5),
  n3 = size$rec_n3, nsub = size$rec_nsub)
rec
```

## 6. Counterfactual shock

In the calibrated world, the tie propensity of half the actors rises by
`delta` from period `t0` on. Two gates come before the estimate is read:
the support check (is the post-shock world structurally poorer, so that a
lower outcome follows arithmetically?) and the real-time random-date placebo
(does the estimator return an effect when exposure is random?).

```{r counterfactual}
delta <- 0.05
t0 <- floor(size$cf_W / 2) + 1
treated <- seq_len(cal$M) <= cal$M %/% 2
set.seed(seed + 4)
B_cf <- world_redraw_panel(cal$M, cal$N, size$cf_W, cal$p, cal$redraw,
                           shock = list(period = t0, actors = which(treated),
                                        p_post = min(1, cal$p + delta)))
dimnames(B_cf) <- list(sprintf("a%03d", seq_len(cal$M)),
                        sprintf("c%03d", seq_len(cal$N)),
                        as.character(seq_len(size$cf_W)))
k_cf <- searchnet_k_readings(B_cf)
panel <- world_actor_panel(k_cf, sprintf("a%03d", which(treated)), t0)

support <- searchnet_shock_support_check(panel, shock_step = t0, warn = FALSE)
support

pl <- searchnet_placebo(panel, estimator = "did_2x2",
                        n_draws = size$placebo_draws, seed = seed + 5,
                        alpha = alpha)
pl
```

## 7. Report: every null with its minimum detectable effect

A null is a bound on an association, not evidence of absence. Each test
below is listed with its MDE: the smallest effect the design detects with
probability `power` at level `alpha`, under a normal approximation.

```{r report}
z <- qnorm(1 - alpha / 2) + qnorm(power)
tests <- data.frame(
  test     = "counterfactual DID (placebo-calibrated)",
  estimate = pl$estimate,
  p_value  = pl$p_value,
  null     = !isTRUE(pl$p_value <= alpha),
  mde      = z * pl$placebo_sd,
  note     = paste("placebo gate:", pl$gate$verdict),
  stringsAsFactors = FALSE)
if (!is.null(rec)) {
  r <- rec$table[rec$table$effect == "inPop", ]
  tests <- rbind(tests, data.frame(
    test = "recovery: inPop (planted 0.15)", estimate = r$mean_est,
    p_value = NA_real_,
    null = if (r$n_converged > 0) r$rejection < power else NA,
    mde = r$mde,
    note = paste0("rejection rate ", signif(r$rejection, 3), ", ",
                  r$n_converged, " converged; verdict ", rec$verdict),
    stringsAsFactors = FALSE))
}
tests
cat("Moment gate:", gate$verdict, "
")
for (i in which(is.na(tests$null)))
  cat(sprintf("NOT ASSESSED: %s (%s).
", tests$test[i], tests$note[i]))
nulls <- tests[tests$null %in% TRUE, , drop = FALSE]
for (i in seq_len(nrow(nulls)))
  cat(sprintf("NULL: %s, estimate %.3g; effects smaller than %.3g cannot be excluded.
",
              nulls$test[i], nulls$estimate[i], nulls$mde[i]))
if (!nrow(nulls)) cat("No nulls among the assessed tests.
")
```

```{r session-info}
sessionInfo()
```
