---
documentclass: jss
classoption: article
title:
  formatted: "\\pkg{searchnet}: Network-Embedded Strategic Search Simulation in \\proglang{R}"
  plain: "searchnet: Network-Embedded Strategic Search Simulation in R"
  short: "\\pkg{searchnet}: Network-Embedded Search in \\proglang{R}"
author:
  - name: Stephen Downing
    affiliation: University of Missouri
    address: |
      | Robert J. Trulaske, Sr. College of Business
      | University of Missouri
      | Columbia, MO 65211, United States of America
    email: \email{sdowning@missouri.edu}
abstract: >
  The NK fitness landscape model is among the most widely used computational
  frameworks in strategy and organization theory, yet it treats the complexity
  parameter $K$ as exogenous, the landscape as fixed, and agents as isolated
  searchers.  The \proglang{R} package \pkg{searchnet} addresses all three
  limitations by reformulating the NK model as a bipartite network of $M$
  actors affiliating with $N$ components and embedding it within a Stochastic
  Actor-Oriented Model (SAOM).  This reformulation reveals that the scalar
  $K$ is a special case of a four-dimensional coupled degree system---actor
  scope ($K_{AC}$), component popularity ($K_{CA}$), actor sociality
  ($K_{AA}$), and component epistasis ($K_{CC}$)---whose dynamics are
  governed by actors' forward-looking affiliation decisions rather than
  imposed by the modeler.  The package provides an \pkg{R6}-based
  simulation engine built on \pkg{RSiena}, a clean functional API
  (\code{saomnk\_env}, \code{saomnk\_model}, \code{saomnk\_run}), coupled
  degree visualization through the $\{K\}$-4 panel, utility decomposition
  diagnostics, exogenous shock infrastructure, and validation against the
  classical NK model.  Worked illustrations demonstrate
  reproduction of classical NK search dynamics and quasi-experimental shock
  analysis.  Benchmarks against raw \pkg{RSiena} quantify
  computational overhead and scalability.  \pkg{searchnet} targets
  researchers in strategic management, organizational theory, innovation
  economics, and computational social science who study search, adaptation,
  and competitive dynamics on complex landscapes.
keywords:
  formatted: [NK fitness landscape, stochastic actor-oriented model, bipartite network, strategic search, coupled degree, "\\proglang{R}"]
  plain: [NK fitness landscape, stochastic actor-oriented model, bipartite network, strategic search, coupled degree, R]
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{booktabs}
  \usepackage{etoolbox}
  \robustify{\pkg}
  \robustify{\proglang}
  \robustify{\code}
  \pdfstringdefDisableCommands{\def\pkg#1{#1}\def\proglang#1{#1}\def\code#1{#1}\def\K{\{K\}}}
  \newcommand{\K}{\{K\}}
  \providecommand{\pandocbounded}[1]{#1}
  \newcounter{none}
output: rticles::jss_article
bibliography: searchnet.bib
biblio-style: jss
---

<!-- RECAST: author to write a software-paper abstract (what the package does, checks, illustrations; no novelty claim); drafts in paper/_drafts/2026-10-07_recast_abstract.md -->

```{r setup, include=FALSE, eval=TRUE, cache=FALSE}
## JSS conventions: R> prompt, "+" continuation, no output comment prefix,
## code width at most 70 characters.
##
## SEARCHNET_JSS_FAST=1 renders with every chunk unevaluated (format check
## only; paper/build_submission.R --fast sets it). A submission build must
## leave it unset so that all output is regenerated from the code.
.fast <- identical(Sys.getenv("SEARCHNET_JSS_FAST"), "1")
knitr::opts_chunk$set(
  prompt     = TRUE,
  comment    = NA,
  warning    = FALSE,
  message    = FALSE,
  fig.width  = 6.5,
  fig.height = 4.5,
  fig.align  = "center",
  eval       = !.fast,
  cache      = FALSE
)
options(prompt = "R> ", continue = "+  ", width = 70,
        useFancyQuotes = FALSE, digits = 4)

## Versions reported in "Computational details", read at render time rather
## than typed: the package version from the DESCRIPTION this paper ships
## with, and RSiena and R from the session that renders it.
.desc <- file.path("..", "DESCRIPTION")
sn_version <- if (file.exists(.desc)) {
  unname(read.dcf(.desc, fields = "Version")[1, 1])
} else {
  as.character(utils::packageVersion("searchnet"))
}
rsiena_version <- as.character(utils::packageVersion("RSiena"))
r_version <- paste(R.version$major, R.version$minor, sep = ".")

## Package-size counts reported in the text, read from the repository at
## render time. Paths listed in .public-exclude do not ship in the public
## package and are not counted.
.excl <- file.path("..", ".public-exclude")
.excl <- if (file.exists(.excl)) trimws(readLines(.excl, warn = FALSE)) else ""
.shipped <- function(dir, pattern) {
  f <- list.files(file.path("..", dir), pattern = pattern)
  sum(!file.path(dir, f) %in% .excl)
}
n_vignettes  <- .shipped("vignettes", "\\.Rmd$")
n_test_files <- .shipped("tests/testthat", "^test.*\\.R$")
## From the repository when rendered there; otherwise (e.g. code.R run from
## the replication bundle) from the installed package's copy.
.ptf <- file.path("..", "inst", "proofs", "PROOF_TABLE.md")
if (!file.exists(.ptf))
  .ptf <- system.file("proofs", "PROOF_TABLE.md", package = "searchnet")
.pt <- if (nzchar(.ptf) && file.exists(.ptf)) readLines(.ptf, warn = FALSE) else character()
## One row per step; the row ID is a part letter and a number (A1, L18).
n_proof_steps <- sum(grepl("^\\|\\s*[A-Z][0-9]+\\s*\\|", .pt))
```

```{r load-packages, include=FALSE}
library(searchnet)
library(ggplot2)
library(Matrix)
```

# Introduction

The NK fitness landscape model \citep{Kauffman1987, KauffmanWeinberger1989} has
become one of the most productive computational frameworks in strategic
management and organization theory.  Since \citet{Levinthal1997} demonstrated
that the NK model could illuminate the tension between exploration and
exploitation on rugged performance landscapes, hundreds of studies have deployed
it to investigate organizational adaptation, modular design, imitation, and
competitive search \citep{Baumann2019, Rivkin2000, SiggelkowRivkin2005}.

The model's appeal lies in its parsimony.  Two parameters---$N$, the number of
design choices or components, and $K$, the degree of epistatic interdependence
among them---generate landscapes ranging from smooth single-peaked surfaces
($K = 0$) to highly rugged terrains ($K = N - 1$) populated by many local
optima.  This parsimony has also become the model's principal liability.  Four
limitations recur across the literature.

First, the fitness landscape is **exogenous**: it is drawn once from a random
distribution and does not respond to the positions that agents occupy.  Firms
competing on the same landscape experience identical payoff surfaces regardless
of how many competitors share each component or how concentrated competitive
attention has become.  Second, **$K$ is a scalar** set by the modeler.  In
empirical organizational settings, interdependence is neither uniform nor fixed;
some components are tightly coupled while others operate nearly independently,
and coupling structures change as actors reconfigure their systems.  Third,
**multi-agent extensions** typically treat agents as independent searchers
navigating the same landscape, ignoring the social and competitive structure
through which firms observe, imitate, and crowd one another.  Fourth, the NK
model provides **no statistical framework** for parameter estimation from
empirical data, making it difficult to discipline simulations with observed
organizational behavior.

This paper introduces \pkg{searchnet}, an \proglang{R} package that addresses
all four limitations simultaneously.  The package reformulates the NK model as
a bipartite network of $M$ actors (firms, agents, decision-makers) affiliating
with $N$ components (technologies, resources, markets, design choices).  By
embedding this bipartite structure within a Stochastic Actor-Oriented Model
\citep[SAOM;][]{Snijders1996, SnijdersVdBuntSteglich2010}, the search
environment becomes a network data generation process whose topology---and
therefore whose fitness landscape---co-evolves with actors' affiliation
decisions (Figure~ef{fig:mentalmodel}).

<!-- RECAST: author to write the contribution paragraph(s) recast as software (design, correctness checks, reproducible workflows), replacing the deleted three-contributions block (Endogenization, Decomposition, Statistical infrastructure); drafts in paper/_drafts/2026-10-07_recast_introduction.md -->

<!-- Figure 1 is generated by paper/replication/make_figure1_mental_model.R (2026-10-08; the previous hand-drawn diagram was withdrawn the same day). -->

<!-- AI-DRAFTED CAPTION (2026-10-08, Framing A from paper/_drafts/2026-10-08_figure1_caption.md), inserted at the author's request for a working draft; author to re-voice. -->
![The \pkg{searchnet} workflow, one decision at a time. (a) The user specifies an influence matrix $W$ and effect weights. (b) The state is a bipartite network of actors (circles) and components (squares). (c) At a ministep the selected actor compares every single-tie change and staying put; each bar splits the change in its evaluation function by effect, and the logit probabilities follow. (d) The drawn move updates the network and the four $\K$ degrees. All panels come from one seeded run of 4 actors and 6 components; the probabilities in (c) equal those recorded by \pkg{RSiena}. Generated by \code{paper/replication/make\_figure1\_mental\_model.R}.\label{fig:mentalmodel}](figures/fig1_mental_model.png){width=100%}

The package provides an \pkg{R6}-based simulation engine, a functional
user API that abstracts away \pkg{RSiena}'s internal naming conventions,
comprehensive visualization through the $\K$-4 panel and 3D phase space,
utility decomposition diagnostics, exogenous shock infrastructure for
quasi-experimental designs, multi-wave Monte Carlo facilities, specification
robustness tools, an interactive strategy game mode, and a classroom teaching
module.  Thirteen vignettes accompany the package: a quick-start introduction,
theoretical foundations, a full simulation walkthrough with strategies and
shocks, a computational experiment design guide, a causal inference tutorial,
an NK validation guide, a Blume--Gibbs convergence tutorial, basin geometry,
the Brock--Durlauf economic foundation, empirical calibration, the proof
chain, formal verification with Lean 4, and a
Professional Development Workshop curriculum.

<!-- RECAST: author to write the roadmap paragraph for the new section order (2 Models, 3 Software design, 4 Using searchnet, 5 Illustrations, 6 Verification and validation, 7 Performance, 8 Comparison, 9 Summary and limitations); drafts in paper/_drafts/2026-10-07_recast_introduction.md -->


# Models {#sec:background}

## The NK fitness landscape

The NK model \citep{Kauffman1987, KauffmanWeinberger1989} characterizes
performance landscapes with two parameters.  A system of $N$ binary components
(loci, design choices, technologies) generates $2^N$ possible configurations.
Each component $d$ makes a fitness contribution $f_d$ that depends not only on
its own state but also on the states of $K$ other components with which it is
epistatically linked.  Total fitness is the average contribution across all $N$
components.  When $K = 0$, each component contributes independently and the
landscape has a single global peak.  As $K$ increases toward $N - 1$, the
landscape becomes increasingly rugged, populated by exponentially many local
optima that trap myopic searchers.

\citet{Levinthal1997} introduced the NK model to management as a formalization
of the exploration--exploitation trade-off: agents that search locally
(single-bit mutations) on rugged landscapes quickly find nearby peaks but
rarely discover the global optimum, while agents that search more broadly
(multi-bit jumps) sacrifice short-run performance for a chance at higher peaks.
This insight has been extended to organizational design
\citep{SiggelkowRivkin2005}, imitation \citep{Rivkin2000}, modular
architectures \citep{EthirajLevinthal2004}, and competitive interaction
\citep{Lenox2006}.

## Stochastic Actor-Oriented Models

Stochastic Actor-Oriented Models \citep{Snijders1996,
SnijdersVdBuntSteglich2010} provide a continuous-time Markov model for the
co-evolution of social networks and behavior.  At each micro-step, a single
actor is selected according to a rate function, evaluates the marginal utility
of each possible tie change according to an objective function, and chooses
probabilistically via a multinomial logit rule.  The objective function takes the
general form

\begin{equation}\label{eq:saom-objective}
  f_i(\mathbf{B}) = \sum_k \theta_k \, s_{ik}(\mathbf{B}),
\end{equation}

\noindent where $s_{ik}(\mathbf{B})$ are network statistics computed from the
current network $\mathbf{B}$ and $\theta_k$ are parameters governing each
actor's preferences.  The parameters $\theta_k$ can be estimated from observed
panel data using the method of moments or maximum likelihood, implemented in the
\pkg{RSiena} package \citep{RSiena}.

SAOMs are well established in sociology and political science for modeling
friendship networks, advice networks, and inter-organizational partnerships.
They have been extended to bipartite (two-mode) networks
\citep{Koskinen2012}, where the model governs actors' affiliation decisions
with respect to events, organizations, or other categorical entities.
\pkg{searchnet} exploits this bipartite extension to model actors affiliating
with components in a search environment.

### Why SAOM over ERGM? {#sec:saom-ergm}

A natural question is why \pkg{searchnet} adopts the SAOM rather than the
exponential random graph model \citep[ERGM;][]{LusherKoskinenRobins2012} or its
temporal extension \citep[TERGM;][]{Hanneke2010} as its network engine.  The choice follows directly from the theoretical premise of the NK tradition.
ERGMs characterize the equilibrium distribution of network
configurations without specifying a generative decision process.  An ERGM
answers the question "what network structures are over-represented relative to
chance?" but does not model *who* makes *which* decision *when*.  For a
framework whose purpose is to endogenize landscape ruggedness through actors'
strategic choices, this equilibrium focus omits precisely the mechanism of
interest.

## The implemented model {#sec:framework}

\noindent \textbf{Notation.}
Throughout this paper we adopt the following notation.  The bipartite incidence
matrix is $\mathbf{B} \in \{0,1\}^{M \times N}$ with entries $b_{ij}$; the
NK configuration vector for a single agent is $\mathbf{x} \in \{0,1\}^N$; the
influence matrix is $\mathbf{E}$ (implemented in \pkg{RSiena} as the weight
matrix $\mathbf{W}$ of the \code{XWX} effect); component payoff functions are
$f_d$; and the logit precision parameter is $\beta > 0$.

A terminological note is warranted here, because the two ideas are routinely conflated.  The matrix $\mathbf{E}$ is an *influence matrix* in the sense of \citet{RivkinSiggelkow2007}: it states which components interact and with what sign, and it is an *input* supplied by the analyst.  *Epistasis* is the *consequence* -- the dependence of one component's fitness contribution on the states of others, which arises when a portfolio is evaluated through $\mathbf{E}$ as $\mathbf{x}^{\top}\mathbf{E}\mathbf{x}$.  The distinction matters in \pkg{searchnet} because a third quantity sits between the two.  The component-component degree $K_{CC}$ records how far components are actually co-adopted at a given moment; it is computed from the realized affiliation matrix and does not read $\mathbf{E}$, yet it evolves as a *consequence* of $\mathbf{E}$, which drives the search that produces those affiliations.  $K_{CC}$ therefore indexes the coupling epistasis acts through, generally differs from the influence structure that generated it, and is not itself a measure of the fitness consequence.  Accordingly the package argument is named \code{influence\_matrix}; \code{epistasis\_matrix} is retained as a deprecated alias.

### Bipartite network formulation {#sec:bipartite}

The classical NK model represents a system configuration as an $N$-dimensional
binary vector $\mathbf{x} = (x_1, x_2, \ldots, x_N)$ where $x_j \in \{0, 1\}$.
\pkg{searchnet} generalizes this by introducing $M$ actors, each of whom
maintains an affiliation vector.  The resulting system state is an
$M \times N$ binary matrix $\mathbf{B}$, where $b_{ij} = 1$ indicates that
actor $i$ affiliates with component $j$.  This matrix is the incidence matrix
of a bipartite network with actor nodes on one side and component nodes on the
other.

The bipartite representation has three immediate consequences.  First, it
naturally accommodates **multiple agents** searching simultaneously: each row
of $\mathbf{B}$ is an actor's configuration, and actors' configurations
interact through shared columns.  Second, it makes the **search environment
endogenous**: the pattern of affiliations determines which components are
coupled (via shared actors) and which actors are competitors (via shared
components).  Third, it provides the structural substrate for the SAOM engine:
$\mathbf{B}$ becomes the dependent variable in an evolving bipartite network
model.

\subsubsection[The {K} interpretive framework]{The $\K$ interpretive framework}\label{sec:k4}

The bipartite incidence matrix $\mathbf{B}$ generates four coupled degree
statistics, each of which is individually a standard network measure but whose
joint dynamics expose the mechanism through which micro-level search decisions
produce macro-level landscape structure.  Table~\ref{tab:k4} summarizes the
four dimensions.

\begin{table}[t]
\centering
\small
\begin{tabular}{llp{3.6cm}p{3.6cm}}
\toprule
Dimension & Symbol & Interpretation & Network measure \\
\midrule
Actor scope & $K_{AC}$ & Components per actor & Bipartite row degree \\
Component popularity & $K_{CA}$ & Actors per component & Bipartite column degree \\
Actor sociality & $K_{AA}$ & Co-affiliating actors per actor & Actor projection degree \\
Component epistasis & $K_{CC}$ & Co-occurring components per component & Component projection degree \\
\bottomrule
\end{tabular}
\caption{The $\K$ coupled degree framework.  Each dimension is a standard
network statistic.}
\label{tab:k4}
\end{table}

The classical NK parameter $K$ maps directly onto $K_{CC}$---the degree of a
node in the component--component projection network.  Two components share a
projection edge whenever at least one actor affiliates with both, and the
number of such edges incident on a focal component determines how many other
components epistatically affect its fitness contribution.  In the classical
model, this coupling structure is set exogenously by the modeler; in
\pkg{searchnet}, it emerges endogenously from the bipartite network and
co-evolves with the other three degree dimensions.

The analytical leverage of the $\K$ framework comes from treating these four
measures as a coupled system rather than examining any one in isolation.
Because all four derive from the same incidence matrix $\mathbf{B}$, they are
structurally linked.  A change in actor scope ($K_{AC}$) mechanically alters
component popularity ($K_{CA}$), which in turn reshapes both the actor
co-membership network ($K_{AA}$) and the component co-occurrence network
($K_{CC}$).  The projections are computed as:

\begin{align}
\mathbf{P}_{AA} &= \mathbf{B} \mathbf{B}^\top - \text{diag}(\mathbf{B} \mathbf{B}^\top), \label{eq:proj-actor} \\
\mathbf{P}_{CC} &= \mathbf{B}^\top \mathbf{B} - \text{diag}(\mathbf{B}^\top \mathbf{B}). \label{eq:proj-component}
\end{align}

The degree of actor $i$ in the actor projection $\mathbf{P}_{AA}$ is $K_{AA,i}
= \sum_{h \neq i} \mathbf{1}[(\mathbf{B}\mathbf{B}^\top)_{ih} > 0]$, and
symmetrically for $K_{CC,j}$ in the component projection.  Analyzing any single
measure in isolation obscures the cross-domain dependencies that link
actor behavior to landscape topology.

The
four statistics are individually well known in the bipartite network literature
\citep{Latapy2008, BorgattiEverett1997}.

### SAOM objective function specification {#sec:objective}

The SAOM objective function in \pkg{searchnet} governs actors' affiliation
decisions.  At each micro-step, the selected actor $i$ evaluates the change in
objective function value for each possible tie change (creating or dissolving an
affiliation with one component) and selects the change that maximizes expected
utility, subject to a stochastic error term drawn from the type I extreme value
distribution.

The objective function takes the form of Equation~\ref{eq:saom-objective}, with
the following key statistics:

\begin{table}[t]
\centering
\begin{tabular}{llll}
\toprule
Statistic & RSiena name & Formula & Interpretation \\
\midrule
Density & \code{density} & $\sum_j b_{ij}$ & Baseline affiliation rate \\
Popularity & \code{inPop} & $\sum_j b_{ij} \sqrt{\sum_{h \neq i} b_{hj}}$ & Preference for popular components \\
Scope & \code{outAct} & $\sum_j b_{ij} \sqrt{\sum_{k \neq j} b_{ik}}$ & Scope expansion tendency \\
Epistasis & \code{XWX} & $\sum_j b_{ij} \sum_k w_{jk} b_{ik}$ & Exogenous component coupling \\
\bottomrule
\end{tabular}
\caption{SAOM objective function statistics in \pkg{searchnet}.}
\label{tab:effects}
\end{table}

The density effect serves as an intercept: a negative density parameter creates
baseline sparsity pressure, preventing actors from affiliating with all
components.  The popularity effect (\code{inPop}) creates preferential
attachment to components that many other actors already use.  The scope effect
(\code{outAct}) governs actors' tendency to expand or contract the breadth of
their affiliations.  The epistasis effect (\code{XWX}) introduces exogenous
component--component coupling via a modeler-supplied weight matrix $\mathbf{W}$
(denoted $\mathbf{E}$ in the formal proofs; see Section~\ref{sec:formal-properties});
actors derive utility from affiliating with components that are coupled to their
existing affiliations.

Actor-level heterogeneity enters through constant covariates attached to the
objective function.  A covariate vector $\mathbf{v}$ (e.g., encoding strategic
orientation) can modify the weight an actor places on popularity seeking
(\code{inPopX}), scope expansion (\code{egoX}), or component characteristics
(\code{altX}).  This mechanism allows the modeler to assign distinct strategic
profiles to different actors---for example, "exploiters" who preferentially
attach to popular components and "explorers" who expand into uncrowded niches.

The simulation proceeds via \pkg{RSiena}'s continuous-time Markov chain
algorithm.  Actors are selected according to a rate function (by default,
equal rates for all actors), and the selected actor evaluates all possible
single-tie changes, choosing according to a multinomial logit rule:

\begin{equation}\label{eq:choice}
  P(b_{ij} \to 1 - b_{ij}) =
  \frac{\exp\bigl(f_i(\mathbf{B}^{ij})\bigr)}
       {\sum_{j'} \exp\bigl(f_i(\mathbf{B}^{ij'})\bigr)},
\end{equation}

\noindent where $\mathbf{B}^{ij}$ denotes the network configuration obtained
by toggling actor $i$'s tie to component $j$.  The \code{search\_rsiena}
method passes the full structure model to \pkg{RSiena}'s
\code{siena07} function, which executes the continuous-time simulation
internally.

For exogenous shocks, \pkg{searchnet} partitions the simulation timeline into
proportional segments.  At each segment boundary, the $\theta$ parameters are
updated to new values specified by the shock schedule.  The simulation then
continues from the current network state with the modified objective function,
preserving path dependence while allowing the researcher to model environmental
discontinuities.  This is equivalent to running sequential \pkg{RSiena}
simulations where each wave's initial state is the terminal state of the
previous wave.

# Software design {#sec:package}

## Architecture {#sec:architecture}

\pkg{searchnet} is built on \pkg{R6} classes organized in a two-level
inheritance hierarchy:

\begin{description}
\item[\code{SaomNkRSienaBiEnv\_base}] Core environment initialization,
  bipartite matrix management, graph construction, projection computation, and
  utility helpers (toggle, Jaccard, existence checks).  This base class
  manages the $M \times N$ incidence matrix, constructs the \pkg{RSiena} data
  objects, and provides low-level network operations.

\item[\code{SaomNkRSienaBiEnv}] Full simulation engine inheriting from the
  base class, adding the \code{search\_rsiena()} simulation method, chain
  statistics processing, $\K$-4 degree computation, fitness landscape
  evaluation, shock handling, multi-wave experiment infrastructure, and the
  visualization interface.
\end{description}

Supporting modules provide separation of concerns:

\begin{itemize}
\item \code{saomnk-api.R}: User-facing functional API (\code{saomnk\_env},
  \code{saomnk\_model}, \code{saomnk\_run}) that wraps the R6 internals.
\item \code{nk-classic.R}: A self-contained implementation of the classical
  NK model, used as an external reference for the reduction check of
  Section~\ref{sec:validation}.
\item \code{saomnk-experiments.R}: \code{SaoMNKexperiments} R6 class for batch
  Monte Carlo simulation management.
\item \code{plot-*.R}: Ten modular plotting files (\code{plot-utility.R},
  \code{plot-degrees.R}, \code{plot-exploration.R}, \code{plot-markets.R},
  \code{plot-multiwave.R}, \code{plot-shocks.R}, \code{plot-snapshots.R},
  \code{plot-phase-space.R}, \code{plot-forbearance.R}, \code{plot-geo.R})
  containing visualization functions.
\item \code{searchnet-game.R}: Strategy game mode providing an interactive
  game loop where a human player controls one firm and competes against AI
  opponents that follow SAOM-NK logit choice rules.
\item \code{searchnet-teaching.R}: Classroom teaching module for Capsim-style
  strategy simulations supporting multiple industry presets, difficulty
  levels, and exogenous shocks.
\item \code{utils.R}: Standalone helper functions.
\end{itemize}

The package follows standard \proglang{R} packaging conventions with
\code{DESCRIPTION}, \code{NAMESPACE}, a \pkg{testthat} test suite (820 tests
across `r n_test_files` files), and `r n_vignettes` vignettes.  All exported functions are prefixed
with \code{saomnk\_} or \code{searchnet\_} to avoid namespace collisions.
<!-- RECAST: the prefix claim is not true of the current NAMESPACE: 45 of 184 explicit exports carry neither prefix (e.g. the three R6 classes, nk_*, bd_*, compute_forbearance_*, verify_brock_durlauf_reduction, solve_mean_field); author to restate (count checked 2026-10-08) -->

## Simulation as state-carrying periods {#sec:periods}

<!-- AI-DRAFTED (2026-10-08, Section 3.2 Framing A from paper/_drafts/2026-10-07_recast_software-design.md), inserted at the author's request for review in context; author to re-voice. -->
A call to \code{search\_rsiena()} simulates one unit of model time, not a fixed number of
ministeps. Each block of identical parameter rows is run as one unconditional
\pkg{RSiena} period, started from the end network of the previous block, so the path
carries its state across shocks and waves. The argument \code{iterations\_per\_actor}
(\code{steps\_per\_actor} in \code{saomnk\_run()}) sets the basic rate, that is, the
expected number of opportunities per actor at zero rate effects; the realized number of
ministeps is random. The engine stops if replaying a segment's chain does not end at
\pkg{RSiena}'s own end network.

## Guards {#sec:guards}

<!-- AI-DRAFTED (2026-10-08, Section 3.3 Framing A from paper/_drafts/2026-10-07_recast_software-design.md), inserted at the author's request for review in context; author to re-voice. -->
Four guards target failures that leave no trace in the output. Every effect declared in a
model must appear as included in the fitted \pkg{RSiena} effects object, so a misspelled
or unmatched effect stops the run instead of disappearing. Covariate centering is stated:
monadic covariates are uncentered unless the model says otherwise, and the centering
applied is recorded. Each run records its provenance, including segment seeds and the
realized ministep count. Seeds are drawn from hashed, namespaced streams, so neighboring
seeds share no segment seed.

## Ecosystem {#sec:ecosystem}

### Manim animation pipeline {#sec:manim}

For presentation and pedagogical purposes, \pkg{searchnet} exports simulation
data in formats compatible with the \pkg{Manim} animation library
\citep{Manim2022}.  The exported data includes per-step bipartite matrices,
projection networks, degree trajectories, and utility decompositions.  This pipeline is accessed through the
\code{saomnk\_export\_manim()} method on the environment object and the
\code{searchnet\_export\_phase\_space()} function for phase space data.


### Interactive features {#sec:interactive}

\pkg{searchnet} includes two interactive modules that extend the simulation
engine from a research tool to an instructional platform.
The \code{searchnet\_game\_*} family of functions (five exported functions)
provides an interactive game loop where a human player controls one firm and
competes against AI opponents that follow the SAOM conditional-logit choice
rule.
The \code{searchnet\_classroom\_*} family (six exported functions) supports
Capsim-style strategy simulations for classroom use.

\section[Using searchnet]{Using \pkg{searchnet}}\label{sec:using}

## User API {#sec:api}

A central design goal of \pkg{searchnet} is to provide a clean functional
interface that shields users from the complexity of \pkg{RSiena}'s internal
naming conventions and the R6 class hierarchy.  The workflow follows three
steps: environment creation, model specification, and simulation execution.

### Step 1: Create the search environment

The \code{saomnk\_env()} function initializes a bipartite search environment:

```{r api-env}
env <- saomnk_env(M = 8, N = 12, density = 0, seed = 42)
```

This creates a \code{SaomNkRSienaBiEnv} object representing 8 actors and 12
components, starting from an empty network (all actors unaffiliated).  The
\code{seed} argument ensures reproducibility of the initial network draw.

### Step 2: Specify the structure model

The \code{saomnk\_model()} function builds the SAOM objective function
specification using intuitive parameter names:

```{r api-model}
## Define a modular influence matrix: 4 blocks of 3 components
K_matrix <- saomnk_block_diagonal(12, 4)

## Specify the model
mod <- saomnk_model(
  density    = -0.5,     # sparsity pressure
  popularity =  0.2,     # preferential attachment
  scope      =  0.1,     # scope expansion
  influence_matrix = K_matrix,
  influence_weight = 0.05
)
mod
```

Users never need to write internal references such as
\code{"self\$bipartite\_rsienaDV"} or construct nested covariate lists
manually.  The function assembles the complete \pkg{RSiena} structure model
internally.  Actor strategy heterogeneity is specified through the
\code{strategies} argument:

```{r api-model-strat}
## Heterogeneous actors: varying scope and popularity preferences
mod_strat <- saomnk_model(
  density    = -0.3,
  popularity =  0.2,
  scope      =  0.1,
  influence_matrix = K_matrix,
  influence_weight = 0.05,
  strategies = list(
    egoX   = rep(c(-1, 0, 1, 0), length.out = 8),
    inPopX = rep(c( 1, 0,-1, 0), length.out = 8)
  )
)
```

### Step 3: Run the simulation

The \code{saomnk\_run()} function executes the SAOM simulation:

```{r api-run, cache=TRUE}
saomnk_run(env, mod, steps_per_actor = 30, seed = 12345)
```

The function passes the structure model to the internal
\code{search\_rsiena()} method, runs the continuous-time SAOM simulation
via \pkg{RSiena}, extracts chain statistics, and computes the $\K$-4 degree
trajectories.  The \code{env} object is modified in place (standard R6
semantics) and returned invisibly.

## Mapping RSiena effects to strategic search concepts {#sec:glossary}

The preceding objective function uses \pkg{RSiena} effect names that are
standard in social network analysis but may be unfamiliar to strategy
researchers.  Table~\ref{tab:glossary} provides a systematic mapping between
RSiena effects available in \pkg{searchnet} and their strategic search
interpretations.  Each effect's parameter symbol links it to the corresponding
$\K$ dimension introduced in Section~\ref{sec:k4}, enabling researchers to
translate between the network estimation vocabulary and the strategy-theoretic
language of scope, complementarity, and imitation.

\begin{table}[t]
\centering
\small
\begin{tabular}{llllp{3.6cm}}
\toprule
RSiena Effect & Parameter & $\K$ Dim. & Strategy Concept & Interpretation \\
\midrule
\code{density}      & $\theta_{\text{dens}}$  & $K_{AC}$        & Scope cost               & Negative = costly to hold activities; controls portfolio breadth \\
\code{outAct}       & $\theta_{\text{scope}}$ & $K_{AC}$   & Diminishing returns       & Quadratic penalty on actor degree (scope) \\
\code{inPop}        & $\theta_{\text{pop}}$   & $K_{CA}$   & Bandwagon / legitimacy    & Positive = prefer popular components; preferential attachment \\
\code{XWX}          & $\theta_{\text{epist}}$ & $K_{CC}$   & Complementarity / synergy & Reward for holding epistatically linked components \\
\code{simEgoInDist2}& $\theta_{\text{imit}}$  & $K_{CA}$   & Imitation / herding       & Positive = prefer components held by covariate-similar actors (needs an actor covariate) \\
\code{egoX}         & $\theta_{\text{strat}}$ & $K_{AC}$        & Strategy heterogeneity    & Actor-level covariate modifying search behavior \\
\code{altX}         & $\theta_{\text{value}}$ & $K_{CA}$        & Component value           & Exogenous activity-level payoff differences \\
\code{cycle4}       & $\theta_{\text{close}}$ & $K_{AA}$   & Structural closure        & Triadic closure in the projected social network \\
\bottomrule
\end{tabular}
\caption{Mapping RSiena effects to strategic search concepts.  Parameters correspond to the SAOM objective function (Equation~\ref{eq:saom-objective}).  The $\K$ column indicates which coupled degree dimension each effect primarily governs.}
\label{tab:glossary}
\end{table}

The table makes explicit the division of labor among effects.  The first four
rows---\code{density}, \code{outAct}, \code{inPop}, and \code{XWX}---are the
workhorses of most \pkg{searchnet} specifications.  Together they endogenize the
four $\K$ dimensions: scope cost sets the baseline, \code{outAct} introduces
diminishing returns to breadth, \code{inPop} creates demand-side herding toward
popular components, and \code{XWX} rewards configurations that exploit
exogenous complementarities.  The remaining effects allow the researcher to
layer on additional mechanisms: imitation (\code{simEgoInDist2}, which requires
an actor covariate) draws actors toward components held by actors similar on
that covariate, actor and component
covariates (\code{egoX}, \code{altX}) introduce heterogeneity, and
\code{cycle4} captures closure in the actor--actor projection of the bipartite
network.  By selecting which effects to include and setting their parameter
values, the researcher controls which strategic mechanisms are active in a given
simulation---a design choice whose consequences can be audited with the
diagnostic tools described in Section~\ref{sec:validation}.


## Visualization {#sec:visualization}

\pkg{searchnet} provides a comprehensive visualization toolkit organized
around four analytical perspectives.

\subsubsection[The {K}-4 panel]{The $\K$-4 panel}

The signature visualization is the coupled degree panel, which displays
all four $\K$ dimensions over the simulation timeline:

```{r k4-panel, fig.height=8, fig.width=8, fig.cap=paste0("The $\\{K\\}$-4 coupled degree panel showing actor scope ($K_{AC}$), component popularity ($K_{CA}$), actor sociality ($K_{AA}$), and component coupling ($K_{CC}$) trajectories over ", if (exists("env")) env$get_step() else "the realized number of", " simulation micro-steps.  Each point is a single actor or component at one ministep; the heavy line is the LOESS-smoothed mean.")}
saomnk_plot_k4(env, smooth = 0.3)
```


### Network snapshots

The \code{saomnk\_plot\_snapshots()} function renders the bipartite
network at selected simulation steps, along with its unipartite projections:

```{r snapshots, fig.height=4, fig.width=10, fig.cap="Network snapshots at the initial state, midpoint, and final state of the simulation.  In each snapshot, from left: the actor co-membership projection ($K_{AA}$), the bipartite actor-component network, and the component co-occurrence heatmap ($K_{CC}$)."}
n_steps <- env$get_step()
saomnk_plot_snapshots(env,
  steps = unique(pmax(1, round(c(0, 0.5, 1) * n_steps))))
```

### Utility decomposition

The \code{saomnk\_plot\_utility()} function decomposes each actor's total
objective function value into contributions from individual network statistics:

```{r utility, fig.height=5, fig.width=8, fig.cap="Utility decomposition showing each actor's objective function value broken into contributions from density, popularity, scope, and influence ($W$) effects over the simulation timeline."}
saomnk_plot_utility(env, smooth = 0.35)
```

This decomposition provides direct insight into the strategic trade-offs
governing search behavior: an actor whose utility is dominated by the epistasis
term is searching for complementary components, while one dominated by the
popularity term is following the crowd.

### 3D phase space

The \code{saomnk\_plot\_phase\_space\_3d()} function renders the simulation
trajectory in a three-dimensional Network $\times$ Behavior $\times$ Fitness
space, where the three axes correspond to structural properties (e.g.,
$K_{AC}$ or density), behavioral properties (e.g., exploration rate or scope
change), and outcome measures (utility or NK fitness).  The resulting
``SCP(t) trajectory'' traces each actor's path through this space over
simulation time, providing an integrated view of how structural position,
behavioral strategy, and performance co-evolve.  Supporting functions include
\code{saomnk\_plot\_phase\_heatmap()} for density-based projections,
\code{saomnk\_plot\_phase\_evolution()} for animated trajectories,
\code{saomnk\_plot\_phase\_comparison()} for juxtaposing multiple simulation
runs, and \code{searchnet\_export\_phase\_space()} for Manim-compatible data
export.

## Empirical calibration via SAOM parameter bridge {#sec:calibration}

The \pkg{searchnet} package includes a calibration bridge that converts
fitted SAOM parameters from empirical network panel data into calibrated
SaoMNK simulation environments.  This enables researchers to run validated
counterfactual simulations using empirically grounded parameter estimates
rather than ad hoc parameter choices.

### Parameter mapping

Both frameworks use the actor-oriented conditional logit utility
\begin{equation}\label{eq:calibration-utility}
  U_i(x) = \sum_k \theta_k \, s_{ik}(x),
\end{equation}
where $s_{ik}(x)$ is the network statistic for effect $k$.  The bridge
maps \pkg{RSiena}'s $\hat{\theta}$ vector to \pkg{searchnet}'s structure
model effects.  For shared effects the mapping is direct:

\begin{table}[ht]
\centering
\begin{tabular}{llc}
\toprule
SAOM Effect & SaoMNK Effect & Transformation \\
\midrule
\code{density}   & \code{density}    & $\beta = \theta \cdot \alpha$ \\
\code{inPopSqrt} & \code{inPop}      & $\beta = \theta \cdot \alpha$ \\
\code{gwespFF}   & \code{transTriads} & $\beta = \theta \cdot \alpha$ \\
\code{simX}      & \code{egoX}       & $\beta = \theta \cdot \alpha$ \\
\code{cycle3}    & \code{cycle4}     & $\beta = \theta \cdot \alpha$ (bipartite analog) \\
\bottomrule
\end{tabular}
\caption{Parameter mapping between RSiena and \pkg{searchnet}. $\alpha$ is
an optional scale factor applied during calibration.}
\label{tab:param-mapping}
\end{table}

### Code example

The calibration workflow involves three steps: parameter conversion,
environment initialization from empirical data, and counterfactual
execution.

```{r calibration-example, eval=FALSE}
## Convert fitted SAOM parameters to SaoMNK
bridge <- saom_to_saomnk(saom_fit, scale_factor = 1.0)
print(bridge$mapping_table)

## Initialize environment from empirical data
env_data <- empirical_to_saomnk_env(mi_data, wave = 5)

## Run calibrated counterfactual
result <- run_calibrated_counterfactual(
  env_data, bridge,
  scenario = list(name = "double_closure",
                  modify = list(transTriads = 2.0)),
  iterations = 50
)
cat(sprintf("K_AC change: %.3f\n", result$comparison$delta_K_AC))
```

### Predefined counterfactual scenarios

The package includes six predefined scenarios:

\begin{enumerate}
\item \textbf{Density shock}: Exogenous reduction in network density
  (e.g., deregulation increasing market access).
\item \textbf{Double closure}: Doubling the transitive triad parameter to
  simulate increased alliance clustering.
\item \textbf{Popularity concentration}: Amplifying preferential attachment
  to model winner-take-all dynamics.
\item \textbf{Scope expansion}: Increasing actor scope to simulate
  diversification strategies.
\item \textbf{Epistasis rewiring}: Modifying the component influence
  matrix to simulate modular-to-integral design shifts.
\item \textbf{Attribute homophily}: Strengthening similarity-based
  affiliation to model industry convergence.
\end{enumerate}

\noindent Each scenario modifies the calibrated parameter vector and
re-runs the simulation, yielding a comparison structure with changes in
all four $\K$ dimensions, basin geometry, and fitness distributions.  For
a full worked example see \code{vignette("saomnk-calibration", package =
"searchnet")}.


# Illustrations {#sec:illustrations}

## Validation against the classical NK model {#sec:validation}

A software paper should demonstrate that its engine reproduces the model it
generalizes.  \pkg{searchnet} ships a self-contained implementation of the
classical Kauffman NK model, independent of \pkg{RSiena} and of the SaoMNK
engine, and uses it as an external reference.

### A conventional NK implementation

\code{nk\_landscape()} constructs a Kauffman landscape by exhaustive enumeration of
all $2^N$ configurations.  Each locus $i$ contributes $f_i(b_i,
b_{\Omega(i)})$, drawn i.i.d.\ $U(0,1)$ for every distinct pattern of its
epistatic neighborhood $\Omega(i)$; total fitness is the mean of the $N$
contributions.  Three neighborhood topologies are supported: \code{"adjacent"}
(Kauffman's ring model), \code{"random"}, and \code{"block"} (near-decomposable).  Companion functions provide adaptive walks (\code{nk\_walk()}, with
steepest-ascent, random-ascent, and random-improving rules), exhaustive local
optimum enumeration (\code{nk\_local\_optima()}), and the canonical ruggedness sweep
(\code{nk\_sweep\_K()}).

These functions are useful in their own right, but their role here is to
supply a reference implementation that shares no code with the SAOM engine.

### Verifying the reduction

Property 1 (Section~\ref{sec:formal-properties}) states that the classical NK model is the $M = 1$,
$\beta \rightarrow \infty$ special case of SaoMNK.  \code{nk\_to\_saomnk()} emits the
environment and structure-model settings that place a single-actor SaoMNK
environment on a supplied NK landscape, making the reduction constructive
rather than merely asserted.

\code{nk\_verify\_reduction()} then checks the construction.  It exploits the
defining property of the NK model: contribution $f_i$ depends only on the loci
in $\Omega(i)$, so any two configurations sharing the same neighborhood
pattern must carry identical $f_i$.  Rebuilding each locus's payoff table keyed
by that pattern and recomputing fitness must return the original values
exactly.  A non-zero difference would indicate an indexing or bit-ordering
error in landscape construction.

```{r nk-verify}
nk <- nk_landscape(N = 10, K = 3, seed = 42)
nk_verify_reduction(N = 10, K = 3, seed = 42)
```

The check passes at machine precision (maximum absolute difference
$0.00 \times 10^{0}$) across the $(N, K)$ combinations exercised in the test
suite.  Configuration encoding is most-significant-bit-first throughout and is
covered by exhaustive round-trip tests for $N = 1, \ldots, 12$; this guards a
class of bit-ordering error that is easy to introduce and silent when it
occurs.

### Recovering classical results

The implementation reproduces the standard qualitative findings of the NK
literature.  With $K = 0$ the landscape is single-peaked, and a steepest-ascent
walk from any starting configuration reaches the global optimum.  As $K$ rises
the number of local optima grows, approaching the familiar $2^N / (K+1)$
scaling, adaptive walks terminate sooner, and terminal fitness falls
\citep{Kauffman1987, KauffmanWeinberger1989, Levinthal1997}.  \code{nk\_sweep\_K()} reproduces this
experiment directly.

### Reproducing Levinthal (1997) {#sec:levinthal}

The first illustration establishes that \pkg{searchnet} can reproduce the
core dynamics of the classical NK model under appropriate parameter settings.
We configure an environment with no endogenous network effects, leaving only
the exogenous influence matrix active.  Setting both \code{popularity} and
\code{scope} to zero removes every term through which one actor's state can
enter another's objective function, so the actors search *independently*: the
model reduces to parallel single-searcher processes on a landscape whose
ruggedness is determined solely by the exogenous coupling matrix, which is the
functional equivalent of the classical NK search process.  We use $M = 2$ rather
than $M = 1$ because \code{search\_rsiena()} requires at least two actors ---
\pkg{RSiena}'s \code{sienaDataCreate()} does not accept a single-actor
bipartite network.  For genuinely single-actor work the landscape methods
(\code{compute\_fitness\_landscape}, \code{verify\_nk\_equivalence}) accept
$M = 1$ directly, and Section~\ref{sec:validation} uses them for exactly that.

```{r levinthal-setup, cache=TRUE}
## Two INDEPENDENT searchers (popularity = scope = 0 removes all
## coupling), 12 components, empty start.  M >= 2 is required by
## RSiena's bipartite data constructor; with the interaction terms
## zeroed the actors do not influence one another, so each is a
## classical NK searcher.
env_nk <- saomnk_env(M = 2, N = 12, density = 0, seed = 1234)

## Block-diagonal epistasis: 3 modules of 4 components (K ~ 3)
K_levinthal <- saomnk_block_diagonal(12, 3)

## Only epistasis active; no endogenous effects
mod_nk <- saomnk_model(
  density    = -0.3,
  popularity = 0,
  scope      = 0,
  influence_matrix = K_levinthal,
  influence_weight = 0.15
)

saomnk_run(env_nk, mod_nk, steps_per_actor = 100, seed = 42)
```

Under this parameterization, the agent's search trajectory exhibits the
signature features of NK dynamics: rapid initial improvement as the agent
discovers complementary component clusters, followed by convergence to a
local optimum where no single-component change improves utility.  The K-4
panel reduces to the $K_{AC}$ trajectory (actor scope) and the $K_{CC}$
trajectory (epistasis), with the latter reflecting the fixed block-diagonal
structure; the $K_{CA}$ and $K_{AA}$ channels are flat by construction, since
nothing couples the two searchers.

```{r levinthal-degrees, fig.height=3.5, fig.width=6, fig.cap="Actor scope ($K_{AC}$) trajectory for independent agents searching on a block-diagonal epistasis landscape.  The trajectory shows rapid scope expansion followed by stabilization---the SAOM equivalent of convergence to a local optimum on a rugged NK landscape."}
env_nk$plot_actor_degrees(loess_span = 0.5)
```

This result provides face validity: when all endogenous network effects are
disabled, \pkg{searchnet} recovers the qualitative behavior of the classical
NK model.  The landscape is effectively exogenous (determined by the fixed
$\mathbf{W}$ matrix), search is local (single-tie changes), and convergence
reflects the trapping property of rugged landscapes.

A deeper convergence analysis is provided in the Blume tutorial vignette
(\code{vignette("saomnk-blume-tutorial")}), which establishes the
statistical-mechanical foundations of the SAOM-NK choice rule by connecting it
to the Blume--Capel model \citep{Blume1993}, Gibbs measures, and the
Ising--logit correspondence.

## Coupled degree dynamics {#sec:coupling}

<!-- RECAST: author to write a lead-in for the coupling illustration (what the figure demonstrates about the package output); drafts in paper/_drafts/2026-10-07_recast_illustrations.md -->

![Simulated demonstration that the four $\K$ dimensions move as a coupled system.  A single seeded run of 20 actors over 24 components with a block-diagonal influence matrix, plotted by \pkg{searchnet} from the simulated chain.  Because all four degrees are read off the same bipartite incidence matrix, they rise together from an empty start; the smallest pairwise correlation across the four series is $0.98$.  This panel is simulated output, not empirical data.  Generated by \code{paper/replication/make\_k\_system\_figures.R}.](figures/fig_k_coupling_sim.png){width=80%}

## Ergodicity: forgetting the initial condition {#sec:ergodicity}

Property~4 (Section~\ref{sec:formal-properties}) asserts that the chain's stationary distribution is unique and
independent of where the simulation starts.  That property is easy to
illustrate badly.  Running the model once from a sparse start and once from a
dense start, then observing that the two final densities are close, establishes
very little: no threshold was declared in advance, a single run per arm supplies
no sampling distribution, and---most importantly---an ergodic chain forgets its
initial condition *at a rate*, which a pair of endpoints discards.

\pkg{searchnet} therefore ships the demonstration as a measurement.
\code{searchnet\_ergodicity\_sweep()} runs the same fixed-coefficient model from
two contrasting starting densities across a range of chain lengths, with
independent replicates at each, and reports how fast the between-arm gap decays:

```{r ergodicity, cache=TRUE, fig.height=5, fig.width=10, fig.cap="Independence from initial conditions, measured rather than asserted. (a) Two arms starting at density 0.10 and 0.80 converge as run length grows; points are replicates. (b) The between-arm gap decays as a power of run length on log--log axes. The dashed line is the declared equivalence margin; the shaded band is the Monte Carlo floor, below which the measured gap reflects replicate noise rather than mixing, and which is excluded from the fitted slope."}
erg <- searchnet_ergodicity_sweep(
  M = 12, N = 15, start_densities = c(0.1, 0.8),
  run_lengths = c(15, 30, 60, 120, 240),
  replicates = 10, equivalence_margin = 0.05, seed = 42
)
erg
plot(erg)
```

Two arms separated by a starting gap of 0.70 close to within 0.01 by 240
iterations per actor, a collapse of more than 99\%, and the gap decays as a
power of run length with slope near $-1$ and $R^2 = 0.617$.  The verdict is
computed rather than printed: at the longest run length the arms are compared by
a two-one-sided-tests equivalence procedure against a margin fixed in the call,
so the function is able to report *non*-equivalence.  This is what makes the sweep evidence rather than illustration.

Two details matter for anyone reproducing this.  The Monte Carlo floor is drawn
explicitly, because once the arms are genuinely equivalent the measured gap
settles at the expected difference between two sample means from the same
distribution rather than falling to zero; fitting a decay slope through those
points would overstate the rate.  And the grid is sized so that density has
resolution: on the 4 $\times$ 5 grid used in earlier versions of this
demonstration a single tie moved the density by 0.05, which is coarser than any
equivalence margin worth declaring.

The vignette derives detailed balance with respect to the Gibbs distribution for
single-flip logit revision.  The multinomial ministep simulated here has no
general Gibbs form for $M > 1$, so the convergence measured here rests on
ergodicity alone.

## Exogenous shocks {#sec:shocks}

This illustration demonstrates \pkg{searchnet}'s shock infrastructure.
We run two simulations from identical initial conditions: a baseline with
stable parameters and a treatment that experiences a density shock at the
simulation midpoint, modeling an environmental discontinuity (e.g., a
regulatory change that increases the cost of broad scope).

```{r shock-setup, cache=TRUE}
env_base <- saomnk_env(M = 6, N = 8, density = 0, seed = 42)
env_shock <- saomnk_env(M = 6, N = 8, density = 0, seed = 42)

K_small <- saomnk_block_diagonal(8, 2)
mod_base <- saomnk_model(
  density = -0.5,
  influence_matrix = K_small,
  influence_weight = 0.5
)

## Baseline: no shocks
saomnk_run(env_base, mod_base, steps_per_actor = 80, seed = 12345)

## Shocked: density drops from -0.5 to -2.0 at midpoint
shock_baseline <- saomnk_shock("density", parameter = -0.5,
                               portion = 1)
shock_event    <- saomnk_shock("density", parameter = -2.0,
                               portion = 1)

saomnk_run(env_shock, mod_base, steps_per_actor = 80, seed = 12345,
           shocks = list(shock_baseline, shock_event))
```

```{r shock-comparison, fig.height=8, fig.width=8, fig.cap="Baseline $\\{K\\}$-4 panel (this figure) and the shocked run (next figure).  The density shock at the midpoint (dashed vertical line) creates a visible structural break in all four degree trajectories.  Increased sparsity pressure compresses actor scope, which propagates through component popularity, actor sociality, and component coupling."}
## Display baseline K-4 panel
saomnk_plot_k4(env_base, smooth = 0.3)
```

```{r shock-treatment, fig.height=8, fig.width=8}
## Display shocked K-4 panel
saomnk_plot_k4(env_shock, smooth = 0.3)
```

The shock at the midpoint creates a visible structural break in all four
$\K$ dimensions.  The increased density penalty (from $-0.5$ to $-2.0$)
intensifies sparsity pressure, causing actors to shed affiliations.  This
compresses $K_{AC}$ (scope), which propagates through the coupled system:
lower scope reduces $K_{CA}$ (fewer actors per component), which decreases
$K_{AA}$ (fewer co-affiliating actors) and $K_{CC}$ (fewer co-occurring
components).  The landscape becomes less rugged as a consequence of the
exogenous shock to a single parameter.

This kind of quasi-experimental comparison---identical starting conditions
with and without a parameter intervention---is the basis for difference-in-differences analyses of search dynamics, a workflow that
\pkg{searchnet} supports natively through its shock infrastructure and
multi-wave simulation capabilities.

![Simulated demonstration that a shock relocates the coupled system.  Three seeded arms share starting conditions and the random seed, so their trajectories coincide exactly until the intervention at the dotted line.  A sparsity shock (density $-5.0 \rightarrow -6.5$) leaves every dimension near zero, as in the control; a popularity shock ($0.15 \rightarrow 1.20$) moves every dimension up.  Terminal mean $K_{CC}$ is $0.0$, $0.1$ and $14.6$ in the sparsity, control and popularity arms.  This panel is simulated output, not empirical data.  Generated by \code{paper/replication/make\_k\_system\_figures.R}.](figures/fig_k_shock_relocation.png){width=80%}


## From shocks to causal inference {#sec:causal}

The shock infrastructure demonstrated above does more than produce visually
compelling before--after comparisons.  Because \code{saomnk\_run} records the
full actor $\times$ step $\times$ treatment history, shocked simulations
generate panel data with the same structure required by standard causal
inference estimators.  The $\theta$-shock partitions the simulation timeline
into a pre-treatment period (steps before the shock) and a post-treatment
period (steps after), while the researcher controls which actors are
treated.

\pkg{searchnet} provides helper functions that extract this panel structure
and interface with the \pkg{did} package \citep{CallawayS2021}, which
implements the \citet{CallawayS2021} estimator for difference-in-differences
with staggered treatment adoption.  The workflow proceeds in three steps:
extract a panel data frame from the shocked and unshocked environments, estimate group--time
average treatment effects, and visualize the results.

```{r causal-panel, cache=TRUE}
## Treated arm: the six actors of the shocked run (env_shock, above).
## Comparison arm: the six actors of the unshocked run (env_base), which
## shares the initial network and the run seed. did::att_gt() needs at
## least five never-treated units; ids 7-12 keep the arms' units distinct.
shock_step <- min(env_shock$theta_shocks[[2]]$chain_step_ids)
panel_tr <- searchnet_causal_panel(env_shock, shock_step = shock_step,
                                   outcome = "utility")
panel_co <- searchnet_causal_panel(env_base, shock_step = shock_step,
                                   outcome = "utility",
                                   treated_actors = integer(0))
panel_co$actor_id <- factor(as.integer(as.character(panel_co$actor_id)) + 6L)

## Keep the steps both arms reach, so the panel is balanced.
common <- seq_len(min(max(panel_tr$step), max(panel_co$step)))
panel <- rbind(panel_tr[panel_tr$step %in% common, ],
               panel_co[panel_co$step %in% common, ])
str(panel)
```

The resulting data frame contains one row per actor--step combination, with
columns for actor identifier, simulation step, a binary treatment indicator
(1 for treated actors in every step, 0 otherwise), and the outcome
variable (here, actor utility).  This is precisely the structure expected by
\code{att\_gt} in the \pkg{did} package.

```{r causal-did, cache=TRUE, fig.cap="Difference-in-differences estimates from shocked and unshocked runs with the same start and seed.  The plot shows event-time average treatment effects on actor utility, with 95\\% simultaneous confidence intervals."}
did_result <- searchnet_did(panel)
did::aggte(did_result, type = "simple")
searchnet_causal_plot(did_result)
```

The connection between simulation shocks and causal inference is not merely
a convenience.  In empirical strategy research, difference-in-differences
designs require a credible source of exogenous variation and a parallel trends
assumption.  In the simulation setting, both are satisfied by construction:
the researcher controls the shock timing and magnitude, and the pre-treatment
trajectories of treated and control actors evolve under the same data-generating
process.  This means that simulated counterfactuals can serve as a *complement*
to empirical estimation---the researcher can first calibrate the SAOM parameters
to match an observed network, then introduce a shock that mirrors a real
environmental discontinuity and compare the simulated treatment effects to
those estimated from observational data.

For a complete treatment of the causal inference workflow, including staggered
adoption designs, sensitivity analysis, and integration with empirical panel
data, see the \code{vignette("causal-inference", package = "searchnet")}.


## Basin geometry {#sec:basins}

The \pkg{searchnet} package provides tools for analyzing the geometry of
fitness landscape basins---regions of the configuration space that converge
to local optima under gradient ascent.  Basin analysis quantifies two
fundamental properties of competitive positions.

**Basin depth** measures the fitness gap between a local peak and its basin
boundary:
\begin{equation}\label{eq:basin-depth}
  \text{Depth}(p) = F(p) - \frac{1}{|\partial B(p)|} \sum_{x \in \partial B(p)} F(x),
\end{equation}
where $\partial B(p)$ is the boundary of basin $B(p)$.

**Basin width** measures the mean Hamming distance of basin members from the
peak:
\begin{equation}\label{eq:basin-width}
  \text{Width}(p) = \frac{1}{|B(p)|} \sum_{x \in B(p)} d_H(x, p).
\end{equation}
Width captures strategic flexibility---the number of distinct configurations
that converge to the same competitive outcome.

### Computing basin geometry

The \code{compute\_fitness\_landscape()} method enumerates configurations
and identifies local peaks:

```{r basin-landscape, eval=FALSE}
env <- SaomNkRSienaBiEnv$new(list(M = 8, N = 8, BI_PROB = 0.4,
                                  rand_seed = 42))
env$compute_fitness_landscape(n_landscapes = 1)

## Extract landscape data
landscape <- env$fitness_landscape[1,,]
N <- env$N
configs  <- landscape[, 1:N]
fitness  <- landscape[, 2*N + 1]
is_peak  <- landscape[, 2*N + 2] == 1
```

Basin assignment proceeds by mapping each configuration to its nearest peak
via Hamming distance, then computing depth and width for each basin.

### Landscape erosion under shocks

When shocks modify the influence matrix $\mathbf{W}$, the landscape
reshapes.  The \pkg{searchnet} shock infrastructure enables systematic
measurement of **landscape erosion**:
\begin{equation}\label{eq:erosion}
  \text{Erosion}(p) = \text{Depth}_{\mathbf{W}_0}(p) - \text{Depth}_{\mathbf{W}_1}(p).
\end{equation}

\noindent The
\code{shock\_theta\_matrix()} function implements regime changes at
designated iterations, enabling pre- vs.\ post-shock basin comparison
within a single simulation run.

## Firm and industry effects {#sec:applications}

We illustrate the package with a worked application built on simulated data.
It states a problematization, the specification that answers it, and a
takeaway, so that a reader can find the closest case to their own problem.

| Application | Domain | Mode of use | $\K$ channels driven |
|---|---|---|---|
| Firm vs.\ industry effects | Variance decomposition | Simulation engine | $K_{AC}$, $K_{CA}$, $K_{CC}$ |

Two channels deserve a warning that applies throughout.  First, $K_{CA}$ is
driven by \code{inPop} (degree popularity), not by similarity-weighted
imitation.  With an actor covariate, \pkg{RSiena}~1.5.0 does define
\code{simEgoInDist2} for a bipartite dependent variable, but \pkg{searchnet}'s
co-holder statistic, \code{saomnk\_coholder\_similarity()}, is a different
quantity.  \pkg{searchnet} supplies it for post-hoc computation
but cannot make it a simulable evaluation term, and the application below does
not use it as one.  Second, $K_{AA}$ is a *projection*, so it is frequently measured as an
outcome rather than entered as an effect; the table records what is driven, not
what is reported.

The variance-decomposition literature asks whether firm or industry effects
explain more of the variance in profitability.  If firms are competitively
coupled, the two are not separable in principle, and a simulation in which the
data-generating process is known can show what a decomposition recovers.

This application is the cleanest illustration of the minimal functional
pipeline, and we present it as such:

```{r app-variance-spec, eval=FALSE}
env <- saomnk_env(M = M, N = N, density = 0.2, seed = seed)
## arch_type: "modular", "hierarchical" or "random"
K   <- make_influence_matrix(N, arch_type)

mod <- saomnk_model(density = -0.5, popularity = coupling,
                    scope = 0.05, influence_matrix = K,
                    influence_weight = 0.1)

saomnk_run(env, mod, steps_per_actor = 15L, seed = seed + 1L)
bi_mat <- saomnk_get_bipartite(env)
```

Coupling strength enters through \code{popularity}, sweeping $\{0, 0.1, 0.3,
0.5, 0.8\}$ across three architectures.  We report the pipeline, not a bias
estimate: in the present implementation the "true" effects are re-derived from
the same realized performance the estimator sees, so the recovered bias is zero
by construction.  Establishing a bias requires taking the truth from the
generating parameters instead, and that work is ongoing.  We state this because
a software paper is the wrong place to launder an unfinished result, and because
the pipeline above is genuinely the shortest path from a specification to a
simulated bipartite panel.

**Takeaway.**  The shortest path from specification to simulated panel is four
calls.  The vignette also shows what a simulation cannot do for you: if the
"true" effects are re-derived from the same realized outcome the estimator sees,
recovered bias is zero by construction, whatever the coupling.

\subsubsection[Failure modes inherited from RSiena]{Failure modes inherited from \pkg{RSiena}}\label{sec:app-failures}

Because \pkg{searchnet} delegates estimation to \code{siena07}, it inherits that
estimator's failure modes, and users should know two of them.

The first is provable.  Under method-of-moments estimation, no effect whose
observed statistic lies exactly on the boundary of its attainable support can
converge: the estimate diverges as $\theta \to \pm\infty$.  In a *balanced*
bipartite panel every actor holds at least one component in every wave, so the
isolate count is exactly zero and \code{outTrunc(1)} is exactly $M$ --- both
saturated.  The screen is one
line --- compute $(\text{observed} - \text{lo}) / (\text{hi} - \text{lo})$ and
refuse exactly $0$ or $1$ --- and it costs nothing to run before estimating.  We
note that a stronger version of this claim, covering *near*-boundary statistics,
was tested and refuted: only exact saturation disqualifies.

<!-- RECAST: boundary-saturation figure and its external-study numbers withdrawn 2026-10-08 (data source not in the repository); author may add a demonstration on synthetic or documented public data. -->

The second is a silent-failure pattern rather than a theorem, and it bit this
package.  \code{RSiena::includeEffects} matches effect names by non-standard
evaluation, so passing a name held in a variable silently selects nothing.  A
version of \pkg{searchnet} passed \code{shortName = eff\$effect} into \code{...},
where the unevaluated expression was deparsed and no effect matched; because the
call was wrapped in \code{tryCatch(..., error = function(e) warning(...))}, the
simulation continued with the covariate effect *absent*.  Every actor-covariate
effect was affected, so results labeled "heterogeneous strategies" were in fact
homogeneous-actor results.  The package now ships a regression test asserting
that every effect declared in a \code{structure\_model} appears with
\code{include == TRUE} in the fitted effects object.  The design lesson
generalizes beyond this package: \code{tryCatch}-to-warning is right for
interactive exploration and wrong for a simulation engine whose output gets
published.


# Verification and validation {#sec:verification}

<!-- RECAST: author to write the opening of this section (the properties below are documented properties of the implemented model, checked by tests and the package Lean library; no novelty claim; statistics pinned to RSiena targets; what is not verified); drafts in paper/_drafts/2026-10-07_recast_verification.md -->

## Formal properties {#sec:formal-properties}

We state the properties here; complete proofs with all lemmas are provided in
\code{inst/proofs/} in the package (access via \code{searchnet\_proof()}).

\begin{description}

\item[Property 1 (reduction to NK).] The standard NK model is a degenerate special case
of the SAOM-NK model.  Specifically, the NK landscape and its adaptive walk are
recovered exactly under five restrictions: (R1)~$M = 1$ (single actor);
(R2)~$\mathbf{E} = \mathbf{A}$ (SAOM-NK influence matrix equals the NK
influence matrix); (R3)~$\boldsymbol{\theta} = \mathbf{0}$ (all SAOM
parameters zero); (R4)~$\beta \to \infty$ (deterministic greedy choice);
(R5)~uniform component selection probability.
\newline Test: \code{tests/testthat/test-nk-equivalence.R}, \code{tests/testthat/test-nk-classic.R}.
\newline Lean: \code{SaomNK.utility\_eq\_nkFitness}, \code{SaomNK.logitProb\_tendsto\_one}, \code{SaomNK.stayProb\_tendsto\_one}.

\item[Property 2 (generalization of NK).] SAOM-NK strictly generalizes the NK model
along three dimensions: (a)~multiple actors ($M > 1$), who affect one another
only through a cross-actor effect such as \code{inPop}, that is, through (b), (b)~endogenous landscape co-evolution through SAOM network
statistics, and (c)~bounded rationality with tunable search precision
($\beta \in (0, \infty)$).  The coupled degree vector
$\K = (K_{AC}, K_{CA}, K_{AA}, K_{CC})$ spans a strictly larger class of
fitness landscapes than the scalar $K$ parameterization.  The classical NK
model is recovered as a proper special case: NK $\subsetneq$ SAOM-NK.
\newline Test: \code{tests/testthat/test-property-2.R}.
\newline Lean: \code{SaomNK.card\_config\_gt}, \code{SaomNK.CoreSpec.landscape\_endogeneity}, \code{SaomNK.logitProb\_pos}.

\item[Property 3 (approximation of NK landscapes).] For any NK landscape
$(N, K, \mathbf{A}, \{f_d\}_{d=1}^N)$, there exists a parameterization
of the SAOM-NK model that reproduces the NK payoff structure to arbitrary
precision.  It holds exactly by (a)~constructive existence via dummy
covariates.  (b)~Actor heterogeneity substitution, with covariates linear in the
actor's own row, reproduces only $K = 0$ landscapes; (c)~universal
approximation via mixed logit is not checked numerically.
\newline Test: \code{tests/testthat/test-property-3.R}.
\newline Lean: \code{SaomNK.nkTermFull\_in\_dummy\_span}.

\item[Property 4 (SAOM--QRE correspondence).] The SAOM multinomial logit choice
rule (Equation~\ref{eq:choice}) is isomorphic to a quantal response
equilibrium (QRE) in the bipartite affiliation game, establishing a
game-theoretic foundation for the statistical model.  The resulting chain has a
unique stationary law; its Gibbs form and zero-noise selection of potential
maximizers are guaranteed for single-flip logit revision, not for this
multinomial rule when $M > 1$.
\newline Test: \code{tests/testthat/test-searchnet-ergodicity.R}, \code{tests/testthat/test-potential-cycles.R}, \code{tests/testthat/test-lean.R}.
\newline Lean: \code{SaomNK.Spec.exact\_potential}, \code{SaomNK.detailed\_balance\_of\_potential}, \code{SaomNK.global\_balance\_of\_potential}, \code{SaomNK.reachable}.

\item[Property 5 (Brock--Durlauf recovery).] The \citet{BrockDurlauf2001} binary
discrete-choice-with-social-interactions model is a degenerate special case of
SaoMNK. Specifically, under restrictions Rb1--Rb5 (Online Appendix~C), the SaoMNK Gibbs stationary
distribution reduces (componentwise) to the rational-expectations
self-consistent equilibrium of \citet{BrockDurlauf2001}.
\newline Test: \code{tests/testthat/test-mean-field.R}.
\newline Lean: \code{SaomNK.MeanField.exists\_nonzero\_root\_iff}.

\end{description}

## Proof registry {#sec:proofs}

The package ships with a complete formal verification chain for the SaoMNK
framework, accessible via \code{searchnet\_proof()}.  The proof registry
(\code{inst/proofs/PROOF\_TABLE.md}) enumerates `r n_proof_steps` logical steps---from axioms
through definitions, lemmas, and propositions to the main properties---organized so that every result can be traced backward through its
dependencies to foundational axioms.  This self-contained proof chain
provides reviewers and users with a verifiable logical scaffold for the
theoretical claims summarized in Section~\ref{sec:formal-properties}.

## Brock--Durlauf harness (Options B and C) {#sec:bd-harness}

<!-- RECAST: author to write the rationale for the two options as the code now stands (since 0.11.0 RSiena's inPop on a two-mode variable is linear and the operational fixed point is the finite-N ministep map F_N, see R/searchnet-brock-durlauf.R); the sqrt-premise sentence, the sqrt rescaling derivation, and the M = 20/50/100 results table with its four-decimal-places paragraph were deleted as no longer true; re-run numbers are needed; drafts in paper/_drafts/2026-10-07_recast_verification.md -->

Lemma 5 (utility reduction, L8) and Lemma 6 (mean-field self-consistency, L9)
together imply that the SaoMNK empirical mean adoption
$m_j(\mathbf{B}^{\text{stationary}})$ should converge to the analytical B\&D
fixed point. The companion verification function \code{verify\_brock\_durlauf\_reduction()}
reports both analyticals.

**Option C (B\&D fixed point, primary).** Default parameters $J_b = 0.5$,
$h_b = -J_b/2 = -0.25$, $\beta = 1$ give the symmetric configuration with
$m^* = 0.5$ (sub-threshold $\beta J_{\sigma} = 1/8 < 1$); this puts the
system inside the regime where the linearization is locally accurate
($|m^* - 0.5| < 0.15$, indicated by \code{in\_BD\_regime = TRUE} in the harness
output). In this regime the empirical converges to B\&D's analytical fixed
point at $O(M^{-1/2})$.

**Option B (saomnk-inPop fixed point, secondary).** For arbitrary
$(\beta, J_b, h_b)$ outside the Option C regime, the simulation converges
to the actual SaoMNK-inPop fixed point
$m = F_N\bigl(\beta(h_b + \theta_{\mathrm{inPop}}(1 + (M-1)m))\bigr)$,
solved via \code{saomnk\_inpop\_self\_consistency()}. The harness reports
\code{abs\_error\_saomnk} as the binding numerical metric in this case.

```{r bd-verification, eval = FALSE}
library(searchnet)

# Option C regime: J_b modest, h_b symmetric -> m* ~= 0.5
result <- verify_brock_durlauf_reduction(
  M_seq = c(20, 50, 100),
  J_b = 0.5, h_b = -0.25,        # symmetric, sub-threshold
  beta = 1,
  n_components = 8,
  n_steps = 100,                  # 100 ministeps per actor
  n_replicates = 3
)
print(result[, c("M", "m_b_empirical", "m_BD_analytical",
                 "m_saomnk_analytical", "in_BD_regime",
                 "abs_error_BD", "abs_error_saomnk")])
# in_BD_regime = TRUE everywhere; abs_error_BD ~ 0.01-0.05
```

**Tradeoffs.** Option C uses B\&D's own concept directly but cannot probe the
multiplicity regime ($\beta J > 1$, requiring $|m^*|$ far from 1/2). Option B
covers all parameter combinations but introduces a non-standard fixed-point
object with no literature precedent. For SAOM/RSiena practitioners, Option C
is the natural setting: standard RSiena estimations have fitted thetas in
$(-2, 2)$ log-odds with equilibrium magnetizations near 1/2, so restricting
to Option C is alignment with the framework's natural regime, not a workaround.

**Tier 1 evidence.** Recovering small empirical-vs-analytical error in the
Option C regime provides Tier-1 evidence for the reduction
(\code{PROOF\_TABLE.md} L14): the limit *is* the B\&D
rational-expectations equilibrium, not an approximation to it. The
multiplicity claim (Corollary 2, L11) is verified analytically via
\code{bd\_equilibrium\_count()}; the *empirical*
verification of the pitchfork bifurcation requires future work (a native
\code{congestion} RSiena effect; see L16 roadmap note).


# Computational performance {#sec:performance}

<!-- RECAST: author to write the replacement numbers and their protocol after the chain-processing optimization lands; every number and growth sentence in this section (overhead, scalability table, landscape timings) is pre-optimization, and the overhead and growth-in-N sentences already fail the 2026-10-07 measurement (reparam_results Section 3); drafts in paper/_drafts/2026-10-07_recast_performance.md -->

This section provides benchmarks addressing computational overhead and
scalability, computed on a laptop with an AMD Ryzen AI 9 HX 370 processor
(12 cores), 32 GB RAM, and \proglang{R} 4.5.3 on Windows 11.

\subsection[Overhead relative to RSiena]{Overhead relative to \pkg{RSiena}}

\pkg{searchnet} adds computational overhead to the core \pkg{RSiena}
simulation through degree computation, projection construction, chain
statistics extraction, and visualization data preparation.  To quantify this
overhead, we time matched simulations using raw \pkg{RSiena} calls versus
\pkg{searchnet}'s \code{saomnk\_run()} for equivalent bipartite SAOM
specifications.

```{r benchmark-overhead, cache=TRUE}
## --- searchnet wrapper timing (M=4, N=6, 30 steps/actor) ---
time_searchnet <- system.time({
  e_bench <- saomnk_env(M = 4, N = 6, seed = 1)
  m_bench <- saomnk_model(
    density = -0.5,
    influence_matrix = saomnk_block_diagonal(6, 2)
  )
  saomnk_run(e_bench, m_bench, steps_per_actor = 30, seed = 1)
})

## --- raw RSiena siena07 call on equivalent bipartite problem ---
time_rsiena <- system.time({
  ## Build the same bipartite SAOM data and model by hand
  net_start <- matrix(0L, nrow = 4, ncol = 6)
  net_end   <- matrix(0L, nrow = 4, ncol = 6)
  dep_var   <- RSiena::sienaDependent(
    array(c(net_start, net_end), dim = c(4, 6, 2)),
    type = "bipartite", nodeSet = c("actors", "components"),
    ## Both waves are empty, so without allowOnly = FALSE RSiena
    ## infers that no change is admissible and siena07 rejects the
    ## model as having no effects.
    allowOnly = FALSE
  )
  actors_set <- RSiena::sienaNodeSet(4, nodeSetName = "actors")
  comp_set   <- RSiena::sienaNodeSet(6, nodeSetName = "components")
  dat_raw    <- RSiena::sienaDataCreate(
    dep_var,
    nodeSets = list(actors_set, comp_set)
  )
  ## getEffects() already includes the rate and density terms for
  ## a bipartite dependent variable: the specification being timed.
  eff_raw <- RSiena::getEffects(dat_raw)
  alg_raw <- RSiena::sienaAlgorithmCreate(
    projname = "bench_raw",
    nsub = 0, n3 = 4 * 30, simOnly = TRUE, seed = 1
  )
  suppressMessages(
    RSiena::siena07(alg_raw, data = dat_raw, effects = eff_raw,
                    batch = TRUE, silent = TRUE, returnDeps = TRUE)
  )
})

overhead_pct <- round(
  100 * (time_searchnet["elapsed"] - time_rsiena["elapsed"]) /
    time_rsiena["elapsed"], 1
)
cat(sprintf(
  "searchnet: %.2fs | raw RSiena: %.2fs | overhead: %.1f%%\n",
  time_searchnet["elapsed"], time_rsiena["elapsed"], overhead_pct
))
```

Typical overhead is 15--25\% relative to a raw \pkg{RSiena} simulation of
equivalent size, with the majority attributable to chain statistics extraction
and $\K$-4 degree computation.  The projection operations
(Equations~\ref{eq:proj-actor}--\ref{eq:proj-component}) are $O(M^2 N)$ and
$O(M N^2)$ respectively, but these are dominated by the \pkg{RSiena} simulation
cost for problem sizes up to $M, N \approx 50$.

## Scalability

Table~\ref{tab:scalability} reports median wall-clock times for
\code{saomnk\_run()} across representative problem sizes.

\begin{table}[t]
\centering
\begin{tabular}{rrr}
\toprule
$M$ & $N$ & Time (s) \\
\midrule
 6  &  8  &  0.230 \\
10  & 15  &  0.260 \\
10  & 30  &  0.410 \\
12  & 20  &  0.295 \\
20  & 15  &  0.570 \\
20  & 30  &  0.510 \\
20  & 40  &  0.645 \\
30  & 50  &  0.985 \\
\bottomrule
\end{tabular}
\caption{Median wall-clock time (seconds) for \code{saomnk\_run()} with
\code{steps\_per\_actor = 30}, based on 10 replications
(\code{paper/replication/reparam/bench\_protocol.R}; seeds $1000 + r$).  Influence matrix is
block-diagonal with $\lfloor N/3 \rfloor$ blocks.}
\label{tab:scalability}
\end{table}

Scaling is approximately quadratic in $N$ (driven by the projection
computations and the size of the \pkg{RSiena} change statistic vectors) and
linear in $M$ (driven by the number of actors evaluated at each micro-step).
For Monte Carlo experiments requiring many replications, the
\code{SaoMNKexperiments} class provides parallel execution via
\code{parallel::mclapply()} on Unix systems or sequential execution with
progress reporting on Windows.

## Fitness landscape computation

The \code{compute\_fitness\_landscape()} method enumerates all $2^N$ possible
component configurations to compute the full fitness surface.  This computation
is $O(2^N)$ and is feasible only for $N \leq 20$ on standard hardware.
For $N = 12$ (4,096 configurations), landscape computation takes approximately
0.3 seconds; for $N = 20$ (1,048,576 configurations), approximately 45
seconds.  For larger $N$, the method issues a warning and offers an MCMC-based
sampling approximation that draws $S$ random configurations (default
$S = 10{,}000$) to estimate landscape statistics (number of local optima,
correlation length, global peak location) without full enumeration.

```{r landscape-scaling, cache=TRUE}
## Landscape timing: full enumeration vs. sampled approximation
landscape_sizes <- c(6, 8, 10, 12)
landscape_times <- data.frame(
  N = integer(), method = character(), elapsed = numeric(),
  stringsAsFactors = FALSE
)

for (n in landscape_sizes) {
  ## Full enumeration
  e_land <- saomnk_env(M = 4, N = n, density = 0.3, seed = 42)
  m_land <- saomnk_model(
    density = -0.5,
    influence_matrix = saomnk_block_diagonal(n, max(2, n %/% 3)),
    influence_weight = 0.05
  )
  saomnk_run(e_land, m_land, steps_per_actor = 10, seed = 1)

  t_full <- system.time(e_land$compute_fitness_landscape())
  landscape_times <- rbind(landscape_times, data.frame(
    N = n, method = "full", elapsed = t_full["elapsed"]
  ))

  ## Sampled approximation (S = 1000)
  t_samp <- system.time(
    e_land$compute_fitness_landscape(sample_size = 1000)
  )
  landscape_times <- rbind(landscape_times, data.frame(
    N = n, method = "sample_1000", elapsed = t_samp["elapsed"]
  ))
}

## Display results
landscape_wide <- reshape2::dcast(landscape_times, N ~ method,
                                  value.var = "elapsed")
landscape_wide$configs <- 2^landscape_wide$N
knitr::kable(
  landscape_wide[, c("N", "configs", "full", "sample_1000")],
  col.names = c("$N$", "Configurations", "Full (s)", "Sample (s)"),
  digits = 3,
  caption = paste("Fitness landscape computation time: full",
                  "enumeration vs.\\ sampled approximation",
                  "($S = 1{,}000$).")
)
```

# Comparison with existing tools {#sec:comparison}

\pkg{searchnet} occupies a unique niche at the intersection of fitness landscape
simulation and social network modeling.  Table~\ref{tab:feature-comparison}
summarizes the feature sets of four relevant tools.

\begin{table}[t]
\centering
\begin{tabular}{lcccc}
\toprule
Feature & \pkg{searchnet} & \pkg{nkmodel} & \pkg{RSiena} & NetLogo \\
\midrule
NK landscape generation           & $\checkmark$              & $\checkmark$  & ---           & (ext.)    \\
Multi-actor search                & $\checkmark$ (endogenous) & ---           & $\checkmark$ (networks) & $\checkmark$ (ABM) \\
Bipartite networks                & $\checkmark$              & ---           & $\checkmark$  & (ext.)    \\
Influence matrix ($\mathbf{B}'\mathbf{W}\mathbf{B}$) & $\checkmark$ & (implicit via $K$) & --- & (manual) \\
Statistical estimation            & $\checkmark$ (via RSiena) & ---           & $\checkmark$  & ---       \\
Causal inference pipeline         & $\checkmark$ (DID/Synth)  & ---           & ---           & ---       \\
$\K$ framework                    & $\checkmark$              & ---           & ---           & ---       \\
\bottomrule
\end{tabular}
\caption{Feature comparison of software tools for fitness landscape and
network-based search simulation.  ``$\checkmark$''~indicates native support;
``---''~indicates no support; parenthetical notes indicate partial or indirect
support.}
\label{tab:feature-comparison}
\end{table}

The three most relevant points of comparison are as follows.

**\pkg{nkmodel}.** The \pkg{nkmodel} package \citep{Baumann2019nk} provides a
faithful implementation of the classical NK model: it generates random fitness
landscapes, computes fitness values for binary configuration vectors, and
supports hill-climbing search.  However, it implements a single-agent model in
which the landscape is exogenously fixed.  There is no network structure, no
multi-actor interaction, and no estimation machinery.  The epistasis structure
is controlled implicitly through the scalar $K$ parameter rather than through
an explicit influence matrix, precluding modular or block-diagonal
specifications.  \pkg{searchnet} subsumes the \pkg{nkmodel} use case as a
special case: setting $M = 1$ and disabling all endogenous network effects
recovers the classical NK model, while the explicit influence matrix
$\mathbf{W}$ (Section~\ref{sec:bipartite}) provides finer-grained control over
interdependence structure.

**\pkg{RSiena}.** The \pkg{RSiena} package \citep{RSiena} is the reference
implementation for stochastic actor-oriented models.  It provides a complete
estimation and simulation framework for network evolution, including bipartite
networks.  However, it contains no fitness landscape machinery---no epistasis
specification, no landscape visualization, and no bridge to the NK
literature.  Using \pkg{RSiena} alone, a researcher must manually construct the
mapping between network statistics and landscape properties, a task that
\pkg{searchnet} automates through the $\K$ framework.  \pkg{searchnet} uses
\pkg{RSiena} as its simulation engine, wrapping it with a domain-specific API
that translates strategy-theory concepts (epistasis, fitness, ruggedness) into
the corresponding SAOM specification.

**NetLogo.** NetLogo \citep{Wilensky1999} is widely used for agent-based
simulation in management research.  While NK landscapes can be implemented via
user-written extensions, NetLogo does not provide native NK support, bipartite
network primitives, or statistical estimation.  Simulations are specified
procedurally rather than declaratively, and reproducibility depends on the
modeler's implementation choices.  \pkg{searchnet}'s declarative specification
(\code{saomnk\_model()}) and validation against the classical NK model provide
stronger guardrails for computational experiments.

In summary, \pkg{nkmodel} provides NK landscapes without networks,
\pkg{RSiena} provides networks without landscapes, and NetLogo provides
neither natively.  \pkg{searchnet} bridges both traditions, adding the $\K$
interpretive framework and a causal inference pipeline that none of the
alternatives offer.

# Summary and limitations {#sec:conclusion}

\pkg{searchnet} provides an \proglang{R} package that bridges two established
research traditions---NK fitness landscapes from strategy and organization
theory, and stochastic actor-oriented models from social network
analysis---through a bipartite network reformulation of the search environment.

<!-- RECAST: author to write the summary and limitations prose (scope, RSiena scale, panel waves needed, what is not verified, roadmap), replacing the deleted contribution paragraphs (endogenizes the landscape; decomposes K); drafts in paper/_drafts/2026-10-07_recast_summary.md -->

It provides a **complete simulation and analysis toolkit**.  The
functional API (\code{saomnk\_env}, \code{saomnk\_model}, \code{saomnk\_run})
lowers the barrier to entry for management scholars unfamiliar with \pkg{RSiena}
syntax.  The $\K$-4 panel, 3D phase space visualization, utility
decomposition, network snapshots, and specification robustness diagnostics
equip researchers with the tools needed to design, execute, and validate
computational experiments.  The package ships with `r n_vignettes` vignettes, 31 Manim
animation scenes, a `r n_proof_steps`-step formal proof chain (Parts A--N, including
Property 5's Brock--Durlauf recovery), 820 unit tests, and
interactive game and classroom modules that make the framework accessible to
students and practitioners.

Several directions for future development are planned.  First, integration with
\pkg{RSiena}'s estimation machinery will allow researchers to fit the SaoMNK
model to observed bipartite network panel data, enabling empirical calibration
of simulation parameters.  Second, the fitness landscape module will be
extended with scalable approximation methods (e.g., Walsh decomposition,
landscape correlation analysis) to support problem sizes beyond $N = 20$.
Third, a \pkg{Shiny} dashboard is under development to provide an interactive
visual interface for simulation design and exploration, lowering the adoption
barrier for researchers without programming experience.  Fourth, a web API via
\pkg{plumber} will enable integration with external frontends and
cross-platform deployment.  Two initially planned future directions have
already been realized in the current release: the strategy game mode
(Section~\ref{sec:interactive}) provides gamified engagement with the
framework, and the classroom teaching module offers a Capsim-style
instructional platform with industry presets, AI opponents, and automated
debrief generation.  The accompanying PDW workshop vignette
(\code{vignette("saomnk-pdw-workshop")}) provides a complete three-hour
curriculum for introducing the framework to strategy and organization
scholars.

\pkg{searchnet} is available from GitHub at
\url{https://github.com/sdownin/searchnet} and is under preparation for CRAN
submission.  The package requires \proglang{R} $\geq$ 4.1.0 and depends on
\pkg{R6}, \pkg{RSiena}, \pkg{igraph}, \pkg{ggplot2}, \pkg{dplyr},
\pkg{Matrix}, \pkg{network}, and \pkg{reshape2}.


# Computational details {#sec:compdetails}

The results in this paper were obtained using \proglang{R}~`r r_version` with the
packages \pkg{searchnet}~`r sn_version` and \pkg{RSiena}~`r rsiena_version`
(\pkg{searchnet} requires \pkg{RSiena}~$\geq$~1.5.0).  \proglang{R} itself and
all packages used other than \pkg{searchnet} are available from the
Comprehensive \proglang{R} Archive Network (CRAN) at
\url{https://CRAN.R-project.org/}.  \pkg{searchnet} is not on CRAN; it is
available from GitHub at \url{https://github.com/sdownin/searchnet}.



