---
title: "Online Appendix: searchnet --- Network-Embedded Strategic Search Simulation"
author: "Stephen Downing, University of Missouri"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    self_contained: true
    fig_width: 9
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  ## SEARCHNET_JSS_FAST=1 (paper/build_submission.R --fast) renders the
  ## appendix with chunks unevaluated, as a format check only.
  eval      = !identical(Sys.getenv("SEARCHNET_JSS_FAST"), "1")
)
options(digits = 4)
```

```{r load-saomnk, include=FALSE}
## paper/build_submission.R loads the package from the repository root
## with pkgload::load_all() before rendering; otherwise use the installed
## package.
if (!"searchnet" %in% loadedNamespaces()) library(searchnet)

library(Matrix)
library(ggplot2)

DV_NAME <- "self$bipartite_rsienaDV"
```

---

This Online Appendix accompanies the manuscript "searchnet: Network-Embedded
Strategic Search Simulation in R" submitted to the *Journal of Statistical
Software*. It provides: (A) a quick-start tutorial with runnable code and figures, NK
validation against Levinthal (1997), and a full simulation walkthrough with
shocks; (B) the formal mathematical framework; (C) the properties checked by
the package and the complete proof registry; and (D) the causal inference
pipeline.

---

# Appendix A: Quick-Start Tutorial {.tabset}

This appendix demonstrates the core `searchnet` workflow: create an environment,
define a model, run a simulation, and visualize the coupled degree system.

## Environment and Model

```{r appendix-a-env}
env_qs <- SaomNkRSienaBiEnv$new(list(
  M = 3, N = 6, BI_PROB = 0, rand_seed = 1234, name = "_quickstart_"
))
cat("Environment:", env_qs$M, "actors x", env_qs$N, "components\n")
```

The structure model defines the SAOM objective function. We specify a
block-diagonal influence matrix (2 blocks of 3 components) via the `XWX`
effect, with a negative `density` parameter creating sparsity pressure:

```{r appendix-a-model}
epistasis_qs <- create_block_diag(6, 2)

structure_qs <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -0.3, dv_name = DV_NAME, fix = TRUE),
      list(effect = "inPop",   parameter =  0.2, dv_name = DV_NAME, fix = TRUE)
    ),
    coDyadCovars = list(
      list(
        effect = "XWX", parameter = 0.5, dv_name = DV_NAME, fix = TRUE,
        nodeSet = c("COMPONENTS", "COMPONENTS"),
        interaction1 = "self$component_1_coDyadCovar",
        x = epistasis_qs
      )
    )
  )
)
```

## Running the Simulation

```{r appendix-a-run}
env_qs$search_rsiena(
  structure_model = structure_qs,
  iterations_per_actor = 20,
  run_seed = 12345,
  verbose = TRUE
)
```

## K-4 Panel Plot

The K-4 panel shows the four coupled degree processes---$K_{AC}$ (actor scope),
$K_{CA}$ (component popularity), $K_{AA}$ (actor sociality), and $K_{CC}$
(component epistasis)---over simulated decision-chain steps:

```{r appendix-a-k4, fig.height=10, fig.width=10}
env_qs$plot_degree_4panel(loess_span = 0.3)
```

## Network Snapshots

Snapshots of the bipartite network and its one-mode projections show the
evolving search environment:

```{r appendix-a-snapshots, fig.height=5, fig.width=11}
n_steps_qs <- env_qs$get_step()
snapshot_ids_qs <- unique(c(1, round(n_steps_qs / 2), n_steps_qs))
env_qs$plot_snapshots(snapshot_ids_qs)
```

## Utility Decomposition

Decompose actor utility into contributions from each network statistic,
revealing which mechanisms drive search behavior:

```{r appendix-a-utility, fig.height=8, fig.width=9}
env_qs$plot_utility_contributions(loess_span = 0.35, use_thetas = TRUE)
```

## NK Validation (Levinthal 1997 Reproduction)

This appendix validates the SaoMNK framework against canonical results from the
NK fitness landscape literature.

### Single-Agent Rugged Landscape

In the classical NK model, ruggedness is governed by $K$. Levinthal (1997)
showed that higher $K$ increases local optima and traps the agent at suboptimal
peaks. SaoMNK reproduces this with $M=2$, $N=8$, and modular epistasis:

```{r nk-val-rugged}
epistasis_nk <- create_block_diag(8, 2)

env_rugged <- SaomNkRSienaBiEnv$new(list(
  M = 2, N = 8, BI_PROB = 0, rand_seed = 42,
  name = "_nk_rugged_"
))

sm_rugged <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -1.5, dv_name = DV_NAME, fix = TRUE)
    ),
    coCovars = list(), varCovars = list(),
    coDyadCovars = list(
      list(
        effect = "XWX", parameter = 1.5, dv_name = DV_NAME, fix = TRUE,
        nodeSet = c("COMPONENTS", "COMPONENTS"),
        interaction1 = "self$component_1_coDyadCovar",
        x = epistasis_nk
      )
    ),
    varDyadCovars = list()
  )
)

env_rugged$search_rsiena(sm_rugged, iterations_per_actor = 30, run_seed = 12345)
```

```{r nk-val-rugged-k4, fig.height=10, fig.width=10}
env_rugged$plot_degree_4panel(loess_span = 0.5)
```

### Multi-Agent Endogenous Landscape

With $M=6$ and endogenous effects activated, $K_{AA}$ (actor sociality) and
$K_{CC}$ (component epistasis) emerge from the coupled bipartite process:

```{r nk-val-endog, fig.height=10, fig.width=10}
env_endog <- SaomNkRSienaBiEnv$new(list(
  M = 6, N = 8, BI_PROB = 0, rand_seed = 99,
  name = "_nk_multi_endog_"
))

sm_endog <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -1.0, dv_name = DV_NAME, fix = TRUE),
      list(effect = "inPop",   parameter = 0.2,  dv_name = DV_NAME, fix = TRUE),
      list(effect = "outAct",  parameter = 0.1,  dv_name = DV_NAME, fix = TRUE)
    ),
    coCovars = list(), varCovars = list(),
    coDyadCovars = list(
      list(
        effect = "XWX", parameter = 1.0, dv_name = DV_NAME, fix = TRUE,
        nodeSet = c("COMPONENTS", "COMPONENTS"),
        interaction1 = "self$component_1_coDyadCovar",
        x = epistasis_nk
      )
    ),
    varDyadCovars = list()
  )
)

env_endog$search_rsiena(sm_endog, iterations_per_actor = 30, run_seed = 12345)
env_endog$plot_degree_4panel(loess_span = 0.3)
```

With endogenous effects active, the K-4 panel shows coupled dynamics: density
pressure constrains $K_{AC}$, preferential attachment generates heterogeneous
$K_{CA}$, and the emergent $K_{CC}$ is no longer a free parameter but a
structural consequence of actors' affiliation choices.

### Parameter Mapping Table

| NK Parameter | SaoMNK Mapping | Effect |
|--------------|----------------|--------|
| $N$ (components) | `N` | Number of binary components |
| $K$ (epistasis) | XWX matrix structure | Block-diagonal $\to$ modular; full $\to$ $K = N-1$ |
| Single agent | `M = 2` (SaoMNK minimum) | Single active searcher |
| Fixed landscape | $\theta = \mathbf{0}$ | No endogenous network effects |
| Greedy search | $\beta \to \infty$ | Deterministic best-response |
| Ruggedness | `density` magnitude | More negative $\to$ costlier exploration |

## Simulation Walkthrough

This appendix demonstrates the full simulation workflow: heterogeneous actor
strategies, exogenous shocks, and multi-wave execution.

### Actor Strategies with Heterogeneous Covariates

Actors can be assigned to different strategy groups using `coCovars`, where each
group has different weights on the network statistics:

```{r sim-strategies}
env_strat <- SaomNkRSienaBiEnv$new(list(
  M = 6, N = 8, BI_PROB = 0, rand_seed = 1234, name = "_strategy_demo_"
))

## Two strategy dimensions assigned cyclically
strategies <- list(
  egoX   = c(-1, 0, 1),
  inPopX = c( 1, 0, -1)
)
actor_strats <- lapply(strategies, function(s) rep(s, length.out = 6))

epistasis_strat <- create_block_diag(8, round(8 / 3))

sm_strat <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -0.3, dv_name = DV_NAME, fix = TRUE),
      list(effect = "inPop",   parameter =  0.2, dv_name = DV_NAME, fix = TRUE),
      list(effect = "outAct",  parameter =  0.1, dv_name = DV_NAME, fix = TRUE)
    ),
    coCovars = list(
      list(effect = "egoX", parameter = 0.2, dv_name = DV_NAME, fix = TRUE,
           interaction1 = "self$strat_1_coCovar",
           x = actor_strats[[1]]),
      list(effect = "inPopX", parameter = 0.2, dv_name = DV_NAME, fix = TRUE,
           interaction1 = "self$strat_2_coCovar",
           x = actor_strats[[2]])
    ),
    varCovars = list(),
    coDyadCovars = list(
      list(effect = "XWX", parameter = 0.02, dv_name = DV_NAME, fix = TRUE,
           nodeSet = c("COMPONENTS", "COMPONENTS"),
           interaction1 = "self$component_1_coDyadCovar",
           x = epistasis_strat)
    ),
    varDyadCovars = list(), interactions = list()
  )
)

env_strat$search_rsiena(sm_strat, iterations = env_strat$M * 30, run_seed = 12345)
```

```{r sim-strategies-k4, fig.height=10, fig.width=10}
env_strat$plot_degree_4panel(loess_span = 0.25)
```

### Exogenous Shocks via `theta_shocks`

The `theta_shocks` mechanism models exogenous events (regulatory changes,
technology disruptions) by modifying SAOM parameters at a designated step:

```{r sim-shocks}
env_base <- SaomNkRSienaBiEnv$new(list(
  M = 6, N = 8, BI_PROB = 0, rand_seed = 42, name = "_baseline_"
))

sm_base <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -0.5, dv_name = DV_NAME, fix = TRUE)
    ),
    coDyadCovars = list(
      list(effect = "XWX", parameter = 0.5, dv_name = DV_NAME, fix = TRUE,
           nodeSet = c("COMPONENTS", "COMPONENTS"),
           interaction1 = "self$component_1_coDyadCovar",
           x = create_block_diag(8, 2))
    )
  )
)

## Baseline (no shock)
env_base$search_rsiena(sm_base, iterations_per_actor = 30, run_seed = 12345)
```

```{r sim-shocks-plot, fig.height=10, fig.width=10}
env_base$plot_degree_4panel(loess_span = 0.3)
```

```{r sim-shock-apply}
## Shocked simulation: density drops from -0.5 to -2.0 at midpoint
env_shocked <- SaomNkRSienaBiEnv$new(list(
  M = 6, N = 8, BI_PROB = 0, rand_seed = 42, name = "_shocked_"
))

## Two segments of equal length: density -0.5, then -2.0. `effect_level`
## names a column of the theta grid; `portion` sets each segment's length.
theta_shocks <- list(
  list(effect_level = "density", parameter = -0.5, portion = 1),
  list(effect_level = "density", parameter = -2.0, portion = 1)
)

env_shocked$search_rsiena(
  sm_base,
  iterations_per_actor = 30,
  run_seed = 12345,
  theta_shocks = theta_shocks
)
```

```{r sim-shock-k4, fig.height=10, fig.width=10}
env_shocked$plot_degree_4panel(loess_span = 0.3)
```

The structural break in the K-4 trajectories at the shock step demonstrates how
exogenous parameter changes propagate through the coupled search system.

# Appendix B: Formal Mathematical Framework

This appendix states the definitions.

## Definition 1: NK Landscape

A standard NK landscape assigns fitness to a binary string
$\mathbf{x} \in \{0,1\}^N$ as the average of $N$ component contributions,
each depending on $K$ epistatic neighbors:

$$W(\mathbf{x}) = \frac{1}{N} \sum_{d=1}^{N} f_d(\mathbf{x}_{N(d)})$$

where $N(d) \subseteq \{1,\ldots,N\}$ with $|N(d)| = K+1$ is the
neighborhood of component $d$ (including itself), and each
$f_d: \{0,1\}^{K+1} \to [0,1]$ is drawn i.i.d. from $\text{Uniform}(0,1)$
(Kauffman, 1993).

## Definition 2: SaoMNK Utility

For actor $i$ with affiliation vector $\mathbf{b}_i \in \{0,1\}^N$
in the context of all other actors' affiliations $\mathbf{B}_{-i}$ and
influence matrix $E \in \{0,1\}^{N \times N}$:

$$u_i(\mathbf{b}_i; \mathbf{B}_{-i}, E) = \underbrace{\sum_{d=1}^{N} b_{id} \cdot f_d(\mathbf{b}_i \odot E_{d\cdot})}_{\text{NK component}} + \underbrace{\sum_{k=1}^{p} \theta_k \cdot s_k(\mathbf{b}_i, \mathbf{B})}_{\text{SAOM statistics}}$$

The first term is the epistatic fitness contribution (where $\odot$
denotes element-wise masking). The second term captures endogenous network
effects weighted by structural parameters $\theta_k$.

## Definition 3: Choice Probability (McFadden Conditional Logit)

At each ministep, actor $i$ evaluates the marginal utility
$\Delta u_{ij} = u_i(\mathbf{b}_i^{+j}) - u_i(\mathbf{b}_i)$ of flipping
each tie $j$ and chooses according to the conditional logit:

$$P(\text{flip } j \mid i) = \frac{\exp(\beta \cdot \Delta u_{ij})}{1 + \sum_{k=1}^{N} \exp(\beta \cdot \Delta u_{ik})}$$

where $\beta > 0$ is the rationality parameter. The $1$ in the denominator
represents the option of making no change.

## The $\{K\}$ Interpretive Framework

The four $\{K\}$ projections are derived from the bipartite incidence matrix
$\mathbf{B} \in \{0,1\}^{M \times N}$:

| Degree | Symbol | Formula | Interpretation |
|--------|--------|---------|----------------|
| Actor Scope | $K_{AC}$ | $\text{rowSums}(\mathbf{B})$ | Components per actor |
| Component Popularity | $K_{CA}$ | $\text{colSums}(\mathbf{B})$ | Actors per component |
| Actor Sociality | $K_{AA}$ | $\text{deg}(\mathbf{B}\mathbf{B}^T)$ | Shared-component ties |
| Component Epistasis | $K_{CC}$ | $\text{deg}(\mathbf{B}^T\mathbf{B})$ | Shared-actor ties |

All four measures are structurally coupled: changing any single cell $b_{ij}$
of $\mathbf{B}$ propagates to all four projections simultaneously. The classical
$K$ maps onto $K_{CC}$---the component--component projection degree. Under
$M=1$, the four-dimensional system collapses to the scalar $K$.


## SAOM Statistics Table

| Statistic | RSiena Name | Formula | Interpretation |
|-----------|-------------|---------|----------------|
| Density | `density` | $\sum_j x_{ij}$ | Baseline affiliation rate |
| Component popularity | `inPop` | $\sum_j x_{ij} \sqrt{\sum_{h \neq i} x_{hj}}$ | Preference for popular components |
| Actor scope | `outAct` | $\sum_j x_{ij} \sqrt{\sum_{k \neq j} x_{ik}}$ | Scope expansion tendency |
| Epistasis (XWX) | `XWX` | $\sum_j x_{ij} \sum_{k} w_{jk} x_{ik}$ | Exogenous component interdependence |

# Appendix C: Properties Checked by the Package {.tabset}

This appendix states the five properties, each with its test file and Lean
declaration, and the proof registry.

<!-- RECAST: author to write a short note that these are documented properties of the implemented model (checked by tests and the package Lean library in inst/lean, not presented as contributions), mirroring the opening of manuscript Section 6; drafts in paper/_drafts/2026-10-07_recast_verification.md -->

## Property 1: Reduction (NK as Special Case)

**Statement.** *Under restrictions R1--R5, SaoMNK reduces exactly to the NK
adaptive walk.*

**Restrictions:**

- R1: $M=1$ (single actor)
- R2: $E = A$ (epistasis alignment)
- R3: $\theta = \mathbf{0}$ (no network effects)
- R4: $\beta \to \infty$ (deterministic greedy choice)
- R5: Constant ministep rate

**Test:** `tests/testthat/test-nk-equivalence.R`, `tests/testthat/test-nk-classic.R`.

**Lean:** `SaomNK.utility_eq_nkFitness`, `SaomNK.logitProb_tendsto_one`, `SaomNK.stayProb_tendsto_one`.

## Property 2: Generalization (Three Extensions)

**Statement.** *SaoMNK strictly extends NK along three dimensions:*

*(a) Multi-actor interaction ($M > 1$).* Multiple actors search simultaneously on
the same landscape. With no cross-actor effect ($\theta_{inPop} = 0$) they do not
affect one another; externalities enter through (b).

*(b) Endogenous landscape co-evolution ($\theta \neq \mathbf{0}$).* Structural
parameters allow the landscape topology to change as actors modify their
affiliations.

*(c) Bounded rationality ($\beta < \infty$).* Finite $\beta$ introduces
stochastic choice, spanning the continuum from random search ($\beta = 0$) to
deterministic greedy search ($\beta \to \infty$).

**Test:** `tests/testthat/test-property-2.R`.

**Lean:** `SaomNK.card_config_gt`, `SaomNK.CoreSpec.landscape_endogeneity`, `SaomNK.logitProb_pos`.

## Property 3: Approximation (Mixed Logit Universality)

**Statement.** *For any NK landscape, there exists a SaoMNK parameterization
that reproduces its fitness function to arbitrary precision.*

**Test:** `tests/testthat/test-property-3.R`.

**Lean:** `SaomNK.nkTermFull_in_dummy_span`.

## Property 4: SAOM--QRE Correspondence

**Statement.** *The stationary distribution of the SaoMNK ministep process is
unique; under single-flip logit revision it is a Quantal Response Equilibrium (QRE).*

**Properties of the equilibrium:**

- **Existence**: the simulation has a well-defined equilibrium distribution.
- **Uniqueness**: there is only one stationary distribution.
- **Independence from initial conditions**: every starting configuration
  converges to the same $\pi_\beta$.
- **Limiting behavior**: under single-flip revision, as $\beta \to \infty$, $\pi_\beta$ concentrates on
  Nash equilibria; as $\beta \to 0$, $\pi_\beta$ approaches the uniform
  distribution.

**Test:** `tests/testthat/test-searchnet-ergodicity.R`, `tests/testthat/test-potential-cycles.R`, `tests/testthat/test-lean.R`.

**Lean:** `SaomNK.Spec.exact_potential`, `SaomNK.detailed_balance_of_potential`, `SaomNK.global_balance_of_potential`, `SaomNK.reachable`.

## Property 5: Brock--Durlauf Recovery

**Property 5 (Brock--Durlauf Recovery).** *The Brock & Durlauf (2001) binary
discrete-choice-with-social-interactions model is a degenerate special case of
SaoMNK. Specifically, under restrictions Rb1--Rb5, the SaoMNK Gibbs stationary
distribution reduces (componentwise) to the rational-expectations
self-consistent equilibrium of B&D2001.*

The five restrictions (verbatim from PROOF_TABLE.md L3--L7) are:

- **Rb1 (many-actor mean-field limit).** $M \to \infty$ with $N$ held fixed;
  rows of $\mathbf{B}$ indexed by an exchangeable population. *Many firms
  compete on a small fixed set of components.*
- **Rb2 (component independence).** Either $N=1$, or $K=0$ in the NK landscape
  *and* all cross-component coupling parameters in the 10-component utility
  ($\beta_F$, $\beta_w$, $\beta_h$, $\beta_{\text{disp}}$, $\beta_{\text{comp}}$,
  $\beta_{\text{clos}}$, $\beta_{\text{riv}}$, $\beta_{\text{leg}}$) vanish.
  *Components must not couple to each other; only density, scope cost, and
  congestion remain active.*
- **Rb3 (mean-field scaling of social interaction).** The congestion
  coefficient is rescaled as $\beta_{\text{cong}} = J_b \cdot M/(M-1)$, the
  standard Curie--Weiss convention.
- **Rb4 (spin-recoding tracking).** The adoption variable $b \in \{0,1\}$ is
  recoded to the spin variable $\sigma = 2b - 1 \in \{-1,+1\}$, with the
  factor-of-4 correction $J_\sigma = J_b/4$, $h_\sigma = h_b/2 + J_b/4$.
- **Rb5 (finite logit precision).** $\beta \in (0, \infty)$ strictly interior;
  neither greedy ($\beta \to \infty$) nor random ($\beta = 0$) limits.

**Test:** `tests/testthat/test-mean-field.R`.

**Lean:** `SaomNK.MeanField.exists_nonzero_root_iff`.

## Proof Registry

The complete proof chain is maintained in `inst/proofs/PROOF_TABLE.md`. The
table below summarizes the logical structure and verification status.

### Logical Structure

The proof chain is organized into 14 parts (104 steps total):

| Part | Content | Steps | Key Results |
|------|---------|-------|-------------|
| A | Foundations (Axioms) | A1--A11 | Binary state space, NK landscape, adaptive walk |
| B | SaoMNK Model | B1--B7 | Utility, statistics, CTMC dynamics, choice set |
| C | Property 1: Reduction | C1--C10 | NK is special case under R1--R5 |
| D | Property 2: Generalization | D1--D6 | Three independent extensions beyond NK |
| E | Property 3: Approximation | E1--E12 | Any NK landscape reproducible in SaoMNK |
| F | Property 4: SAOM-QRE | F1--F9 | Stationary distribution = QRE |
| G | $\{K\}$ Framework | G1--G6 | Four coupled degree measures |
| H | Extended Utility | H1--H11 | 10-component mechanism catalog |
| I | Policy/Shock Formalization | I1 | theta_shocks, search corridors |
| J | SCP Connection | J1--J2 | Structure-Conduct-Performance as coupled CTMC |
| K | Existence and Ising | K1--K2 | Nash existence, stat. mech. correspondence |
| L | Property 5: Brock--Durlauf Recovery | L1--L18 | B&D (2001) discrete choice with social interactions as degenerate special case in $M \to \infty$ mean-field limit |
| M | Two-Layer Integrability (Property 6) | M1--M6 | Two-layer process, cycle affinity, Property 6 |
| N | Creation/Endowment Boundary (Property 7) | N1--N3 | Two-cycle and four-cycle sums, Property 7 |

### Dependency Graph

```
AXIOMS (A1-A11)
    |
SAOM-NK MODEL (B1-B7)
    |
    +---> PROPERTY 1: REDUCTION (C1-C10)
    |        Lemma 1: Payoff Equivalence
    |        Lemma 2: Dynamic Equivalence
    |
    +---> PROPERTY 2: GENERALIZATION (D1-D6)
    |        Lemma 3: Landscape Endogeneity
    |        Lemma 4: Search Precision Spectrum
    |
    +---> PROPERTY 3: APPROXIMATION (E1-E12)
    |        Prop 1: Dummy Covariates (exact, exponential)
    |        Prop 2: Actor Heterogeneity (exact, polynomial)
    |        Prop 3: Mixed Logit (approximate, universal)
    |
    +---> PROPERTY 4: SAOM-QRE (F1-F8)
    |        Blume's Theorem: Logit-Gibbs convergence
    |        QRE identification
    |
    +---> {K} FRAMEWORK (G1-G6)
    +---> EXTENDED UTILITY (H1-H11)
    +---> POLICY/SHOCK (I1-I6)
    +---> SCP CONNECTION (J1-J2)
    +---> EXISTENCE & ISING (K1-K2)
```

### Cross-Reference Index

| Result | Step(s) | Source | Verification |
|--------|---------|--------|-------------|
| NK Landscape (Def 1) | A4 | EqProof, Def 1 | Standard |
| NK Adaptive Walk (Def 2) | A11 | EqProof, Def 2 | Standard |
| SaoMNK Model (Def 3) | B1 | EqProof, Def 3 | Implemented |
| SaoMNK Utility (Def 4) | B2 | EqProof, Def 4 | Implemented |
| SaoMNK Dynamics (Def 5) | B5 | EqProof, Def 5 | Implemented |
| Lemma 1: Payoff Equivalence | C6 | EqProof, Lemma 1 | Computational |
| Lemma 2: Dynamic Equivalence | C8 | EqProof, Lemma 2 | Computational |
| **Property 1: Reduction** | C9 | EqProof, Property 1 | Algebraic + Computational |
| Lemma 3: Landscape Endogeneity | D1 | EqProof, Lemma 3 | Constructive |
| Lemma 4: Search Precision | D2 | EqProof, Lemma 4 | Algebraic |
| **Property 2: Generalization** | D6 | EqProof, Property 2 | Three examples |
| Prop 1: Dummy Covariates | E2 | EqProof, Prop 1 | Constructive |
| Prop 2: Actor Heterogeneity | E5 | EqProof, Prop 2 | Parameter counting |
| Prop 3: Mixed Logit | E8 | EqProof, Prop 3 | McFadden-Train (2000) |
| **Property 3: Approximation** | E1 | EqProof, Property 3 | Three strategies |
| Blume's Theorem | F5 | Blume (1993) | External |
| SAOM is Potential Game | F3 | Snijders (2001) | Structural |
| **Property 4: SAOM-QRE** | F7 | Theory, Property 4 | Chain of identification |
| $\{K\}$ Structural Coupling | G5 | Theory | Algebraic |
| $\{K\}$ degenerates to $K$ | G6 | Theory | Algebraic |

# Appendix D: Causal Inference Pipeline

The `theta_shocks` mechanism generates panel data suitable for quasi-experimental
analysis. The `searchnet_causal_panel()` function extracts actor-level outcome
trajectories indexed by `(actor_id, chain_step_id)`, mapping directly to three
causal inference designs.

| Design | Package | When to Use |
|--------|---------|-------------|
| Difference-in-Differences | `did` | Multiple treated and control actors |
| Synthetic Control | `Synth` | Single treated actor; data-driven counterfactual |
| Regression Discontinuity | `rdrobust` | Sharp change at the shock step |

## Setting Up a Shocked Simulation

```{r causal-setup, eval=knitr::opts_chunk$get("eval") && requireNamespace("ggplot2", quietly=TRUE)}
env_ci <- SaomNkRSienaBiEnv$new(list(
  M = 6, N = 8, BI_PROB = 0, rand_seed = 42, name = "_causal_"
))

K_ci <- create_block_diag(8, 2)

sm_ci <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -0.5, dv_name = DV_NAME, fix = TRUE)
    ),
    coDyadCovars = list(
      list(effect = "XWX", parameter = 0.5, dv_name = DV_NAME, fix = TRUE,
           nodeSet = c("COMPONENTS", "COMPONENTS"),
           interaction1 = "self$component_1_coDyadCovar",
           x = K_ci)
    )
  )
)

## Two segments of equal portion: the first half of the run keeps the
## baseline density (-0.5); the shock (-2.0) covers the second half only.
ci_shocks <- list(
  list(
    effect_level = c("density"),
    parameter = c(-0.5),
    portion = 1
  ),
  list(
    effect_level = c("density"),
    parameter = c(-2.0),
    portion = 1
  )
)

env_ci$search_rsiena(sm_ci, iterations_per_actor = 20, run_seed = 12345,
                     theta_shocks = ci_shocks)

## The shock step is the first ministep simulated under the shocked value.
ci_shock_step <- min(env_ci$theta_shocks[[2]]$chain_step_ids)
```

## DID Example

```{r causal-did, eval=FALSE}
## Requires the 'did' package (Callaway & Sant'Anna, 2021)
## Every actor in env_ci faces the shock, so the controls are the six actors
## of an unshocked comparison run with the same start and run seed; att_gt()
## needs at least five never-treated units.
env_ci_ctrl <- SaomNkRSienaBiEnv$new(list(
  M = 6, N = 8, BI_PROB = 0, rand_seed = 42, name = "_causal_ctrl_"
))
env_ci_ctrl$search_rsiena(sm_ci, iterations_per_actor = 20, run_seed = 12345)

panel_tr <- searchnet_causal_panel(env_ci, shock_step = ci_shock_step,
                                   outcome = "utility")
panel_co <- searchnet_causal_panel(env_ci_ctrl, shock_step = ci_shock_step,
                                   outcome = "utility",
                                   treated_actors = integer(0))
panel_co$actor_id <- factor(as.integer(as.character(panel_co$actor_id)) + 6L)
common <- seq_len(min(max(panel_tr$step), max(panel_co$step)))
panel_did <- rbind(panel_tr[panel_tr$step %in% common, ],
                   panel_co[panel_co$step %in% common, ])

att <- searchnet_did(panel_did)
did::aggte(att, type = "simple")
searchnet_causal_plot(att, type = "did")
```

## Synthetic Control Example

```{r causal-synth, eval=FALSE}
## Requires the 'Synth' package
panel_sc <- searchnet_causal_panel(
  env_ci, shock_step = ci_shock_step,
  outcome = "utility", treated_actors = c(1)
)

sc_result <- searchnet_synth(panel_sc, treated_unit = 1)
p <- searchnet_causal_plot(sc_result, type = "synth")
print(p)
```

## Regression Discontinuity Example

```{r causal-rd, eval=FALSE}
## Requires the 'rdrobust' package
panel_rd <- searchnet_causal_panel(
  env_ci, shock_step = ci_shock_step,
  outcome = "utility"
)

rd_result <- searchnet_rd(panel_rd)
summary(rd_result$rd)
p <- searchnet_causal_plot(rd_result, type = "rd")
print(p)
```

All three methods also work with K-4 outcomes (`K_AC`, `K_CA`, `K_AA`, `K_CC`),
enabling researchers to analyze the causal effect of shocks on specific dimensions
of the coupled degree system.


# References {-}

- Blume, L. E. (1993). The statistical mechanics of strategic interaction.
  *Games and Economic Behavior*, 5(3), 387--424.
- Brock, W. A., & Durlauf, S. N. (2001). Discrete choice with social
  interactions. *Review of Economic Studies*, 68(2), 235--260.
- Durlauf, S. N. (1999). How can statistical mechanics contribute to social
  science? *Proceedings of the National Academy of Sciences*, 96(19),
  10582--10584.
- Ellis, R. S. (1985). *Entropy, Large Deviations, and Statistical Mechanics*.
  Springer.
- Kauffman, S. A. (1993). *The Origins of Order*. Oxford University Press.
- Krapivsky, P. L., Redner, S., & Ben-Naim, E. (2010). *A Kinetic View of
  Statistical Physics*. Cambridge University Press.
- Kauffman, S. A., & Levin, S. (1987). Towards a general theory of adaptive
  walks on rugged landscapes. *Journal of Theoretical Biology*, 128(1), 11--45.
- Levinthal, D. A. (1997). Adaptation on rugged landscapes.
  *Management Science*, 43(7), 934--950.
- Manski, C. F. (1993). Identification of endogenous social effects: The
  reflection problem. *Review of Economic Studies*, 60(3), 531--542.
- McFadden, D. (1974). Conditional logit analysis of qualitative choice
  behavior. In P. Zarembka (Ed.), *Frontiers in Econometrics* (pp. 105--142).
- McFadden, D., & Train, K. (2000). Mixed MNL models for discrete response.
  *Journal of Applied Econometrics*, 15(5), 447--470.
- McKelvey, R. D., & Palfrey, T. R. (1995). Quantal response equilibria for
  normal form games. *Games and Economic Behavior*, 10(1), 6--38.
- Monderer, D., & Shapley, L. S. (1996). Potential games. *Games and Economic
  Behavior*, 14(1), 124--143.
- Snijders, T. A. B. (2001). The statistical evaluation of social network
  dynamics. *Sociological Methodology*, 31(1), 361--395.


```{r session-info}
sessionInfo()
```
