---
title: "How SaoMNK Generalizes the NK Model: A Tutorial"
subtitle: "Online Appendix — Proof Walkthrough with Worked Examples"
author: "Stephen Downing"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 3
    number_sections: true
    code_folding: hide
    fig_width: 9
    fig_height: 6
    css: !expr NULL
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", cache = TRUE
)
library(knitr)
library(kableExtra)
```

# What This Document Proves {.unnumbered}

This tutorial walks through three formal results connecting the **SaoMNK** model
(Stochastic Actor-Oriented Model on NK Landscapes) to the **standard NK** model
from complexity theory (Kauffman, 1993; Levinthal, 1997).

| Result | Plain-English Statement |
|--------|------------------------|
| **Theorem 1 (Reduction)** | Standard NK is a *special case* of SaoMNK. Set five switches to their extreme positions, and SaoMNK collapses exactly to NK. |
| **Theorem 2 (Generalization)** | SaoMNK can do things NK *cannot*: multiple actors, endogenous landscape co-evolution, and bounded rationality. |
| **Theorem 3 (Approximation)** | For any NK landscape you can dream up, there exists an SaoMNK parameterization that reproduces it. |

The takeaway: **every NK result you already know is automatically a special case of SaoMNK**, and SaoMNK opens doors that NK cannot.

---

# Preliminaries: The Two Models Side by Side

Before diving into proofs, let's build intuition by comparing the two models on paper, then in code.

## The Standard NK Model

An NK landscape is defined by two integers:

- **N** = number of binary components (think: decisions a firm makes)
- **K** = number of *other* components that affect each component's payoff (epistasis)

```{r nk-example-landscape}
# A tiny NK landscape: N=4, K=1
# Each component depends on itself + 1 other component
N <- 4; K <- 1

# Interaction matrix: component d depends on columns where E[d,] == 1
set.seed(7)
E <- diag(N)
for (d in 1:N) {
  others <- sample(setdiff(1:N, d), K)
  E[d, others] <- 1
}

cat("Epistasis matrix E (rows = components, cols = which bits matter):\n")
E
```

Each component $d$ has a payoff table with $2^{K+1}$ entries --- one for every
possible configuration of its $K+1$ relevant bits:

```{r nk-payoff-tables}
# Random payoff tables (the "landscape")
payoffs <- lapply(1:N, function(d) round(runif(2^(K+1)), 3))

# Show the payoff table for component 1
cat("Component 1 depends on bits:", which(E[1,] == 1), "\n")
cat("Payoff table for component 1 (indexed by the 2-bit configuration):\n")

epistatic_bits_1 <- which(E[1,] == 1)
config_labels <- expand.grid(replicate(length(epistatic_bits_1), 0:1, simplify = FALSE))
names(config_labels) <- paste0("Bit_", epistatic_bits_1)
config_labels$Payoff <- payoffs[[1]]

kable(config_labels, caption = "Component 1's payoff table") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

**Total fitness** for a configuration $\mathbf{x} \in \{0,1\}^N$ is the average
of all $N$ component payoffs:

$$W(\mathbf{x}) = \frac{1}{N} \sum_{d=1}^{N} f_d\!\bigl(\mathbf{x}_{N(d)}\bigr)$$

```{r nk-fitness-function}
# NK fitness function
nk_fitness <- function(x, payoffs, E) {
  N <- length(x)
  fit <- numeric(N)
  for (d in 1:N) {
    epistatic_bits <- x[which(E[d, ] == 1)]
    idx <- sum(epistatic_bits * 2^((length(epistatic_bits) - 1):0)) + 1
    fit[d] <- payoffs[[d]][idx]
  }
  mean(fit)
}

# Example: fitness of x = (1, 0, 1, 0)
x_example <- c(1, 0, 1, 0)
cat("Fitness of x =", x_example, "is:", nk_fitness(x_example, payoffs, E), "\n")
```

## The SaoMNK Model

SaoMNK uses the **same** random landscape but stores it differently: a single
$2^N \times N$ matrix instead of $N$ separate tables.

```{r saomnk-landscape}
# SaoMNK stores the ENTIRE landscape as one matrix
# Rows correspond to all possible configurations (indexed by Power_key)
# Columns correspond to components

powerkey <- function(N) 2^((N - 1):0)
PK <- powerkey(N)
cat("Power key for N=4:", PK, "\n")
cat("  x = (1,0,1,0) maps to row:", sum(c(1,0,1,0) * PK) + 1, "\n")
cat("  x = (0,0,0,0) maps to row:", sum(c(0,0,0,0) * PK) + 1, "\n")
cat("  x = (1,1,1,1) maps to row:", sum(c(1,1,1,1) * PK) + 1, "\n")
```

The key operation is **epistasis masking**: for component $d$, SaoMNK zeros
out irrelevant bits before computing the row index:

$$r_d(\mathbf{x}) = \sum_{j=1}^{N} x_j \cdot E_{dj} \cdot \text{PK}_j + 1$$

Because $E_{dj} = 0$ for non-epistatic bits, those bits contribute nothing to
the index --- exactly like the standard NK lookup.

```{r saomnk-calc-fit}
# SaoMNK calc_fit: the ACTUAL function from the R6 class
saomnk_calc_fit <- function(N, NK_land, inter_m, x, Power_key) {
  Fit_vector <- numeric(N)
  for (d in 1:N) {
    Fit_vector[d] <- NK_land[sum(x * inter_m[d, ] * Power_key) + 1, d]
  }
  Fit_vector
}
```

### SaoMNK adds three things NK lacks

| Feature | NK | SaoMNK |
|---------|:--:|:------:|
| Multiple actors (M firms on same landscape) | 1 | 1, 2, 5, 10, ... |
| Cross-actor effects (crowding, structural coupling) | None | inPop, cycle4, ... |
| Bounded rationality (noisy search) | Greedy only | $\beta \in [0, \infty]$ |
| Estimable from data | No | Yes (RSiena MLE) |

---

# Theorem 1: NK is a Special Case of SaoMNK

## The Five Restrictions

To turn SaoMNK into plain NK, set five switches:

```{r restrictions-table, echo=FALSE}
restrictions <- data.frame(
  Switch = c("R1", "R2", "R3", "R4", "R5"),
  Setting = c("M = 1", "E = A (full epistasis matrix)",
              "theta = 0 (no network effects)",
              "beta -> infinity (greedy search)",
              "Constant rate function"),
  `What.It.Eliminates` = c(
    "Multiple actors",
    "Bipartite projection differences",
    "Crowding (inPop), structural coupling (cycle4)",
    "Stochastic choice (bounded rationality)",
    "Heterogeneous selection rates"
  )
)
kable(restrictions, col.names = c("Switch", "Setting", "What It Eliminates"),
      caption = "The five restrictions that reduce SaoMNK to NK") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Worked Example: Payoff Equivalence (Lemma 1)

Let's verify this numerically. We build both an NK landscape and its SaoMNK
equivalent, then check that every possible configuration gets the same fitness.

```{r payoff-equivalence}
set.seed(42)
N <- 6; K <- 2

# Step 1: Generate shared interaction matrix
E <- diag(N)
for (d in 1:N) {
  if (K > 0) {
    others <- sample(setdiff(1:N, d), K)
    E[d, others] <- 1
  }
}

# Step 2: Generate standard NK payoff tables
payoffs_std <- lapply(1:N, function(d) runif(2^(K + 1)))

# Step 3: Build equivalent SaoMNK landscape
PK <- powerkey(N)
NK_land <- matrix(0, nrow = 2^N, ncol = N)

all_configs <- expand.grid(replicate(N, 0:1, simplify = FALSE))
for (d in 1:N) {
  epistatic_indices <- which(E[d, ] == 1)
  for (c_idx in 1:nrow(all_configs)) {
    x <- as.integer(all_configs[c_idx, ])
    # SaoMNK row index
    saomnk_row <- sum(x * E[d, ] * PK) + 1
    # Standard NK index
    epistatic_bits <- x[epistatic_indices]
    nk_idx <- sum(epistatic_bits * 2^((length(epistatic_bits) - 1):0)) + 1
    NK_land[saomnk_row, d] <- payoffs_std[[d]][nk_idx]
  }
}

# Step 4: Compare fitness for EVERY configuration
max_diff <- 0
n_configs_checked <- 0
for (c_idx in 1:nrow(all_configs)) {
  x <- as.integer(all_configs[c_idx, ])

  # NK fitness (standard)
  f_nk <- nk_fitness(x, payoffs_std, E)

  # SaoMNK fitness (M=1)
  f_saomnk <- mean(saomnk_calc_fit(N, NK_land, E, x, PK))

  max_diff <- max(max_diff, abs(f_nk - f_saomnk))
  n_configs_checked <- n_configs_checked + 1
}

cat(sprintf("Checked all %d configurations for N=%d, K=%d\n",
            n_configs_checked, N, K))
cat(sprintf("Maximum |f_NK - f_SaoMNK| = %.2e\n", max_diff))
cat(sprintf("Result: %s\n",
            ifelse(max_diff < 1e-12, "IDENTICAL (machine precision)", "MISMATCH")))
```

### How the index masking works (step by step)

Let's trace through one component to see exactly why the payoffs match:

```{r trace-masking}
# Pick component d=1 and a specific configuration
d <- 1
x <- c(1, 0, 1, 1, 0, 1)
cat("Configuration x =", x, "\n")
cat("Component d =", d, "\n")
cat("E[d,] =", E[d,], " (bits", which(E[d,] == 1), "are epistatic)\n\n")

# NK method: extract epistatic bits, compute local index
epistatic_bits <- x[which(E[d,] == 1)]
nk_idx <- sum(epistatic_bits * 2^((length(epistatic_bits) - 1):0)) + 1
cat("NK method:\n")
cat("  Epistatic bits =", epistatic_bits, "\n")
cat("  Local index =", nk_idx, "\n")
cat("  Payoff = payoffs[[1]][", nk_idx, "] =", payoffs_std[[1]][nk_idx], "\n\n")

# SaoMNK method: mask with E[d,], compute global index
masked <- x * E[d,]
saomnk_row <- sum(masked * PK) + 1
cat("SaoMNK method:\n")
cat("  x * E[d,] =", masked, " (non-epistatic bits zeroed out)\n")
cat("  Row index = sum(", paste(masked, "*", PK, sep = ""), ") + 1 =", saomnk_row, "\n")
cat("  Payoff = NK_land[", saomnk_row, ",", d, "] =", NK_land[saomnk_row, d], "\n\n")

cat("Same payoff?", abs(payoffs_std[[1]][nk_idx] - NK_land[saomnk_row, d]) < 1e-15, "\n")
```

**Insight:** The epistasis mask `E[d,]` zeros out irrelevant bits in the power-key
sum. Different configurations that agree on the epistatic bits of component $d$
map to the **same row** in `NK_land`. This is algebraically identical to looking
up a smaller payoff table indexed only by the epistatic bits.

## Worked Example: Dynamic Equivalence (Lemma 2)

Under $\beta \to \infty$, the logit choice rule becomes **greedy best-response** ---
identical to the NK adaptive walk:

$$P(\text{flip bit } d) = \frac{e^{\beta \cdot \Delta u_d}}{\sum_{d'} e^{\beta \cdot \Delta u_{d'}}} \xrightarrow{\beta \to \infty} \begin{cases} 1 & \text{if } d = \arg\max \Delta u \\ 0 & \text{otherwise} \end{cases}$$

```{r dynamic-equivalence}
set.seed(99)
N <- 8; K <- 3

# Build shared landscape
E <- diag(N)
for (d in 1:N) {
  others <- sample(setdiff(1:N, d), K)
  E[d, others] <- 1
}
payoffs_std <- lapply(1:N, function(d) runif(2^(K + 1)))

# Build SaoMNK equivalent
PK <- powerkey(N)
NK_land <- matrix(0, nrow = 2^N, ncol = N)
all_configs <- expand.grid(replicate(N, 0:1, simplify = FALSE))
for (d in 1:N) {
  ei <- which(E[d,] == 1)
  for (c_idx in 1:nrow(all_configs)) {
    x <- as.integer(all_configs[c_idx, ])
    saomnk_row <- sum(x * E[d,] * PK) + 1
    nk_idx <- sum(x[ei] * 2^((length(ei)-1):0)) + 1
    NK_land[saomnk_row, d] <- payoffs_std[[d]][nk_idx]
  }
}

# NK adaptive walk
nk_walk <- function(x, payoffs, E) {
  trajectory <- list(x)
  repeat {
    N <- length(x)
    current_fit <- nk_fitness(x, payoffs, E)
    best_flip <- NA; best_fit <- current_fit
    for (d in 1:N) {
      x_new <- x; x_new[d] <- 1 - x_new[d]
      f_new <- nk_fitness(x_new, payoffs, E)
      if (f_new > best_fit) { best_fit <- f_new; best_flip <- d }
    }
    if (is.na(best_flip)) break
    x[best_flip] <- 1 - x[best_flip]
    trajectory <- c(trajectory, list(x))
  }
  list(path = trajectory, fitness = sapply(trajectory, nk_fitness, payoffs, E))
}

# SaoMNK adaptive walk (beta -> infinity = greedy)
saomnk_walk <- function(x, NK_land, E) {
  N <- length(x); PK <- powerkey(N)
  trajectory <- list(x)
  repeat {
    current_fit <- mean(saomnk_calc_fit(N, NK_land, E, x, PK))
    best_flip <- NA; best_fit <- current_fit
    for (d in 1:N) {
      x_new <- x; x_new[d] <- 1 - x_new[d]
      f_new <- mean(saomnk_calc_fit(N, NK_land, E, x_new, PK))
      if (f_new > best_fit) { best_fit <- f_new; best_flip <- d }
    }
    if (is.na(best_flip)) break
    x[best_flip] <- 1 - x[best_flip]
    trajectory <- c(trajectory, list(x))
  }
  list(path = trajectory, fitness = sapply(trajectory, function(x)
    mean(saomnk_calc_fit(N, NK_land, E, x, PK))))
}

# Run both from the same start
start <- sample(0:1, N, replace = TRUE)
res_nk <- nk_walk(start, payoffs_std, E)
res_saomnk <- saomnk_walk(start, NK_land, E)

cat("Start:", start, "\n\n")
cat("NK walk:\n")
cat("  Steps:", length(res_nk$fitness) - 1, "\n")
cat("  Fitness trajectory:", round(res_nk$fitness, 4), "\n")
cat("  Final config:", res_nk$path[[length(res_nk$path)]], "\n\n")

cat("SaoMNK walk (beta -> inf):\n")
cat("  Steps:", length(res_saomnk$fitness) - 1, "\n")
cat("  Fitness trajectory:", round(res_saomnk$fitness, 4), "\n")
cat("  Final config:", res_saomnk$path[[length(res_saomnk$path)]], "\n\n")

cat("Trajectories identical?",
    identical(res_nk$path, res_saomnk$path), "\n")
```

## Visual: Many Walks, Perfect Agreement

```{r scatter-walks, fig.width=7, fig.height=6}
set.seed(123)
n_walks <- 200
nk_finals <- numeric(n_walks)
saomnk_finals <- numeric(n_walks)

for (w in 1:n_walks) {
  start <- sample(0:1, N, replace = TRUE)
  nk_finals[w] <- nk_walk(start, payoffs_std, E)$fitness |> tail(1)
  saomnk_finals[w] <- saomnk_walk(start, NK_land, E)$fitness |> tail(1)
}

plot(nk_finals, saomnk_finals,
     pch = 16, cex = 0.8, col = adjustcolor("steelblue", 0.6),
     xlab = "NK Final Fitness", ylab = "SaoMNK Final Fitness",
     main = "200 Adaptive Walks: NK vs SaoMNK (M=1)",
     sub = sprintf("Max difference: %.2e", max(abs(nk_finals - saomnk_finals))))
abline(0, 1, col = "red", lty = 2, lwd = 2)
legend("topleft", "Perfect agreement (y = x)", col = "red", lty = 2, lwd = 2, bty = "n")
```

Every point sits exactly on the 45-degree line. The two models produce identical
results. **Theorem 1 is verified computationally.**

---

# Theorem 2: SaoMNK Strictly Extends NK

SaoMNK extends NK along three **independent** dimensions. Each one introduces
dynamics that are provably impossible in standard NK.

## Dimension A: Multiple Actors ($M > 1$)

NK has one searcher on a landscape of $2^N$ configurations.
SaoMNK has $M$ searchers on a landscape of $2^{MN}$ configurations.

```{r state-space-table, echo=FALSE}
state_space <- data.frame(
  N = c(10, 10, 10, 20),
  M = c(1, 5, 10, 10),
  `NK.States` = c("1,024", "1,024", "1,024", "1,048,576"),
  `SaoMNK.States` = c("1,024", "~10^15", "~10^30", "~10^60")
)
kable(state_space, col.names = c("N", "M", "NK State Space", "SaoMNK State Space"),
      caption = "State space explosion with multiple actors") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

This isn't just "running NK in parallel." The actors **interact**.

## Dimension B: Endogenous Landscape Co-Evolution

In standard NK, the landscape is fixed. In SaoMNK, actor $j$ changing a component
alters the landscape that actor $i$ faces. This happens through **network statistics**:

- **inPop (crowding):** How many other actors hold the same component? If many
  firms adopt the same technology, its value changes.
- **cycle4 (structural coupling):** Are there 4-cycle patterns in the bipartite
  network? This captures indirect competitive interdependence.

```{r crowding-demo, fig.width=8, fig.height=5}
# Simple crowding demonstration: 3 actors, 5 components
# When actors overlap on components, their fitness is affected

N_demo <- 5
M_demo <- 3

# Three actors with partially overlapping component choices
actors <- matrix(c(
  1, 1, 0, 0, 1,   # Actor 1: components 1, 2, 5
  0, 1, 1, 0, 1,   # Actor 2: components 2, 3, 5  (overlaps on 2 & 5)
  0, 0, 0, 1, 1    # Actor 3: components 4, 5      (overlaps on 5)
), nrow = 3, byrow = TRUE)

# Overlap structure (projected search network)
overlap <- actors %*% t(actors)
diag(overlap) <- 0

cat("Bipartite matrix B (actors x components):\n")
print(actors)
cat("\nOverlap matrix (how many components each pair shares):\n")
print(overlap)

# inPop: for each component, count how many actors hold it
inpop <- colSums(actors)
cat("\ninPop (popularity of each component):", inpop, "\n")
cat("  Component 5 is held by all 3 actors -- high crowding\n")
cat("  Component 4 is held by only 1 actor -- no crowding\n\n")

# Show how Actor 2 flipping component 5 changes the landscape for Actors 1 & 3
cat("If Actor 2 DROPS component 5:\n")
actors_new <- actors
actors_new[2, 5] <- 0
inpop_new <- colSums(actors_new)
cat("  inPop changes from", inpop, "to", inpop_new, "\n")
cat("  Actors 1 & 3 now face LESS crowding on component 5\n")
cat("  Their utility for component 5 increases -- landscape shifted!\n")
cat("  This is impossible in standard NK (no actor interactions)\n")
```

## Dimension C: Bounded Rationality ($\beta < \infty$)

NK uses greedy search: always pick the best improving flip. SaoMNK uses a
**logit choice** where the precision parameter $\beta$ controls rationality:

```{r bounded-rationality, fig.width=8, fig.height=5}
# Demonstrate how beta affects choice probabilities
delta_u <- c(0.05, 0.02, -0.01, -0.03, 0.00)  # utility changes for 5 possible flips
component_labels <- paste("Flip bit", 1:5)

betas <- c(0.1, 1, 5, 20, 100, Inf)
prob_matrix <- matrix(NA, nrow = length(betas), ncol = length(delta_u))

for (b_idx in seq_along(betas)) {
  beta <- betas[b_idx]
  if (is.infinite(beta)) {
    probs <- rep(0, length(delta_u))
    probs[which.max(delta_u)] <- 1
  } else {
    log_probs <- beta * delta_u
    log_probs <- log_probs - max(log_probs)  # numerical stability
    probs <- exp(log_probs) / sum(exp(log_probs))
  }
  prob_matrix[b_idx, ] <- probs
}

rownames(prob_matrix) <- paste0("beta=", betas)
colnames(prob_matrix) <- paste0("Flip ", 1:5, " (du=", delta_u, ")")

cat("Choice probabilities under different rationality levels:\n\n")
cat("  Utility changes: du =", delta_u, "\n")
cat("  Best flip: Flip 1 (du = +0.05)\n\n")

kable(round(prob_matrix, 3),
      caption = "How beta controls search precision") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  column_spec(2, bold = TRUE)
```

| $\beta$ value | Behavior | NK equivalent? |
|:---:|---|:---:|
| $\beta = 0$ | Random search (all flips equally likely) | No |
| $\beta \in (0, \infty)$ | Noisy best-response (bounded rationality) | No |
| $\beta \to \infty$ | Greedy best-response (**this is NK**) | **Yes** |

NK lives at **one extreme** of a continuous spectrum. SaoMNK nests the whole spectrum.

At finite $\beta$, actors can make **fitness-decreasing** moves with positive
probability. This lets them escape local optima --- addressing the competency trap
that is central to the organizational learning literature.

---

# Theorem 3: Approximation

**Claim:** For *any* NK(N,K) landscape, there exists an SaoMNK parameterization
that reproduces it exactly (or to arbitrary precision).

This is important because it means SaoMNK doesn't just contain NK as a special
case --- it can *reconstruct* any specific NK landscape.

## Mechanism 1: Dummy Covariates (Exact Reconstruction)

The idea: create $2^{K+1}$ binary "dummy" covariates, one for each possible
configuration of the $K+1$ epistatic bits. Set each coefficient equal to the
NK payoff value.

```{r dummy-covariates}
# Example: N=4, K=1
# Component 1 depends on bits {1, 3} (itself + one other)
# The K+1 = 2 epistatic bits create 2^2 = 4 configurations
# We create 4 dummy covariates, one per configuration

N_ex <- 4; K_ex <- 1
epistatic_bits_ex <- c(1, 3)  # component 1 depends on bits 1 and 3
payoff_values <- c(0.23, 0.67, 0.45, 0.91)  # the NK payoff table

cat("NK payoff table for component 1:\n")
dummy_table <- data.frame(
  `Bit_1` = c(0, 1, 0, 1),
  `Bit_3` = c(0, 0, 1, 1),
  `NK_Payoff` = payoff_values,
  `Dummy_Covariate` = paste0("Z_", 1:4),
  `SaoMNK_Coefficient` = payoff_values
)
kable(dummy_table,
      col.names = c("Bit 1", "Bit 3", "NK Payoff", "Dummy Covariate", "SaoMNK Coeff."),
      caption = "Dummy covariate construction: one covariate per epistatic configuration") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

cat("\nFor x = (1, ?, 0, ?), only Z_2 is 'active' (bit1=1, bit3=0)\n")
cat("SaoMNK utility contribution = 1 * 0.67 = 0.67 = NK payoff. Exact match.\n")
```

**Cost:** This needs $N \times 2^{K+1}$ dummy covariates --- exponentially many.
But it proves *existence*: the SaoMNK model class contains every NK landscape.

## Mechanism 2: Actor Heterogeneity (Practical Sufficiency)

With $M$ actors and $Q$ continuous covariates, SaoMNK has $Q \times M \times N$
effective parameters. We need $Q \cdot M \geq 2^{K+1}$:

```{r actor-heterogeneity, echo=FALSE}
het_table <- data.frame(
  M = c(5, 10, 10, 20),
  K = c(2, 3, 5, 7),
  `Needed` = c("8", "16", "64", "256"),
  Q_needed = c(2, 2, 7, 13),
  Interpretation = c(
    "2 covariates (e.g., size + experience)",
    "2 covariates suffice for moderate epistasis",
    "7 covariates for high epistasis",
    "13 covariates (may need richer specification)"
  )
)
kable(het_table,
      col.names = c("M (actors)", "K (epistasis)", "2^(K+1) params needed",
                     "Q (covariates needed)", "Interpretation"),
      caption = "How many covariates Q does SaoMNK need to match NK dimensionality?") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

**Key insight:** Actor heterogeneity is not a trick --- it's a *natural, empirically
grounded* source of landscape complexity. Real firms differ in size, capabilities,
and strategic positioning. These differences generate the landscape richness
that NK achieves through random numbers.

## Mechanism 3: Universal Approximation (Strongest Guarantee)

McFadden & Train (2000) proved that **mixed logit** can approximate *any* random
utility model to arbitrary precision. Since:

1. SAOM is a conditional logit model
2. Adding random effects to SAOM gives mixed logit
3. NK payoffs define a discrete-choice utility

...it follows that mixed SaoMNK can approximate any NK landscape to any desired
precision. This is the strongest theoretical guarantee --- infinite-dimensional
and non-constructive.

---

# Putting It All Together: The Multi-Actor Difference

```{r multi-actor-demo, fig.width=10, fig.height=8}
set.seed(42)
N_ma <- 8; K_ma <- 2

# Build a shared landscape
E_ma <- diag(N_ma)
for (d in 1:N_ma) {
  others <- sample(setdiff(1:N_ma, d), K_ma)
  E_ma[d, others] <- 1
}
PK_ma <- powerkey(N_ma)
NK_land_ma <- matrix(runif(2^N_ma * N_ma), nrow = 2^N_ma, ncol = N_ma)

# Simulate M=1 vs M=5 with simple crowding
simulate_search <- function(M, n_steps = 30) {
  positions <- matrix(sample(0:1, M * N_ma, replace = TRUE), nrow = M)
  fitness_history <- matrix(NA, nrow = n_steps + 1, ncol = M)

  for (m in 1:M) {
    fitness_history[1, m] <- mean(saomnk_calc_fit(N_ma, NK_land_ma, E_ma,
                                                   positions[m,], PK_ma))
  }

  for (step in 1:n_steps) {
    for (m in sample(1:M)) {
      x <- positions[m,]
      current_fit <- mean(saomnk_calc_fit(N_ma, NK_land_ma, E_ma, x, PK_ma))

      # Add crowding penalty: reduce fitness for popular components
      if (M > 1) {
        overlap_penalty <- sum(colSums(positions[-m, , drop = FALSE]) * x) * 0.01
        current_fit <- current_fit - overlap_penalty
      }

      best_flip <- NA; best_fit <- current_fit
      for (d in 1:N_ma) {
        x_new <- x; x_new[d] <- 1 - x_new[d]
        f_new <- mean(saomnk_calc_fit(N_ma, NK_land_ma, E_ma, x_new, PK_ma))
        if (M > 1) {
          pen <- sum(colSums(positions[-m, , drop = FALSE]) * x_new) * 0.01
          f_new <- f_new - pen
        }
        if (f_new > best_fit) { best_fit <- f_new; best_flip <- d }
      }

      if (!is.na(best_flip)) positions[m, best_flip] <- 1 - positions[m, best_flip]
    }
    for (m in 1:M) {
      fitness_history[step + 1, m] <- mean(saomnk_calc_fit(N_ma, NK_land_ma, E_ma,
                                                            positions[m,], PK_ma))
    }
  }
  fitness_history
}

# Run comparisons
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))

# M = 1 (this IS NK)
traj_1 <- simulate_search(1)
plot(0:30, traj_1[,1], type = "l", lwd = 2, col = "steelblue",
     xlab = "Step", ylab = "Fitness", main = "M = 1 (Standard NK)",
     ylim = c(0.3, 0.8))

# M = 3
traj_3 <- simulate_search(3)
matplot(0:30, traj_3, type = "l", lwd = 2, lty = 1,
        col = c("steelblue", "coral", "seagreen"),
        xlab = "Step", ylab = "Fitness", main = "M = 3 (Competition)",
        ylim = c(0.3, 0.8))
legend("bottomright", paste("Actor", 1:3), col = c("steelblue", "coral", "seagreen"),
       lwd = 2, bty = "n", cex = 0.8)

# M = 5
traj_5 <- simulate_search(5)
matplot(0:30, traj_5, type = "l", lwd = 2, lty = 1,
        col = rainbow(5),
        xlab = "Step", ylab = "Fitness", main = "M = 5 (Crowded)",
        ylim = c(0.3, 0.8))

# Final fitness comparison
finals <- data.frame(
  M = factor(rep(c("M=1", "M=3", "M=5"), c(1, 3, 5))),
  Fitness = c(tail(traj_1, 1), tail(traj_3, 1), tail(traj_5, 1))
)
boxplot(Fitness ~ M, data = finals, col = c("steelblue", "coral", "seagreen"),
        main = "Final Fitness by # Actors", ylab = "Final Fitness")
```

As $M$ increases, competitive crowding pushes actors away from the global optimum.
Trajectories become more volatile and final fitness drops. **None of this is possible in standard NK.**

---

# Summary: The Three Theorems at a Glance

```{r summary-table, echo=FALSE}
summary <- data.frame(
  Theorem = c("1. Reduction", "2. Generalization", "3. Approximation"),
  Statement = c(
    "Set M=1, theta=0, beta=Inf, E=A, constant rate => SaoMNK = NK exactly",
    "M>1 gives multi-actor search; theta!=0 gives landscape co-evolution; finite beta gives bounded rationality",
    "Any NK(N,K) landscape can be reconstructed in SaoMNK via dummy covariates, actor heterogeneity, or mixed logit"
  ),
  Proof.Type = c(
    "Algebraic + computational verification",
    "Constructive examples + impossibility argument",
    "Existence (constructive) + universal approximation (McFadden-Train 2000)"
  ),
  Implication = c(
    "All existing NK results carry over as special cases",
    "SaoMNK can model phenomena NK cannot (competition, crowding, noisy search)",
    "SaoMNK is at least as expressive as NK in every dimension"
  )
)
kable(summary, col.names = c("Theorem", "Statement", "Proof Method", "Implication"),
      caption = "The three theorems establishing SaoMNK as a strict generalization of NK") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  column_spec(1, bold = TRUE, width = "12em") %>%
  column_spec(2, width = "20em") %>%
  column_spec(3, width = "18em") %>%
  column_spec(4, width = "18em")
```

---

# Why This Matters for Strategy Research

| NK Limitation | SaoMNK Solution |
|---|---|
| One searcher in isolation | $M$ firms competing on the same landscape |
| Fixed landscape | Landscape co-evolves as firms act |
| Only greedy (perfectly rational) search | Bounded rationality via $\beta$ parameter |
| Uninterpretable random payoff tables | Payoffs from observable covariates |
| No estimation from data | Full MLE via RSiena |
| Simulation only | Simulation *and* empirical estimation |
| $N \times 2^{K+1}$ ad hoc parameters | Parsimonious: scope, popularity, clustering |

The proof establishes that SaoMNK is not merely *inspired by* NK --- it formally
*contains* NK as a degenerate special case. Every theoretical insight from the NK
literature (landscape ruggedness, local optima traps, modularity advantages,
exploration-exploitation tradeoffs) remains valid within the SaoMNK framework.
But SaoMNK adds the capacity to bring these insights to empirical data and to
model the competitive dynamics that are central to strategy research.

---

# References {.unnumbered}

- Blume, L. E. (1993). The statistical mechanics of strategic interaction. *Games and Economic Behavior*, 5(3), 387--424.
- Kauffman, S. A. (1993). *The Origins of Order*. Oxford University Press.
- Levinthal, D. A. (1997). Adaptation on rugged landscapes. *Management Science*, 43(7), 934--950.
- McFadden, D. (1974). Conditional logit analysis of qualitative choice behavior. In *Frontiers in Econometrics*.
- McFadden, D. & Train, K. (2000). Mixed MNL models for discrete response. *Journal of Applied Econometrics*, 15(5), 447--470.
- Rivkin, J. W. (2000). Imitation of complex strategies. *Management Science*, 46(6), 824--844.
- Siggelkow, N. & Levinthal, D. A. (2003). Temporarily divide to conquer. *Organization Science*, 14(6), 650--669.
- Siggelkow, N. & Rivkin, J. W. (2005). Speed and search. *Organization Science*, 16(2), 101--122.
- Snijders, T. A. B. (2001). The statistical evaluation of social network dynamics. *Sociological Methodology*, 31(1), 361--395.
- Snijders, T. A. B., van de Bunt, G. G., & Steglich, C. E. G. (2010). Introduction to stochastic actor-based models. *Social Networks*, 32(1), 44--60.
