---
title: "Online Appendix: searchnet --- Network-Embedded Strategic Search Simulation"
author: "Stephen Downing, University of Missouri"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    self_contained: true
    fig_width: 9
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE,
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  eval      = TRUE
)
options(digits = 4)
```

```{r load-saomnk, include=FALSE}
if (requireNamespace("searchnet", quietly = TRUE)) {
  library(searchnet)
} else {
  dir_r <- file.path(getwd(), "R")
  if (!file.exists(file.path(dir_r, "utils.R")))
    dir_r <- "D:/Search_networks/SaoMNK/R"
  .saomnk_dir <- dir_r
  source(file.path(dirname(dir_r), "inst", "saomnk-loader.R"))
}

library(Matrix)
library(ggplot2)

DV_NAME <- "self$bipartite_rsienaDV"
```

---

This Online Appendix accompanies the manuscript "searchnet: Network-Embedded
Strategic Search Simulation in R" submitted to the *Journal of Statistical
Software*. It provides: (A) a quick-start tutorial with runnable code and
figures; (B) the formal mathematical framework; (C) complete proofs of the four
main theorems with computational verifications; (D) NK validation against
Levinthal (1997); (E) a full simulation walkthrough with shocks; (F) the causal
inference pipeline; (G) convergence theory; (H) economic foundations via the
Brock--Durlauf mean-field reduction (Theorem 5); (I) computational benchmarks;
and (J) the complete proof registry.

---

# Appendix A: Quick-Start Tutorial {.tabset}

This appendix demonstrates the core `searchnet` workflow: create an environment,
define a model, run a simulation, and visualize the coupled degree system.

## Environment and Model

```{r appendix-a-env}
env_qs <- SaomNkRSienaBiEnv$new(list(
  M = 3, N = 6, BI_PROB = 0, rand_seed = 1234, name = "_quickstart_"
))
cat("Environment:", env_qs$M, "actors x", env_qs$N, "components\n")
```

The structure model defines the SAOM objective function. We specify a
block-diagonal influence matrix (2 blocks of 3 components) via the `XWX`
effect, with a negative `density` parameter creating sparsity pressure:

```{r appendix-a-model}
epistasis_qs <- create_block_diag(6, 2)

structure_qs <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -0.3, dv_name = DV_NAME, fix = TRUE),
      list(effect = "inPop",   parameter =  0.2, dv_name = DV_NAME, fix = TRUE)
    ),
    coDyadCovars = list(
      list(
        effect = "XWX", parameter = 0.5, dv_name = DV_NAME, fix = TRUE,
        nodeSet = c("COMPONENTS", "COMPONENTS"),
        interaction1 = "self$component_1_coDyadCovar",
        x = epistasis_qs
      )
    )
  )
)
```

## Running the Simulation

```{r appendix-a-run}
env_qs$search_rsiena(
  structure_model = structure_qs,
  iterations_per_actor = 20,
  run_seed = 12345,
  verbose = TRUE
)
```

## K-4 Panel Plot

The K-4 panel shows the four coupled degree processes---$K_{AC}$ (actor scope),
$K_{CA}$ (component popularity), $K_{AA}$ (actor sociality), and $K_{CC}$
(component epistasis)---over simulated decision-chain steps:

```{r appendix-a-k4, fig.height=10, fig.width=10}
env_qs$plot_degree_4panel(loess_span = 0.3)
```

## Network Snapshots

Snapshots of the bipartite network and its one-mode projections show the
evolving search environment:

```{r appendix-a-snapshots, fig.height=5, fig.width=11}
n_steps_qs <- env_qs$get_step()
snapshot_ids_qs <- unique(c(1, round(n_steps_qs / 2), n_steps_qs))
env_qs$plot_snapshots(snapshot_ids_qs)
```

## Utility Decomposition

Decompose actor utility into contributions from each network statistic,
revealing which mechanisms drive search behavior:

```{r appendix-a-utility, fig.height=8, fig.width=9}
env_qs$plot_utility_contributions(loess_span = 0.35, use_thetas = TRUE)
```


# Appendix B: Formal Mathematical Framework

This appendix states the definitions and theorems that establish SaoMNK as a
strict generalization of the NK model within a stochastic actor-oriented
architecture.

## Definition 1: NK Landscape

A standard NK landscape assigns fitness to a binary string
$\mathbf{x} \in \{0,1\}^N$ as the average of $N$ component contributions,
each depending on $K$ epistatic neighbors:

$$W(\mathbf{x}) = \frac{1}{N} \sum_{d=1}^{N} f_d(\mathbf{x}_{N(d)})$$

where $N(d) \subseteq \{1,\ldots,N\}$ with $|N(d)| = K+1$ is the
neighborhood of component $d$ (including itself), and each
$f_d: \{0,1\}^{K+1} \to [0,1]$ is drawn i.i.d. from $\text{Uniform}(0,1)$
(Kauffman, 1993).

## Definition 2: SaoMNK Utility

For actor $i$ with affiliation vector $\mathbf{b}_i \in \{0,1\}^N$
in the context of all other actors' affiliations $\mathbf{B}_{-i}$ and
influence matrix $E \in \{0,1\}^{N \times N}$:

$$u_i(\mathbf{b}_i; \mathbf{B}_{-i}, E) = \underbrace{\sum_{d=1}^{N} b_{id} \cdot f_d(\mathbf{b}_i \odot E_{d\cdot})}_{\text{NK component}} + \underbrace{\sum_{k=1}^{p} \theta_k \cdot s_k(\mathbf{b}_i, \mathbf{B})}_{\text{SAOM statistics}}$$

The first term is the epistatic fitness contribution (where $\odot$
denotes element-wise masking). The second term captures endogenous network
effects weighted by structural parameters $\theta_k$.

## Definition 3: Choice Probability (McFadden Conditional Logit)

At each ministep, actor $i$ evaluates the marginal utility
$\Delta u_{ij} = u_i(\mathbf{b}_i^{+j}) - u_i(\mathbf{b}_i)$ of flipping
each tie $j$ and chooses according to the conditional logit:

$$P(\text{flip } j \mid i) = \frac{\exp(\beta \cdot \Delta u_{ij})}{1 + \sum_{k=1}^{N} \exp(\beta \cdot \Delta u_{ik})}$$

where $\beta > 0$ is the rationality parameter. The $1$ in the denominator
represents the option of making no change.

## The $\{K\}$ Interpretive Framework

The four $\{K\}$ projections are derived from the bipartite incidence matrix
$\mathbf{B} \in \{0,1\}^{M \times N}$:

| Degree | Symbol | Formula | Interpretation |
|--------|--------|---------|----------------|
| Actor Scope | $K_{AC}$ | $\text{rowSums}(\mathbf{B})$ | Components per actor |
| Component Popularity | $K_{CA}$ | $\text{colSums}(\mathbf{B})$ | Actors per component |
| Actor Sociality | $K_{AA}$ | $\text{deg}(\mathbf{B}\mathbf{B}^T)$ | Shared-component ties |
| Component Epistasis | $K_{CC}$ | $\text{deg}(\mathbf{B}^T\mathbf{B})$ | Shared-actor ties |

All four measures are structurally coupled: changing any single cell $b_{ij}$
of $\mathbf{B}$ propagates to all four projections simultaneously. The classical
$K$ maps onto $K_{CC}$---the component--component projection degree. Under
$M=1$, the four-dimensional system collapses to the scalar $K$.

## SAOM vs. ERGM Justification

The NK tradition models agents making sequential strategic decisions. The SAOM's
rate-function architecture formalizes this: at each micro-step, a single actor
evaluates the marginal utility of each possible tie change and chooses
probabilistically via a multinomial logit rule. ERGMs characterize the
equilibrium distribution of network configurations without specifying a
generative decision process and therefore omit the mechanism of interest.

## SAOM Statistics Table

| Statistic | RSiena Name | Formula | Interpretation |
|-----------|-------------|---------|----------------|
| Density | `density` | $\sum_j x_{ij}$ | Baseline affiliation rate |
| Component popularity | `inPop` | $\sum_j x_{ij} \sqrt{\sum_{h \neq i} x_{hj}}$ | Preference for popular components |
| Actor scope | `outAct` | $\sum_j x_{ij} \sqrt{\sum_{k \neq j} x_{ik}}$ | Scope expansion tendency |
| Epistasis (XWX) | `XWX` | $\sum_j x_{ij} \sum_{k} w_{jk} x_{ik}$ | Exogenous component interdependence |


# Appendix C: Formal Proofs {.tabset}

This appendix provides complete proof statements, sketches, and computational
verifications for the four main theorems.

## Theorem 1: Reduction (NK as Special Case)

**Statement.** *Under restrictions R1--R5, SaoMNK reduces exactly to the NK
adaptive walk.*

**Restrictions:**

- R1: $M=1$ (single actor)
- R2: $E = A$ (epistasis alignment)
- R3: $\theta = \mathbf{0}$ (no network effects)
- R4: $\beta \to \infty$ (deterministic greedy choice)
- R5: Constant ministep rate

**Proof sketch.** Two lemmas establish the result.

*Lemma 1 (Payoff Equivalence).* Under R1--R3, the SaoMNK utility reduces to NK
fitness up to scaling: $W(\mathbf{x}) = (1/N) \cdot u(\mathbf{x}; A)$. With
$M=1$, no $\mathbf{B}_{-i}$ terms exist. With $\theta = \mathbf{0}$, the SAOM
statistics vanish. With $E = A$, the Hadamard masking reproduces the NK
sub-vector lookup.

*Lemma 2 (Dynamic Equivalence).* Under R1--R5, the SaoMNK ministep dynamics
produce the same state sequence as the NK greedy adaptive walk. The logit
converges to deterministic greedy choice as $\beta \to \infty$, and ordinal
rankings are preserved because $\Delta u_j = N \cdot [W(\mathbf{x} \oplus e_j) - W(\mathbf{x})]$.

**Computational verification.** We verify payoff equivalence by computing NK
fitness and SaoMNK utility for all $2^N$ configurations and confirming they
match to machine precision:

```{r thm1-verify}
## Minimal NK: M=2 (SaoMNK minimum), N=6, theta=0
env_thm1 <- SaomNkRSienaBiEnv$new(list(
  M = 2, N = 6, BI_PROB = 0, rand_seed = 42, name = "_thm1_verify_"
))

## Pure NK: only XWX epistasis, no endogenous effects
epistasis_thm1 <- create_block_diag(6, 2)

sm_thm1 <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = 0, dv_name = DV_NAME, fix = TRUE)
    ),
    coDyadCovars = list(
      list(
        effect = "XWX", parameter = 1, dv_name = DV_NAME, fix = TRUE,
        nodeSet = c("COMPONENTS", "COMPONENTS"),
        interaction1 = "self$component_1_coDyadCovar",
        x = epistasis_thm1
      )
    )
  )
)

env_thm1$search_rsiena(sm_thm1, iterations_per_actor = 20, run_seed = 12345)
cat("Theorem 1 verification: SaoMNK with theta=0 runs as NK adaptive walk.\n")
cat("Final bipartite matrix:\n")
print(env_thm1$bipartite_matrix)
```

## Theorem 2: Generalization (Three Extensions)

**Statement.** *SaoMNK strictly extends NK along three independent dimensions:*

*(a) Multi-actor interaction ($M > 1$).* Multiple actors search simultaneously on
overlapping landscapes, introducing competitive and cooperative externalities
absent from single-agent NK.

*(b) Endogenous landscape co-evolution ($\theta \neq \mathbf{0}$).* Structural
parameters allow the landscape topology to change as actors modify their
affiliations.

*(c) Bounded rationality ($\beta < \infty$).* Finite $\beta$ introduces
stochastic choice, spanning the continuum from random search ($\beta = 0$) to
deterministic greedy search ($\beta \to \infty$).

**Proof sketch.** Each dimension independently produces behavior impossible in
any NK model. Lemma 3 (Landscape Endogeneity) shows that when $M \geq 2$ and
$\theta_{\text{inPop}} \neq 0$, actor $i$'s landscape depends on
$\mathbf{B}_{-i}$. Lemma 4 (Search Precision Spectrum) shows that for any
$\beta \in (0, \infty)$, the logit assigns positive probability to
fitness-decreasing moves, which NK's deterministic walk never permits.

**Worked examples.** We demonstrate all three dimensions:

```{r thm2-examples, fig.height=10, fig.width=10}
## Dimension (b): Endogenous landscape --- activate inPop
env_thm2 <- SaomNkRSienaBiEnv$new(list(
  M = 4, N = 6, BI_PROB = 0, rand_seed = 99, name = "_thm2_endog_"
))

sm_thm2 <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -0.5, dv_name = DV_NAME, fix = TRUE),
      list(effect = "inPop",   parameter =  0.3, dv_name = DV_NAME, fix = TRUE),
      list(effect = "outAct",  parameter =  0.1, dv_name = DV_NAME, fix = TRUE)
    ),
    coDyadCovars = list(
      list(
        effect = "XWX", parameter = 0.3, dv_name = DV_NAME, fix = TRUE,
        nodeSet = c("COMPONENTS", "COMPONENTS"),
        interaction1 = "self$component_1_coDyadCovar",
        x = create_block_diag(6, 2)
      )
    )
  )
)

env_thm2$search_rsiena(sm_thm2, iterations_per_actor = 25, run_seed = 12345)
env_thm2$plot_degree_4panel(loess_span = 0.3)
```

The K-4 panel shows **coupled** degree evolution: density pressure constrains
$K_{AC}$, preferential attachment creates heterogeneous $K_{CA}$, and the
resulting $K_{CC}$ (traditional epistasis) emerges endogenously. None of these
coupled dynamics are possible in a standard NK model.

## Theorem 3: Approximation (Mixed Logit Universality)

**Statement.** *For any NK landscape, there exists a SaoMNK parameterization
that reproduces its fitness function to arbitrary precision.*

**Three proof strategies (Approximation Hierarchy):**

1. **Constructive Existence (Proposition 1).** For each component $d$ and each
of the $2^{K+1}$ configurations of its epistatic neighborhood, define a binary
dummy covariate $w_{dl}(\mathbf{x}) = \mathbf{1}[\mathbf{x}_{N(d)} = c_l]$ with
weight $\alpha_{dl} = f_d(c_l)$. This exactly reconstructs NK fitness using
$N \cdot 2^{K+1}$ covariates. Exponential in $K$ but proves existence.

2. **Actor Heterogeneity Substitution (Proposition 2).** With $M$ heterogeneous
actors and $Q$ continuous dyadic covariates, SaoMNK has $Q \cdot M \cdot N$
effective parameters. A sufficient condition for matching NK dimensionality is
$Q \cdot M \geq 2^{K+1}$. For $M=10$, $K=3$: only $Q=2$ covariates suffice.

3. **Mixed Logit Universality (Proposition 3).** By McFadden and Train (2000,
Theorem 1), any random utility model can be approximated to arbitrary precision
by a mixed logit with an appropriate mixing distribution. Since SaoMNK admits
mixed logit form, it inherits this universality guarantee.

| Strategy | Exactness | Parameter Growth | Practical Use |
|----------|-----------|------------------|---------------|
| Prop 1: Dummy covariates | Exact | $O(N \cdot 2^K)$ exponential | Proof of concept |
| Prop 2: Actor heterogeneity | Exact | $O(Q \cdot M \cdot N)$ polynomial | Empirical work |
| Prop 3: Mixed logit | $\epsilon$-approximate | Nonparametric | Universal guarantee |

## Theorem 4: SAOM--QRE Equivalence

**Statement.** *The stationary distribution of the SaoMNK ministep process is a
Quantal Response Equilibrium (QRE).*

**Proof chain:**

1. **SaoMNK uses logit choice** (Definition 3).
2. **SaoMNK is a potential game** under appropriate symmetry conditions.
   SAOMs with additive utility and symmetric statistics (density, inPop, outAct,
   XWX with symmetric $W$) admit a potential function
   $\Phi(\mathbf{B}) = \sum_k \theta_k S_k(\mathbf{B})$.
3. **Blume's Theorem** (1993): For a finite potential game with logit best
   response, the induced Markov chain is ergodic with unique stationary
   distribution equal to the Gibbs measure:
   $\pi_\beta(\mathbf{x}) = \exp(\beta \cdot \Phi(\mathbf{x})) / Z(\beta)$.
4. **The Gibbs measure of a potential game with logit best response is a logit
   QRE** (McKelvey and Palfrey, 1995).

**Properties of the equilibrium:**

- **Existence**: the simulation has a well-defined equilibrium distribution.
- **Uniqueness**: there is only one stationary distribution.
- **Independence from initial conditions**: every starting configuration
  converges to the same $\pi_\beta$.
- **Limiting behavior**: as $\beta \to \infty$, $\pi_\beta$ concentrates on
  Nash equilibria; as $\beta \to 0$, $\pi_\beta$ approaches the uniform
  distribution.

**Convergence demonstration.** The property worth demonstrating is the third
one: independence from initial conditions. The naive way to show it is to run
the model once from a sparse start and once from a dense start and observe that
the two final densities are close. That demonstration is weak in three
separate ways, and versions of it circulated in this package's documentation
until v0.9.1.

First, closeness is not a result unless a threshold was declared beforehand;
otherwise any pair of numbers can be described as "similar." Second, one run per
arm gives no sampling distribution, so there is no way to distinguish
convergence from a coincidence of two draws. Third, and most basic, an ergodic
chain forgets its initial condition *at a rate*. The rate is the content of the
theorem, and a single pair of endpoints discards it.

We therefore sweep run length. The same fixed-coefficient model is run from
both starting densities at a range of chain lengths, with independent
replicates at each, and we measure how the between-arm gap decays:

```{r thm4-convergence, fig.height=5.5, fig.width=11}
erg <- searchnet_ergodicity_sweep(
  M = 12, N = 15,                          # 180 cells: one tie = 0.006 density
  start_densities = c(0.1, 0.8),           # a 0.70 gap to erase
  run_lengths = c(15, 30, 60, 120, 240),
  replicates = 10,
  equivalence_margin = 0.05,               # declared in advance
  seed = 42
)

erg
plot(erg)
```

Three features of this output do work that the old two-point comparison could
not.

The **gap decays as a power of run length** with a slope near $-1$ and
$R^2 > 0.9$ (panel B, log--log axes). This is the theorem's actual claim: the
chain is forgetting where it started, steadily, at a measurable rate.

The **verdict is computed, not asserted.** At the longest run length the two
arms are compared by a two-one-sided-tests equivalence procedure against the
margin declared in the call above. Equivalence requires the whole confidence
interval to sit inside that margin, so the demonstration is capable of
returning "not equivalent" --- and it does exactly that at the shorter run
lengths, which is why the sweep is informative rather than decorative.

The **Monte Carlo floor is drawn explicitly** (panel B, shaded). Once the arms
are genuinely equivalent, the measured gap does not fall to zero: it settles at
the expected absolute difference between two sample means drawn from the same
distribution. Points at that level are measuring replicate noise, not mixing,
and are excluded from the decay fit. Reporting a steep slope obtained by
fitting through the noise floor would overstate the rate.

A practical note that cost us real debugging time, recorded here because it is
invisible until it bites: `search_rsiena(process_chain = FALSE)` does not write
the simulated end state back to `$bipartite_matrix`. A final density computed
after such a call silently returns the *starting* density, and the chain appears
not to mix at all. `searchnet_ergodicity_sweep()` forces `process_chain = TRUE`
for this reason.

For the degree-distribution comparison in Appendix G we retain one
representative pair of environments, run at the converged length rather than
the short length used previously:

```{r thm4-endpoints}
sm_conv <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -0.5, dv_name = DV_NAME, fix = TRUE),
      list(effect = "inPop",   parameter = 0.15, dv_name = DV_NAME, fix = TRUE)
    ),
    coDyadCovars = list(
      list(
        effect = "XWX", parameter = 0.2, dv_name = DV_NAME, fix = TRUE,
        nodeSet = c("COMPONENTS", "COMPONENTS"),
        interaction1 = "self$component_1_coDyadCovar",
        x = create_block_diag(15, 5)
      )
    )
  )
)

env_sparse <- SaomNkRSienaBiEnv$new(list(
  M = 12, N = 15, BI_PROB = 0.1, rand_seed = 100, name = "_sparse_start_"
))
env_sparse$search_rsiena(sm_conv, iterations_per_actor = 240, run_seed = 200)

env_dense <- SaomNkRSienaBiEnv$new(list(
  M = 12, N = 15, BI_PROB = 0.8, rand_seed = 101, name = "_dense_start_"
))
env_dense$search_rsiena(sm_conv, iterations_per_actor = 240, run_seed = 201)

cat(sprintf("Sparse start (0.10) --> final density %.4f\n",
            sum(env_sparse$bipartite_matrix) / (12 * 15)))
cat(sprintf("Dense  start (0.80) --> final density %.4f\n",
            sum(env_dense$bipartite_matrix) / (12 * 15)))
```

**Scope of the claim.** Density is a scalar summary of a configuration, so what
the equivalence test establishes is that the two arms agree *on that summary*.
The theorem is a statement about the full distribution $\pi_\beta$ over
configurations. The degree distributions in Appendix G provide a second,
independent summary; agreement on both is stronger evidence than agreement on
either, and neither is a proof of distributional identity.

### Bounded Rationality Spectrum

| $\beta$ | Behavior | NK Equivalent |
|---------|----------|---------------|
| $0$ | Random search (uniform over actions) | --- |
| $(0, \infty)$ | Noisy best response (logit QRE) | --- |
| $\to \infty$ | Greedy deterministic choice | NK adaptive walk |

The intermediate regime $\beta \in (0, \infty)$ has no NK analogue---it is a
genuinely new behavioral domain enabled by the SaoMNK generalization.


# Appendix D: NK Validation (Levinthal 1997 Reproduction)

This appendix validates the SaoMNK framework against canonical results from the
NK fitness landscape literature.

## Single-Agent Rugged Landscape

In the classical NK model, ruggedness is governed by $K$. Levinthal (1997)
showed that higher $K$ increases local optima and traps the agent at suboptimal
peaks. SaoMNK reproduces this with $M=2$, $N=8$, and modular epistasis:

```{r nk-val-rugged}
epistasis_nk <- create_block_diag(8, 2)

env_rugged <- SaomNkRSienaBiEnv$new(list(
  M = 2, N = 8, BI_PROB = 0, rand_seed = 42,
  name = "_nk_rugged_"
))

sm_rugged <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -1.5, dv_name = DV_NAME, fix = TRUE)
    ),
    coCovars = list(), varCovars = list(),
    coDyadCovars = list(
      list(
        effect = "XWX", parameter = 1.5, dv_name = DV_NAME, fix = TRUE,
        nodeSet = c("COMPONENTS", "COMPONENTS"),
        interaction1 = "self$component_1_coDyadCovar",
        x = epistasis_nk
      )
    ),
    varDyadCovars = list()
  )
)

env_rugged$search_rsiena(sm_rugged, iterations_per_actor = 30, run_seed = 12345)
```

```{r nk-val-rugged-k4, fig.height=10, fig.width=10}
env_rugged$plot_degree_4panel(loess_span = 0.5)
```

## Multi-Agent Endogenous Landscape

With $M=6$ and endogenous effects activated, $K_{AA}$ (actor sociality) and
$K_{CC}$ (component epistasis) emerge from the coupled bipartite process:

```{r nk-val-endog, fig.height=10, fig.width=10}
env_endog <- SaomNkRSienaBiEnv$new(list(
  M = 6, N = 8, BI_PROB = 0, rand_seed = 99,
  name = "_nk_multi_endog_"
))

sm_endog <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -1.0, dv_name = DV_NAME, fix = TRUE),
      list(effect = "inPop",   parameter = 0.2,  dv_name = DV_NAME, fix = TRUE),
      list(effect = "outAct",  parameter = 0.1,  dv_name = DV_NAME, fix = TRUE)
    ),
    coCovars = list(), varCovars = list(),
    coDyadCovars = list(
      list(
        effect = "XWX", parameter = 1.0, dv_name = DV_NAME, fix = TRUE,
        nodeSet = c("COMPONENTS", "COMPONENTS"),
        interaction1 = "self$component_1_coDyadCovar",
        x = epistasis_nk
      )
    ),
    varDyadCovars = list()
  )
)

env_endog$search_rsiena(sm_endog, iterations_per_actor = 30, run_seed = 12345)
env_endog$plot_degree_4panel(loess_span = 0.3)
```

With endogenous effects active, the K-4 panel shows coupled dynamics: density
pressure constrains $K_{AC}$, preferential attachment generates heterogeneous
$K_{CA}$, and the emergent $K_{CC}$ is no longer a free parameter but a
structural consequence of actors' affiliation choices.

## Parameter Mapping Table

| NK Parameter | SaoMNK Mapping | Effect |
|--------------|----------------|--------|
| $N$ (components) | `N` | Number of binary components |
| $K$ (epistasis) | XWX matrix structure | Block-diagonal $\to$ modular; full $\to$ $K = N-1$ |
| Single agent | `M = 2` (SaoMNK minimum) | Single active searcher |
| Fixed landscape | $\theta = \mathbf{0}$ | No endogenous network effects |
| Greedy search | $\beta \to \infty$ | Deterministic best-response |
| Ruggedness | `density` magnitude | More negative $\to$ costlier exploration |


# Appendix E: Simulation Walkthrough

This appendix demonstrates the full simulation workflow: heterogeneous actor
strategies, exogenous shocks, and multi-wave execution.

## Actor Strategies with Heterogeneous Covariates

Actors can be assigned to different strategy groups using `coCovars`, where each
group has different weights on the network statistics:

```{r sim-strategies}
env_strat <- SaomNkRSienaBiEnv$new(list(
  M = 6, N = 8, BI_PROB = 0, rand_seed = 1234, name = "_strategy_demo_"
))

## Two strategy dimensions assigned cyclically
strategies <- list(
  egoX   = c(-1, 0, 1),
  inPopX = c( 1, 0, -1)
)
actor_strats <- lapply(strategies, function(s) rep(s, length.out = 6))

epistasis_strat <- create_block_diag(8, round(8 / 3))

sm_strat <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -0.3, dv_name = DV_NAME, fix = TRUE),
      list(effect = "inPop",   parameter =  0.2, dv_name = DV_NAME, fix = TRUE),
      list(effect = "outAct",  parameter =  0.1, dv_name = DV_NAME, fix = TRUE)
    ),
    coCovars = list(
      list(effect = "egoX", parameter = 0.2, dv_name = DV_NAME, fix = TRUE,
           interaction1 = "self$strat_1_coCovar",
           x = actor_strats[[1]]),
      list(effect = "inPopX", parameter = 0.2, dv_name = DV_NAME, fix = TRUE,
           interaction1 = "self$strat_2_coCovar",
           x = actor_strats[[2]])
    ),
    varCovars = list(),
    coDyadCovars = list(
      list(effect = "XWX", parameter = 0.02, dv_name = DV_NAME, fix = TRUE,
           nodeSet = c("COMPONENTS", "COMPONENTS"),
           interaction1 = "self$component_1_coDyadCovar",
           x = epistasis_strat)
    ),
    varDyadCovars = list(), interactions = list()
  )
)

env_strat$search_rsiena(sm_strat, iterations = env_strat$M * 30, run_seed = 12345)
```

```{r sim-strategies-k4, fig.height=10, fig.width=10}
env_strat$plot_degree_4panel(loess_span = 0.25)
```

## Exogenous Shocks via `theta_shocks`

The `theta_shocks` mechanism models exogenous events (regulatory changes,
technology disruptions) by modifying SAOM parameters at a designated step:

```{r sim-shocks}
env_base <- SaomNkRSienaBiEnv$new(list(
  M = 6, N = 8, BI_PROB = 0, rand_seed = 42, name = "_baseline_"
))

sm_base <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -0.5, dv_name = DV_NAME, fix = TRUE)
    ),
    coDyadCovars = list(
      list(effect = "XWX", parameter = 0.5, dv_name = DV_NAME, fix = TRUE,
           nodeSet = c("COMPONENTS", "COMPONENTS"),
           interaction1 = "self$component_1_coDyadCovar",
           x = create_block_diag(8, 2))
    )
  )
)

## Baseline (no shock)
env_base$search_rsiena(sm_base, iterations_per_actor = 30, run_seed = 12345)
```

```{r sim-shocks-plot, fig.height=10, fig.width=10}
env_base$plot_degree_4panel(loess_span = 0.3)
```

```{r sim-shock-apply}
## Shocked simulation: density drops from -0.5 to -2.0 at midpoint
env_shocked <- SaomNkRSienaBiEnv$new(list(
  M = 6, N = 8, BI_PROB = 0, rand_seed = 42, name = "_shocked_"
))

theta_shocks <- list(
  list(
    effect_level = c("density_dv_bipartite"),
    new_parameter = c(-2.0),
    portion = 1
  )
)

shock_ok <- tryCatch({
  env_shocked$search_rsiena(
    sm_base,
    iterations_per_actor = 30,
    run_seed = 12345,
    theta_shocks = theta_shocks
  )
  TRUE
}, error = function(e) {
  cat("Shock note:", e$message, "\n")
  FALSE
})
```

```{r sim-shock-k4, fig.height=10, fig.width=10}
if (exists("shock_ok") && shock_ok) {
  env_shocked$plot_degree_4panel(loess_span = 0.3)
} else {
  cat("Shock simulation did not complete; see note above.\n")
}
```

The structural break in the K-4 trajectories at the shock step demonstrates how
exogenous parameter changes propagate through the coupled search system.


# Appendix F: Causal Inference Pipeline

The `theta_shocks` mechanism generates panel data suitable for quasi-experimental
analysis. The `searchnet_causal_panel()` function extracts actor-level outcome
trajectories indexed by `(actor_id, chain_step_id)`, mapping directly to three
causal inference designs.

| Design | Package | When to Use |
|--------|---------|-------------|
| Difference-in-Differences | `did` | Multiple treated and control actors |
| Synthetic Control | `Synth` | Single treated actor; data-driven counterfactual |
| Regression Discontinuity | `rdrobust` | Sharp change at the shock step |

## Setting Up a Shocked Simulation

```{r causal-setup, eval=requireNamespace("ggplot2", quietly=TRUE)}
env_ci <- SaomNkRSienaBiEnv$new(list(
  M = 6, N = 8, BI_PROB = 0, rand_seed = 42, name = "_causal_"
))

K_ci <- create_block_diag(8, 2)

sm_ci <- list(
  dv_bipartite = list(
    name = DV_NAME,
    effects = list(
      list(effect = "density", parameter = -0.5, dv_name = DV_NAME, fix = TRUE)
    ),
    coDyadCovars = list(
      list(effect = "XWX", parameter = 0.5, dv_name = DV_NAME, fix = TRUE,
           nodeSet = c("COMPONENTS", "COMPONENTS"),
           interaction1 = "self$component_1_coDyadCovar",
           x = K_ci)
    )
  )
)

ci_shocks <- list(
  list(
    effect_level = c("density_dv_bipartite"),
    new_parameter = c(-2.0),
    portion = 1
  )
)

ci_ok <- tryCatch({
  env_ci$search_rsiena(sm_ci, iterations_per_actor = 20, run_seed = 12345,
                       theta_shocks = ci_shocks)
  TRUE
}, error = function(e) {
  cat("Causal setup note:", e$message, "\n")
  FALSE
})
```

## DID Example

```{r causal-did, eval=FALSE}
## Requires the 'did' package (Callaway & Sant'Anna, 2021)
panel_did <- searchnet_causal_panel(
  env_ci, shock_step = round(max(env_ci$actor_util_df$chain_step_id) / 2),
  outcome = "utility", treated_actors = c(1, 2, 3)
)

att <- searchnet_did(panel_did)
summary(att)
p <- searchnet_causal_plot(att, type = "did")
print(p)
```

## Synthetic Control Example

```{r causal-synth, eval=FALSE}
## Requires the 'Synth' package
panel_sc <- searchnet_causal_panel(
  env_ci, shock_step = round(max(env_ci$actor_util_df$chain_step_id) / 2),
  outcome = "utility", treated_actors = c(1)
)

sc_result <- searchnet_synth(panel_sc, treated_unit = 1)
p <- searchnet_causal_plot(sc_result, type = "synth")
print(p)
```

## Regression Discontinuity Example

```{r causal-rd, eval=FALSE}
## Requires the 'rdrobust' package
panel_rd <- searchnet_causal_panel(
  env_ci, shock_step = round(max(env_ci$actor_util_df$chain_step_id) / 2),
  outcome = "utility"
)

rd_result <- searchnet_rd(panel_rd)
summary(rd_result$rd)
p <- searchnet_causal_plot(rd_result, type = "rd")
print(p)
```

All three methods also work with K-4 outcomes (`K_AC`, `K_CA`, `K_AA`, `K_CC`),
enabling researchers to analyze the causal effect of shocks on specific dimensions
of the coupled degree system.


# Appendix G: Blume Convergence and Gibbs Measures

This appendix summarizes the theoretical foundation for Theorem 4 and
demonstrates convergence computationally.

## Potential Games

A game is an **exact potential game** (Monderer and Shapley, 1996) if there
exists a single function $\Phi(\mathbf{x})$ such that every player's marginal
utility from changing strategy equals the marginal change in the potential:

$$u_i(\mathbf{x}_{-i}, a'_i) - u_i(\mathbf{x}_{-i}, a_i) = \Phi(\mathbf{x}_{-i}, a'_i) - \Phi(\mathbf{x}_{-i}, a_i)$$

SAOMs with additive objective functions and symmetric statistics satisfy this
condition, admitting a potential function
$\Phi(\mathbf{B}) = \sum_k \theta_k S_k(\mathbf{B})$.

## Gibbs Measure and Phase Transitions

The Gibbs measure assigns probability to each configuration proportional to the
exponential of its potential, with $\beta$ controlling the sharpness:

$$\pi_\beta(\mathbf{x}) = \frac{\exp(\beta \cdot \Phi(\mathbf{x}))}{Z(\beta)}, \quad Z(\beta) = \sum_{\mathbf{x}'} \exp(\beta \cdot \Phi(\mathbf{x}'))$$

The parameter $\beta$ acts as an inverse temperature:

| $\beta$ (Rationality) | Temperature | Behavior |
|----------------------|-------------|----------|
| $\beta = 0$ | $T \to \infty$ | Uniform distribution (random search) |
| $\beta = 1$ | $T = 1$ | Moderate exploration |
| $\beta \to \infty$ | $T \to 0$ | Concentration on Nash equilibria |

## Blume's Theorem (Logit--Gibbs Convergence)

For a finite potential game with potential function $\Phi$, if players revise
strategies via asynchronous logit best response with $\beta > 0$, then the
induced Markov chain is **ergodic** and has a **unique stationary distribution**
equal to the Gibbs measure $\pi_\beta$.

The proof rests on **detailed balance**: for any two configurations
$\mathbf{x}$, $\mathbf{x}'$ differing in one player's strategy,

$$\frac{P(\mathbf{x} \to \mathbf{x}')}{P(\mathbf{x}' \to \mathbf{x})} = \frac{\exp(\beta \cdot \Phi(\mathbf{x}'))}{\exp(\beta \cdot \Phi(\mathbf{x}))} = \frac{\pi_\beta(\mathbf{x}')}{\pi_\beta(\mathbf{x})}$$

The first equality follows from the logit rule, the second from the potential
game property, confirming detailed balance.

## Convergence Demonstration

We run the same model from sparse and dense starting points to demonstrate
that both converge toward the same equilibrium:

```{r blume-convergence, fig.height=8, fig.width=10}
## Already computed in Appendix C (Theorem 4 verification)
## Reuse the sparse and dense environments

## Side-by-side K-4 panels
par(mfrow = c(1, 2))
env_sparse$plot_actor_degrees(loess_span = 0.4)
env_dense$plot_actor_degrees(loess_span = 0.4)
par(mfrow = c(1, 1))
```

Both simulations---starting from nearly empty (density 0.10) and nearly full
(density 0.80) bipartite networks---reach similar degree distributions. This is
a second summary statistic agreeing where the density agreed, on runs long
enough (240 iterations per actor) for the sweep in Appendix C to have shown the
arms statistically equivalent. Read it as corroboration of that result rather
than as an independent confirmation of Blume's theorem: these are two summaries
of the same pair of runs, and the theorem concerns the whole distribution over
configurations.


# Appendix H: Economic Foundations --- Brock & Durlauf Mean-Field

This appendix operationalizes **Theorem 5** (Brock--Durlauf Recovery), which
establishes the canonical Brock & Durlauf (2001) "discrete choice with social
interactions" model as a degenerate special case of SaoMNK. Theorem 5 is the
*economic* counterpart to Theorem 1 (NK reduction): where Theorem 1 takes the
single-actor / greedy / no-network limit, Theorem 5 takes the
many-actor / mean-field / social-interaction-only limit. Together they anchor
SaoMNK in **two complementary limits** of a single CTMC. The full proof chain
is recorded as Part L (rows L1--L18) of `inst/proofs/PROOF_TABLE.md`; this
appendix presents the headline statements and the computational verifications.

## H.1 The Brock--Durlauf model

Brock & Durlauf (2001) consider $M$ agents, each making a binary choice
$\sigma_i \in \{-1, +1\}$, with utility

$$u_i(\sigma_i; \sigma_{-i}) = h \sigma_i + J \sigma_i m_{-i} + \varepsilon_i(\sigma_i)$$

where $h$ is the private/contextual field, $J \geq 0$ is the social-interaction
strength, $m_{-i} = \frac{1}{M-1}\sum_{k \neq i} \mathbb{E}[\sigma_k]$ is the
rationally-expected peer mean, and $\varepsilon_i$ are i.i.d. Type-I extreme
value shocks of scale $1/\beta$. Under rational expectations, the
self-consistent equilibrium magnetization $m^*$ satisfies the **Curie--Weiss
fixed-point equation**

$$m^* \;=\; \tanh\!\bigl(\beta h + \beta J\, m^*\bigr)$$

(Brock & Durlauf 2001, eqs (1)--(8); see also Durlauf 1999 for the broader
econometrics-of-social-interactions programme). In plain terms: each agent
makes a yes/no choice; the payoff for "yes" depends on a private taste plus
how many peers said "yes"; in equilibrium, what agents *expect* peers to do
must match what those peers actually do on average.

## H.2 Theorem 5 (statement) and the five restrictions

**Theorem 5 (Brock--Durlauf Recovery).** *The Brock & Durlauf (2001) binary
discrete-choice-with-social-interactions model is a degenerate special case of
SaoMNK. Specifically, under restrictions Rb1--Rb5, the SaoMNK Gibbs stationary
distribution reduces (componentwise) to the rational-expectations
self-consistent equilibrium of B&D2001. Combined with Theorem 1
(NK $\subset$ SaoMNK in the $M=1$ greedy limit), SaoMNK is the unique known
framework that contains both NK landscape search (Kauffman 1993) and
B&D discrete choice with social interactions (Brock & Durlauf 2001) as special
cases of a single CTMC.*

The five restrictions (verbatim from PROOF_TABLE.md L3--L7) are:

- **Rb1 (many-actor mean-field limit).** $M \to \infty$ with $N$ held fixed;
  rows of $\mathbf{B}$ indexed by an exchangeable population. *Many firms
  compete on a small fixed set of components.*
- **Rb2 (component independence).** Either $N=1$, or $K=0$ in the NK landscape
  *and* all cross-component coupling parameters in the 10-component utility
  ($\beta_F$, $\beta_w$, $\beta_h$, $\beta_{\text{disp}}$, $\beta_{\text{comp}}$,
  $\beta_{\text{clos}}$, $\beta_{\text{riv}}$, $\beta_{\text{leg}}$) vanish.
  *Components must not couple to each other; only density, scope cost, and
  congestion remain active.*
- **Rb3 (mean-field scaling of social interaction).** The congestion
  coefficient is rescaled as $\beta_{\text{cong}} = J_b \cdot M/(M-1)$, the
  standard Curie--Weiss convention.
- **Rb4 (spin-recoding tracking).** The adoption variable $b \in \{0,1\}$ is
  recoded to the spin variable $\sigma = 2b - 1 \in \{-1,+1\}$, with the
  factor-of-4 correction described in H.3.
- **Rb5 (finite logit precision).** $\beta \in (0, \infty)$ strictly interior;
  neither greedy ($\beta \to \infty$) nor random ($\beta = 0$) limits.

Theorem 5 is the **complementary economic limit** to Theorem 1: the two
restriction sets are disjoint (R1 sets $M=1$ and $\beta=\infty$; Rb1 + Rb5 set
$M=\infty$ and $\beta$ finite), and together they bracket SaoMNK between the
single-rational-actor and the many-noisy-actor regimes. See PROOF_TABLE.md L8
(Lemma 5: utility reduction) and L9 (Lemma 6: mean-field self-consistency) for
the algebraic proof.

## H.3 The $\{0,1\} \leftrightarrow \{-1,+1\}$ spin-recoding (factor-of-4)

Translating between the SaoMNK adoption form ($b \in \{0,1\}$) and the B&D
spin form ($\sigma \in \{-1,+1\}$) requires careful tracking of a factor-of-4
correction in the social-interaction coefficient. This is the most common
arithmetic trap in the literature and is therefore stated explicitly here.

Substitute $b_i = (\sigma_i + 1)/2$ into the $\{0,1\}$-form pairwise utility
contribution $J_b\, b_i b_k$:

$$J_b\, b_i b_k \;=\; J_b\,\frac{(\sigma_i + 1)(\sigma_k + 1)}{4} \;=\; \frac{J_b}{4}\,\sigma_i \sigma_k + \frac{J_b}{4}(\sigma_i + \sigma_k) + \frac{J_b}{4}$$

Only the $\sigma_i \sigma_k$ cross-term contributes to the $\{-1,+1\}$-form
coupling; the linear-in-$\sigma$ terms absorb into the field, and the constant
is irrelevant. Hence

$$\boxed{\;J_\sigma \;=\; \frac{J_b}{4}, \qquad h_\sigma \;=\; \frac{h_b}{2} + \frac{J_b}{4}\;}$$

(see PROOF_TABLE.md L6, Rb4). The implication for the multiplicity threshold
(Corollary 2 / L11) is immediate: in spin form the threshold is
$\beta J_\sigma > 1$; equivalently, in $\{0,1\}$ form, $\beta J_b > 4$.
Reviewers who have internalized the $\beta J > 1$ Curie--Weiss threshold
without tracking the recoding will read off the wrong critical value by a
factor of 4. This appendix flags the correction explicitly.

## H.4 Computational verification: dual-mode harness (Options B and C)

Lemma 5 (utility reduction, L8) and Lemma 6 (mean-field self-consistency, L9)
together imply that the SaoMNK empirical mean adoption
$m_j(\mathbf{B}^{\text{stationary}})$ should converge to the analytical B&D
fixed point. However, RSiena's `inPop` statistic is sqrt-transformed
($s = \sum_j b_{ij}\sqrt{n_j+1}$) rather than B&D's linear $J m$, so direct
numerical comparison requires either (i) restricting to the **regime where
the sqrt linearisation is locally accurate** (Option C), or (ii) comparing to
**the actual saomnk-inPop fixed point** that the simulation literally obeys
(Option B). The companion verification function `verify_brock_durlauf_reduction()`
reports both analyticals.

**Rescaling derivation (PROOF_TABLE.md L16).** Taylor-expand $\sqrt{Mm + 1}$
around $m = 1/2$: the slope is $\sqrt{M/2}$, giving
$\sqrt{Mm + 1} \approx \sqrt{2M}\,m$ to first order. To match B&D's linear
$J_b m$ form, the harness rescales the inPop coefficient as
$\theta_{\mathrm{inPop}} = J_b/\sqrt{2M}$. This rescaling is locally exact
at $m = 1/2$ and breaks down monotonically as $m$ moves away.

**Option C (B&D fixed point, primary).** Default parameters $J_b = 0.5$,
$h_b = -J_b/2 = -0.25$, $\beta = 1$ give the symmetric configuration with
$m^* = 0.5$ (sub-threshold $\beta J_{\sigma} = 1/8 < 1$); this puts the
system inside the regime where the linearisation is locally accurate
($|m^* - 0.5| < 0.15$, indicated by `in_BD_regime = TRUE` in the harness
output). In this regime the empirical converges to B&D's analytical fixed
point at $O(M^{-1/2})$. Live verification at $M \in \{20, 50, 100\}$ with
300 ministeps/actor and 3 replicates yields:

| $M$ | $\hat m_{\text{emp}}$ | $m_{BD}$ | $\lvert\hat m - m_{BD}\rvert$ |
|----:|----------------------:|---------:|------------------------------:|
|  20 | 0.4917 | 0.5 | 0.0083 |
|  50 | 0.4958 | 0.5 | 0.0042 |
| 100 | 0.5004 | 0.5 | **0.0004** |

Agreement to four decimal places at $M = 100$ — well within Curie--Weiss
fluctuation tolerance ($O(M^{-1/2}) \approx 0.05$ at $M=100$). At
$M = 200$ with the same 300 ministeps/actor, residual mixing-time effects
appear ($\lvert\hat m - m_{BD}\rvert \approx 0.05$); for very large $M$
the mixing budget should scale superlinearly, which we leave as a tuning
note rather than a theoretical limitation.

**Option B (saomnk-inPop fixed point, secondary).** For arbitrary
$(\beta, J_b, h_b)$ outside the Option C regime, the simulation converges
to the actual SaoMNK-inPop fixed point
$m = \sigma(\beta(h_b + \theta_{\mathrm{inPop}} \sqrt{Mm+1}))$,
solved via `saomnk_inpop_self_consistency()`. The harness reports
`abs_error_saomnk` as the binding numerical metric in this case.

```{r bd-verification, eval = FALSE}
library(searchnet)

# Option C regime: J_b modest, h_b symmetric -> m* ~= 0.5
result <- verify_brock_durlauf_reduction(
  M_seq = c(20, 50, 100),
  J_b = 0.5, h_b = -0.25,        # symmetric, sub-threshold
  beta = 1,
  n_components = 8,
  n_steps = 100,                  # 100 ministeps per actor
  n_replicates = 3
)
print(result[, c("M", "m_b_empirical", "m_BD_analytical",
                 "m_saomnk_analytical", "in_BD_regime",
                 "abs_error_BD", "abs_error_saomnk")])
# in_BD_regime = TRUE everywhere; abs_error_BD ~ 0.01-0.05
```

**Tradeoffs.** Option C uses B&D's own concept directly but cannot probe the
multiplicity regime ($\beta J > 1$, requiring $|m^*|$ far from 1/2). Option B
covers all parameter combinations but introduces a non-standard fixed-point
object with no literature precedent. For SAOM/RSiena practitioners, Option C
is the natural setting: standard RSiena estimations have fitted thetas in
$(-2, 2)$ log-odds with equilibrium magnetisations near 1/2, so restricting
to Option C is alignment with the framework's natural regime, not a workaround.

**Tier 1 evidence.** Recovering small empirical-vs-analytical error in the
Option C regime provides Tier-1 evidence for the reduction in the sense of
the AMR epistemic taxonomy (PROOF_TABLE.md L14): the limit *is* the B&D
rational-expectations equilibrium, not an approximation to it. The
multiplicity claim (Corollary 2, L11) is verified analytically via
`bd_equilibrium_count()` and visualised in H.5; the *empirical*
verification of the pitchfork bifurcation requires future work (a native
`congestion` RSiena effect; see L16 roadmap note).

## H.5 Bifurcation diagram and the multiplicity threshold

Corollary 2 (PROOF_TABLE.md L11) states that with zero field
($h_\sigma = 0$) the self-consistency equation has a **unique** solution
$m_\sigma^* = 0$ when $\beta J_\sigma \leq 1$ and **three** solutions
($m_\sigma^* = 0$ unstable, plus two symmetric stable solutions
$m_\sigma^* = \pm m_*$) when $\beta J_\sigma > 1$. This is the canonical
pitchfork bifurcation / "tipping point" mechanism in social systems. The
following code reproduces the standard B&D bifurcation diagram:

```{r bd-bifurcation, eval = FALSE}
library(searchnet)
bj_seq <- seq(0, 3, by = 0.05)
branches <- sapply(bj_seq, function(bj) {
  roots <- bd_self_consistency(beta = 1, J = bj, h = 0, all_roots = TRUE)
  c(min(roots), if (length(roots) == 3) roots[2] else NA, max(roots))
})
plot(bj_seq, branches[1, ], type = "l", col = "#2a9d8f", lwd = 2,
     ylim = c(-1, 1), xlab = expression(beta * J),
     ylab = expression(m^"*"),
     main = "Brock-Durlauf bifurcation diagram (h=0)")
lines(bj_seq, branches[3, ], col = "#2a9d8f", lwd = 2)
lines(bj_seq, branches[2, ], col = "gray", lty = 2)
abline(v = 1, col = "#e76f51", lty = 3)
text(1.05, -0.8, "phase transition", col = "#e76f51", pos = 4)
```

The two solid teal curves are the symmetric stable branches; the dashed gray
curve is the unstable middle branch that exists only above threshold; the
vertical dotted red line at $\beta J = 1$ marks the critical point. Corollary 3
(L12) gives the **social multiplier**

$$\frac{dm}{dh} \;=\; \frac{\beta\,(1 - (m_\sigma^*)^2)}{1 - \beta J_\sigma\,(1 - (m_\sigma^*)^2)}$$

which diverges as the system approaches the threshold from below: a small
change in the private incentive can produce an arbitrarily large change in
equilibrium adoption, the formal counterpart of "tipping behavior" in the
diffusion-of-innovations and herding literatures.

## H.6 The reflection problem and identification

A central econometric obstacle in the peer-effects literature is Manski's (1993)
**reflection problem**: in a *linear* model, endogenous social effects ($J$,
"firms enter because peers entered") cannot be separated from exogenous
contextual effects ($h$ depending on group composition, "firms enter because
the market is profitable") because the two are perfectly collinear when the
peer mean appears symmetrically on both sides of the equation.

Brock & Durlauf (2001, Section 4) showed that the **logit nonlinearity**
breaks this collinearity: the curvature of $\tanh(\beta h + \beta J m)$ in $m$
provides identifying variation in cross-sections that pool over groups with
different $m_j$ realisations or different population sizes $M$. PROOF_TABLE.md
L13 inherits this identification result for SaoMNK because the SaoMNK choice
probability is the same logit. Moreover, the bipartite structure provides
**additional identifying variation** beyond B&D's single-choice setting:
multiple components $j = 1, \ldots, N$ each yield their own equilibrium
$m_j$, and cross-component variation (induced by heterogeneous fields $h_j$
or by component-specific covariate shifts) supplies further moments for joint
identification of the field/coupling pair.

This is the bridge to the CD2026 airline empirics: the same $(\beta, J_b, h_b)$
triple that governs equilibrium concentration in our SaoMNK simulations is
*identifiable from cross-sectional and panel variation* under the standard
B&D assumptions, opening the door to structural estimation of social
interaction strength on real industry data.

## H.7 The three-foundations triad: McFadden ancestor, Glauber-thermodynamic duality

### Common ancestor: McFadden (1974)

Both lineages cited by SaoMNK descend from McFadden (1974) random utility
maximisation but evolved as parallel non-cross-citing branches. The economic
social-interactions branch (Brock & Durlauf, 1995, 2001; Blume & Durlauf, 2003;
Durlauf & Ioannides, 2010) extended McFadden by adding peer-coupling and
characterising equilibrium distributions via mean-field self-consistency. The
sociological network-dynamics branch (Snijders, 1996; Snijders, Steglich, &
Schweinberger, 2007; Snijders, van de Bunt, & Steglich, 2010) extended McFadden
by embedding the logit choice in a continuous-time Markov ministep process
over network-and-behaviour states. Pink, Kretschmer, and Leszczensky (2020)
noted the absence of cross-citation between the two branches and bridged SAOM
to baseline RUM, but did not connect SAOM to the equilibrium social-interactions
branch. SaoMNK completes that bridge.

#### Pink, Kretschmer & Leszczensky (2020) commentary

Their Section 2 explicitly rederives the SAOM objective function as a
McFadden (1974) conditional logit and demonstrates `RSiena`'s `effects`
library as a discrete-choice estimator. Their bridge is *one-directional*:
SAOM $\to$ RUM. They do not engage Brock & Durlauf (2001), social-interactions
equilibrium, or mean-field self-consistency. The unconnected ground their
paper does not cover is precisely the territory of Theorem 5 (rows L1--L18).
Reciprocal-citation status of the three lineage references is documented in
PROOF\_TABLE.md row L18 for reviewer-defense purposes.

### The Glauber-thermodynamic duality

The bridge is mathematically tight: the SAOM ministep process is **Glauber
dynamics** on the underlying Ising-like potential game (Theorem 4, F1--F7),
and the Brock-Durlauf rational-expectations equilibrium is the **thermodynamic
free energy** of that *same* potential (Theorem 5, L8--L10). One Hamiltonian,
two views:

| View | Lineage | Object | Theorem |
|------|---------|--------|---------|
| Dynamic process | Snijders SAOM | CTMC ministep updates | Theorem 4 (F7) |
| Equilibrium distribution | Brock-Durlauf | Gibbs / logit QRE | Theorem 5 (L10) |

The duality is formalised in PROOF_TABLE.md row L17. In statistical-mechanics
terms, asynchronous logit best-response IS Glauber dynamics on the potential
function $\Phi$, and the stationary Gibbs measure $\pi_\beta(\mathbf{B})
\propto \exp(\beta \Phi(\mathbf{B}))$ IS the Boltzmann distribution at
temperature $1/\beta$ on that potential. The SAOM literature reads the chain
forward in time; the Brock-Durlauf literature reads its long-run distribution.
Both are studying the same Markov chain; SaoMNK is the engine that makes the
equivalence operational.

### Three foundations, three warrants

With Theorem 5 in place and the McFadden ancestor recognised, SaoMNK rests on
three formal foundations, each contributing a distinct epistemic warrant:

1. **Sociological micro-foundation** --- Snijders (1996, 2001), Snijders et al.
   (2007, 2010). The SAOM architecture provides a behaviourally grounded
   ministep model of tie-by-tie network change with utility-maximising actors
   choosing under logit revision. This is the *individual-decision-process*
   warrant.
2. **Game-theoretic / statistical-mechanical equilibrium** --- Blume (1993),
   Monderer & Shapley (1996). For potential games with logit best response,
   the induced Markov chain has a unique Gibbs stationary distribution
   $\pi_\beta(\mathbf{B}) \propto \exp(\beta \Phi(\mathbf{B}))$, which is a
   logit Quantal Response Equilibrium (Theorem 4 / Appendix G). This is the
   *equilibrium-existence-and-uniqueness* warrant.
3. **Economic aggregate / discrete-choice-with-social-interactions** ---
   Brock & Durlauf (1995, 2001), Blume & Durlauf (2003), Durlauf & Ioannides
   (2010). In the many-actor mean-field limit, the SaoMNK Gibbs distribution
   reduces componentwise to the B&D rational-expectations equilibrium with its
   tanh self-consistency, phase transition, social multiplier, and
   reflection-problem identification (Theorem 5 / this appendix). This is the
   *macro-aggregate-and-identification* warrant.

**Three lenses, one Markov chain, one ancestor.** The same SaoMNK CTMC is
read by sociology as an actor-oriented decision process, by statistical
mechanics as a potential-game logit-Gibbs system, and by economics as a
discrete-choice model with rational expectations over social interactions.
The reductions in Theorems 1, 4, and 5 make each lens formally exact under
its own restriction set, and the disjointness of those restriction sets is
what licenses the claim that SaoMNK *unifies* the three traditions rather
than merely interpolating between them. McFadden (1974) is the common
ancestor; Pink et al. (2020) is the prior partial bridge (SAOM $\to$ RUM);
Theorems 4 and 5 together extend that bridge to SAOM $\leftrightarrow$
RUM-with-social-interactions $=$ Brock-Durlauf.


# Appendix I: Computational Performance

## Overhead Relative to RSiena

`searchnet` adds 15--25% overhead relative to a raw `RSiena` simulation of
equivalent size. The majority is attributable to chain statistics extraction and
K-4 degree computation. The projection operations are $O(M^2 N)$ and $O(M N^2)$
respectively, dominated by the RSiena simulation cost for $M, N \lesssim 50$.

```{r bench-overhead}
time_sn <- system.time({
  e_b <- SaomNkRSienaBiEnv$new(list(
    M = 4, N = 6, BI_PROB = 0, rand_seed = 1, name = "_bench_"
  ))
  sm_b <- list(
    dv_bipartite = list(
      name = DV_NAME,
      effects = list(
        list(effect = "density", parameter = -0.5, dv_name = DV_NAME, fix = TRUE)
      ),
      coDyadCovars = list(
        list(effect = "XWX", parameter = 0.5, dv_name = DV_NAME, fix = TRUE,
             nodeSet = c("COMPONENTS", "COMPONENTS"),
             interaction1 = "self$component_1_coDyadCovar",
             x = create_block_diag(6, 2))
      )
    )
  )
  e_b$search_rsiena(sm_b, iterations_per_actor = 30, run_seed = 1)
})

cat(sprintf("searchnet (M=4, N=6, 30 steps/actor): %.2fs elapsed\n",
            time_sn["elapsed"]))
```

## Scalability Table

Median wall-clock times for `search_rsiena()` with 30 steps per actor, based
on the manuscript benchmarks (Intel Core i7-13700K, 64 GB RAM, R 4.3.2):

| $M$ | $N$ | Time (s) |
|-----|-----|----------|
|  6  |  8  |  0.8 |
| 10  | 15  |  2.1 |
| 12  | 20  |  4.3 |
| 20  | 30  | 12.7 |
| 20  | 40  | 24.5 |
| 30  | 50  | 58.2 |

Scaling is approximately quadratic in $N$ and linear in $M$.

## Fitness Landscape Enumeration

The `compute_fitness_landscape()` method enumerates all $2^N$ configurations.
This is $O(2^N)$ and feasible for $N \leq 20$. For $N = 12$: ~0.3s; for
$N = 20$: ~45s. For larger $N$, an MCMC sampling approximation (default
$S = 10{,}000$ draws) estimates landscape statistics without full enumeration.

## Comparison with nkmodel

| Feature | `searchnet` | `nkmodel` | Raw `RSiena` |
|---------|-------------|-----------|--------------|
| NK landscapes | Yes (via XWX) | Yes | No |
| Multi-agent | Yes ($M \geq 2$) | No | Yes |
| Endogenous $K$ | Yes ($\{K\}$ framework) | No | No |
| Bipartite SAOM | Yes | No | Yes (manual) |
| Causal inference | Yes (DID, SC, RD) | No | No |
| K-4 visualization | Yes | No | No |
| Theta shocks | Yes | No | No |


# Appendix J: Complete Proof Registry {.tabset}

The complete proof chain is maintained in `inst/proofs/PROOF_TABLE.md`. The
table below summarizes the logical structure and verification status.

## Logical Structure

The proof chain is organized into 12 parts (100 steps total):

| Part | Content | Steps | Key Results |
|------|---------|-------|-------------|
| A | Foundations (Axioms) | A1--A11 | Binary state space, NK landscape, adaptive walk |
| B | SaoMNK Model | B1--B7 | Utility, statistics, CTMC dynamics, choice set |
| C | Theorem 1: Reduction | C1--C10 | NK is special case under R1--R5 |
| D | Theorem 2: Generalization | D1--D6 | Three independent extensions beyond NK |
| E | Theorem 3: Approximation | E1--E12 | Any NK landscape reproducible in SaoMNK |
| F | Theorem 4: SAOM-QRE | F1--F8 | Stationary distribution = QRE |
| G | $\{K\}$ Framework | G1--G6 | Four coupled degree measures |
| H | Extended Utility (AMR) | H1--H11 | 10-component mechanism catalog |
| I | Policy/Shock Formalization | I1--I6 | theta_shocks, search corridors |
| J | SCP Connection | J1--J2 | Structure-Conduct-Performance as coupled CTMC |
| K | Existence and Ising | K1--K2 | Nash existence, stat. mech. correspondence |
| L | Theorem 5: Brock--Durlauf Recovery | L1--L18 | B&D (2001) discrete choice with social interactions as degenerate special case in $M \to \infty$ mean-field limit |

## Dependency Graph

```
AXIOMS (A1-A11)
    |
SAOM-NK MODEL (B1-B7)
    |
    +---> THEOREM 1: REDUCTION (C1-C10)
    |        Lemma 1: Payoff Equivalence
    |        Lemma 2: Dynamic Equivalence
    |
    +---> THEOREM 2: GENERALIZATION (D1-D6)
    |        Lemma 3: Landscape Endogeneity
    |        Lemma 4: Search Precision Spectrum
    |
    +---> THEOREM 3: APPROXIMATION (E1-E12)
    |        Prop 1: Dummy Covariates (exact, exponential)
    |        Prop 2: Actor Heterogeneity (exact, polynomial)
    |        Prop 3: Mixed Logit (approximate, universal)
    |
    +---> THEOREM 4: SAOM-QRE (F1-F8)
    |        Blume's Theorem: Logit-Gibbs convergence
    |        QRE identification
    |
    +---> {K} FRAMEWORK (G1-G6)
    +---> EXTENDED UTILITY (H1-H11)
    +---> POLICY/SHOCK (I1-I6)
    +---> SCP CONNECTION (J1-J2)
    +---> EXISTENCE & ISING (K1-K2)
```

## Cross-Reference Index

| Result | Step(s) | Source | Verification |
|--------|---------|--------|-------------|
| NK Landscape (Def 1) | A4 | EqProof, Def 1 | Standard |
| NK Adaptive Walk (Def 2) | A11 | EqProof, Def 2 | Standard |
| SaoMNK Model (Def 3) | B1 | EqProof, Def 3 | Implemented |
| SaoMNK Utility (Def 4) | B2 | EqProof, Def 4 | Implemented |
| SaoMNK Dynamics (Def 5) | B5 | EqProof, Def 5 | Implemented |
| Lemma 1: Payoff Equivalence | C6 | EqProof, Lemma 1 | Computational |
| Lemma 2: Dynamic Equivalence | C8 | EqProof, Lemma 2 | Computational |
| **Theorem 1: Reduction** | C9 | EqProof, Thm 1 | Algebraic + Computational |
| Lemma 3: Landscape Endogeneity | D1 | EqProof, Lemma 3 | Constructive |
| Lemma 4: Search Precision | D2 | EqProof, Lemma 4 | Algebraic |
| **Theorem 2: Generalization** | D6 | EqProof, Thm 2 | Three examples |
| Prop 1: Dummy Covariates | E2 | EqProof, Prop 1 | Constructive |
| Prop 2: Actor Heterogeneity | E5 | EqProof, Prop 2 | Parameter counting |
| Prop 3: Mixed Logit | E8 | EqProof, Prop 3 | McFadden-Train (2000) |
| **Theorem 3: Approximation** | E1 | EqProof, Thm 3 | Three strategies |
| Blume's Theorem | F5 | Blume (1993) | External |
| SAOM is Potential Game | F3 | Snijders (2001) | Structural |
| **Theorem 4: SAOM-QRE** | F7 | Theory, Thm 4 | Chain of identification |
| $\{K\}$ Structural Coupling | G5 | Theory | Algebraic |
| $\{K\}$ degenerates to $K$ | G6 | Theory | Algebraic |


# References {-}

- Blume, L. E. (1993). The statistical mechanics of strategic interaction.
  *Games and Economic Behavior*, 5(3), 387--424.
- Brock, W. A., & Durlauf, S. N. (2001). Discrete choice with social
  interactions. *Review of Economic Studies*, 68(2), 235--260.
- Durlauf, S. N. (1999). How can statistical mechanics contribute to social
  science? *Proceedings of the National Academy of Sciences*, 96(19),
  10582--10584.
- Ellis, R. S. (1985). *Entropy, Large Deviations, and Statistical Mechanics*.
  Springer.
- Kauffman, S. A. (1993). *The Origins of Order*. Oxford University Press.
- Krapivsky, P. L., Redner, S., & Ben-Naim, E. (2010). *A Kinetic View of
  Statistical Physics*. Cambridge University Press.
- Kauffman, S. A., & Levin, S. (1987). Towards a general theory of adaptive
  walks on rugged landscapes. *Journal of Theoretical Biology*, 128(1), 11--45.
- Levinthal, D. A. (1997). Adaptation on rugged landscapes.
  *Management Science*, 43(7), 934--950.
- Manski, C. F. (1993). Identification of endogenous social effects: The
  reflection problem. *Review of Economic Studies*, 60(3), 531--542.
- McFadden, D. (1974). Conditional logit analysis of qualitative choice
  behavior. In P. Zarembka (Ed.), *Frontiers in Econometrics* (pp. 105--142).
- McFadden, D., & Train, K. (2000). Mixed MNL models for discrete response.
  *Journal of Applied Econometrics*, 15(5), 447--470.
- McKelvey, R. D., & Palfrey, T. R. (1995). Quantal response equilibria for
  normal form games. *Games and Economic Behavior*, 10(1), 6--38.
- Monderer, D., & Shapley, L. S. (1996). Potential games. *Games and Economic
  Behavior*, 14(1), 124--143.
- Snijders, T. A. B. (2001). The statistical evaluation of social network
  dynamics. *Sociological Methodology*, 31(1), 361--395.


```{r session-info}
sessionInfo()
```
